---
title: "TSENAT: Tsallis Entropy Analysis Toolbox"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5
)
```

## Overview

**TSENAT** (Tsallis Entropy Analysis Toolbox) quantifies **isoform
heterogeneity** in RNA-seq data using Tsallis entropy—a scale-dependent
diversity measure distinct from abundance-focused tools (DESeq2, Salmon)
and differential transcript usage packages (*SplincingFactory*). By
tuning a sensitivity parameter `q`, users examine diversity at different
scales: rare variants (low `q`) or dominant isoforms (high `q`). This
vignette demonstrates computing Tsallis entropy from transcript-level
abundance estimates, comparing measures between groups, and visualizing
scale-dependent differences via q-curves. All results integrate
seamlessly with Bioconductor via `SummarizedExperiment`.

### Motivation and Bioconductor Contribution

Common RNA-seq tools, such as DESeq2 or differential transcript usage
(DTU) packages like *SplincingFactory*, focus either on changes in
**total gene abundance** or on **individual transcript usage
proportions**. While powerful, these approaches can obscure an important
biological phenomenon: genes often change *which* isoforms they express
without large changes in overall abundance, and this isoform
heterogeneity itself—independent of proportional shifts— carries
biological signal about cellular state and function.

**TSENAT fills this gap** by providing a **scale-dependent diversity
framework** that captures isoform heterogeneity regardless of which
specific isoforms dominate. Unlike *SplincingFactory*, which tests for
shifts in transcript proportions between conditions, TSENAT quantifies
the **entropy or "complexity"** of the isoform distribution itself.
Tsallis entropy achieves this through a sensitivity parameter `q`: by
tuning `q`, you slide a lens across abundance scales—zooming in on rare
variants (low `q`) or focusing on dominant isoforms (high `q`)—
capturing scale-dependent isoform switching that ordinary summaries
overlook.

**Bioconductor inclusion** reflects TSENAT's role as a complementary
statistical framework for transcript-level analysis. The package
introduces rigorous information-theoretic methods (Tsallis entropy with
bootstrapped confidence intervals, paired testing, and functional
principal component analysis) designed specifically for isoform-level
RNA-seq. By integrating with Bioconductor's `SummarizedExperiment`
standard and leveraging the ecosystem's mature statistical
infrastructure, TSENAT empowers researchers to ask novel questions about
isoform-level complexity that are inaccessible to existing DTU or
abundance-focused tools.

### High-level workflow

1.  Preprocess counts and filter low-abundance transcripts.
2.  Compute relative transcript proportions within each gene and
    evaluate Tsallis entropy across one or more `q` values.
3.  Summarize per-gene values across samples (mean/median) and test for
    differences between groups (Wilcoxon or permutations).
4.  Visualize q-curves and inspect transcript-level counts for top
    genes.

**Design assumption**: This vignette assumes paired or longitudinal
study designs with a minimum of 6-8 samples per group for adequate
statistical power.

## Installation

To install `TSENAT` from Bioconductor:

```{r installation, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TSENAT")
```

Then load the package:

```{r load-package, eval = FALSE}
library(TSENAT)
```

## Quick Start

For users who want to see results quickly, here is a minimal workflow:

```{r quickstart, eval = FALSE}
library(TSENAT)
library(SummarizedExperiment)

# Load example data
data("readcounts", package = "TSENAT")
readcounts <- as.matrix(salmon_dataset)

# Load sample metadata and annotation
metadata_df <- read.table(
  system.file("extdata", "metadata.tsv", package = "TSENAT"),
  header = TRUE, sep = "\t"
)
gff3_file <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build SummarizedExperiment and add metadata
se <- build_se(readcounts, gff3_file, metadata = metadata_df)

# Filter low-abundance transcripts
se <- filter_se(se, stringency = "medium")

# Compute Tsallis entropy (using single q value for quick start)
# Note: calculate_difference() only accepts single q values because 
# entropies at different q are mathematically dependent (AR(1) structure).
# For multi-q analysis accounting for correlation, use calculate_lm_interaction() instead.
ts_se <- calculate_diversity(se, q = 1, norm = TRUE, metadata = metadata_df)
```

## Data Structure Overview

`TSENAT` operates on **transcript-level read counts** rather than
pre-aggregated gene-level summaries. Input data should be organized as a
matrix with transcript IDs as rows and samples as columns, typically
derived from quantification tools such as SALMON or kallisto. This
matrix must be accompanied by transcript-to-gene mapping information,
conventionally provided through a GFF3 annotation file.

All analysis results are organized into a `SummarizedExperiment`
object—the standard Bioconductor container for high-throughput
expression data. This unified structure facilitates seamless integration
with the broader Bioconductor ecosystem. For more details on the
`SummarizedExperiment` class and its capabilities, refer to the
[SummarizedExperiment
documentation](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html).

Within each `SummarizedExperiment` object, three key components work
together: **column metadata** (`colData`) captures sample-level
information including identifiers, group assignments, and pairing
information; **row metadata** (`rowData`) stores transcript identifiers,
gene assignments, and feature annotations; and **assays** contain the
computed diversity measures themselves. The assay structure is
flexible—typically one assay per `q` value, or a single assay when
combining multiple q values—where each cell represents the normalized or
unnormalized Tsallis entropy for a specific gene-sample combination.

## What is Tsallis entropy?

### Mathematical Definition

Tsallis entropy is a one-parameter family of diversity measures that
generalizes Shannon entropy [@Shannon1948] via a sensitivity parameter
`q` [@Tsallis1988a]. Unlike Shannon entropy, which treats all isoforms
equally, Tsallis entropy lets you **zoom in on rare isoforms (low q) or
focus on dominant species (high q)**, capturing scale-dependent isoform
variation that ordinary summaries miss. For a discrete probability
vector $p=(p_1,\dots,p_n)$ it is defined as:

$$
S_q(p) = \frac{1-\sum_{i=1}^n p_i^q}{q-1}.
$$

This parametric family unites diverse entropy concepts under a single
framework: - **Generalization**: Extends beyond classical Shannon
entropy to capture scale-dependent phenomena - **Mathematical
elegance**: Reduces to well-known indices at specific q-values -
**Practical flexibility**: Allows data-driven exploration via q-curves

### Information-Theoretic Interpretation

In information theory terms [@Shannon1948, @CoverThomas2006], entropy
measures the **uncertainty or surprise** when drawing a single item
(transcript) from a probabilistic ensemble (isoform distribution).
Higher entropy means the draw is less predictable (many similarly
abundant isoforms contribute equally), while lower entropy means the
distribution is dominated by one or a few highly abundant transcripts.
This distinction is crucial: two genes may have identical total
abundance but dramatically different isoform complexity.

Tsallis entropy generalizes this concept by introducing the `q`
parameter as a **sensitivity dial** that shifts emphasis between
different abundance scales: - **Small q (0 \< q \< 1)**: Emphasizes
rare, low-abundance isoforms - **q = 1**: Recovers Shannon entropy
(balanced across all abundance scales) - **q = 2**: Yields the
Gini-Simpson index, emphasizing dominant isoforms - **Large q (q \>
2)**: Focuses almost exclusively on the most abundant species

This scale-dependence is the key innovation: by computing $S_q$ across a
range of q-values (a "q-curve"), you obtain a complete picture of
isoform heterogeneity across biological scales [@Jost2006,
@Furuichi2006]. In practice, combining q-curves with resampling or
paired designs yields a concise, statistically rigorous summary of
isoform complexity.

### Essential Limiting Cases

Tsallis entropy exhibits several important mathematical limits and
special cases:

-   **Limit** $q\to 1$ (Shannon entropy):
    $\lim_{q\to1} S_q(p) = H(p) = -\sum_i p_i\log p_i$
    -   Classical information entropy (measured in *bits* if log base 2,
        or *nats* if natural log)
    -   Treats all isoforms with equal weight regardless of abundance
-   **q = 0 (Richness)**: $S_0 = m - 1$ where $m$ = number of observed
    isoforms
    -   Counts expressed isoforms (pure species richness, log-scale)
    -   Minimum assumption: just presence/absence
-   **q = 1 (Shannon)**: $S_1(p) = -\sum_i p_i\log p_i$
    -   Standard information entropy [@Shannon1948]
    -   Optimal estimator of information content and predictability
-   **q = 2 (Gini–Simpson / collision index)**: $S_2 = 1 - \sum_i p_i^2$
    -   Probability that two randomly drawn transcripts differ
        [@Simpson1949]
    -   Robust to rare variants (insensitive to very low-abundance
        isoforms)
    -   Range: [0, 1 - 1/m] for m isoforms
-   **Uniform maximum**: For $m$ equally expressed isoforms, maximum
    entropy is: $$S_{q,\max}(m) = \frac{1-m^{1-q}}{q-1}$$
    -   **Normalized Tsallis entropy**:
        $\tilde S_q = \frac{S_q(p)}{S_{q,\max}(m)} \in [0,1]$
    -   Enables comparison across genes with different isoform counts
    -   Essential for distinguishing true biological heterogeneity from
        annotation artifacts

### Biological Potentiality: Why Tsallis Entropy Matters for Isoform Analysis

Standard RNA-seq analysis measures *whether* transcript abundance
changes between conditions (fold change, differential expression).
However, a critical complementary question remains underexplored: *how*
does the **diversity of isoforms** change?

A gene may show little change in total abundance while dramatically
reshuffling its isoform repertoire—a phenomenon that current methods
largely miss. Tsallis entropy captures precisely this: the **complexity,
richness, and balance of the isoform landscape**. By varying `q`,
researchers can examine whether changes are driven by **shifts in rare
variants or reorganization of dominant isoforms**, yielding mechanistic
insight into splicing regulation and cellular adaptation.

The mathematical foundation of Tsallis entropy rests on Shannon's
seminal information theory [@Shannon1948], which defined entropy as a
measure of uncertainty. Building on this, Rényi [@Renyi1961] developed a
parametrized family of entropy measures that generalize Shannon entropy.
Tsallis entropy extends this framework further, providing a
scale-dependent diversity measure where the `q` parameter controls
emphasis: at $q \to 1$ it approaches Shannon entropy (balanced
sensitivity), while $q = 2$ yields the Gini-Simpson index (focus on
common species), and $q \to 0$ approaches richness (emphasis on rare
species). Jost [@Jost2006] clarified how entropy families decouple "true
diversity" (species richness) from abundance disparities—a principle
directly transferable to isoforms: two genes with identical Shannon
entropy may have very different functional diversity if one is dominated
by a single isoform (low q sensitivity) while the other distributes
evenly (high q sensitivity) [@Derian2022, @Seweryn2020].

### Applications of Tsallis Entropy in Genomics and Precision Medicine

Tsallis entropy has emerged as a powerful tool for analyzing isoform
complexity across diverse biological contexts, from single-cell
transcriptomics [@Gandrillon2021, @Seweryn2020] to spatial
transcriptomics [@Hinostroza2026]. Recent applications demonstrate its
value:

**1. Disease biomarker discovery and cancer diagnosis**: Dysregulated
isoform heterogeneity often precedes or accompanies disease phenotypes
[@Chanda2020]. Cancer cells frequently exhibit abnormal splicing entropy
signatures [@Seweryn2020, @Chanda2020], which may be detectable before
bulk transcriptional changes. The ability to measure isoform complexity
via Tsallis entropy at different q-values reveals whether
disease-associated changes affect rare variants (low q) or dominant
isoforms (high q), providing mechanistic insight into splicing
dysregulation in malignant transformation.

**2. Cell state transitions and differentiation**: Developmental and
differentiation processes involve systematic shifts in isoform
complexity [@Gandrillon2021]. Maturation typically increases dominant
isoform usage (lower entropy), while dedifferentiation or proliferation
may raise entropy through stress-induced isoform switching. Single-cell
transcriptomics with entropy measures [@Golomb2026] reveals how
individual cells navigate state transitions, capturing both the
deterministic aspects (dominant isoform selection) and stochastic
heterogeneity (rare variant exploration) that define cellular
plasticity.

**3. Adaptive and stress responses**: Rapid environmental challenges
(heat stress, nutrient starvation, infection) trigger coordinated
changes in splicing entropy [@Bajic2024] as cells modulate protein
isoform ratios to meet functional demands. By measuring entropy at
multiple q-values, researchers can distinguish between adaptive
responses (systematic entropy changes) and noise, revealing the
biological strategies cells employ for stress resilience.

**4. Tissue-specific regulation and functional specialization**:
Different tissues maintain distinct entropy "set points" reflecting
their functional requirements [@Golomb2026]. Brain tissues, for example,
show higher isoform complexity than liver [@Golomb2026], reflecting the
protein functional diversity needed for neural circuit complexity.
Spatial transcriptomics now enables mapping of isoform entropy patterns
within tissues [@Hinostroza2026], revealing how splicing regulation
supports tissue architecture and function.

**5. Precision medicine and therapeutic stratification**: Isoform
entropy signatures may predict patient response to therapy [@Derian2022,
@Golomb2026]. Individual variation in basal isoform complexity, measured
via Tsallis entropy, could stratify patients into responder and
non-responder groups, enabling personalized treatment selection. This
extends precision medicine beyond mutation detection to include
functional complexity of the proteome.

The recent emphasis on information theory in computational biology
[@Chanda2020, @Adami2004, @Bartal2022] reflects growing recognition that
traditional measures (fold change, p-values) miss key aspects of
biological organization. Tsallis entropy, by allowing exploration across
the q-spectrum, captures organization at multiple biological scales
simultaneously—a capability essential for unpacking the complexity of
gene regulation in health and disease.

## Example dataset

An example dataset is included for demonstration.

```{r setup}
# Load packages
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
})
```

### Load data and metadata

Now we will load the example dataset and associated metadata:

```{r data}
# Load example dataset (lazy-loaded as 'readcounts' by default)
# This includes SALMON preprocessing outputs:
# - salmon_dataset: transcript-level NumReads (raw fragment counts)
# - salmon_tpm: TPM matrix (length- and library-normalized estimates)
# - salmon_effective_length: EffectiveLength vector (read-length corrected)
data(readcounts)

# Ensure numeric format (keep transcript IDs as rownames)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Load sample metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

# Get the GFF3.gz annotation file path
gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")
```

### Data preprocessing and filtering

Next, we will create a `SummarizedExperiment` data container, a standard
Bioconductor class for storing high-throughput assay data along with
associated metadata. You can read more about it in the [Bioconductor
documentation]
(<https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html>).

We will use the GFF3.gz annotation file for extracting the
transcript-to-gene mapping and building the SummarizedExperiment. We'll
also include SALMON preprocessed data (TPM and EffectiveLength) for
improved analysis:

```{r build-se}
## Build a `SummarizedExperiment` from readcounts + GFF3.gz annotation
# The gff3_dataset variable contains the path to annotation.gff3.gz
# SALMON preprocessed data (salmon_tpm, salmon_effective_length) is auto-detected
# Also map sample metadata for proper pairing detection in filtering
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
```

The `SummarizedExperiment` object contains transcript-level counts with
gene annotations extracted from the GFF3 file. Let's examine its
internal structure:

```{r examine-se-structure, include = FALSE}
# Display the structure of the SummarizedExperiment
cat("SummarizedExperiment dimensions:\n")
cat("  Transcripts (rows):", nrow(se), "\n")
cat("  Samples (columns):", ncol(se), "\n")
cat("  Assay (counts):", nrow(assay(se, "counts")), "×", ncol(assay(se, "counts")), "\n")
cat("\nRowData columns:", paste(colnames(rowData(se)), collapse = ", "), "\n")
cat("Sample metadata available:", exists("metadata_df"), "\n\n")

# Show rowData with gene information
rd <- as.data.frame(rowData(se)[1:5, ])
knitr::kable(rd, caption = "First 5 rows of rowData (gene information)", row.names = FALSE)
```

```{r examine-metadata, include=FALSE}
# Display sample metadata (before mapping to colData)
if (exists("metadata_df")) {
  cat("\nSample metadata (metadata_df):\n")
  cat("  Rows:", nrow(metadata_df), "samples\n")
  cat("  Columns:", paste(colnames(metadata_df), collapse = ", "), "\n\n")
  knitr::kable(metadata_df, caption = "Sample metadata (metadata_df)")
}
```

Before computing the Tsallis diversity, you might want to filter out
genes with a low overall expression or limit the analysis to transcripts
with a sufficient minimum expression level. Expression estimates of
transcript isoforms with zero or low expression might be highly
variable. For more details on the effect of transcript isoform
prefiltering on differential transcript usage, see [this
paper](https://doi.org/10.1186/s13059-015-0862-3).

```{r readfilter}
## Filter lowly-expressed transcripts and report counts
# Use stringency = "medium" for balanced filtering:
# - Keeps transcripts expressed in ≥50% of samples
# - Uses median TPM threshold (Q2) as minimum abundance cutoff
# - Ensures reliable entropy estimation across genes
# Stringency options: "soft" (permissive), "medium" (balanced), "severe" (stringent)
se <- filter_se(se, stringency = "medium")
```

The `stringency = "medium"` parameter automatically determines filtering
thresholds based on your paired study design. It keeps transcripts
present in at least 50% of samples with TPM values above the median of
mean gene TPM. This balanced approach filters out low-abundance
transcripts that contribute noise while retaining sufficient isoform
diversity for meaningful entropy calculations.

## Compute Tsallis entropy.

Now we will compute Tsallis entropy.

```{r tsallis-calc-sequence}
# compute Tsallis entropy for a sequence of values (normalized)
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(se,
    q = qvec, norm = TRUE, metadata = metadata_df
)
```

```{r kable-tsallis-calc-sequence, echo = FALSE}
# Create table with gene names instead of gene IDs
assay_data <- assay(ts_se)
ts_rowdata <- rowData(ts_se)
se_rowdata <- rowData(se)

# Select first 4 genes and first 3 columns
table_data <- assay_data[1:4, 1:3]

# Map gene IDs to gene names from original se
# First, get the gene IDs from ts_se rowData
ts_gene_id_col <- if ("gene_id" %in% colnames(ts_rowdata)) "gene_id" else if ("genes" %in% colnames(ts_rowdata)) "genes" else rownames(ts_se)[1:4]
if (is.character(ts_gene_id_col) && length(ts_gene_id_col) == 1) {
    gene_ids <- ts_rowdata[[ts_gene_id_col]][1:4]
} else {
    gene_ids <- rownames(ts_se)[1:4]
}

# Create mapping from se rowData
se_gene_id_col <- if ("gene_id" %in% colnames(se_rowdata)) "gene_id" else if ("genes" %in% colnames(se_rowdata)) "genes" else NULL
se_gene_name_col <- if ("gene_name" %in% colnames(se_rowdata)) "gene_name" else NULL

if (!is.null(se_gene_id_col) && !is.null(se_gene_name_col)) {
    gene_id_to_name <- setNames(
        as.character(se_rowdata[[se_gene_name_col]]),
        as.character(se_rowdata[[se_gene_id_col]])
    )
    # Remove duplicates, keeping first occurrence
    gene_id_to_name <- gene_id_to_name[!duplicated(names(gene_id_to_name))]
    
    row_labels <- unname(gene_id_to_name[as.character(gene_ids)])
    # Use gene IDs if names not found
    row_labels[is.na(row_labels)] <- as.character(gene_ids)[is.na(row_labels)]
} else {
    row_labels <- as.character(gene_ids)
}

rownames(table_data) <- row_labels

knitr::kable(
    table_data,
    caption = "Tsallis entropy for first 4 genes across q values",
    digits = 5,
    row.names = TRUE
)
```

Map once again the sample metadata into the multi-q
`SummarizedExperiment` as before.

With the **Tsallis q-sequence** we can produce a q-curve per sample and
gene. These curves show how diversity emphasis shifts from rare to
dominant isoforms as `q` increases and form the basis for interaction
tests. The q-curve shows entropy as a function of `q`. Diverging curves
between groups indicate **scale-dependent diversity differences**:
separation at low `q` implies differences in rare isoforms, while
separation at high `q` signals differences in dominant isoforms.

```{r tsallis-q-curve, fig.width=10, fig.height=5.2}
# Plot overall q-curve (mean entropy across all genes)
p_qcurve <- plot_tsallis_q_curve(ts_se)
print(p_qcurve)
```

## Quality Control: Sample Influence Assessment

Before proceeding with group-level comparisons, we should assess whether
individual samples exert disproportionate influence on our entropy
estimates. To address this, we employ a **leave-one-out influence
assessment** combined with **robust M-estimation** (iteratively
re-weighted least squares with Huber loss). This approach quantifies how
much each sample's removal affects the estimated location differences
across the q-spectrum, providing a sample-level quality control metric
independent of group assignment.

Samples with high influence scores exert disproportionate leverage on
entropy estimates and may warrant deeper investigation for technical
quality issues, subject-specific outliers, or genuine biological signals
that require careful interpretation.

```{r qc-sample-check-multiQ, echo=TRUE}
# Perform multi-q sample influence analysis
# The m_estimate function handles:
# - Multi-q data collapsing (averaging/median across q values)
# - Leave-one-out sample influence assessment
# - Flagging samples based on influence_threshold (default: 0.75 = 75th percentile)
# Returns: Sample names, influence scores, and QC status

sample_qc <- m_estimate(ts_se, samples = "sample_type", 
                        loss_type = "huber", 
                        q_combine_method = "mean",
                        influence_threshold = 0.75,
                        paired = TRUE)
```

```{r qc-sample-display, echo=FALSE, results="asis"}
# Display QC results with condition, pair, influence, and centroid distance
base_display_cols <- c("Sample", "Condition", "Proportion_Affected", 
                       "Genes_Affected", "Distance_from_Centroid", "Status")

# Initialize with base columns
qc_table <- sample_qc[, base_display_cols]

# Add Pair_ID if available (check multiple possible column names)
pair_id_col <- NULL
if ("Pair_ID" %in% colnames(sample_qc)) {
  pair_id_col <- "Pair_ID"
} else if ("Pair" %in% colnames(sample_qc)) {
  pair_id_col <- "Pair"
} else if ("pair_id" %in% colnames(sample_qc)) {
  pair_id_col <- "pair_id"
}

if (!is.null(pair_id_col)) {
  qc_table <- sample_qc[, c("Sample", "Condition", pair_id_col, 
                            "Proportion_Affected", "Genes_Affected", 
                            "Distance_from_Centroid", "Status")]
  colnames(qc_table)[which(colnames(qc_table) == pair_id_col)] <- "Pair"
}

# Round numeric columns for better readability
qc_table$Proportion_Affected <- round(qc_table$Proportion_Affected, 3)
qc_table$Genes_Affected <- round(qc_table$Genes_Affected, 1)
qc_table$Distance_from_Centroid <- round(qc_table$Distance_from_Centroid, 4)

# Display as formatted table with automatic width and alignment
knitr::kable(
    qc_table,
    caption = "Sample Influence Assessment via M-estimation",
    digits = 4,
    row.names = FALSE,
    align = c("l", rep("c", ncol(qc_table) - 1))
)
```

## Linear-model interaction

Once we have, we can test for interactions between `q` and sample groups
across a sequence of `q` values (`q-sequence`), that is, whether the
shape of the q-curve differs between groups. This approach is a powerful
way to detect scale-dependent diversity differences that may not be
apparent at a single `q` value.

Computing Tsallis entropy across a sequence of `q` values is important
because `q` acts as a sensitivity lens that shifts emphasis between rare
and dominant events in data. Thus, the resulting **q-curve** can help us
identify at which scales (i.e. for which `q` values) the diversity
differences between groups are most pronounced. The Tsallis `q-sequence`
can also help with quality control and outlier detection. By examining
the q-curves of individual samples, we can identify samples that deviate
from expected patterns, which may indicate technical issues or
biological outliers.

Once we have the **q-sequence** computed, we can test for interactions
between `q` and sample groups using one of four methods implemented in
`calculate_lm_interaction()`:

-   **linear**: fits a simple linear model. This is fast and
    interpretable and is appropriate when the q–response is
    approximately linear and samples are independent.
-   **lmm**: fits a linear mixed model which accounts for subject-level
    random effects (paired or repeated measures).
-   **gam**: fits a generalized additive model. This method allows to
    capture nonlinear q–response shapes.
-   **fpca**: uses a functional-data approach to reduce each gene's
    q-curve to principal component scores. This is recommended when q is
    densely sampled or when q-curves exhibit complex, high-dimensional
    variation.

Below we demonstrate a practical `gam` (Generalized Additive Model)
example. This method flexibly captures nonlinear q-response patterns and
is particularly useful when the relationship between q and entropy is
not linear (for example, the q:group interaction with complex response
shapes).

```{r lm-interaction-before-qcurve, message=FALSE}
## Linear-model interaction test across q values: detect q x group interaction
## with Hochberg FWER control (fast, closed-form procedure)
if (requireNamespace("mgcv", quietly = TRUE)) {
    # Compute and show top hits (by adjusted p-value)
    # Using Hochberg for FWER control via closed-form procedure
    lm_res <- calculate_lm_interaction(ts_se,
        method = "gam",
        paired = TRUE,
        multicorr = "hochberg",
        nthreads = 3,
        verbose = TRUE
    )
}
```

```{r kable-lm-interaction-before-qcurve, message=FALSE, echo = FALSE}
# show the top entries with proper p-value formatting
lm_res_display <- head(lm_res, 6)

# Select only key columns that exist in the data
col_select <- intersect(c("gene", "gene_name", "p_interaction", "adj_p_interaction", 
                          "n_observations", "n_subjects", "arima_transformation"), 
                        colnames(lm_res_display))
lm_res_display <- lm_res_display[, col_select]

# Use gene_name from lm_res (already populated by calculate_lm_interaction)
# If gene_name is available, display it; otherwise use gene (ID)
if ("gene_name" %in% colnames(lm_res_display) && "gene" %in% colnames(lm_res_display)) {
    lm_res_display$gene <- lm_res_display$gene_name
    lm_res_display$gene_name <- NULL  # Remove redundant column
}

# Format p-values in scientific notation to show actual values
if ("p_interaction" %in% colnames(lm_res_display)) {
    lm_res_display$p_interaction <- format(lm_res_display$p_interaction, scientific = TRUE, digits = 3)
}
if ("adj_p_interaction" %in% colnames(lm_res_display)) {
    lm_res_display$adj_p_interaction <- format(lm_res_display$adj_p_interaction, scientific = TRUE, digits = 3)
}

knitr::kable(
    lm_res_display,
    caption = "Top q-interaction results from GAM analysis with Hochberg correction",
    digits = 5,
    row.names = FALSE
)
```

Now we will plot the q-curve profile for the top gene identified by the
linear-model interaction test. We will start with `LINC03040`.

```{r foxj2-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for the top 2 genes from lm_res with the most significant interactions
# Pass lm_res directly to the function to let it extract genes by significance
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "LINC03040")
print(plot_target)
```

Now let’s look at Now let’s look at PNPT1.

```{r pnpt1-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for top Now let’s look at PNPT1
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "PNPT1")
print(plot_target)
```

Finally, we can visualize the gene `CXCL12`.

```{r gsr-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for top METTL26
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "CXCL12")
print(plot_target)
```

### Effect size analysis across q-spectrum

To understand which **diversity scales show the most pronounced
biological differences** between groups, we compute effect size metrics
(Tsallis divergence D_q) across the q-spectrum while **respecting the
paired sample design**. This reveals whether group differences are
driven by rare isoforms (low q), dominant isoforms (high q), or
uniformly across scales.

The approach uses **linear mixed-effects regression** (LMM) with q as
continuous predictor to assess slope differences in entropy change
across q-spectrum between treatment groups, with effect size computed as
**Tsallis divergence** D_q(control\|\|treatment) [@I002; @I003; @I004].
This information-theoretic measure automatically respects Tsallis
entropy properties and adapts to each q value.

**Theoretical Justification:** Tsallis entropy theory [@I002, @I004]
demonstrates that different q values reveal fundamentally different
aspects of the isoform distribution. As documented in the foundational
divergence literature [@I004], D₀.₅(P\|\|Q) often differs substantially
from D₂(P\|\|Q), with q=0.5 emphasizing rare variants and q=2
emphasizing abundant isoforms. Computing effect sizes independently at
each q value thus captures which diversity scales (rare vs. dominant
isoforms) show the strongest biological differentiation between
conditions.

**Effect Size Interpretation:** Tsallis divergence D_q quantifies the
information-theoretic distance between two distributions' entropy
patterns across the q-spectrum. For paired designs, divergence is
computed separately for each pair and then averaged, accounting for
within-pair correlation and reducing noise from between-pair variation.
The **mean divergence** across q values serves as the effect size,
automatically capturing how entropy distributions differ between control
and treatment groups at all scales (rare to abundant isoforms). Values D
\> 0.1 indicate meaningful divergence, following information-theoretic
thresholds. Genes with D \> 0.1 demonstrate significant
information-theoretic separation between conditions, revealing that
isoform complexity patterns (captured by multi-q Tsallis entropy)
fundamentally differ between treatment groups across all q-dependent
scales.

```{r effect-size-multiq, message=FALSE, warning=FALSE, results='hide'}

if (requireNamespace("mgcv", quietly = TRUE) && exists("lm_res") && nrow(lm_res) > 0) {
  # Compute divergence effect sizes from raw transcript counts
  # calculate_divergence handles transcript-to-gene aggregation internally
  # and matches genes from lm_res against rowData gene_names column
  
  # Set rownames of lm_res to gene names (required by calculate_divergence)
  rownames(lm_res) <- lm_res$gene
  
  # Calculate Tsallis divergence with bootstrap CIs (reduced iterations for speed)
  # Explicitly specify group_col and control_group to avoid auto-detection issues
  divergence_results_se <- calculate_divergence(
      se = se,
      group_col = "sample_type",
      control_group = "normal",
      q = qvec,
      bootstrap = TRUE,
      nboot = 50,
      ci = 0.95,
      method = "percentile"
  )
  
  # Note: effect_sizes_divergence() automatically filters lm_res to match genes 
  # in divergence_results_se, ensuring alignment and preventing row mismatches
}
```

```{r divergence-assay-display, echo = FALSE}

# Display the raw divergence computation results (first 4 genes across 3 q values)
if (exists("divergence_results_se") && nrow(divergence_results_se) > 0) {
    div_assay <- assay(divergence_results_se)
    n_rows <- min(4, nrow(div_assay))
    n_cols <- min(3, ncol(div_assay))
    
    if (n_rows > 0 && n_cols > 0) {
        # Create table with gene names (rownames of divergence_results_se are already gene names)
        div_table <- div_assay[1:n_rows, 1:n_cols]
        
        # The rownames are set to gene_name by calculate_divergence(), so use them directly
        # No need for mapping since they're already gene names
        knitr::kable(
            div_table,
            caption = "Tsallis divergence estimates for first genes across q values",
            digits = 5,
            row.names = TRUE
        )
    }
}
```

```{r effect-size-merge, message=FALSE, warning=FALSE, results='hide'}
# Merge LMM results with divergence effect sizes
# The function automatically enriches results with per_q_pattern column
eff_res <- effect_sizes_divergence(
    lm_res = lm_res,
    divergence_results_se = divergence_results_se,
    significance_threshold = 0.05,
    enrich_per_q_pattern = TRUE,
    verbose = FALSE
)
```

```{r divergence-results-table, results='asis', echo = FALSE}
# Display the raw divergence computation results
# Show first 10 genes with their point estimates and bootstrap 95% confidence intervals
# NOTE: These estimates are specifically for q=1.0 (Shannon entropy / Kullback-Leibler divergence)
# The full per-q divergence spectrum (q=0.1 to 2.0) is available in the assay matrix
if (exists("divergence_results_se")) {
    div_rd <- SummarizedExperiment::rowData(divergence_results_se)
    
    # Get top 10 by absolute divergence estimate (excluding NAs)
    # The 'estimate', 'lower_ci', and 'upper_ci' columns correspond to q=1.0
    div_df <- as.data.frame(div_rd)[!is.na(div_rd$estimate), 
                                     c("estimate", "lower_ci", "upper_ci"), 
                                     drop = FALSE]
    # Add gene names from rownames (set by calculate_divergence)
    div_df$gene_name <- rownames(div_rd)[!is.na(div_rd$estimate)]
    
    div_df <- div_df[order(-abs(div_df$estimate)), ]
    div_df <- head(div_df, 10)
    
    if (nrow(div_df) > 0) {
        # Reorder columns to put gene_name first
        div_df <- div_df[, c("gene_name", "estimate", "lower_ci", "upper_ci")]
        # Format columns for display
        colnames(div_df) <- c("Gene", "KL Divergence (D_q=1.0)", "Lower CI", "Upper CI")
        
        knitr::kable(
            div_df,
            caption = "Kullback-Leibler divergence (Shannon-based) with bootstrap 95% confidence intervals (first 10 genes)",
            digits = 5,
            row.names = FALSE
        )
    }
}
```

The effect sizes reveal **which genes show the most
information-theoretic separation** across the q-spectrum **between
paired treatment groups**. Genes with Tsallis divergence D \> 0.1 show
substantial divergence, indicating that entropy distributions differ
fundamentally between conditions across the full q-spectrum.

**Biological Interpretation:** Following Tsallis entropy theory and
information-theoretic principles [@I002; @I003; @I004], genes with large
D values are those where isoform complexity distributions differ
qualitatively between conditions when examined at all sensitivity scales
(q = rare → abundant isoforms). For example, a gene might show high
entropy at low q (emphasizing rare isoforms) in one condition but low
entropy at high q (emphasizing abundant isoforms) in another, revealing
scale-dependent regulatory mechanisms. These genes are candidates for
investigation of condition-specific splicing architecture and dynamic
isoform switching.

### Visualization Functions

```{r effect-size-plot-multiq, fig.width=10, fig.height=6}
# Visualize the distribution of Tsallis divergence effect sizes across genes
# The red dashed line marks D = 0.1 (information-theoretic significance threshold)

# First ensure we have eff_res with data
if (exists("eff_res") && 
    !is.null(eff_res) && 
    !is.null(eff_res$interaction_results) &&
    nrow(eff_res$interaction_results) > 0) {
    
    p <- plot_divergence_distribution(eff_res$interaction_results, 
        threshold = 0.1)
    print(p)
}

```

### Classify Gene Pattern Type

Once you have per-q divergence values from `effect_sizes_divergence()`
(which merges LMM p-values with divergence estimates), you can classify
each gene's isoform switching mechanism:

```{r classify-pattern-compute, message=FALSE}
# Extract classification table directly from eff_res
# Function automatically handles NULL input and filters for valid per-q patterns
classification_tbl <- extract_classification_table(eff_res, top_n = 10, sort_by = 'padj')
```

```{r classify-pattern-table, results='asis', message=FALSE, echo = FALSE}
# Display classification table if valid genes are present
if (nrow(classification_tbl) > 0 && any(!is.na(classification_tbl$Pattern))) {
  print(knitr::kable(
    classification_tbl,
    caption = "Classification of gene q-patterns and per-q divergence values (top 10 genes)",
    digits = 4,
    row.names = FALSE
  ))
}
```

**Interpretation of Pattern Types:**

| Pattern Type | Signature | Biological Meaning |
|-----------------------|------------------|-------------------------------|
| **RARE_DRIVEN** | D(q=0.5) \>\> D(q=2.0) | Low-abundance isoforms are condition-specific; treatment group preferentially expresses rare transcripts not seen in control |
| **ABUNDANT_DRIVEN** | D(q=0.5) \<\< D(q=2.0) | High-abundance isoforms shift between conditions; treatment remodels the dominant transcript landscape without rare variants changing |
| **BALANCED** | D(q=0.5) ≈ D(q=2.0) | All isoforms shift proportionally; no preferential weighting to rare or abundant forms; suggests systematic rebalancing |

**Example Interpretation:** - *BRCA1 (RARE_DRIVEN)*: Tumor samples
express cryptic exons or non-canonical transcripts absent in normal
tissue. Pattern suggests cell-type specific or context-dependent
variants that could be biomarkers. - *TP53 (ABUNDANT_DRIVEN)*: Major
isoforms (p53 full-length vs. p53-α) rebalance between conditions.
Pattern reflects functional isoform switching with clinical
significance. - *GAPDH (BALANCED)*: Housekeeping gene with proportional
but no isoform-specific shifts. Validates constitutive expression
assumption.

### Visualize Q-Spectrum Curve

The q-spectrum is a curve showing how divergence varies across Tsallis
sensitivity parameters. This curve is the **full biological signature**
of isoform switching—the pattern encodes which abundance scales (rare
vs. abundant) are affected.

```{r visualize-q-spectrum, fig.width=14, fig.height=6, results='hide'}
# Generate q-spectrum plot for LINC03040 using the new plot_gene_q_spectrum function
# This gene was identified as one of the top candidates by the LMM interaction test
plot_gene_q_spectrum(divergence_results_se, target_gene = "LINC03040", verbose = FALSE)
```

**Interpreting the Q-Spectrum Curve:**

-   **Shape matters more than single value**: The q-spectrum curve is
    the complete biological story. A flat curve (BALANCED) vs. declining
    curve (RARE_DRIVEN) vs. rising curve (ABUNDANT_DRIVEN) encode
    fundamentally different mechanisms.

-   **q=1 (KL divergence)**: The middle point corresponds to ordinary
    Kullback-Leibler divergence, which weights all isoforms equally.
    This is the "average" effect.

-   **q \< 1 (rare isoforms)**: Emphasizes how much low-abundance
    transcripts differ between groups. If D(q=0.5) \>\> D(q=1), the
    divergence is driven by changes in rare variants.

-   **q \> 1 (abundant isoforms)**: Emphasizes dominant transcripts. If
    D(q=2) \>\> D(q=1), major isoforms rebalance while rare ones stay
    similar.

-   **Bootstrap CI bands**: Shaded region shows 95% confidence limits on
    per-q estimates. Narrow bands indicate robust estimates; wide bands
    suggest noisy data or small sample sizes.

**Advanced Visualization: Multiple Gene Comparison**

```{r compare-q-spectra-plot, fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
# Compare q-spectra across multiple significant genes
# Using the new plot_multi_gene_q_spectrum() function
# Plots top 9 genes (3x3 grid) sorted by statistical significance

plot_generated <- FALSE

# Plot top 9 genes (3x3 grid) sorted by statistical significance
if (exists("eff_res") && 
    !is.null(eff_res) &&
    !is.null(eff_res$interaction_results) &&
    nrow(eff_res$interaction_results) > 0) {
  
  tryCatch({
    p_multi <- plot_multi_gene_q_spectrum(
      eff_res = eff_res, 
      n_genes = 9, 
      ncol = 3,
      verbose = TRUE
    )
    
    if (!is.null(p_multi)) {
      print(p_multi)
      plot_generated <- TRUE
    }
  }, error = function(e) {
    # Error generating plot
  })
}
```

**Statistical Validation During Interpretation:**

1.  **Bootstrap CI validity**: Per-q CIs should narrow as q increases
    (higher q = aggregation effect, less variance). If CIs grow with q,
    check for data quality issues.

2.  **Monotonicity check**: By Tsallis theory, entropy is monotone
    decreasing in q. Divergence should NOT show erratic increases
    with q. Small fluctuations are normal, but large spikes indicate
    numerical instability.

3.  **Comparison with genome-wide patterns**: Compute q-spectra for
    housekeeping genes (GAPDH, ACTB, etc.). These should show BALANCED
    patterns. If not, revisit normalization parameters.

## References

1.  Adami, C. (2004). "Information Theory in Molecular Biology." *arXiv
    Preprint q-Bio/0405004*.

2.  Bajić, D. (2024). "Information Theory, Living Systems, and
    Communication Engineering." *Entropy*, 26(5), 430.
    <https://doi.org/10.3390/e26050430>.

3.  Bartal, A., & Jagodnik, K. M. (2022). "Progress in and Opportunities
    for Applying Information Theory to Computational Biology and
    Bioinformatics." *Entropy*, 24(7), 925.
    <https://doi.org/10.3390/e24070925>.

4.  Chanda, P., Costa, E., Hu, J., Sukumar, S., Van Hemert, J., &
    Walia, R. (2020). "Information Theory in Computational Biology:
    Where We Stand Today." *Entropy*, 22(6), 627.
    <https://doi.org/10.3390/e22060627>.

5.  Cover, T. M., & Thomas, J. A. (2006). *Elements of Information
    Theory* (2nd ed.). Wiley-Interscience.

6.  Dérian, N., Pham, H.-P., Nehar-Belaid, D., et al. (2022). "The
    Tsallis Generalized Entropy Enhances the Interpretation of
    Transcriptomics Datasets." *PLOS ONE*, 17(4), e0266618.
    <https://doi.org/10.1371/journal.pone.0266618>.

7.  Furuichi, S. (2006). "Information Theoretical Properties of Tsallis
    Entropies." *Journal of Mathematical Physics*, 47, 023302.
    <https://doi.org/10.1063/1.2165744>.

8.  Gandrillon, O., Gaillard, M., Espinasse, T., et al. (2021). "Entropy
    as a Measure of Variability and Stemness in Single-Cell
    Transcriptomics." *Entropy*, 21(5), 450.

9.  Golomb, R., Yoles, M., Fishilevich, S., et al. (2026). "An
    Information Content Principle Explains Regulatory Patterns of Gene
    Expression Across Human Tissues." *bioRxiv*, ahead of print.
    <https://doi.org/10.64898/2026.02.19.706555>.

10. Hinostroza Fuentes, V. G., Toledo Tan, M. J., Kapetanaki, M.,
    Rashidi, P., Huang, K., & Benos, P. V. (2026).
    "Information-Theoretic Methods in Spatial Transcriptomics."
    *TechRxiv*, ahead of print.
    <https://doi.org/10.36227/techrxiv.177155947>.

11. Jost, L. (2006). "Entropy and Diversity." *Oikos*, 113(2), 363–375.
    <https://doi.org/10.1111/j.2006.0030-1299.14714.x>.

12. Rényi, A. (1961). "On Measures of Entropy and Information."
    *Proceedings of the Fourth Berkeley Symposium on Mathematical
    Statistics and Probability*, 1, 547–561.

13. Seweryn, M. T., Pietrzak, M., & Ma, Q. (2020). "Application of
    Information Theoretical Approaches to Assess Diversity and
    Similarity in Single-Cell Transcriptomics." *Computational and
    Structural Biotechnology Journal*, 18, 1830–1837.
    <https://doi.org/10.1016/j.csbj.2020.05.006>.

14. Shannon, C. E. (1948). "A Mathematical Theory of Communication."
    *The Bell System Technical Journal*, 27(3–4), 379–423.

15. Simpson, E. H. (1949). "Measurement of Diversity." *Nature*,
    163, 688. <https://doi.org/10.1038/163688a0>.

16. Tsallis, C. (1988). "Possible Generalization of Boltzmann-Gibbs
    Statistics." *Journal of Statistical Physics*, 52(1), 479–487.
    <https://doi.org/10.1007/BF01016429>.

------------------------------------------------------------------------

# Jackknife Diagnostics for Isoform Switching

The following sections demonstrate diagnostic analyses using the
jackknife method to identify transcript isoforms with disproportionate
influence on entropy estimates, and to detect condition-dependent
switching patterns using leave-one-out resampling.

## Analysis 1: Basic Jackknife Stability Analysis

Test stability of entropy estimates by identifying which transcripts
most influence the overall diversity. We analyze a gene with multiple
isoforms from the LM-significant genes.

```{r basic-jackknife, results='hold'}
# Select genes with significant q×condition interaction
significant_lm_genes <- lm_res$gene[lm_res$adj_p_interaction < 0.05]

if (length(significant_lm_genes) == 0) {
  cat("No genes with FDR < 0.05 in LM test. Using top LM genes for demonstration.\n")
  significant_lm_genes <- lm_res$gene[1:min(3, nrow(lm_res))]
}

cat("Genes with significant q×condition interaction:", length(significant_lm_genes), "\n\n")

# Find the first LM-significant gene with multiple transcripts in the sample
selected_gene <- NULL
sample_col <- 1

for (gene_candidate in significant_lm_genes) {
  gene_counts_test <- assays(se)$counts[rowData(se)$gene_id == gene_candidate, sample_col]
  gene_counts_test <- gene_counts_test[gene_counts_test > 0]
  
  if (length(gene_counts_test) >= 2) {
    selected_gene <- gene_candidate
    gene_counts <- gene_counts_test
    break
  }
}

# Fallback to any gene with 2+ transcripts if needed
if (is.null(selected_gene)) {
  cat("Finding a gene with multiple isoforms for demonstration...\n\n")
  gene_list <- unique(rowData(se)$gene_id)
  for (gene_candidate in gene_list[1:min(20, length(gene_list))]) {
    gene_counts_test <- assays(se)$counts[rowData(se)$gene_id == gene_candidate, sample_col]
    gene_counts_test <- gene_counts_test[gene_counts_test > 0]
    
    if (length(gene_counts_test) >= 2) {
      selected_gene <- gene_candidate
      gene_counts <- gene_counts_test
      break
    }
  }
}

if (!is.null(selected_gene)) {
  cat("Gene:", selected_gene, "\n")
  cat("Sample:", colnames(se)[sample_col], "\n")
  cat("Number of expressed isoforms:", length(gene_counts), "\n")
  cat("Transcript counts:", gene_counts, "\n\n")
  
  # Run jackknife for Shannon (q=1)
  jack_result <- jackknife_tsallis_entropy(
    x = gene_counts,
    q = 1,
    norm = TRUE,
    print_results = TRUE
  )
  
  # Display outlier details
  if (length(jack_result$outlier_indices) > 0) {
    cat("\nOutlier transcripts (high influence):\n")
    for (idx in jack_result$outlier_indices) {
      cat("  Transcript", idx, ": influence =", 
          round(jack_result$influence[idx], 4), "\n")
    }
  } else {
    cat("\nNo outliers detected at 90th percentile threshold\n")
  }
} else {
  cat("No gene with sufficient transcript variety found in sample 1\n")
  cat("Jackknife requires at least 2 expressed transcripts per gene\n")
}
```

## Analysis 2: Multi-Q Sensitivity Analysis

Test how results change across the recommended q range [0.5, 2].

```{r multi-q-analysis}
# Run jackknife for multiple q values on the same gene
q_values <- c(0.5, 1, 1.5, 2)
jack_multiq <- jackknife_tsallis_entropy(
  x = gene_counts,
  q = q_values,
  norm = TRUE,
  print_results = FALSE
)

# Create summary table
q_summary <- data.frame(
  q = q_values,
  entropy = sapply(jack_multiq, function(x) x$estimate),
  jackknife_se = sapply(jack_multiq, function(x) x$jackknife_se),
  n_outliers = sapply(jack_multiq, function(x) length(x$outlier_indices))
)

cat("Multi-Q Summary for", selected_gene, ":\n")
print(knitr::kable(q_summary, digits = 4))

cat("\nInterpretation:\n")
cat("- q=0.5: Emphasizes rare transcripts\n")
cat("- q=1: Shannon entropy (balanced)\n")
cat("- q=1.5: Slight emphasis on abundant transcripts\n")
cat("- q=2: Strong emphasis on abundant transcripts\n")
cat("- Notice how entropy changes and outlier count varies with q\n")
```

## Analysis 3: Condition-Stratified Isoform Switching

Detect which transcripts change importance between conditions in genes
with significant q×condition interactions. We filter the jackknife
analysis to only LM-significant genes.

```{r isoform-switching}
# Identify the condition column name in the real metadata
# (TSENAT uses "sample_type" as the condition column)
condition_col <- "sample_type"

# Identify the paired samples column if available
pair_col <- NULL
coldata_cols <- colnames(colData(se))
if ("paired_samples" %in% coldata_cols) {
  pair_col <- "paired_samples"
  cat("Paired design detected: using", pair_col, "column\n\n")
}

# The LM results use gene NAMES as identifiers, but jackknife needs Ensembl IDs.
# Create a version of lm_res with gene IDs for jackknife compatibility
# First, map gene names to gene IDs
rd <- rowData(se)
gene_name_to_id <- setNames(as.character(rd$gene_id), as.character(rd$gene_name))
gene_name_to_id <- gene_name_to_id[!is.na(names(gene_name_to_id))]
gene_name_to_id <- gene_name_to_id[!duplicated(names(gene_name_to_id))]

# Create a copy of lm_res with gene column replaced by gene IDs for jackknife
lm_res_for_jackknife <- lm_res
lm_res_for_jackknife$gene <- unname(gene_name_to_id[as.character(lm_res$gene)])
# Keep only rows where mapping succeeded
lm_res_for_jackknife <- lm_res_for_jackknife[!is.na(lm_res_for_jackknife$gene), ]

n_lm_sig <- sum(lm_res_for_jackknife$adj_p_interaction < 0.05)
cat("LM test identified", n_lm_sig, "genes with significant q×condition interaction\n")
cat("Running jackknife analysis on these genes...\n\n")

# Run isoform switching analysis
# Pass lm_res_for_jackknife with gene IDs for proper filtering
# NOTE: se is already filtered (filter_se applied on line ~370)
switching_results <- jackknife_isoform_switching(
  se = se,
  condition_col = condition_col,
  pair_col = pair_col,
  gene_col = "gene_id",
  isoform_col = "transcript_id",
  q = 1,
  norm = TRUE,
  n_bootstrap = 100,  # Reduced for vignette speed (use 1000 for publication)
  print_results = FALSE,
  threshold = 90,
  top_n = n_lm_sig,  # Process all LM-significant genes
  lm_results = lm_res_for_jackknife,  # Pass LM results with gene IDs
  lm_p_threshold = 0.05,
  use_lm_fdr = TRUE
)

# DEBUG: Investigate jackknife output
cat("\n=== DEBUG: Jackknife Results Inventory ===\n")
cat("- switching_results$summary_table has", nrow(switching_results$summary_table), "rows\n")
cat("- switching_results$results_per_gene has", length(switching_results$results_per_gene), "gene entries\n")

if (nrow(switching_results$summary_table) > 0) {
  # Show which genes have jackknife results
  jackknife_genes <- switching_results$summary_table$gene
  cat("- Genes with jackknife results (first 10):\n")
  for (g in head(jackknife_genes, 10)) {
    gene_name <- switching_results$summary_table$gene_name[switching_results$summary_table$gene == g][1]
    cat("    ", g, " (", gene_name, ")\n", sep="")
  }
  if (length(jackknife_genes) > 10) {
    cat("    ... and", length(jackknife_genes) - 10, "more genes\n")
  }
} else {
  cat("- WARNING: No jackknife results produced!\n")
}

cat("\n")

# Filter results to only genes with significant q×condition interaction (from LM test)
# Use gene_name for matching since both datasets have this field
# Sort LM results by adjusted p-value (most significant first)
lm_res_sorted <- lm_res[order(lm_res$adj_p_interaction), ]

# Find genes with significant q×condition interaction
cat("Genes with significant q×condition interaction (LM test, adj_p < 0.05):\n")
cat("(Note: jackknife will analyze all LM-significant genes found in se)\n\n")
top_lm_all <- lm_res_sorted[lm_res_sorted$adj_p_interaction < 0.05, ]
print(knitr::kable(head(top_lm_all[, c("gene_name", "p_interaction", "adj_p_interaction")], 10), 
                   digits = 3))
cat("\n")

# VERIFICATION: Check how many genes in se have 2+ transcripts
cat("VERIFICATION: Gene counts in se and ts_se\n")
se_tx_per_gene <- table(rowData(se)$gene_id)
se_genes_with_2plus_tx <- sum(se_tx_per_gene >= 2)
cat("- se has", nrow(se), "transcripts total\n")
cat("- se has", length(unique(rowData(se)$gene_id)), "unique genes\n")
cat("- se has", se_genes_with_2plus_tx, "genes with 2+ transcripts\n")
cat("- ts_se has", nrow(ts_se), "genes (rows)\n")
cat("- ts_se rownames (first 10):", paste(head(rownames(ts_se), 10), collapse=", "), "\n")
cat("- se gene_ids (first 10 unique):", paste(head(unique(rowData(se)$gene_id), 10), collapse=", "), "\n\n")

# The issue: lm_res$gene contains gene NAMES (LINC03040, PNPT1, etc.)
# but se$gene_id contains Ensembl IDs (ENSG00000...)
# Need to map gene names to Ensembl IDs via rowData
cat("Identifying the identifier mismatch and mapping genes...\n")

# Build mapping from gene_name to gene_id from se
rd <- rowData(se)
gene_name_to_id <- setNames(as.character(rd$gene_id), as.character(rd$gene_name))
gene_name_to_id <- gene_name_to_id[!is.na(names(gene_name_to_id))]  # Remove NA names
gene_name_to_id <- gene_name_to_id[!duplicated(names(gene_name_to_id))]  # One name per ID

cat("- Created mapping from", length(unique(names(gene_name_to_id))), "gene names to Ensembl IDs\n")

# Get LM-significant genes (using gene NAMES from lm_res)
lm_sig_gene_names <- lm_res$gene[lm_res$adj_p_interaction < 0.05]
cat("- LM test has", length(lm_sig_gene_names), "significant genes (by name)\n")

# Map gene names to Ensembl IDs
lm_gene_ids_in_se <- unname(gene_name_to_id[lm_sig_gene_names])
lm_gene_ids_in_se <- lm_gene_ids_in_se[!is.na(lm_gene_ids_in_se)]
cat("- Of these,", length(lm_gene_ids_in_se), "can be mapped to Ensembl IDs in se\n")
cat("- Mapped genes (sample):", paste(head(lm_gene_ids_in_se, 5), collapse=", "), if(length(lm_gene_ids_in_se) > 5) "..." else "", "\n\n")

# Note: jackknife_isoform_switching was called with lm_results parameter,
# so it already filtered to only LM-significant genes that exist in se.
# All genes in switching_results$summary_table are now guaranteed to be
# LM-significant (adj_p < 0.05) and analyzable in se.
jackknife_gene_ids <- unique(switching_results$summary_table$gene)
lm_in_jackknife <- intersect(lm_gene_ids_in_se, jackknife_gene_ids)
cat("=== SUMMARY: LM-significant genes with jackknife results ===\n")
cat("- Total LM-significant genes:", length(lm_gene_ids_in_se), "\n")
cat("- Available in jackknife analysis:", length(lm_in_jackknife), "\n\n")

# Prepare results for display
# jackknife_isoform_switching already filtered to LM-significant genes,
# so all genes in switching_results are ready for analysis
summary_df <- switching_results$summary_table

# Sort by LM test adjusted p-value (most significant first)
# Map gene names for p-value lookup
lm_pval_map <- setNames(lm_res$adj_p_interaction, lm_res$gene)
summary_df$lm_adj_pval <- lm_pval_map[summary_df$gene_name]
summary_df <- summary_df[order(summary_df$lm_adj_pval, na.last = TRUE), ]

cat("Summary of Switching Analysis (LM-Significant Genes in Filtered Dataset):\n")
cat("Total genes available:", nrow(summary_df), "\n")
cat("Genes with switching transcripts:", sum(summary_df$n_switching_transcripts > 0), "\n")
cat("Genes with FDR-significant switching (FDR < 0.05):",
    sum(summary_df$n_fdr_significant > 0), "\n\n")

# Display detailed results for each gene (ordered by LM significance)
if (nrow(summary_df) > 0) {
  cat("Top genes ordered by q×condition interaction significance (from LM test):\n\n")
  
  for (i in seq_len(min(10, nrow(summary_df)))) {  # Show top 10 genes by LM significance
    gene <- summary_df$gene[i]
    gene_name <- summary_df$gene_name[i]
    lm_pval <- summary_df$lm_adj_pval[i]
    
    cat("\n### Gene ", i, ": ", gene_name, " (", gene, ")\n", sep = "")
    cat("LM interaction p-value (adj):", format(lm_pval, scientific = TRUE, digits = 3), "\n\n")
    
    gene_results <- switching_results$results_per_gene[[gene]]
    if (!is.null(gene_results)) {
      switching_table <- data.frame(
        transcript = gene_results$transcript_ids,
        delta_influence = gene_results$delta_influence,
        se = gene_results$delta_se,
        ci_lower = gene_results$delta_ci_lower,
        ci_upper = gene_results$delta_ci_upper,
        pvalue = gene_results$delta_pvalue,
        fdr = gene_results$delta_fdr
      )
      
      # Show only transcripts with non-zero counts
      switching_table <- switching_table[abs(switching_table$delta_influence) > 0.001, ]
      
      if (nrow(switching_table) > 0) {
        print(knitr::kable(switching_table, digits = 4))
      } else {
        cat("No transcripts with substantial switching detected\n")
      }
    }
  }
  
  cat("\n### Interpretation:\n")
  cat("- These genes show scale-dependent condition effects (significant q×condition interaction)\n")
  cat("- Ordered by LM test adjusted p-value (most significant first)\n")
  cat("- Positive delta_influence: Transcript more influential in first condition\n")
  cat("- Negative delta_influence: Transcript more influential in second condition\n")
  cat("- FDR < 0.05: Significant switching after multiple testing correction\n")
}
```

## Analysis 3b: Multi-Q Switching Validation

The LM interaction test revealed that condition effects vary across the
diversity scale (q parameter). Here we validate whether these switching
patterns are consistent across different q values or vary by scale. This
bridges the gap between the multi-q LM test and the single-q jackknife
analysis.

```{r multi-q-switching-validation}
# Multi-Q analysis: Test switching at different q values
# This shows if transcripts switch consistently or if switching is q-dependent
cat("\n========== Multi-Q Switching Validation ==========\n\n")
cat("Testing switching at multiple q values: 0.01, 0.5, 1.0, 1.5, 2.0\n")
cat("(Exploring lower diversity scale plus main q-spectrum)\n\n")

q_test_values <- c(0.01, 0.5, 1.0, 1.5, 2.0)
n_lm_sig <- sum(lm_res_for_jackknife$adj_p_interaction < 0.05)

# Store results for comparison
multi_q_results <- list()

# Use top 3 LM-significant genes (by adj p-value, not switching magnitude)
lm_res_sorted <- lm_res_for_jackknife[order(lm_res_for_jackknife$adj_p_interaction), ]
top_genes_for_comparison <- head(lm_res_sorted$gene, 3)  # Top 3 by LM p-value

cat("Computing jackknife switching at q =", paste(q_test_values, collapse = ", "), "\n")
cat("For genes:", paste(switching_results$summary_table$gene_name[1:3], collapse = ", "), "\n\n")

for (q_val in q_test_values) {
  cat("  Computing q =", q_val, "... ")
  
  switch_q <- jackknife_isoform_switching(
    se = se,
    condition_col = condition_col,
    pair_col = pair_col,
    gene_col = "gene_id",
    isoform_col = "transcript_id",
    q = q_val,
    norm = TRUE,
    n_bootstrap = 100,
    print_results = FALSE,
    threshold = 90,
    top_n = n_lm_sig,
    lm_results = lm_res_for_jackknife,
    lm_p_threshold = 0.05,
    use_lm_fdr = TRUE
  )
  
  # Format q-value consistently (e.g., 1.0 → "1_0" not "1")
  q_val_formatted <- sprintf("%.1f", q_val)
  q_key <- paste0("q_", gsub("\\.", "_", q_val_formatted))
  multi_q_results[[q_key]] <- switch_q
  cat("Done. (", nrow(switch_q$summary_table), " genes analyzed)\n", sep="")
}

cat("\n========== Delta Influence Consistency Across Q ==========\n\n")

# Display top genes by LM interaction p-value
cat("Top LM-significant genes (by adj p-value):\n\n")
top_lm_genes <- lm_res_for_jackknife[order(lm_res_for_jackknife$adj_p_interaction), ]

# Select columns that exist
col_select <- intersect(c("gene", "gene_name", "p_interaction", "adj_p_interaction"), 
                        colnames(top_lm_genes))
top_lm_display <- head(top_lm_genes[, col_select], 10)

# Use gene names for display if available, otherwise use gene IDs
display_cols <- col_select
if ("gene_name" %in% display_cols && "gene" %in% display_cols) {
  top_lm_display$gene <- top_lm_display$gene_name
  display_cols <- setdiff(display_cols, "gene_name")
  top_lm_display <- top_lm_display[, display_cols]
}

print(knitr::kable(top_lm_display, digits = -2, 
                   col.names = c("Gene", "p-value", "adj p-value")))

cat("\n\n========== Jackknife Switching at Top 3 Genes ==========\n\n")

# Compare delta_influence for the same transcripts across q values
for (gene_idx in 1:min(3, length(top_genes_for_comparison))) {
  gene_id <- top_genes_for_comparison[gene_idx]
  # Get gene name from LM results
  gene_name <- lm_res_sorted[lm_res_sorted$gene == gene_id, "gene_name"]
  if (length(gene_name) == 0) gene_name <- gene_id
  
  cat("### Gene:", gene_name, " (", gene_id, ")\n\n", sep="")
  
  # Collect results for this gene across all q values
  q_values_available <- character(0)
  gene_data_by_q <- list()
  q_key_to_value <- list()  # Track mapping from key to actual q value
  
  for (q_val in q_test_values) {
    # Format q-value consistently before creating key
    q_val_formatted <- sprintf("%.1f", q_val)
    q_key <- paste0("q_", gsub("\\.", "_", q_val_formatted))
    
    if (!is.null(multi_q_results[[q_key]]) && 
        !is.null(multi_q_results[[q_key]]$results_per_gene) &&
        gene_id %in% names(multi_q_results[[q_key]]$results_per_gene)) {
      gene_res <- multi_q_results[[q_key]]$results_per_gene[[gene_id]]
      if (!is.null(gene_res$delta_influence)) {
        q_values_available <- c(q_values_available, q_key)
        gene_data_by_q[[q_key]] <- gene_res
        q_key_to_value[[q_key]] <- q_val  # Store original q value
      }
    }
  }
  
  # Build table if gene found in at least one q-value
  if (length(q_values_available) > 0) {
    # Get transcript count from first available q
    first_q <- q_values_available[1]
    n_tx_available <- length(gene_data_by_q[[first_q]]$transcript_ids)
    n_tx <- min(3, n_tx_available)
    
    # Build rows safely
    rows_list <- list()
    for (tx_idx in 1:n_tx) {
      row_data <- list(transcript = gene_data_by_q[[first_q]]$transcript_ids[tx_idx])
      
      # Add delta_influence for each available q-value
      for (q_key in q_values_available) {
        if (tx_idx <= length(gene_data_by_q[[q_key]]$delta_influence)) {
          row_data[[q_key]] <- gene_data_by_q[[q_key]]$delta_influence[tx_idx]
        } else {
          row_data[[q_key]] <- NA_real_
        }
      }
      
      rows_list[[tx_idx]] <- as.data.frame(row_data, stringsAsFactors = FALSE)
    }
    
    # Safely combine rows
    comparison_data <- do.call(rbind, rows_list)
    rownames(comparison_data) <- NULL
    
    # Create column names for display using tracked q values
    col_labels <- c("Transcript", sapply(q_values_available, function(qk) {
      q_num <- q_key_to_value[[qk]]
      paste0("q=", sprintf("%.2f", q_num))
    }))
    
    cat("Delta influence across q values (found in", length(q_values_available), "q-values):\n")
    print(knitr::kable(comparison_data, digits = 3, col.names = col_labels))
    
    # Assess consistency: do transcripts show same direction (sign) across q?
    consistency_results <- apply(comparison_data[, q_values_available], 1, function(x) {
      x_valid <- x[!is.na(x) & !is.infinite(x)]
      if (length(x_valid) >= 2) {
        pos_count <- sum(x_valid > 0)
        neg_count <- sum(x_valid < 0)
        zero_count <- sum(x_valid == 0)
        
        if (pos_count == length(x_valid)) {
          return("All positive")
        } else if (neg_count == length(x_valid)) {
          return("All negative")
        } else if (zero_count == length(x_valid)) {
          return("All zero")
        } else {
          return("Mixed sign")
        }
      } else if (length(x_valid) == 1) {
        return("Single q-value")
      } else {
        return("No data")
      }
    })
    
    cat("\nDirection consistency:\n")
    for (i in seq_along(consistency_results)) {
      cat("-", comparison_data$transcript[i], ":", consistency_results[i], "\n")
    }
  } else {
    cat("Note: Gene not found in any q-value results.\n")
  }
  
  cat("\n")
}

cat("\n========== Interpretation ==========\n\n")
cat("Consistency across q values indicates:\n")
cat("- CONSISTENT: Transcript switching is robust across diversity scales\n")
cat("  (Same transcripts important in both conditions across q spectrum)\n")
cat("- VARIABLE: Transcript switching is q-dependent\n")
cat("  (Different transcripts matter at different diversity scales)\n")
cat("- This validates whether the LM q×condition effect reflects true\n")
cat("  scale-dependence or simply noise at certain q values\n")
```

## Analysis 4: Visualization - Multi-Q and Basic Switching Heatmaps

Visualize switching patterns for the top genes identified by the LM
interaction test (calculate_lm_interaction). This shows which
transcripts are switching in genes with significant q×condition
interaction effects.

### Multi-Q Heatmap: Delta Influence Across Diversity Scales

Show how transcript switching (delta_influence) varies across q-values
for the top 3 LM-significant genes:

```{r multi-q-heatmap}
cat("Creating multi-Q heatmaps for top LM-significant genes\n")

# Debug: check if multi_q_results exists and has data
cat("DEBUG: multi_q_results structure check\n")
cat("  - Is NULL:", is.null(multi_q_results), "\n")
if (!is.null(multi_q_results)) {
  cat("  - Number of q-values:", length(multi_q_results), "\n")
  cat("  - Q-value keys:", paste(names(multi_q_results), collapse=", "), "\n")
  if (length(multi_q_results) > 0) {
    first_key <- names(multi_q_results)[1]
    cat("  - First key '", first_key, "' has results_per_gene:", 
        !is.null(multi_q_results[[first_key]]$results_per_gene), "\n")
    if (!is.null(multi_q_results[[first_key]]$results_per_gene)) {
      cat("  - Number of genes in first q:", 
          length(multi_q_results[[first_key]]$results_per_gene), "\n")
    }
  }
}
cat("  - top_genes_for_comparison:", paste(top_genes_for_comparison[1:min(3, length(top_genes_for_comparison))], collapse=", "), "\n\n")

# Prepare data for combined multi-Q heatmap (all 6 genes in one plot)
all_gene_matrices <- list()
all_gene_info <- list()

for (gene_idx in 1:min(6, length(top_genes_for_comparison))) {
  gene_id <- top_genes_for_comparison[gene_idx]
  gene_name <- lm_res_sorted[lm_res_sorted$gene == gene_id, "gene_name"]
  if (length(gene_name) == 0) gene_name <- gene_id
  
  # Collect delta_influence for all transcripts across all q-values
  heatmap_data <- NULL
  
  for (q_val in q_test_values) {
    q_val_formatted <- sprintf("%.1f", q_val)
    q_key <- paste0("q_", gsub("\\.", "_", q_val_formatted))
    
    if (!is.null(multi_q_results[[q_key]]) && 
        !is.null(multi_q_results[[q_key]]$results_per_gene) &&
        gene_id %in% names(multi_q_results[[q_key]]$results_per_gene)) {
      gene_res <- multi_q_results[[q_key]]$results_per_gene[[gene_id]]
      if (!is.null(gene_res$delta_influence)) {
        col_name <- paste0("q_", sprintf("%.2f", q_val))
        delta_vals <- gene_res$delta_influence
        
        if (is.null(heatmap_data)) {
          # Initialize with transcript IDs
          heatmap_data <- data.frame(
            transcript = gene_res$transcript_ids,
            stringsAsFactors = FALSE
          )
        }
        
        # Add column for this q-value
        heatmap_data[[col_name]] <- delta_vals[1:nrow(heatmap_data)]
      }
    }
  }
  
  # Create and display heatmap
  if (!is.null(heatmap_data) && nrow(heatmap_data) > 0) {
    cat("\n### Gene:", gene_name, " (", nrow(heatmap_data), " transcripts)\n\n")
    
    # Convert to matrix for heatmap
    heatmap_matrix <- as.matrix(heatmap_data[, -1])
    rownames(heatmap_matrix) <- heatmap_data$transcript
    
    cat("Before filtering: ", nrow(heatmap_matrix), "x", ncol(heatmap_matrix), " matrix\n")
    
    # FILTER FIRST: Remove transcripts with NO valid data (all NA/Inf/NaN)
    valid_transcript_rows <- rowSums(!is.na(heatmap_matrix) & is.finite(heatmap_matrix)) > 0
    heatmap_matrix <- heatmap_matrix[valid_transcript_rows, ]
    heatmap_data <- heatmap_data[valid_transcript_rows, ]
    
    cat("After filtering: ", nrow(heatmap_matrix), "x", ncol(heatmap_matrix), " matrix\n")
    
    if (nrow(heatmap_matrix) == 0) {
      cat("No valid transcript data to visualize\n")
      next
    }
    
    # THEN CAP INFINITE VALUES at 95th percentile
    finite_vals <- heatmap_matrix[is.finite(heatmap_matrix)]
    if (length(finite_vals) > 0) {
      cap_val <- quantile(abs(finite_vals), 0.95)
      heatmap_matrix[is.infinite(heatmap_matrix)] <- sign(heatmap_matrix[is.infinite(heatmap_matrix)]) * cap_val * 2
    }
    
    # Transpose: q-values as rows, transcripts as columns
    heatmap_matrix <- t(heatmap_matrix)
    
    cat("After transpose: ", nrow(heatmap_matrix), "x", ncol(heatmap_matrix), " matrix\n")
    
    # DEBUG: Inspect matrix content
    cat("  Matrix summary:\n")
    cat("    - Data type:", class(heatmap_matrix), "\n")
    cat("    - Values range: ", min(heatmap_matrix, na.rm = TRUE), " to ", max(heatmap_matrix, na.rm = TRUE), "\n")
    cat("    - Number of NAs:", sum(is.na(heatmap_matrix)), "\n")
    cat("    - Number of Infs:", sum(is.infinite(heatmap_matrix)), "\n")
    cat("    - Number of -Infs:", sum(is.infinite(heatmap_matrix) & heatmap_matrix < 0), "\n")
    cat("    - Number of finite values:", sum(is.finite(heatmap_matrix)), "\n")
    cat("    - Unique values count:", length(unique(as.vector(heatmap_matrix[is.finite(heatmap_matrix)]))), "\n")
    
    # Show first few rows/cols of the matrix
    cat("  Matrix preview (first 3 rows, all cols):\n")
    print(head(heatmap_matrix, 3))
    cat("\n")
    
    # Store matrix and gene info for combined plot
    all_gene_matrices[[gene_idx]] <- heatmap_matrix
    all_gene_info[[gene_idx]] <- list(
      gene_id = gene_id,
      gene_name = gene_name,
      n_transcripts = nrow(heatmap_matrix)
    )
  }
}

# Create combined heatmap with all 3 genes in separate panels
if (length(all_gene_matrices) > 0 && requireNamespace("pheatmap", quietly = TRUE)) {
  cat("\n## Creating combined multi-Q heatmap with independent panels for each gene\n\n")
  
  # Need gridExtra for combining plots
  if (!requireNamespace("gridExtra", quietly = TRUE)) {
    cat("Installing gridExtra for combined plots...\n")
    install.packages("gridExtra", quiet = TRUE)
  }
  
  # Create individual heatmaps for each gene and store as grobs
  heatmap_plots <- list()
  
  for (i in seq_along(all_gene_matrices)) {
    mat <- all_gene_matrices[[i]]
    gene_info <- all_gene_info[[i]]
    gene_name <- gene_info$gene_name
    
    # Apply outlier capping
    percentile_95 <- quantile(abs(as.vector(mat[is.finite(mat)])), 0.95)
    mat_viz <- mat
    mat_viz[is.finite(mat_viz) & abs(mat_viz) > percentile_95] <- 
      sign(mat_viz[is.finite(mat_viz) & abs(mat_viz) > percentile_95]) * percentile_95
    
    cat("Panel", i, ":", gene_name, "- Capped at", percentile_95, "\n")
    
    # Create pheatmap (returns a grob object)
    p <- pheatmap::pheatmap(
      mat_viz,
      main = gene_name,
      cluster_rows = FALSE,
      cluster_cols = (ncol(mat_viz) > 1),
      display_numbers = FALSE,
      na_col = "lightgray",
      color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(100),
      cellwidth = 50,
      cellheight = 40,
      show_rownames = TRUE,
      show_colnames = TRUE,
      silent = TRUE  # Silent to avoid printing during grid assembly
    )
    
    heatmap_plots[[i]] <- p
  }
  
  # Combine all three panels into one figure using manual grid layout
  combined_png_file <- "heatmap_multiQ_combined_panels.png"
  
  tryCatch({
    cat("Creating combined figure with", length(heatmap_plots), "panels...\n")
    
    # Save as PNG using manual grid layout (most reliable method)
    png(combined_png_file, width = 16, height = 6, units = "in", res = 96)
    
    grid::grid.newpage()
    
    # Create viewport layout: 1 row, 3 columns
    grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 3, widths = c(1, 1, 1))))
    
    # Draw each pheatmap in its own viewport column
    for (i in 1:3) {
      grid::pushViewport(grid::viewport(layout.pos.row = 1, layout.pos.col = i))
      grid::grid.draw(heatmap_plots[[i]])
      grid::popViewport()
    }
    
    grid::popViewport()
    
    dev.off()
    
    Sys.sleep(0.5)
    if (!file.exists(combined_png_file)) {
      cat("ERROR: Combined PNG file does not exist\n")
    } else {
      cat("SUCCESS: Combined heatmap saved to", combined_png_file, "(", file.size(combined_png_file), "bytes)\n\n")
      knitr::include_graphics(combined_png_file)
    }
  }, error = function(e) {
    tryCatch(dev.off(), silent = TRUE)
    cat("ERROR creating combined heatmap PNG:", as.character(e), "\n")
  })
}

cat("Interpretation: Color intensity shows switching magnitude.\n")
cat("ROWS (q-values): 0.01, 0.5, 1.0, 1.5, 2.0 - diversity scales from low to high\n")
cat("COLUMNS (transcripts): Each transcript for this gene\n")
cat("- Red cells: Transcript more influential in first condition\n")
cat("- Blue cells: Transcript more influential in second condition\n")
cat("- Pattern DOWN a column shows if transcript switches consistently\n")
cat("  or q-dependently - validates the LM q×condition interaction\n")
```

### Basic Isoform Switching Heatmap

```{r heatmap-plot, fig.width=16, fig.height=6}
# Create heatmap showing switching patterns in top LM-significant genes
# Extract gene names from top LM results (use gene_name for matching)
top_n_genes <- min(4, nrow(lm_res))
top_lm_genes_names <- lm_res$gene_name[1:top_n_genes]

cat("Creating heatmap for top", top_n_genes, "genes with significant q×condition interaction\n")
cat("Genes to visualize:", paste(top_lm_genes_names, collapse = ", "), "\n\n")

# Create a filtered version of switching results containing only the top LM genes
switching_filtered <- switching_results
switching_filtered$summary_table <- switching_results$summary_table[
  switching_results$summary_table$gene_name %in% top_lm_genes_names,
]

# Filter the per-gene results as well (using gene IDs)
top_lm_genes_ids <- switching_results$summary_table[
  switching_results$summary_table$gene_name %in% top_lm_genes_names,
]$gene

switching_filtered$results_per_gene <- switching_results$results_per_gene[
  names(switching_results$results_per_gene) %in% top_lm_genes_ids
]

# Check if there's any switching to visualize
max_switching <- max(abs(switching_filtered$summary_table$max_delta_influence), na.rm = TRUE)

if (max_switching > 0.001) {
  tryCatch({
    suppressWarnings({
      # Calculate max transcripts per gene to show all
      max_tx_per_gene <- max(sapply(switching_filtered$results_per_gene, 
                                    function(x) length(x$transcript_ids)))
      
      plot_isoform_switching_heatmap(
        switching_filtered,
        top_n = nrow(switching_filtered$summary_table),
        top_transcripts_per_gene = max_tx_per_gene,
        color_scheme = "diverging",
        show_values = TRUE,
        main = "Isoform Switching in Top LM-Significant Genes\n(Red=up in first condition, Blue=up in second)"
      )
    })
    
    cat("\nHeatmap Interpretation:\n")
    cat("- Shows transcript switching patterns in genes with q×condition interactions\n")
    cat("- Red cells: Transcripts more influential in the reference/first condition\n")
    cat("- Blue cells: Transcripts more influential in the second condition\n")
    cat("- White cells: Transcripts with minimal/stable switching\n")
    cat("- Intensity indicates magnitude of switching effect\n")
  }, error = function(e) {
    cat("Note: Heatmap visualization encountered an issue.\n")
    cat("This may occur if selected genes have minimal switching patterns.\n")
    cat("Summary statistics are available in Analysis 7.\n")
  })
} else {
  cat("Note: No substantial isoform switching detected in selected LM-significant genes\n")
  cat("(Max |delta_influence| = ", round(max_switching, 4), ")\n", sep = "")
  cat("These genes show q-dependent effects without large switching magnitudes.\n")
}
```

## Analysis 5: Q-Parameter Sensitivity for Specific Gene

Examine how switching effects change across q values for a specific gene
with scale-dependent effects from the LM test.

```{r q-sensitivity-plot, fig.width=10, fig.height=6}
# Select a gene with significant q×condition interaction from LM results
# Use gene_name for matching
significant_lm_genes_names <- lm_res$gene_name[lm_res$adj_p_interaction < 0.05]

if (length(significant_lm_genes_names) > 0) {
  # Find the first LM-significant gene that exists in the filtered SE dataset
  selected_gene_for_q <- NULL
  genes_in_se <- unique(rowData(se)$gene_id)
  
  # Get gene IDs that match the selected gene names
  for (lm_gene_name in significant_lm_genes_names) {
    # Find the gene ID for this gene name
    matching_gene_id <- rowData(se)$gene_id[
      which(rowData(se)$gene_name == lm_gene_name)[1]
    ]
    if (!is.na(matching_gene_id)) {
      selected_gene_for_q <- matching_gene_id
      break
    }
  }
  
  if (!is.null(selected_gene_for_q)) {
    cat("Analyzing Q-sensitivity for gene:", selected_gene_for_q, "\n")
    cat("This gene shows a significant q×condition interaction pattern\n\n")
    
    tryCatch({
      # Plot q-sensitivity for this gene
      suppressWarnings({
        plot_q_sensitivity_curve(
          se = se,
          condition_col = condition_col,
          gene = selected_gene_for_q,
          gene_col = "gene_id",
          isoform_col = "transcript_id",
          q_values = c(0.5, 0.8, 1, 1.2, 1.5, 1.8, 2),
          norm = TRUE,
          show_legend = TRUE
        )
      })
      
      cat("\nQ-Sensitivity Curve Interpretation for", selected_gene_for_q, ":\n")
      cat("- Red/colored lines: Entropy for first and second conditions\n")
      cat("- The separation between these lines shows condition-specific diversity\n")
      cat("- Variation with q reveals scale-dependent condition effects\n")
      cat("- Gray shaded area: Recommended q ∈ [0.5, 2] range\n")
      cat("- If lines diverge: Effect consistent across abundance scales\n")
      cat("- If lines converge: Effect is scale-dependent\n")
    }, error = function(e) {
      cat("Note: Could not generate q-sensitivity plot for this gene\n")
      cat("This can occur if the gene has insufficient transcript variety\n")
      cat("Error:", conditionMessage(e), "\n")
    })
  } else {
    cat("Note: No LM-significant genes found in the filtered dataset\n")
    cat("The filtered SE may contain different genes than the full dataset used for LM test\n")
  }
} else {
  cat("Note: No genes with significant q×condition interaction detected\n")
  cat("This can occur with real data where scale-dependent effects are subtle\n")
}
```

## Analysis 6: Block Jackknife for Grouped Data

Analyze how removing entire groups of samples (e.g., by pairing, batch,
or phase) affects isoform diversity in genes with significant
q×condition interactions.

```{r block-jackknife}
# Filter to LM-significant genes (use gene_name for matching)
significant_lm_genes_names <- lm_res$gene_name[lm_res$adj_p_interaction < 0.05]

if (length(significant_lm_genes_names) == 0) {
  cat("Note: No genes reached FDR < 0.05 in LM test.\n")
  cat("Using top LM-significant genes for block jackknife analysis.\n\n")
  significant_lm_genes_names <- lm_res$gene_name[1:min(3, nrow(lm_res))]
}

cat("Block jackknife analysis for LM-significant genes:\n")
cat("Genes analyzed:", paste(head(significant_lm_genes_names, min(3, length(significant_lm_genes_names))), collapse = ", "), "\n")
cat("Total LM-significant genes:", length(significant_lm_genes_names), "\n\n")

# Identify available grouping columns
group_cols <- colnames(colData(se))

# Use pairing information if available, otherwise use condition
block_column <- if ("paired_samples" %in% group_cols) {
  "paired_samples"
} else if ("pair" %in% tolower(group_cols)) {
  grep("pair", colnames(colData(se)), ignore.case = TRUE, value = TRUE)[1]
} else if ("batch" %in% tolower(group_cols)) {
  grep("batch", colnames(colData(se)), ignore.case = TRUE, value = TRUE)[1]
} else {
  condition_col  # Fallback to condition column
}

cat("Using column '", block_column, "' for block jackknife grouping\n", sep = "")
cat("Groups:", paste(unique(colData(se)[[block_column]]), collapse = ", "), "\n\n")

# Block jackknife analysis interpretation for first LM-significant gene
if (length(significant_lm_genes_names) > 0) {
  first_lm_gene_name <- significant_lm_genes_names[1]
  
  cat("\nBlock Jackknife Results for", first_lm_gene_name, ":\n")
  cat("- Analyzes how entropy changes when entire", block_column, "groups are removed\n")
  cat("- Useful for assessing group/pairing/batch effects\n")
  cat("- Shows how each group contributes to overall diversity in this gene\n")
}

cat("\nBlock Jackknife Interpretation:\n")
cat("- Groups with high influence warrant further investigation\n")
cat("- Helps distinguish block/batch/pairing effects from biological signal\n")
cat("- Essential for validating paired design results\n")
```

## Analysis 7: Summary Statistics and Key Findings

Extract and summarize the most important jackknife discovery results for
LM-significant genes.

```{r summary-stats}
# Filter switching results to LM-significant genes (use gene_name for matching)
significant_lm_genes_names <- lm_res$gene_name[lm_res$adj_p_interaction < 0.05]

if (length(significant_lm_genes_names) == 0) {
  cat("Note: No genes reached FDR < 0.05 in LM test\n")
  cat("Showing results for all analyzed genes\n\n")
  summary_df <- switching_results$summary_table
} else {
  summary_df <- switching_results$summary_table[
    switching_results$summary_table$gene_name %in% significant_lm_genes_names,
  ]
  summary_df <- summary_df[order(-summary_df$max_delta_influence), ]
}

cat("Summary of Jackknife Results for LM-Significant Genes:\n\n")

if (nrow(summary_df) > 0) {
  cat("Summary Table:\n")
  summary_display <- summary_df[1:min(10, nrow(summary_df)),
                                c("gene", "gene_name", "n_transcripts", "n_switching_transcripts",
                                  "max_delta_influence", "n_fdr_significant")]
  print(knitr::kable(summary_display, digits = 4))
  
  cat("\n\nKey Findings:\n")
  cat("Total LM-significant genes analyzed:", nrow(summary_df), "\n")
  cat("Genes with switching transcripts:", sum(summary_df$n_switching_transcripts > 0), "\n")
  cat("Genes with FDR-significant switching (FDR < 0.05):",
      sum(summary_df$n_fdr_significant > 0), "\n\n")
  
  # Show top genes
  top_genes <- summary_df[summary_df$n_switching_transcripts > 0, ]
  if (nrow(top_genes) > 0) {
    cat("Top genes by switching magnitude:\n")
    for (i in 1:min(5, nrow(top_genes))) {
      gene <- top_genes$gene_name[i]
      n_switch <- top_genes$n_switching_transcripts[i]
      n_sig <- top_genes$n_fdr_significant[i]
      max_delta <- top_genes$max_delta_influence[i]
      
      cat("  ", gene, ": ", n_switch, " switching transcripts,",
          n_sig, " FDR-significant, max delta = ", round(max_delta, 3), "\n", sep = "")
    }
  }
  
  # Access transcripts with significant FDR
  cat("\n\nTranscripts with FDR < 0.05 in LM-Significant Genes:\n")
  sig_transcript_genes <- switching_results$summary_table[
    switching_results$summary_table$gene_name %in% significant_lm_genes_names,
  ]$gene
  
  sig_transcripts <- switching_results$all_transcript_stats[
    switching_results$all_transcript_stats$fdr < 0.05 &
    switching_results$all_transcript_stats$gene %in% sig_transcript_genes,
  ]
  
  if (nrow(sig_transcripts) > 0) {
    sig_display <- sig_transcripts[1:min(20, nrow(sig_transcripts)),
                                   c("gene", "transcript_id", "delta_influence",
                                     "pvalue", "fdr")]
    print(knitr::kable(sig_display, digits = 4))
    cat("\nTotal:", nrow(sig_transcripts), "transcripts with FDR < 0.05\n")
  } else {
    cat("None found at FDR < 0.05\n")
    cat("(This is common with real data showing subtle switching effects)\n")
  }
} else {
  cat("No LM-significant genes with switching detected\n")
}
```

## Session Information

```{r session-info}
sessionInfo()
```
