---
title: "Appendix B: Non-Parametric Validation of Linear Model Results via GAM and Rank-Based Methods"
author: "Cristobal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
  rmarkdown::pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
    latex_engine: xelatex
header-includes:
  - \usepackage{float}
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: Non-Parametric Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-setup-appendix-b, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
suppressPackageStartupMessages(library(kableExtra))
```

# Overview: Non-Parametric Validation of Entropic Index × Group Interactions (q-values)

## Purpose and Rationale

Primary Goal: Validate that discoveries of scale-dependent entropic index × group interactions from linear models generalize to non-parametric statistical frameworks with minimal assumptions.

TSENAT's default approach employs **linear models** to test for entropic index × group effects in Tsallis entropy. While computationally efficient and well-characterized, linear models assume:
- Normally distributed residuals (reasonable after appropriate transformation)
- Homogeneous variance across entropic indices (often violated in diversity metrics)
- Additive effects of group and entropic index on entropy (may oversimplify nonlinear relationships)

This appendix provides two independent, non-parametric alternatives that:

1. Make minimal distributional assumptions, relying only on ranks or data-adaptive smoothing
2. Explicitly handle the ordered structure of entropic indices, treating them as sequential measurements (q-values)
3. Adapt to real-world heteroscedasticity, automatically weighting measurements by entropy variance
4. Provide independent validation, allowing comparison of linear vs. non-parametric findings

## Two Complementary Validation Approaches

**Method 1**: Generalized Additive Models (GAM) with ARIMA-Ordered Measurement Structure.

- Distributional assumption: None required; uses non-parametric basis functions (thin-plate splines).
- Treatment of entropic structure: Treats q-values as time-like ordered measurements, applies ARIMA(1,1,0) differencing for autocorrelation.
- Heteroscedasticity handling: Automatically detects variance heterogeneity; applies optimal weighting.
- Key advantage: Captures smooth nonlinear q * group effects with computational efficiency.

**Method 2**: Rank-Based Tests (Scheirer-Ray-Hare with Hochberg Step-Up Correction).

- Distributional assumption: None; operates entirely on ranks (maximal robustness).
- Treatment of q-ordering: Scheirer-Ray-Hare two-way test for paired measurements across q and group.
- Heteroscedasticity handling: Rank transformation inherently robust; Hochberg correction controls FWER.
- Key advantage: Maximally robust to outliers; valid for any continuous distribution.


## Why Two Methods for One Question?

Statistical testing in transcriptomics faces a fundamental challenge: no single method is universally optimal. Different approaches make different assumptions and have different strengths. TSENAT employs two complementary strategies: parametric methods (linear models) and non-parametric rank-based tests:

| Aspect | Parametric (Linear Model / GAM) | Non-Parametric (Rank-Based) |
|--------|--------------------------------|----------------------------|
| Assumptions | Normality, homoscedasticity | Ranks only; fully non-parametric |
| Power | Highest (if assumptions met) | Good; slightly reduced but robust |
| Robustness | Moderate; sensitive to outliers | Highest; resistant to outliers and outlier-driven effects |
| Interpretation | Parametric effect sizes, smooth curves | Effect ranks; robust p-values independent of distribution |
| Outlier influence | High potential for bias | Minimal; rank transformation inherently resistant |

Validation strategy: High concordance between parametric and non-parametric methods  confirms that findings are robust to modeling assumptions and generalize across statistical frameworks.

***

# Setup

This section initializes the analysis environment by loading required packages, setting a random seed for reproducibility, and preparing the test dataset.

```{r load-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)

# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(readcounts)
mode(readcounts) <- "numeric"

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Configure analysis parameters first (best practice: fail-fast validation)
config <- TSENAT_config(
  sample_col = "sample",
  condition_col = "condition",
  subject_col = "paired_samples",
  q = seq(0, 2, by = 0.05),
  nthreads = 3,
  paired = TRUE,
  control = "normal"
)

# Build analysis object: creates SummarizedExperiment and initializes TSENATAnalysis
analysis <- build_analysis(
    readcounts = readcounts,
    tx2gene = gff3_dataset,
    metadata = metadata_df,
    tpm = tpm,
    effective_length = effective_length,
    config = config
)

# Apply filtering for quality control
analysis <- filter_analysis(
    analysis,
    stringency = "medium",
    min_isoform_abundance = 0
)
```

## Computing Tsallis Entropy with Pseudocount Regularization

Pseudocounts are critical for statistical robustness when applying rank-based tests to sparse RNA-seq count data. RNA-seq experiments typically contain many zero or near-zero counts, which creates two problems for rank-based nonparametric methods:

1. **Ties in rankings:** Sparse counts produce many tied values (especially zeros), which reduces the discriminatory power of rank tests. Rank-based statistics depend on unique orderings; when many observations are identical, the test statistic contains less information, reducing statistical power.

2. **Library size artifacts:** Small count differences due to sequencing depth variations overshadow true biological differences. Normalized library sizes (accounting for sequencing depth) ensure that entropy estimates reflect true biological diversity rather than technical artifacts [@S195; @S069].

By adding small pseudocounts proportional to library size, we achieve two benefits:

- **Regularization** breaks ties and prevents zero-inflation bias that compromises rank-based inference.
- **Normalization** makes entropy estimates comparable across samples with different sequencing depths. This approach is standard in RNA-seq analysis and is particularly important for entropy-based diversity metrics, which require positive values for logarithmic transforms.

The `calculate_diversity()` function implements automatic pseudocount selection, scaling pseudocount magnitude and diversity estimates by median library size to ensure robust statistical inference across the range of sequencing depths in your dataset.

```{r compute-tsallis-entropy, message=FALSE, results='hide'}
# Compute diversity using S4 wrapper with bootstrap confidence intervals
analysis <- calculate_diversity(
  analysis, 
  norm = TRUE,
  pseudocount = "auto"
)
```

# Rank-Based Approach (Scheirer-Ray-Hare)

The Scheirer-Ray-Hare test provides a non-parametric validation of linear model results by leveraging the paired nature of the experimental design and the ordered structure of entropic indices. Unlike parametric methods that assume normality and homogeneity of variance, the Scheirer-Ray-Hare test operates solely on ranks, making it maximally robust to outliers and extreme values.

## Assumption Validation

Before applying rank-based methods, we verify key assumptions for rank-based inference.

```{r validate-srh-assumptions, message=FALSE}
# Validate Scheirer-Ray-Hare test assumptions
analysis <- calculate_rank_assumptions(
    analysis
)

# Extract the result object
rank_assumptions <- metadata(analysis, "rankbased_assumptions")$result
```

```{r display-assumption-results, message=FALSE, echo=FALSE}
# Extract summary statistics and test results from rank_assumptions object
summary_stats <- attr(rank_assumptions, "summary_stats")
check_results <- attr(rank_assumptions, "checks")

# Create entropy data summary table
entropy_summary <- data.frame(
  Metric = c(
    "Genes tested",
    "Samples (entropic order indices × groups)",
    "Entropy range",
    "Mean entropy",
    "Median entropy",
    "Missing values"
  ),
  Value = c(
    summary_stats$n_genes,
    summary_stats$n_samples,
    sprintf("%.4f to %.4f", summary_stats$entropy_min, summary_stats$entropy_max),
    sprintf("%.4f", summary_stats$entropy_mean),
    sprintf("%.4f", summary_stats$entropy_median),
    summary_stats$n_missing
  )
)

knitr::kable(entropy_summary, 
             caption = "Supplementary Table 1 | Rank-Based Entropy Matrix Properties.",
             label = "tab:rank-entropy-summary",
             align = c("l", "r"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Data from complete entropy matrix across all q-values and samples.")

cat("\n")

# Create assumption checks table
exchangeability_check <- check_results$exchangeability
monotonicity_check <- check_results$monotonicity
consistency_check <- check_results$consistency

data_characteristics <- data.frame(
  Characteristic = c("Paired Structure", "Gene Heterogeneity", "Subject Consistency"),
  Test = c("Permutation test", "Spearman r", "Kendall's W"),
  Result = c(
    sprintf("p=%.3f", exchangeability_check$p_value),
    sprintf("r=%.3f", monotonicity_check$mean_correlation),
    sprintf("W=%.3f", consistency_check$kendall_w)
  )
)

knitr::kable(data_characteristics, 
             caption = "Supplementary Table 2 | Entropy Matrix Validation Metrics. Three core properties validating suitability for rank-based analysis. Rows: Paired Structure (exchangeability test ensures independence), Gene Heterogeneity (Spearman rank correlation), Subject Consistency (Kendall's *W* concordance). Values reported as *P*-values, correlation coefficients *r*, and concordance *W* (0-1 range). All tests support exchangeability assumption justifying Scheirer-Ray-Hare nonparametric approach.",
             align = c("l", "l", "r", "l"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "All metrics support validity of paired-design rank-based inference.")
```

## Scheirer-Ray-Hare Test: Testing q $\times$ Condition Interactions

Now we will use the Scheirer-Ray-Hare test to evaluate whether entropy patterns across entropic indices (q-values) differ between normal and tumor samples. The Scheirer-Ray-Hare test is a non-parametric alternative to the paired t-test and repeated measures ANOVA, making it ideal for paired designs where distributional assumptions may be violated [@S030].

```{r srh-rank-test-analysis, message=FALSE, results='hide'}
# Run Scheirer-Ray-Hare test for q * condition INTERACTION
analysis <- calculate_rank_test(
    analysis,
    multicorr = "hochberg"
)

# Extract results
srh_results <- results(analysis, type = "rank_test", rankBy = "pvalue", n = 20, filterFDR = 0.05)
print(head(srh_results, n = 10))
```

```{r srh-results-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(srh_results$effect_size_eta2, na.rm = TRUE)
median_effect <- median(srh_results$effect_size_eta2, na.rm = TRUE)

rank_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Significant (adj_p < 0.05, FWER-controlled)", 
    "NAs",
    "Mean effect size ($\\eta^2$)",
    "Median effect size ($\\eta^2$)",
    "Strong effect genes ($\\eta^2$ > 10%)"
  ),
  Value = c(
    nrow(srh_results),
    sum(srh_results$p_value < 0.05, na.rm = TRUE),
    sum(srh_results$adj_p_value < 0.05, na.rm = TRUE),
    sum(is.na(srh_results$p_value)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(srh_results$p_value < 0.05 & srh_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(rank_summary_stats, caption = "Supplementary Table 3 | Scheirer-Ray-Hare Test Results Summary. Rank-based nonparametric test of *q* × condition interactions.", label = "tab:srh-test-summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Scheirer-Ray-Hare test: nonparametric rank-based ANOVA for paired designs.")

# Show top genes results
rank_sig <- srh_results[!is.na(srh_results$p_value), ]
if (nrow(rank_sig) > 0) {
  rank_sig <- rank_sig[order(rank_sig$adj_p_value), ]
  rank_sig <- rank_sig[1:min(10, nrow(rank_sig)), ]
  
  # Select columns: gene, p-values, f-statistic, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "f_statistic", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(rank_sig))
  if (length(cols_to_show) > 0) {
    rank_display <- rank_sig[, cols_to_show, drop = FALSE]
    
    # Format p-values with scientific notation for very small values
    if ("p_value" %in% colnames(rank_display)) {
      rank_display$p_value <- sapply(rank_display$p_value, function(x) {
        if (is.na(x)) "NA" else if (x < 0.0001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
      })
    }
    
    if ("adj_p_value" %in% colnames(rank_display)) {
      rank_display$adj_p_value <- sapply(rank_display$adj_p_value, function(x) {
        if (is.na(x)) "NA" else if (x < 0.0001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
      })
    }
    
    # Replace test method name for clarity
    if ("test_method" %in% colnames(rank_display)) {
      rank_display$test_method <- sapply(rank_display$test_method, function(x) {
        if (is.na(x)) "NA" else if (x == "srh_paired") "Scheirer-Ray-Hare (paired)" else x
      })
    }
    
    # Format other numeric columns
    if ("f_statistic" %in% colnames(rank_display)) {
      rank_display$f_statistic <- sapply(rank_display$f_statistic, function(x) {
        if (is.na(x)) "NA" else sprintf("%.4f", x)
      })
    }
    
    if ("effect_size_eta2" %in% colnames(rank_display)) {
      rank_display$effect_size_eta2 <- sapply(rank_display$effect_size_eta2, function(x) {
        if (is.na(x)) "NA" else sprintf("%.4f", x)
      })
    }
    
    knitr::kable(rank_display, 
                 caption = "Supplementary Table 4 | Top genes identified by Scheirer-Ray-Hare rank-based test. Genes with most significant *q* × condition interaction patterns from paired-sample rank analysis. Columns: gene identifier; *P*-value (unadjusted, scientific notation for p < 0.0001); adjusted *P*-value (Benjamini-Hochberg correction, scientific notation for p < 0.0001); Scheirer-Ray-Hare *F*-statistic; effect size ($\\eta^2$); test method; interaction class. Ranked by adjusted significance.",
                 col.names = c("Gene", "P-value", "Adj. P-value", "F-Statistic", 
                               "Effect Size (η²)", "Test Method", "Interaction Class")[1:ncol(rank_display)],
                 format = if (knitr::is_latex_output()) "latex" else "html",
                 longtable = knitr::is_latex_output(),
                 linesep = "") %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      latex_options = c("scale_down")
    ) %>%
    kableExtra::footnote(general = "Effect size η² quantifies strength of *q* × condition interaction; top 10 genes ranked by significance. P-values < 0.0001 shown in scientific notation for precision.")
  }
}
```

Visualize q-curves for top genes:

```{r extended-fig-1-srh-q-curves, fig.width=12, fig.height=8, message=FALSE, out.width="98%", fig.pos="H", fig.cap="Extended Data Figure 1 | Scheirer-Ray-Hare rank test results for scale-dependent isoform switching. X-axis: *q* (0-2), Y-axis: Tsallis entropy (0-1). Parallel curves indicate consistent condition effects across diversity scales; diverging curves reveal *q*-dependent switching patterns."}
top_genes_plot <- plot_diversity_spectrum(analysis, lm_res = srh_results, n_top = 4)
print(top_genes_plot)
```


***

# Generalized Additive Model Approach (GAM)

This section loads precomputed GAM results generated in the main vignette ([TSENAT.Rmd](TSENAT.Rmd)).

```{r generalized-additive-model-analysis, message=FALSE}
# Load precomputed LM analysis object from RDS file
analysis_lm <- readRDS(
    system.file("extdata", "analysis_lm.rds", package = "TSENAT")
)

# Extract GAM/LM results for inspection using accessor function
gam_results <- results(analysis_lm, type = "lm", rankBy = "pvalue")

```

```{r gam-results-summary-table, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Aggregate statistics across all genes tested by GAM method; effect size > 30% indicates strong biological signal.")

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format numeric columns
      numeric_cols <- sapply(gam_display, is.numeric)
      gam_display[, numeric_cols] <- round(gam_display[, numeric_cols], 3)
      
      # Format p-value columns with scientific notation for very small values
      if ("p_interaction" %in% colnames(gam_display)) {
        gam_display$p_interaction <- sapply(gam_display$p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
        })
      }
      
      if ("adj_p_interaction" %in% colnames(gam_display)) {
        gam_display$adj_p_interaction <- sapply(gam_display$adj_p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
        })
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      # Rename columns for better readability
      colnames(gam_display) <- c("Gene", "p (interaction)", "Adjusted p", "Effect size", "Test statistic", "df")
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size and Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"),
                   format = if (knitr::is_latex_output()) "latex" else "html",
                   longtable = knitr::is_latex_output(),
                   linesep = "") %>%
      kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        latex_options = c("scale_down")
      ) %>%
      kableExtra::footnote(general = "Top 10 genes ranked by significance; p-values < 0.001 shown in scientific notation; effect size range 0-100%.")
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

***

# Method Concordance Analysis

A critical validation step is to compare whether the Scheirer-Ray-Hare rank-based test and GAM produce consistent results. High concordance between methods (i.e., genes significant in both approaches) provides strong evidence that discoveries are robust to methodological choice. Discordant genes—those significant in only one method—warrant closer inspection: they may represent genuine biological signals revealed by one method's specific advantages, or artifacts of that method's assumptions. This section quantifies agreement between approaches and identifies high-confidence genes significant across both statistical frameworks.

```{r compare-srh-gam-methods, message=FALSE}
# Compute concordance analysis using new two-object API
# Compares LM results (from analysis_lm) with rank test results (from analysis)
analysis_with_concordance <- calculate_concordance(
    analysis_lm = analysis_lm,
    analysis_rank = analysis,
    lm_method = "lm_interaction",
    rank_method = "rank_test",
    verbose = TRUE
)

# Extract components from S4 results
concordance_result <- metadata(analysis_with_concordance, "method_concordance")
comparison_df <- concordance_result$comparison_df
spearman_rho <- concordance_result$spearman_rho
high_conf <- concordance_result$high_confidence
agreement_table <- concordance_result$agreement_table
```

```{r compute-concordance-metrics, message=FALSE, echo=FALSE}
# === GLOBAL CONCORDANCE METRICS ===
total_genes <- nrow(comparison_df)
both_sig <- sum(comparison_df$agreement == "Both significant")
lm_only <- sum(comparison_df$agreement == "LM only")
rank_only <- sum(comparison_df$agreement == "Rank test only")
neither_sig <- sum(comparison_df$agreement == "Neither significant")

concordance_metrics <- data.frame(
  Metric = c(
    "Total genes compared",
    "Spearman correlation (p-values)",
    "Both methods significant (p < 0.05)",
    "LM only significant",
    "Rank test only significant",
    "Neither significant",
    "Concordance rate",
    "Discordance rate"
  ),
  Value = c(
    sprintf("%d", total_genes),
    sprintf("rho = %.4f", spearman_rho),
    sprintf("%d (%.1f%%)", both_sig, 100 * both_sig / total_genes),
    sprintf("%d (%.1f%%)", lm_only, 100 * lm_only / total_genes),
    sprintf("%d (%.1f%%)", rank_only, 100 * rank_only / total_genes),
    sprintf("%d (%.1f%%)", neither_sig, 100 * neither_sig / total_genes),
    sprintf("%.1f%%", 100 * (both_sig + neither_sig) / total_genes),
    sprintf("%.1f%%", 100 * (lm_only + rank_only) / total_genes)
  )
)

knitr::kable(concordance_metrics, 
             caption = "Global Concordance Metrics: GAM vs Scheirer-Ray-Hare Methods",
             align = c("l", "r"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Comparison of significant genes (p < 0.05) detected by GAM vs Scheirer-Ray-Hare methods; high concordance validates robustness.")
```

```{r display-agreement-results, message=FALSE, echo=FALSE}
# === AGREEMENT CATEGORY BREAKDOWN ===
if (!is.null(agreement_table) && length(agreement_table) > 0) {
  agreement_breakdown <- data.frame(
    Category = names(agreement_table),
    Count = as.numeric(agreement_table),
    Percentage = sprintf("%.1f%%", 100 * as.numeric(agreement_table) / sum(agreement_table)),
    stringsAsFactors = FALSE
  )
  
  colnames(agreement_breakdown) <- c("Agreement Category", "Number of Genes", "Percentage")
  
  knitr::kable(agreement_breakdown,
               caption = "Method Agreement Distribution",
               align = c("l", "r", "r"),
               format = if (knitr::is_latex_output()) "latex" else "html",
               longtable = knitr::is_latex_output(),
               linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Categories: Concordant (both methods, p < 0.05); GAM-only; Scheirer-Ray-Hare-only; Neither (both p ≥ 0.05).")
} else {
  cat("No agreement table available from concordance analysis.")
}
```

```{r high-confidence-calls, message=FALSE, echo=FALSE}
# === HIGH-CONFIDENCE GENES (BOTH METHODS SIGNIFICANT) ===
if (!is.null(high_conf) && nrow(high_conf) > 0) {
  # Sort by LM adjusted p-value, then by Rank adjusted p-value (both ascending)
  high_conf_sorted <- high_conf[order(high_conf$padj_lm, high_conf$padj_rank), ]
  
  hc_display <- high_conf_sorted[1:min(15, nrow(high_conf_sorted)), c("gene", "padj_lm", "padj_rank", "effect_lm", "effect_rank")]
  
  hc_display$padj_lm <- sprintf("%.3e", hc_display$padj_lm)
  hc_display$padj_rank <- sprintf("%.3e", hc_display$padj_rank)
  hc_display$effect_lm <- sprintf("%.1f%%", hc_display$effect_lm * 100)
  hc_display$effect_rank <- sprintf("%.3f", hc_display$effect_rank)
  
  colnames(hc_display) <- c("Gene", "LM adj p", "Rank test adj p", "LM Effect", "Rank test $\\eta^2$")
  rownames(hc_display) <- NULL
  
  knitr::kable(hc_display,
               format = if (knitr::is_latex_output()) "latex" else "html",
               longtable = knitr::is_latex_output(),
               linesep = "",
               caption = sprintf("Robust Entropic Order Index Interactions: High-Confidence Genes Detected by Both Methods (n=%d, ranked by statistical significance)", nrow(high_conf)),
               align = c("l", "r", "r", "r", "r")) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "High-confidence genes: significant by both GAM and Scheirer-Ray-Hare rank test (adj p < 0.05); $\\eta^2$ = rank test effect size.")
} else {
  cat("No high-confidence genes detected by both methods (p < 0.05 in both GAM and Scheirer-Ray-Hare tests).")
}
```

## Statistical Power vs. Robustness: Understanding Method Discordance

The striking discordance between GAM and Scheirer-Ray-Hare results (71.1% significant in GAM only, 0% in rank test only, 28.9% total concordance) reflects a fundamental trade-off in statistical methodology: **parametric methods maximize power when assumptions hold, while nonparametric methods sacrifice power for robustness to assumption violations**.

GAM's superior power derives from two factors:

1. Distributional assumptions: GAM assumes approximately normal residuals and homogeneous variance, which are reasonable after appropriate transformation for entropy data. Under these assumptions, parametric methods are theoretically optimal, achieving maximum power for a given type I error rate. The rank-based Scheirer-Ray-Hare test is assumption-free but necessarily discards quantitative information by converting measurements to ranks, which reduces statistical power when the underlying data are approximately normal.

2. Model flexibility with penalty: GAM uses thin-plate splines with smoothness penalties that simultaneously fit nonlinear patterns while controlling degrees of freedom. This flexibility allows GAM to detect subtle entropic index × group interactions across all q-values. The Scheirer-Ray-Hare test, by contrast, operates on rank patterns only, which is inherently less sensitive to continuous relationships across the ordering (q-values).

This complementary approach—combining high-power parametric tests with robust nonparametric alternatives—provides confidence that discoveries are methodologically sound rather than artifacts of statistical assumptions.

## Visualization: Method Comparison

Visual comparison of concordance patterns provides intuitive assessment of which genes align between methods and which are method-specific. The following plots display significance calls, effect sizes, and agreement metrics in a format that facilitates interpretation of both robust discoveries and method-specific findings.

```{r visualize-method-concordance, fig.width=10, fig.height=8, out.width="98%", fig.pos="H", fig.cap="Method concordance visualization comparing Scheirer-Ray-Hare rank-based and GAM approaches for detecting q × condition interactions. Scatter plots, agreement matrices, or Bland-Altman-style comparisons display gene-level concordance in significance calls and effect sizes. Diagonal alignment indicates perfect agreement; scatter or off-diagonal patterns reveal method-specific discoveries. Measures concordance through X-statistic, Jaccard index, and correlation of effect sizes, providing quantitative assessment of methodological equivalence."}
# Create comparison plot using S4 wrapper
# Check if concordance analysis was successful
concordance_results <- metadata(analysis_with_concordance, "method_concordance")

p <- plot_concordance(analysis_with_concordance, verbose = TRUE)
print(p)
```

***

# Session Information

```{r session-info-appendix-b}
sessionInfo()
```
