---
title: "TSENAT Appendix B: Validating Results with Rank-Based Methods"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Parametric and Rank-Based Methods

## Purpose

This appendix validates the `calculate_lm_interaction()` parametric results (with improvements: conditional family selection, heteroscedasticity detection, bounded support handling) by comparing with rank-based (non-parametric) alternatives. 

**Key questions:**
1. Do parametric (GAM with improvements) and rank-based methods identify the same significant genes?
2. Does conditional family selection actually change results compared to baseline approaches?
3. Which genes are truly robust (significant in BOTH methods)?

**Expected differences:**
- **GAM with improvements** includes: conditional family selection (Gaussian/Gamma), heteroscedasticity correction, bounded support handling
- **Kruskal-Wallis (rank-based)** includes: no distributional assumptions, but also no special handling for bounded entropy

If results agree, confidence increases. If GAM identifies additional genes, it indicates family selection improved detection. If they diverge strongly, investigate data characteristics.

## Why This Matters

- **Parametric methods** (GAM/LMM) are efficient but assume approximate normality and homoscedasticity
- **Rank-based methods** require only exchangeability—a much weaker assumption
- **Agreement across methods** validates findings without invoking questionable assumptions
- **Genes significant in BOTH approaches** = highest confidence results

## Validation Strategy

1. Run parametric approach: `calculate_lm_interaction()` with GAM
2. Run rank-based alternative: Kruskal-Wallis test (non-parametric)
3. Compare significant genes and effect sizes
4. Identify robust genes (significant in both methods)

---

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Create SummarizedExperiment
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
se <- filter_se(se, stringency = "medium")

# Calculate Tsallis entropy
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(se, q = qvec, norm = TRUE, metadata = metadata_df)

cat("Data Summary:\n")
cat("  Genes:", nrow(ts_se), "\n")
cat("  Unique samples:", length(unique(colData(ts_se)$samples)), "\n")
cat("  Q-values:", length(unique(colData(ts_se)$q)), "\n")
cat("  Total observations:", ncol(ts_se), "\n")
```

---

# Part 1: Parametric Approach (GAM)

## Run interaction test

```{r gam-analysis, message=FALSE}
# Run parametric pipeline (FULL DATASET for consistency with main vignette)
lm_res <- calculate_lm_interaction(
    ts_se,  # Use all genes for consistent comparison
    method = "gam",
    paired = TRUE,
    multicorr = "hochberg",
    nthreads = 3,
    verbose = FALSE
)

cat("GAM Results Summary:\n")
cat("  Genes tested:", nrow(lm_res), "\n")
cat("  Significant (p < 0.05):", sum(lm_res$p_interaction < 0.05), "\n")
cat("  Significant (adjusted p < 0.05):", sum(lm_res$adj_p_interaction < 0.05), "\n")

# Top genes
cat("\n\nTop 10 Genes by Parametric P-Value:\n")
top_gam <- head(lm_res[order(lm_res$p_interaction), ], 10)
print(knitr::kable(
    top_gam[, c("gene", "p_interaction", "adj_p_interaction", "family_used", "n_observations")],
    caption = "Parametric (GAM) - Top Significant Genes",
    digits = 6
))
```

---

# Part 2: Rank-Based Approach (Kruskal-Wallis)

## Non-parametric alternative

```{r rankbased-analysis, message=FALSE}
# Prepare data for rank-based testing (FULL DATASET for consistency)
entropy_matrix <- assay(ts_se)  # Use all genes
ts_coldata <- colData(ts_se)

# Create long-format data
test_genes <- rownames(ts_se)  # Use all genes for consistent comparison
n_genes <- length(test_genes)
n_cols <- ncol(ts_se)

entropy_long <- data.frame(
    entropy = as.numeric(entropy_matrix),
    gene = rep(test_genes, n_cols),
    q_value = rep(ts_coldata$q, each = n_genes),
    sample_type = rep(ts_coldata$sample_type, each = n_genes)
)

# Kruskal-Wallis test for each gene
# Tests: Does entropy differ by sample_type?
kw_results <- list()

for (gene in test_genes) {
    gene_data <- entropy_long[entropy_long$gene == gene, ]
    
    kw_test <- kruskal.test(entropy ~ sample_type, data = gene_data)
    
    kw_results[[gene]] <- data.frame(
        gene = gene,
        kw_statistic = kw_test$statistic,
        kw_p_value = kw_test$p.value,
        n_obs = nrow(gene_data)
    )
}

kw_res <- do.call(rbind, kw_results)
rownames(kw_res) <- NULL

cat("Kruskal-Wallis Results Summary:\n")
cat("  Genes tested:", nrow(kw_res), "\n")
cat("  Significant (p < 0.05):", sum(kw_res$kw_p_value < 0.05), "\n")

# Top genes
cat("\n\nTop 10 Genes by Rank-Based P-Value:\n")
top_kw <- head(kw_res[order(kw_res$kw_p_value), ], 10)
print(knitr::kable(
    top_kw,
    caption = "Rank-Based (Kruskal-Wallis) - Top Significant Genes",
    digits = 6
))
```

---

# Part 3: Comparing Methods

## Concordance

```{r concordance-analysis, message=FALSE}
# Merge results
comparison <- merge(
    lm_res[, c("gene", "p_interaction", "adj_p_interaction")],
    kw_res[, c("gene", "kw_p_value")],
    by = "gene"
)

comparison <- comparison[order(comparison$p_interaction), ]

# Correlation
p_corr <- cor(comparison$p_interaction, comparison$kw_p_value, 
              use = "complete.obs", method = "spearman")

cat("=== METHODOLOGICAL CONCORDANCE ===\n\n")
cat("Spearman correlation of p-values:", sprintf("%.4f\n", p_corr))

if (p_corr > 0.85) {
    cat("✓ HIGH CONCORDANCE: Methods strongly agree\n")
    cat("  Conclusion: Results are ROBUST across different statistical assumptions\n")
} else if (p_corr > 0.70) {
    cat("✓ MODERATE CONCORDANCE: Methods generally agree\n")
    cat("  Conclusion: Some variation expected; core findings consistent\n")
} else {
    cat("⚠ LOW CONCORDANCE: Methods disagree substantially\n")
    cat("  Recommendation: Investigate data characteristics\n")
}

# Agreement at p < 0.05
both_sig <- sum(comparison$p_interaction < 0.05 & comparison$kw_p_value < 0.05)
gam_only <- sum(comparison$p_interaction < 0.05 & comparison$kw_p_value >= 0.05)
kw_only <- sum(comparison$p_interaction >= 0.05 & comparison$kw_p_value < 0.05)
neither <- sum(comparison$p_interaction >= 0.05 & comparison$kw_p_value >= 0.05)

cat("\n\nThreshold Agreement (p < 0.05):\n")
cat(sprintf("  Both methods agree SIGNIFICANT:     %3d genes\n", both_sig))
cat(sprintf("  GAM only significant:               %3d genes\n", gam_only))
cat(sprintf("  Rank-based only significant:        %3d genes\n", kw_only))
cat(sprintf("  Both methods agree NOT significant: %3d genes\n", neither))

pct_agreement <- 100 * (both_sig + neither) / nrow(comparison)
cat(sprintf("\n  Overall agreement rate: %.1f%%\n", pct_agreement))
```

## Side-by-Side Comparison

```{r comparison-table}
cat("\n\nDetailed Comparison (first 15 genes):\n")
print(knitr::kable(
    head(comparison, 15),
    caption = "Parametric vs. Rank-Based P-Values - Direct Comparison",
    digits = 6
))
```

---

# Part 4: Robust Gene List

## High-Confidence Findings

```{r robust-genes, message=FALSE}
# Genes significant in BOTH methods
robust_genes <- comparison[
    comparison$p_interaction < 0.05 & 
    comparison$kw_p_value < 0.05, 
]

cat("=== ROBUST GENE LIST ===\n")
cat("Genes with strong evidence in BOTH methods\n")
cat("(Highest confidence for publication)\n\n")

if (nrow(robust_genes) > 0) {
    robust_display <- robust_genes[order(robust_genes$p_interaction), ]
    
    print(knitr::kable(
        head(robust_display, 15),
        caption = sprintf("High-Confidence Genes (%d total significant in both methods)", nrow(robust_genes)),
        digits = 6
    ))
    
    cat("\n\nInterpretation:\n")
    cat("  • These genes show q×group interactions with robust evidence\n")
    cat("  • Results validated across TWO independent statistical methods\n")
    cat("  • Recommended for detailed biological interpretation\n")
} else {
    cat("  (No genes significant at p < 0.05 in both methods)\n")
}
```

---

# Part 5: Visualization

## P-Value Scatter Plot

```{r scatter-plot, fig.width=10, fig.height=8}
plot_data <- comparison
plot_data$log10_gam <- -log10(pmax(plot_data$p_interaction, 1e-300))
plot_data$log10_kw <- -log10(pmax(plot_data$kw_p_value, 1e-300))
plot_data$agreement <- case_when(
    plot_data$p_interaction < 0.05 & plot_data$kw_p_value < 0.05 ~ "Both Significant",
    plot_data$p_interaction < 0.05 ~ "GAM Only",
    plot_data$kw_p_value < 0.05 ~ "KW Only",
    TRUE ~ "Neither"
)

p1 <- ggplot(plot_data, aes(x = log10_gam, y = log10_kw, 
                            color = agreement, size = agreement)) +
    geom_point(alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50", linewidth = 1) +
    geom_vline(xintercept = -log10(0.05), linetype = "dotted", color = "red", alpha = 0.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "red", alpha = 0.5) +
    scale_color_manual(
        values = c("Both Significant" = "#2ecc71", "GAM Only" = "#3498db", 
                  "KW Only" = "#e74c3c", "Neither" = "#bdc3c7")
    ) +
    scale_size_manual(values = c("Both Significant" = 5, "GAM Only" = 3, 
                                 "KW Only" = 3, "Neither" = 2)) +
    labs(
        title = "Parametric (GAM) vs. Rank-Based (Kruskal-Wallis) Comparison",
        x = "-log10(GAM p-value)",
        y = "-log10(Kruskal-Wallis p-value)",
        color = "Method Agreement",
        size = "Method Agreement"
    ) +
    theme_minimal() +
    theme(legend.position = "right", plot.title = element_text(face = "bold"))

print(p1)
```

## P-Value Distributions

```{r distributions, fig.width=12, fig.height=5}
p2 <- ggplot(plot_data, aes(x = log10_gam)) +
    geom_histogram(fill = "#3498db", alpha = 0.7, bins = 20) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "red", linewidth = 1) +
    labs(title = "GAM P-Value Distribution",
         x = "-log10(p-value)", y = "Number of Genes") +
    theme_minimal()

p3 <- ggplot(plot_data, aes(x = log10_kw)) +
    geom_histogram(fill = "#e74c3c", alpha = 0.7, bins = 20) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "red", linewidth = 1) +
    labs(title = "Kruskal-Wallis P-Value Distribution",
         x = "-log10(p-value)", y = "Number of Genes") +
    theme_minimal()

gridExtra::grid.arrange(p2, p3, ncol = 2)
```

---

# Part 6: Recommendations

## Publication Strategy

### When Methods Strongly Agree (r > 0.85)

✓ **Confidence Level**: VERY HIGH  
✓ **Recommendation**: Results ready for publication  
✓ **Presentation**:
  - Report both methods in main text
  - Highlight robust genes (significant in both)
  - Include comparison table in supplementary materials

### When Methods Show Moderate Agreement (r = 0.70−0.85)

✓ **Confidence Level**: MODERATE TO HIGH  
✓ **Recommendation**: Good support; investigate discordant genes  
✓ **Presentation**:
  - Lead with parametric results (main analysis)
  - Report rank-based as sensitivity analysis
  - Focus publication on robust gene list

### When Methods Disagree (r < 0.70)

⚠ **Confidence Level**: LOWER  
⚠ **Recommendation**: Investigate before publication  
⚠ **Diagnostic checks**:
  - Are there influential outliers?
  - Non-normal distributions?
  - Heteroscedasticity issues?
  - Sample size or power problems?

## Best Practices

```{r best-practice, eval=FALSE}
# Recommended workflow:

# 1. Run parametric method (primary analysis)
lm_res <- calculate_lm_interaction(ts_se, method = "gam", paired = TRUE)

# 2. Run rank-based validation (sensitivity check)  
kw_res <- kruskal_wallis_comparison(entropy_data)

# 3. Calculate concordance
correlation <- cor(lm_res$p_value, kw_res$p_value, method = "spearman")

# 4. If high agreement:
#    Report robust genes (significant in both methods)
#    Include both methods in supplementary table

# 5. If low agreement:
#    Investigate outliers and data quality
#    Consider data transformations
#    Visualize residuals and distributions
```

---

# Session Information

```{r session-info}
sessionInfo()
```
