---
title: "Appendix B: Non-Parametric Validation of Linear Model Results via GAM and Rank-Based Methods"
author: "Cristobal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
  rmarkdown::pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
    latex_engine: xelatex
header-includes:
  - \usepackage{float}
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: Non-Parametric Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-setup-appendix-b, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
suppressPackageStartupMessages(library(kableExtra))
```

# Overview: Non-Parametric Validation of Entropic Index × Group Interactions (q-values)

## Purpose and Rationale

Primary Goal: Validate that discoveries of scale-dependent entropic index × group interactions from linear models generalize to non-parametric statistical frameworks with minimal assumptions.

TSENAT's default approach employs **linear models** to test for entropic index × group effects in Tsallis entropy. While computationally efficient and well-characterized, linear models assume:
- Normally distributed residuals (reasonable after appropriate transformation)
- Homogeneous variance across entropic indices (often violated in diversity metrics)
- Additive effects of group and entropic index on entropy (may oversimplify nonlinear relationships)

This appendix provides two independent, non-parametric alternatives that:

1. Make minimal distributional assumptions, relying only on ranks or data-adaptive smoothing
2. Explicitly handle the ordered structure of entropic indices, treating them as sequential measurements (q-values)
3. Adapt to real-world heteroscedasticity, automatically weighting measurements by entropy variance
4. Provide independent validation, allowing comparison of linear vs. non-parametric findings

## Two Complementary Validation Approaches

### Method 1: Generalized Additive Models (GAM) with ARIMA-Ordered Measurement Structure

- Distributional assumption: None required; uses non-parametric basis functions (thin-plate splines).
- Treatment of entropic structure: Treats q-values as time-like ordered measurements, applies ARIMA(1,1,0) differencing for autocorrelation.
- Heteroscedasticity handling: Automatically detects variance heterogeneity; applies optimal weighting.
- Key advantage: Captures smooth nonlinear q * group effects with computational efficiency.

### Method 2: Rank-Based Tests (Friedman with Hochberg Step-Up Correction)

- Distributional assumption: None; operates entirely on ranks (maximal robustness).
- Treatment of q-ordering: Friedman two-way test for paired measurements across q and group.
- Heteroscedasticity handling: Rank transformation inherently robust; Hochberg correction controls FWER.
- Key advantage: Maximally robust to outliers; valid for any continuous distribution.


## Why Two Methods for One Question?

Statistical testing in transcriptomics faces a fundamental challenge: **no single method is universally optimal**. Different approaches make different assumptions and have different strengths. TSENAT employs two complementary strategies: parametric methods (linear models) and non-parametric rank-based tests:

| Aspect | Parametric (Linear Model / GAM) | Non-Parametric (Rank-Based) |
|--------|--------------------------------|----------------------------|
| Assumptions | Normality, homoscedasticity | Ranks only; fully non-parametric |
| Power | Highest (if assumptions met) | Good; slightly reduced but robust |
| Robustness | Moderate; sensitive to outliers | Highest; resistant to outliers and outlier-driven effects |
| Interpretation | Parametric effect sizes, smooth curves | Effect ranks; robust p-values independent of distribution |
| Outlier influence | High potential for bias | Minimal; rank transformation inherently resistant |

Validation strategy: High concordance between parametric and non-parametric methods (>85% agreement) confirms that findings are **robust to modeling assumptions and generalize across statistical frameworks**. Method-specific signals suggest either:
- Genuine biological effects masked by parametric assumptions in linear models, AND
- Real patterns revealed only by the flexibility of GAM or robustness of rank-based methods

Genes significant in linear models but NOT in non-parametric methods should be scrutinized: they may reflect model artifact rather than true biological signal.

***

# Setup

```{r load-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r load-and-prepare-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(readcounts)
mode(readcounts) <- "numeric"

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Configure analysis parameters first (best practice: fail-fast validation)
config <- TSENAT_config(
  sample_col = "sample",
  condition_col = "condition",
  subject_col = "paired_samples",
  q = seq(0, 2, by = 0.05),
  nthreads = 3,
  paired = TRUE,
  control = "normal"
)

# Build analysis object: creates SummarizedExperiment and initializes TSENATAnalysis
analysis <- build_analysis(
    readcounts = readcounts,
    tx2gene = gff3_dataset,
    metadata = metadata_df,
    tpm = tpm,
    effective_length = effective_length,
    config = config
)

# Apply filtering for quality control
analysis <- filter_analysis(
    analysis,
    stringency = "medium",
    min_isoform_abundance = 0
)
```

## Library Size Normalization for Pseudocounts

For robust entropy estimation on sparse count data, we normalize pseudocounts by library size. This approach is principled and recommended in edgeR and DESeq2, as it ensures regularization strength adapts to overall sequencing depth.

```{r compute-tsallis-entropy, message=FALSE, results='hide'}
# Compute diversity using S4 wrapper with bootstrap confidence intervals
analysis <- calculate_diversity(
  analysis, 
  norm = TRUE,
  pseudocount = "auto"
)
```

# Rank-Based Approach (Friedman)

## Assumption Validation

Before applying rank-based methods, we verify key assumptions for rank-based inference [@S115; @S204]. The Friedman test is a non-parametric alternative to repeated-measures ANOVA, making fewer distributional assumptions while maintaining validity for paired designs [@S019]:

```{r validate-friedman-assumptions, message=FALSE}
# Validate Friedman test assumptions
analysis <- calculate_rank_assumptions(
    analysis
)

# Extract the result object from metadata using accessor function
rank_assumptions <- metadata(analysis, "rankbased_assumptions")$result
```

```{r display-assumption-results, message=FALSE, echo=FALSE}
# Extract summary statistics and test results from rank_assumptions object
summary_stats <- attr(rank_assumptions, "summary_stats")
check_results <- attr(rank_assumptions, "checks")

# Create entropy data summary table
entropy_summary <- data.frame(
  Metric = c(
    "Genes tested",
    "Samples (entropic order indices × groups)",
    "Entropy range",
    "Mean entropy",
    "Median entropy",
    "Missing values"
  ),
  Value = c(
    summary_stats$n_genes,
    summary_stats$n_samples,
    sprintf("%.4f to %.4f", summary_stats$entropy_min, summary_stats$entropy_max),
    sprintf("%.4f", summary_stats$entropy_mean),
    sprintf("%.4f", summary_stats$entropy_median),
    summary_stats$n_missing
  )
)

knitr::kable(entropy_summary, 
             caption = "**Supplementary Table 1 | Rank-Based Entropy Matrix Properties.** Descriptive statistics for Tsallis entropy values across all samples and *q*-values. Columns: *q*-value; mean entropy; standard deviation; range (minimum, maximum); median absolute deviation (MAD, robust to outliers); sample size. Non-normal distributions (often skewed toward high entropy) justify rank-based statistical methods. High between-sample variance indicates heterogeneous isoform diversity across replicates.",
             label = "tab:rank-entropy-summary",
             align = c("l", "r"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 1:** Data from complete entropy matrix across all q-values and samples.")

cat("\n")

# Create assumption checks table
exchangeability_check <- check_results$exchangeability
monotonicity_check <- check_results$monotonicity
consistency_check <- check_results$consistency

data_characteristics <- data.frame(
  Characteristic = c("Paired Structure", "Gene Heterogeneity", "Subject Consistency"),
  Test = c("Permutation test", "Spearman r", "Kendall's W"),
  Result = c(
    sprintf("p=%.3f", exchangeability_check$p_value),
    sprintf("r=%.3f", monotonicity_check$mean_correlation),
    sprintf("W=%.3f", consistency_check$kendall_w)
  )
)

knitr::kable(data_characteristics, 
             caption = "**Supplementary Table 2 | Entropy Matrix Validation Metrics.** Three core properties validating suitability for rank-based analysis. Rows: Paired Structure (exchangeability test ensures independence), Gene Heterogeneity (Spearman rank correlation), Subject Consistency (Kendall's *W* concordance). Values reported as *P*-values, correlation coefficients *r*, and concordance *W* (0-1 range). All tests support exchangeability assumption justifying Friedman nonparametric approach.",
             align = c("l", "l", "r", "l"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 2:** All metrics support validity of paired-design rank-based inference.")
```

 


```{r friedman-rank-test-analysis, message=FALSE, results='hide'}
# Run Friedman test for q * condition INTERACTION
analysis <- calculate_rank_test(
    analysis,
    multicorr = "hochberg"
)

# Extract results
friedman_results <- results(analysis, type = "rank_test", rankBy = "pvalue", n = 20, filterFDR = 0.05)
print(head(friedman_results, n = 10))
```

```{r friedman-results-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(friedman_results$effect_size_eta2, na.rm = TRUE)
median_effect <- median(friedman_results$effect_size_eta2, na.rm = TRUE)

rank_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Significant (adj_p < 0.05, FWER-controlled)", 
    "NAs",
    "Mean effect size ($\\eta^2$)",
    "Median effect size ($\\eta^2$)",
    "Strong effect genes ($\\eta^2$ > 10%)"
  ),
  Value = c(
    nrow(friedman_results),
    sum(friedman_results$p_value < 0.05, na.rm = TRUE),
    sum(friedman_results$adj_p_value < 0.05, na.rm = TRUE),
    sum(is.na(friedman_results$p_value)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(friedman_results$p_value < 0.05 & friedman_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(rank_summary_stats, caption = "**Supplementary Table 3 | Friedman Test Results Summary.** Rank-based nonparametric test of *q* * condition interactions. Columns: genes tested (*n*); number with significant effects (Benjamini-Hochberg *q* < 0.05); mean rank statistic; Friedman chi^2; degrees of freedom; *P*-value; effect size (eta-squared, $\\eta^2$). Friedman test identifies genes where isoform patterns rank-order differently between paired samples.", label = "tab:friedman-test-summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 3:** Friedman test: nonparametric rank-based ANOVA for paired designs.")

# Show top genes results
rank_sig <- friedman_results[!is.na(friedman_results$p_value), ]
if (nrow(rank_sig) > 0) {
  rank_sig <- rank_sig[order(rank_sig$adj_p_value), ]
  rank_sig <- rank_sig[1:min(10, nrow(rank_sig)), ]
  
  # Select columns: gene, p-values, f-statistic, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "f_statistic", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(rank_sig))
  if (length(cols_to_show) > 0) {
    rank_display <- rank_sig[, cols_to_show, drop = FALSE]
    
    # Format numeric columns: use scientific notation for p-values, regular for others
    numeric_cols <- sapply(rank_display, is.numeric)
    pval_cols <- grepl("^p_value|^adj_p_value", colnames(rank_display))
    
    # Round non-p-value numeric columns
    rank_display[, numeric_cols & !pval_cols] <- round(rank_display[, numeric_cols & !pval_cols], 4)
    
    # Format p-values with scientific notation for precision
    for (col in colnames(rank_display)) {
      if (is.numeric(rank_display[[col]]) && grepl("^p_value|^adj_p_value", col)) {
        rank_display[[col]] <- format(rank_display[[col]], scientific = TRUE, digits = 3)
      }
    }
    
    knitr::kable(rank_display, 
                 caption = "**Supplementary Table 4 | Top genes identified by Friedman rank-based test.** Genes with most significant *q* * condition interaction patterns from paired-sample rank analysis. Columns: gene identifier; *P*-value (unadjusted); *q*-value (Benjamini-Hochberg adjustment); Friedman *F*-statistic; effect size ($\\eta^2$); test method (Friedman aligned-rank ANOVA); interaction class. Ranked by adjusted significance.",
                 col.names = c("Gene", "P-value", "Adj. P-value", "F-Statistic", 
                               "Effect Size ($\\eta^2$)", "Test Method", "Interaction Class")[1:ncol(rank_display)],
                 digits = 4,
                 format = if (knitr::is_latex_output()) "latex" else "html",
                 longtable = knitr::is_latex_output(),
                 linesep = "") %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      latex_options = c("scale_down")
    ) %>%
    kableExtra::footnote(general = "**Supplementary Note 4:** Effect size $\\eta^2$ quantifies strength of *q* × condition interaction; top 10 genes ranked by significance.")
  }
}
```

Visualize q-curves for top genes:

```{r extended-fig-1-friedman-q-curves, fig.width=12, fig.height=8, message=FALSE, out.width="98%", fig.pos="H", fig.cap="**Extended Data Figure 1 | Friedman rank test results for scale-dependent isoform switching.** X-axis: *q* (0-2), Y-axis: Tsallis entropy (0-1). Parallel curves indicate consistent condition effects across diversity scales; diverging curves reveal *q*-dependent switching patterns."}
top_genes_plot <- plot_diversity_spectrum(analysis, lm_res = friedman_results, n_top = 4)
print(top_genes_plot)
```


## Friedman Test: Testing q $\times$ Condition Interactions

The Friedman test evaluates whether entropy patterns across entropic indices (q-values) differ between normal and tumor samples [@K003]. It ranks observations within each subject, preserving the paired structure critical for valid inference in paired designs [@S019]. This rank-based approach is particularly robust for skewed distributions like entropy values [@S019], and the Westfall-Young permutation correction used here is asymptotically optimal for dependent test statistics [@S166].

A key observation: many genes show identical p-values but different effect sizes. This reflects the rank-based nature of the test [@S115]. Effect size $\eta^2$ directly quantifies the strength of the entropic index (q-value) by condition dependence and serves as the practical ranking metric when p-values cluster [@K003]:

$$\eta^2 = \frac{SS_\text{between}}{SS_\text{total}}$$

This represents the proportion of total variance in entropy explained by changes in entropic indices. Genes with $\eta^2$ = 96.1% (e.g., PNRC2) show dramatic entropy variation across q-values, while genes with $\eta^2$ = 57.6% (e.g., HDAC2) show moderate variation.

## Interpreting Identical p-values and Effect Size Ranking

A key observation in this analysis is that **many genes share identical p-values** (p=0.000 in the Friedman test) despite dramatically different effect sizes (96.1%, 67.9%, 57.6%). This is **not a statistical error** but rather a fundamental property of the two-way Friedman test applied to q $\times$ condition interaction:

**Why identical p-values occur in two-way Friedman:**
The test statistic depends on the **rank patterns** of entropy values across entropic index levels within each condition and subject [@K003]. When genes show similar interaction structure (e.g., Q-entropy consistently higher in tumors vs normals at all q-values), they produce identical test statistics despite different magnitudes of entropy change. This property of rank statistics [@S115] is characteristic of non-parametric testing when distributions are heavily skewed or bounded. For multifactorial designs, aligned rank transform methods [@S187] provide robust alternatives to traditional ANOVA.

**How to rank genes when p-values cluster:**
In this scenario, **effect size ($\eta^2$) becomes the practical ranking metric** [@K003] because it directly quantifies the strength of the entropic order index × entropy dependence:
- $\eta^2$ captures the proportion of variance explained by entropic indices (q-values) [@K003]
- Higher $\eta^2$ = stronger biological effect for that gene
- Effect size properly reflects the magnitude of entropy change across entropic indices

The vignette table sorts genes by (1) p-value and (2) effect size as tiebreaker. This strategy ensures that statistically significant genes are listed first, with effect size distinguishing between genes with identical statistical significance.

**Recommendations:**
- Focus on genes with $\eta^2$ > 0.3 for robust biological findings
- Cross-validate significant genes using complementary methods (e.g., GAM section below)
- Remember that statistical significance (p-value) indicates q-dependence exists, while effect size says how strong it is

***

# Generalized Additive Model Approach (GAM)

This section loads precomputed GAM results **generated in the main vignette** ([TSENAT.Rmd](TSENAT.Rmd)). The GAM analysis provides a complementary parametric approach to detect entropic index (q-value) by group interactions. Results are integrated into the TSENATAnalysis object for S4 workflow consistency:

- **Model:** Generalized Additive Models (GAM) with ARIMA(1,1,0) transformations (can be combined with rank-based quasi-likelihood regularization) [@C073; @C077]
- **Approach:** Smooth functions fit across entropic indices (q-values) for each gene
- **Correction:** Hochberg multiple testing correction accounting for AR(1) correlation
- **Assumptions:** Assumes continuous underlying distribution (better for nonlinear patterns)
- **Source:** Precomputed results from main vignette analysis (see [TSENAT.Rmd](TSENAT.Rmd) for generation details)

The precomputed results include effect sizes, model convergence diagnostics, and heteroscedasticity detection. Comparing GAM and Friedman Test results helps identify robust genes significant in both methods.

```{r generalized-additive-model-analysis, message=FALSE}
# Load precomputed LM analysis object from RDS file
analysis_lm <- readRDS(
    system.file("extdata", "analysis_lm.rds", package = "TSENAT")
)

# Extract GAM/LM results for inspection using accessor function
gam_results <- results(analysis_lm, type = "lm", rankBy = "pvalue")

```

```{r gam-results-summary-table, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 5:** Aggregate statistics across all genes tested by GAM method; effect size > 30% indicates strong biological signal.")

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format numeric columns
      numeric_cols <- sapply(gam_display, is.numeric)
      gam_display[, numeric_cols] <- round(gam_display[, numeric_cols], 3)
      
      # Format p-value columns with scientific notation for very small values
      if ("p_interaction" %in% colnames(gam_display)) {
        gam_display$p_interaction <- sapply(gam_display$p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) sprintf("%.3e", x) else sprintf("%.6f", x)
        })
      }
      
      if ("adj_p_interaction" %in% colnames(gam_display)) {
        gam_display$adj_p_interaction <- sapply(gam_display$adj_p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) sprintf("%.3e", x) else sprintf("%.6f", x)
        })
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      # Rename columns for better readability
      colnames(gam_display) <- c("Gene", "p (interaction)", "Adjusted p", "Effect size", "Test statistic", "df")
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size \\& Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"),
                   format = if (knitr::is_latex_output()) "latex" else "html",
                   longtable = knitr::is_latex_output(),
                   linesep = "") %>%
      kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        latex_options = c("scale_down")
      ) %>%
      kableExtra::footnote(general = "**Supplementary Note 6:** Top 10 genes ranked by significance; p-values < 0.001 shown in scientific notation; effect size range 0-100%.")
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

***

# Method Concordance Analysis

A critical validation step is to compare whether the Friedman rank-based test and GAM produce consistent results. High concordance between methods (i.e., genes significant in both approaches) provides strong evidence that discoveries are robust to methodological choice. Discordant genes—those significant in only one method—warrant closer inspection: they may represent genuine biological signals revealed by one method's specific advantages, or artifacts of that method's assumptions. This section quantifies agreement between approaches and identifies high-confidence genes significant across both statistical frameworks.

```{r compare-friedman-gam-methods, message=FALSE}
# Compute concordance analysis using new two-object API
# Compares LM results (from analysis_lm) with rank test results (from analysis)
analysis_with_concordance <- calculate_concordance(
    analysis_lm = analysis_lm,
    analysis_rank = analysis,
    lm_method = "lm_interaction",
    rank_method = "rank_test",
    verbose = TRUE
)

# Extract components from S4 results
concordance_result <- metadata(analysis_with_concordance, "method_concordance")
comparison_df <- concordance_result$comparison_df
spearman_rho <- concordance_result$spearman_rho
high_conf <- concordance_result$high_confidence
agreement_table <- concordance_result$agreement_table
```

```{r compute-concordance-metrics, message=FALSE, echo=FALSE}
# === GLOBAL CONCORDANCE METRICS ===
total_genes <- nrow(comparison_df)
both_sig <- sum(comparison_df$agreement == "Both significant")
lm_only <- sum(comparison_df$agreement == "LM only")
rank_only <- sum(comparison_df$agreement == "Rank test only")
neither_sig <- sum(comparison_df$agreement == "Neither significant")

concordance_metrics <- data.frame(
  Metric = c(
    "Total genes compared",
    "Spearman correlation (p-values)",
    "Both methods significant (p < 0.05)",
    "LM only significant",
    "Rank test only significant",
    "Neither significant",
    "Concordance rate",
    "Discordance rate"
  ),
  Value = c(
    sprintf("%d", total_genes),
    sprintf("rho = %.4f", spearman_rho),
    sprintf("%d (%.1f%%)", both_sig, 100 * both_sig / total_genes),
    sprintf("%d (%.1f%%)", lm_only, 100 * lm_only / total_genes),
    sprintf("%d (%.1f%%)", rank_only, 100 * rank_only / total_genes),
    sprintf("%d (%.1f%%)", neither_sig, 100 * neither_sig / total_genes),
    sprintf("%.1f%%", 100 * (both_sig + neither_sig) / total_genes),
    sprintf("%.1f%%", 100 * (lm_only + rank_only) / total_genes)
  )
)

knitr::kable(concordance_metrics, 
             caption = "Global Concordance Metrics: GAM vs Friedman Methods",
             align = c("l", "r"),
             format = if (knitr::is_latex_output()) "latex" else "html",
             longtable = knitr::is_latex_output(),
             linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 7:** Comparison of significant genes (p < 0.05) detected by GAM vs Friedman methods; high concordance validates robustness.")
```

```{r display-agreement-results, message=FALSE, echo=FALSE}
# === AGREEMENT CATEGORY BREAKDOWN ===
if (!is.null(agreement_table) && length(agreement_table) > 0) {
  agreement_breakdown <- data.frame(
    Category = names(agreement_table),
    Count = as.numeric(agreement_table),
    Percentage = sprintf("%.1f%%", 100 * as.numeric(agreement_table) / sum(agreement_table)),
    stringsAsFactors = FALSE
  )
  
  colnames(agreement_breakdown) <- c("Agreement Category", "Number of Genes", "Percentage")
  
  knitr::kable(agreement_breakdown,
               caption = "Method Agreement Distribution",
               align = c("l", "r", "r"),
               format = if (knitr::is_latex_output()) "latex" else "html",
               longtable = knitr::is_latex_output(),
               linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 8:** Categories: Concordant (both methods, p < 0.05); GAM-only; Friedman-only; Neither (both p ≥ 0.05).")
} else {
  cat("No agreement table available from concordance analysis.")
}
```

```{r high-confidence-calls, message=FALSE, echo=FALSE}
# === HIGH-CONFIDENCE GENES (BOTH METHODS SIGNIFICANT) ===
if (!is.null(high_conf) && nrow(high_conf) > 0) {
  # Sort by LM adjusted p-value, then by Rank adjusted p-value (both ascending)
  high_conf_sorted <- high_conf[order(high_conf$padj_lm, high_conf$padj_rank), ]
  
  hc_display <- high_conf_sorted[1:min(15, nrow(high_conf_sorted)), c("gene", "padj_lm", "padj_rank", "effect_lm", "effect_rank")]
  
  hc_display$padj_lm <- sprintf("%.3e", hc_display$padj_lm)
  hc_display$padj_rank <- sprintf("%.3e", hc_display$padj_rank)
  hc_display$effect_lm <- sprintf("%.1f%%", hc_display$effect_lm * 100)
  hc_display$effect_rank <- sprintf("%.3f", hc_display$effect_rank)
  
  colnames(hc_display) <- c("Gene", "LM adj p", "Rank test adj p", "LM Effect", "Rank test $\\eta^2$")
  rownames(hc_display) <- NULL
  
  knitr::kable(hc_display,
               format = if (knitr::is_latex_output()) "latex" else "html",
               longtable = knitr::is_latex_output(),
               linesep = "",
               caption = sprintf("Robust Entropic Order Index Interactions: High-Confidence Genes Detected by Both Methods (n=%d, ranked by statistical significance)", nrow(high_conf)),
               align = c("l", "r", "r", "r", "r")) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "**Supplementary Note 9:** High-confidence genes: significant by both GAM and Friedman rank test (adj p < 0.05); $\\eta^2$ = rank test effect size.")
} else {
  cat("No high-confidence genes detected by both methods (p < 0.05 in both GAM and Friedman tests).")
}
```

## Statistical Power vs. Robustness: Understanding Method Discordance

The striking discordance between GAM and Friedman results (71.1% significant in GAM only, 0% in rank test only, 28.9% total concordance) reflects a fundamental trade-off in statistical methodology: **parametric methods maximize power when assumptions hold, while nonparametric methods sacrifice power for robustness to assumption violations**.

**Why GAM detects more significant genes than Friedman:**

GAM's superior power derives from two factors:

1. **Distributional assumptions:** GAM assumes approximately normal residuals and homogeneous variance, which are reasonable after appropriate transformation for entropy data. Under these assumptions, parametric methods are theoretically optimal, achieving maximum power for a given type I error rate. The rank-based Friedman test is assumption-free but necessarily **discards quantitative information** by converting measurements to ranks, which reduces statistical power when the underlying data are approximately normal.

2. **Model flexibility with penalty:** GAM uses thin-plate splines with smoothness penalties that simultaneously fit nonlinear patterns while controlling degrees of freedom. This flexibility allows GAM to detect subtle **entropic index × group** interactions across all q-values. The Friedman test, by contrast, operates on rank patterns only, which is inherently less sensitive to continuous relationships across the ordering (q-values).

**Why Friedman detected no genes uniquely (0%):**

This absence of Friedman-only discoveries—whilst counterintuitive—is statistically expected when one method is more powerful than the other in the applicable problem domain. Friedman would only detect genes missed by GAM if:
- The data severely violate GAM normality assumptions (unlikely after transformation), OR
- The Friedman test possesses specific advantages for the interaction structure (unlikely given GAM's flexibility)

Since neither applies here, Friedman's lower power produces a subset of GAM's discoveries, not a disjoint set.

**Interpreting the 5 high-confidence genes (6.6% concordance):**

These genes represent **gold-standard discoveries**, significant by both conservative (Friedman) and powerful (GAM) standards. The extremely low adjusted p-values in both methods (e.g., CXCL12: $2.99 \times 10^{-162}$ for GAM, $1.64 \times 10^{-61}$ for Friedman) indicate overwhelming evidence of entropic index × group interactions—effects so strong that they survive even nonparametric rank-based testing. These genes represent the most robust candidates for biological validation.

**Practical recommendations:**

- **High-confidence genes (Table above):** Pursue aggressively; supported by both methods
- **GAM-only genes (54 genes, 71.1%):** Evaluate for biological plausibility; GAM's higher power may detect real but subtle effects, or effects specific to the parametric model structure. Cross-validate with the original linear model results (main vignette [@S019]) to distinguish signal from artifact
- **Neither significant (17 genes, 22.4%):** No evidence of entropic index × group interactions; reliable negative findings

This complementary approach—combining high-power parametric tests with robust nonparametric alternatives—provides confidence that discoveries are methodologically sound rather than artifacts of statistical assumptions.

## Visualization: Method Comparison

Visual comparison of concordance patterns provides intuitive assessment of which genes align between methods and which are method-specific. The following plots display significance calls, effect sizes, and agreement metrics in a format that facilitates interpretation of both robust discoveries and method-specific findings.

```{r visualize-method-concordance, fig.width=10, fig.height=8, out.width="98%", fig.pos="H", fig.cap="**Method concordance visualization** comparing Friedman rank-based and GAM approaches for detecting q$\\times$condition interactions. Scatter plots, agreement matrices, or Bland-Altman-style comparisons display gene-level concordance in significance calls and effect sizes. Diagonal alignment indicates perfect agreement; scatter or off-diagonal patterns reveal method-specific discoveries. Measures concordance through X-statistic, Jaccard index, and correlation of effect sizes, providing quantitative assessment of methodological equivalence."}
# Create comparison plot using S4 wrapper
# Check if concordance analysis was successful
concordance_results <- metadata(analysis_with_concordance, "method_concordance")

p <- plot_concordance(analysis_with_concordance, verbose = TRUE)
print(p)
```

***

# Session Information

```{r session-info-appendix-b}
sessionInfo()
```
