---
title: "TSENAT Appendix B: Rank-Based Methods for Robust Multi-Q Analysis"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Rank-Based Methods for Entropy Analysis

## Motivation

While TSENAT's linear mixed model (LMM) and generalized additive model (GAM) approaches are powerful for detecting q×group interactions, they rely on parametric assumptions that may not hold for entropy data. **Tsallis entropy** distributions are often:

- **Non-normally distributed** (bounded between 0 and log(m), skewed)
- **Heteroscedastic** (variance depends on number of isoforms per gene)
- **Dependent across q** (autocorrelated q-values violate independence assumptions)

This appendix presents **rank-based (non-parametric) alternatives** that complement the main TSENAT pipeline. These methods:

1. **Require no distributional assumptions** beyond exchangeability
2. **Handle dependence between q-values naturally** via ranking
3. **Provide robust inference even under outliers and skewness**
4. **Enable permutation-based exact p-values** for strict Type I control

## When to Use Rank-Based Methods

Use rank-based methods in these scenarios:

- **Diagnostic validation**: Confirm LMM/GAM results using non-parametric alternatives
- **Non-normal data**: When entropy distributions violate normality
- **Outlier sensitivity**: When a few samples exert disproportionate influence
- **Conservative inference**: When Type I error control is critical (e.g., regulatory biomarkers)
- **Multi-testing robustness**: When accounting for dependence between q-values matters

## Key Differences from Parametric Methods

| Aspect | Parametric (GAM/LMM) | Rank-Based |
|--------|----------------------|-----------|
| **Assumptions** | Normality + independence | Exchangeability only |
| **Outlier sensitivity** | High (mean-based) | Low (rank-based) |
| **Multi-q dependence** | Not explicitly handled | Naturally accommodated |
| **FWER control** | Approximation (closed-form) | Exact (permutation) |
| **Computational cost** | Fast (< 1 min for 5000 genes) | Moderate (minutes for permutation) |

---

# Setup and Data Preparation

## Load packages and data

We'll build on the TSENAT pipeline from the main vignette. Start by loading the preprocessed data and entropy calculations:

```{r setup-packages}
# Load packages
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
})

# For reproducibility
set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed example dataset (from main TSENAT.Rmd)
data(readcounts)

# Build SummarizedExperiment with preprocessing
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Create SummarizedExperiment
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)

# Filter low-abundance transcripts
se <- filter_se(se, stringency = "medium")

# Compute Tsallis entropy across q-spectrum
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(se,
    q = qvec, norm = TRUE, metadata = metadata_df
)

cat("Data loaded and preprocessed:\n")
cat("  Genes:", nrow(se), "\n")
cat("  Samples:", ncol(se), "\n")
cat("  Q-values:", length(qvec), "\n")
```

---

# Part 1: Aligned Rank Transform (ART) for Multi-Factor Designs

## Rationale

The Aligned Rank Transform (ART) converts non-normal data to a rank scale suitable for parametric-style testing without assuming normality. This is particularly useful when entropy data violates assumptions of ANOVA or linear regression.

**Two-stage process:**
1. **Align**: Fit and subtract factor effects (e.g., q-value effects)
2. **Rank**: Convert residuals to ranks, then to normal scores

This enables robust ANOVA-type tests on entropy data while respecting multi-factor structure.

## Apply ART to entropy data

```{r art-basic, message=FALSE}
# Extract entropy assay and colData
entropy_matrix <- assay(ts_se)
ts_coldata_full <- colData(ts_se)

# For q × sample_type interaction test, we need both sample types at multiple q-values
# Select columns strategically to ensure we have both normal AND tumor samples

# Get unique q-values (should be 40, from 0.1 to 2)
unique_qs <- unique(ts_coldata_full$q)
n_qs_available <- length(unique_qs)

# Select 3 representative q-values: low, mid, high
selected_q_values <- c(unique_qs[1], 
                       unique_qs[ceiling(n_qs_available/2)], 
                       unique_qs[n_qs_available])

# Find columns matching these q-values with varying sample types
selected_cols <- integer()
for (q_val in selected_q_values) {
    # Get ALL columns with this q-value
    cols_with_q <- which(ts_coldata_full$q == q_val)
    
    # Select 2 normal and 2 tumor from these columns
    cols_normal <- cols_with_q[ts_coldata_full$sample_type[cols_with_q] == "normal"]
    cols_tumor <- cols_with_q[ts_coldata_full$sample_type[cols_with_q] == "tumor"]
    
    # Take first 2 from each type (if available)
    selected_cols <- c(selected_cols, 
                      head(cols_normal, 2),
                      head(cols_tumor, 2))
}

# Extract data and create factor table
subset_genes <- 1:10  # First 10 genes (for compact visualization)
entropy_subset <- entropy_matrix[subset_genes, selected_cols]

selected_coldata <- ts_coldata_full[selected_cols, ]
factors_subset <- data.frame(
    q_value = factor(selected_coldata$q),
    sample_type = factor(selected_coldata$sample_type),
    pair_id = factor(selected_coldata$paired_samples)
)

# Verify we have both sample types in the design
cat("Data for q × sample_type interaction ART:\n")
cat("  Entropy matrix shape:", nrow(entropy_subset), "genes ×", ncol(entropy_subset), "columns\n")
cat("  Factors shape:", nrow(factors_subset), "rows\n")
cat("  Q-values tested:", paste(round(unique(as.numeric(as.character(factors_subset$q_value))), 2), collapse = ", "), "\n")
cat("  Sample types:", paste(unique(as.character(factors_subset$sample_type)), collapse = ", "), "\n")
cat("  Design:", length(unique(factors_subset$q_value)), "q-values ×", 
    length(unique(factors_subset$sample_type)), "sample types\n")
cat("  Sample type breakdown:\n")
print(table(factors_subset[, c("q_value", "sample_type")]))
cat("\n  Alignment check (factors rows == entropy columns):", 
    nrow(factors_subset) == ncol(entropy_subset), "\n\n")

# Apply ART with formula testing q × sample_type interaction (with pairing adjustment)
# This is the rank-based parallel to calculate_lm_interaction()
art_result <- apply_aligned_rank_transform(
    data = entropy_subset,
    factors = factors_subset,
    formula = ~ q_value * sample_type + pair_id  # Test q × condition interaction, adjust for pairing
)

cat("\n")
print(art_result)
```

## Interpretation: ART for q × sample_type Interactions

The ART output shows:

- **aligned_ranks**: Rank-transformed entropy values after adjusting for q-effects, sample-type effects, and pairing
- **normal_scores**: Van der Waerden normal scores (now suitable for parametric ANOVA)
- **alignment_effects**: Estimated effects (q, sample_type, pairing) subtracted during alignment step
- **summary**: Quality metrics for the alignment process

**Key insight:** ART allows testing whether the difference between normal and tumor samples varies across the q-spectrum. This parallels the **q × sample_type interaction** from `calculate_lm_interaction()`, but using rank-based methods that don't assume normality.

After ART alignment and ranking, downstream tests can use standard parametric procedures on the aligned ranks/normal scores with confidence that rank-based robustness is preserved.

```{r art-diagnostic}
# Check alignment quality
art_summary <- art_result$summary
cat("\nART Diagnostic Summary:\n")
cat(art_summary, "\n")

# Get dimensions
n_genes <- nrow(art_result$aligned_ranks)
n_samples <- ncol(art_result$aligned_ranks)

cat("\nART Results:\n")
cat("  Genes processed:", n_genes, "\n")
cat("  Samples/observations:", n_samples, "\n")

# Show sample ranks for first gene
cat("\nAligned ranks for Gene 1:\n")
print(art_result$aligned_ranks[1, ])

cat("\nNormal scores (van der Waerden) for Gene 1:\n")
print(art_result$normal_scores[1, ])
```

---

# Part 2: Rank Correlation Across Q-Values

## Rationale

A key question in multi-q analysis: **Are gene rankings consistent across q-values?**

- **High correlation (r > 0.85)**: Genes rank similarly regardless of q → robust effects
- **Low correlation (r < 0.50)**: Gene rankings shift dramatically with q → scale-dependent effects

## Compute rank correlations

```{r rank-corr-basic, message=FALSE}
# For demonstration: subsample q-values for faster computation
# (Use full q-spectrum in actual analysis)
entropy_matrix_full <- assay(ts_se)
subset_qs <- seq(1, ncol(entropy_matrix_full), length.out = 10)  # Sample 10 from 40
entropy_matrix <- entropy_matrix_full[, subset_qs]

# For visualization: use only first 10 genes
subset_genes <- 1:min(10, nrow(entropy_matrix))
entropy_matrix <- entropy_matrix[subset_genes, ]

n_genes <- nrow(entropy_matrix)
n_qvalues <- ncol(entropy_matrix)

cat("Rank correlation analysis dimensions:", n_genes, "genes ×", n_qvalues, "q-values (subsampled)\n\n")

# KEY: For each q-value, rank the genes (not samples)
# This tests: "Are gene rankings consistent across q-values?"
# Create a list where each element is the ranked genes for one q-value
pvals_list <- list()
for (i in seq_len(n_qvalues)) {
    # Get entropy values for this q-value (across all genes)
    q_entropy <- entropy_matrix[, i]
    # Rank genes by their entropy at this q-value
    pvals_list[[i]] <- q_entropy
}
names(pvals_list) <- colnames(entropy_matrix)

# Compute Spearman rank correlations between q-value-specific gene rankings
rank_corr_result <- compute_rank_correlation_multiq(
    pvalues_list = pvals_list,
    method = "spearman",
    use_ranks = TRUE
)

cat("\n")
print(rank_corr_result)
```

## Visualization

```{r rank-corr-visualization}
# Plot correlation heatmap
# Each cell shows: correlation between gene rankings at two different q-values
# High correlation (red): genes rank similarly at both q-values (robust)
# Low correlation (blue): gene order changes with q-value (scale-dependent)

p_rank_corr <- plot_rank_correlation_heatmap(
    rank_corr_result,
    title = "Gene Ranking Consistency Across Q-Values\n(Spearman correlation of gene-specific entropy rankings)"
)
print(p_rank_corr)
```

## Interpretation Table

```{r rank-corr-interpretation}
# Create interpretation summary
# What we're measuring: For each q-value pair, how similar are gene rankings?

corr_matrix <- rank_corr_result$correlation_matrix

# Extract key statistics
upper_tri_corrs <- corr_matrix[upper.tri(corr_matrix)]
mean_corr <- mean(upper_tri_corrs, na.rm = TRUE)
min_corr <- min(upper_tri_corrs, na.rm = TRUE)
max_corr <- max(upper_tri_corrs, na.rm = TRUE)

cat("Question: Do genes rank consistently across different q-values?\n\n")

summary_df <- data.frame(
    Metric = c("Mean Correlation", "Min Correlation", "Max Correlation", 
               "Interpretation"),
    Value = c(
        sprintf("%.4f", mean_corr),
        sprintf("%.4f", min_corr),
        sprintf("%.4f", max_corr),
        if (mean_corr > 0.80) "Very stable rankings (high q-robustness)"
        else if (mean_corr > 0.65) "Moderate stability (moderate q-effects)"
        else "Variable rankings (strong q-dependence)"
    )
)

knitr::kable(summary_df, caption = "Gene Ranking Consistency Across Q-Values")

cat("\n\nInterpretation:\n")
cat("  > 0.80 : Same top/bottom genes regardless of q (robust, universal effects)\n")
cat("  0.65-0.80 : Generally consistent but some q-dependent shifts (scale effects present)\n")
cat("  < 0.65 : Gene rankings change dramatically with q (scale-dependent biology)\n")
```
---

# Part 3: Rank Correlation Confidence Intervals

## Rationale

Point estimates (correlations) don't capture **uncertainty**. Bootstrap confidence intervals reveal:

- **Wide CI**: Uncertain gene ranking → results may not replicate
- **Narrow CI**: Confident gene ranking → robust, reproducible effects

## Compute bootstrap CIs

```{r rank-ci-basic, message=FALSE}
# Compute 95% bootstrap percentile CIs
# Note: n_bootstrap = 200 for vignette speed; use 1000+ for publication
rank_ci <- rank_correlation_bootstrap_ci(
    pvalues_or_ranks = pvals_list,
    method = "spearman",
    ci = "percentile",
    ci_level = 0.95,
    n_bootstrap = 200,  # Reduced from 1000 for vignette
    seed = 42,
    return_distribution = FALSE
)

cat("\n")
print(rank_ci)
```

## CI Interpretation

```{r rank-ci-interpretation}
# Extract and display interpretation table
ci_interp <- rank_ci$interpretation
knitr::kable(
    ci_interp,
    caption = "Rank Correlation Confidence Intervals and Stability Assessment",
    digits = 4
)
```

**Key insights for each pairwise comparison:**

- **Includes zero**: No significant rank correlation → q-values give independent rankings
- **Narrow CI + high r**: Robust, reliable ranking consistency across q
- **Wide CI + medium r**: Uncertain ranking (need more data or biological replication)
- **Narrow CI + low r**: Stable but weak correlation (q-values select different genes)

---

# Part 4: Detecting Gene-Q Interactions via Rank-Based Testing

## Rationale

Extends the LMM/GAM approach from the main vignette using **Kruskal-Wallis test** (rank-based ANOVA equivalent). Answers: **Do gene rankings differ significantly across q-values?**

This is a non-parametric alternative to `calculate_lm_interaction()`, robust to non-normality and outliers.

## Prepare data for interaction testing

```{r q-interaction-prep, message=FALSE}
# Create long-format data for interaction testing
# Use available genes (min of 50 or total available)
n_test_genes <- min(50, nrow(ts_se))
test_gene_names <- rownames(ts_se)[1:n_test_genes]

# Check colData structure (already has pairing info!)
cat("colData(ts_se) columns:", paste(colnames(colData(ts_se)), collapse = ", "), "\n")
cat("ts_se dimensions:", nrow(ts_se), "genes ×", ncol(ts_se), "(samples × q-values)\n")
cat("Expected entropy matrix rows:", n_test_genes, "\n\n")

# Create long-format data (one row per gene-column combination)
# ts_se has one column per (sample, q-value) combination
# colData(ts_se) already contains sample_type and paired_samples metadata
entropy_values <- as.numeric(assay(ts_se[test_gene_names, ]))
n_cols <- ncol(ts_se)
ts_coldata <- colData(ts_se)

# Create long-format data using colData directly
entropy_long <- data.frame(
    entropy = entropy_values,
    gene = rep(test_gene_names, n_cols),
    sample_id = rep(ts_coldata$samples, each = n_test_genes),
    q = rep(ts_coldata$q, each = n_test_genes),
    sample_type = rep(ts_coldata$sample_type, each = n_test_genes),
    pair_id = rep(ts_coldata$paired_samples, each = n_test_genes)
)

cat("Long-format data created (WITH pairing from colData):\n")
cat("  Rows:", nrow(entropy_long), "\n")
cat("  Expected:", n_test_genes, "genes ×", n_cols, "columns =", n_test_genes * n_cols, "\n")
cat("  Unique sample_type:", paste(unique(entropy_long$sample_type), collapse = ", "), "\n")
cat("  Unique pair_id:", paste(unique(entropy_long$pair_id), collapse = ", "), "\n")
head(entropy_long, 15)
```

## Run interaction test

```{r q-interaction-test, message=FALSE}
# Use subset of genes for faster demo
n_test_genes <- min(40, nrow(ts_se))
test_gene_names <- rownames(ts_se)[1:n_test_genes]

# Create long-format data for testing
n_cols <- ncol(ts_se)
entropy_values <- as.numeric(assay(ts_se[test_gene_names, ]))
ts_coldata <- colData(ts_se)

# Create long-format data using colData directly (already has metadata)
entropy_long <- data.frame(
    entropy = entropy_values,
    gene = rep(test_gene_names, n_cols),
    sample_id = rep(ts_coldata$samples, each = n_test_genes),
    q = rep(ts_coldata$q, each = n_test_genes),
    sample_type = rep(ts_coldata$sample_type, each = n_test_genes),
    pair_id = rep(ts_coldata$paired_samples, each = n_test_genes)
)

# Note: detect_q_gene_interactions() tests q-effect within each gene
# Pairing structure is preserved in the data, though KW test doesn't
# explicitly model random intercepts. For paired analysis, see main vignette.

# Test for q-gene interactions (rank-based)
q_interactions <- detect_q_gene_interactions(
    data = entropy_long,
    entropy_col = "entropy",
    q_col = "q",
    gene_col = "gene",
    method = "kruskal.test"  # Rank-based
)

# Display top significant results
q_inter_sig <- q_interactions[order(q_interactions$p_value), ]
top_genes <- head(q_inter_sig, 10)

knitr::kable(
    top_genes[, c("gene", "f_statistic", "p_value", "effect_size_eta2", "interaction_class")],
    caption = "Top 10 Genes with Significant Q-Gene Interactions (Kruskal-Wallis)",
    digits = 6
)
```

## Classify genes by Q-dependency

```{r q-dependency-classification}
# Classify each gene's q-dependency pattern
classifications <- classify_q_dependency(
    q_interactions,
    p_threshold = 0.05,
    eta2_threshold_moderate = 0.01,
    eta2_threshold_strong = 0.10
)

# Summary table
class_table <- table(classifications)
knitr::kable(
    data.frame(
        Classification = names(class_table),
        Count = as.numeric(class_table),
        Percentage = sprintf("%.1f%%", 100 * class_table / sum(class_table))
    ),
    caption = "Gene Classification by Q-Dependency"
)
```

## Recommend Q-Value Range

```{r q-range-recommendation}
# Get data-driven q-value recommendations
q_recommendation <- recommend_q_range(
    q_interactions,
    robust_threshold = 0.70,
    strong_threshold = 0.05
)

cat("\n=== Q-VALUE RANGE RECOMMENDATION ===\n\n")
cat("Recommendation:", q_recommendation$recommendation, "\n\n")
cat("Rationale:\n", q_recommendation$rationale, "\n\n")
cat("Sample composition:\n")
cat("  Robust genes:", q_recommendation$sample_sizes$robust, 
    sprintf("(%.1f%%)\n", 100 * q_recommendation$robust_pct))
cat("  Moderate q-dependency:", q_recommendation$sample_sizes$moderate,
    sprintf("(%.1f%%)\n", 100 * q_recommendation$moderate_pct))
cat("  Strong q-dependency:", q_recommendation$sample_sizes$strong,
    sprintf("(%.1f%%)\n", 100 * q_recommendation$strong_pct))
cat("\nSuggested q-values:", 
    sprintf("%.2f", q_recommendation$suggested_q_values), "\n")
```

---

# Part 5: Rank-Based FWER Control via Permutation

## Rationale

Traditional multiple testing corrections (Hochberg, Benjamini-Hochberg) assume independence. When testing across multiple q-values (dependent structure), permutation-based FWER control provides **exact Type I error rates** without assumptions.

## Run permutation test

```{r fwer-perm-test, message=FALSE}
# Prepare data: entropy differences between sample types
# Note: Permutation respects original pairing structure in entropy matrix
# (samples are paired by column position from metadata_df)

sample_groups <- factor(colData(ts_se)$sample_type)

# Use available genes (safe subset)
n_fwer_genes <- min(30, nrow(ts_se))  # Reduced from 50
entropy_subset_fwer <- assay(ts_se)[1:n_fwer_genes, ]

# Document pairing structure
cat("FWER Permutation Test\n")
cat("  Genes tested:", n_fwer_genes, "\n")
cat("  Sample groups:", paste(unique(metadata_df$sample_type), collapse=" vs "), "\n")
cat("  Paired samples:", length(unique(metadata_df$paired_samples)), "\n")
cat("  Note: Samples maintain pairing structure during permutation\n\n")

# Run rank-based FWER control
# Note: n_permutations = 250 for vignette speed; use 1000+ for publication
fwer_result <- rank_based_fwer_control(
    data = entropy_subset_fwer,
    groups = sample_groups,
    n_permutations = 250,  # Reduced from 500
    test_statistic = "maxT"
)

cat("\n")
print(fwer_result)
```

## Comparison: Adjusted P-values

```{r fwer-pvalues-display}
# Display adjusted p-values
n_display <- min(20, n_fwer_genes)
fwer_df <- data.frame(
    Gene = rownames(entropy_subset_fwer)[1:n_display],
    Gene_Statistic = fwer_result$gene_level_statistics[1:n_display],
    FWER_Adjusted_P = fwer_result$fwer_adjusted_p[1:n_display]
)

# Identify significant genes
fwer_df$Significant_FWER = fwer_df$FWER_Adjusted_P < 0.05

knitr::kable(
    head(fwer_df[order(fwer_df$FWER_Adjusted_P), ], 10),
    caption = "Genes Significant After Permutation-Based FWER Control (p < 0.05)",
    digits = 6
)
```

---

# Part 6: Diagnostic - Assumptions Validation

## Why check assumptions?

Rank-based methods require **exchangeability** (samples are independent, interchangeable). This diagnostic confirms data fits that assumption.

```{r rank-assumptions-test}
# Validate rank-based method assumptions
assumptions <- test_rankbased_assumptions(
    data = entropy_matrix,
    checks = c("exchangeability", "monotonicity", "consistency")
)

cat("\n")
print(assumptions)
```

---

# Part 7: Comparison - Parametric vs. Rank-Based Results

## Motivation

Do parametric and non-parametric methods agree? Concordance validates results; discrepancies suggest data violations.

## Compare LMM vs. Kruskal-Wallis

```{r comparison-lmm-kw, message=FALSE}
# Run both parametric (LMM) and rank-based (KW) tests on same data
# This serves as validation

# Parametric: linear mixed model (from main vignette)
if (requireNamespace("mgcv", quietly = TRUE)) {
    lm_res <- calculate_lm_interaction(
        ts_se[1:30, ],  # Reduced from 50 genes
        method = "gam",
        paired = TRUE,
        multicorr = "hochberg",
        nthreads = 1,
        verbose = FALSE
    )
    lm_pvals <- data.frame(
        Gene = lm_res$gene,
        LMM_P = lm_res$p_interaction,
        LMM_Padj = lm_res$adj_p_interaction
    )
} else {
    lm_pvals <- NULL
}

# Rank-based: Kruskal-Wallis (already run above)
kw_pvals <- data.frame(
    Gene = q_interactions$gene[1:50],
    KW_P = q_interactions$p_value[1:50]
)

# Merge and compare
if (!is.null(lm_pvals)) {
    comparison <- merge(lm_pvals, kw_pvals, by = "Gene", all = TRUE)
    
    # Correlation of p-values
    p_corr <- cor(comparison$LMM_P, comparison$KW_P, use = "complete.obs")
    
    cat("\n=== PARAMETRIC vs. RANK-BASED COMPARISON ===\n\n")
    cat("Correlation of p-values:", sprintf("%.4f", p_corr), "\n")
    cat("Interpretation: ")
    cat(if (p_corr > 0.85) "High concordance (methods agree)\n"
        else if (p_corr > 0.70) "Moderate concordance (substantial agreement)\n"
        else "Low concordance (methods may identify different genes)\n")
    
    cat("\nTop 10 genes by LMM p-value:\n")
    print(knitr::kable(
        head(comparison[order(comparison$LMM_P), ], 10),
        digits = 6
    ))
}
```

---

# Part 8: Practical Recommendations

## When to use each approach

| Scenario | Recommendation |
|----------|-----------------|
| **Normal entropy data** | GAM/LMM (faster, parametric efficient) |
| **Non-normal or skewed** | Rank-based (robust, no assumptions) |
| **Few outliers** | GAM/LMM (outlier-robust with m-estimation) |
| **Many outliers** | Rank-based (insensitive to extreme values) |
| **High heterogeneity** | Rank-based (no variance homogeneity needed) |
| **Strict Type I control** | Permutation FWER (exact p-values) |
| **Q-dependence critical** | Rank-based (naturally accounts for dependence) |
| **Publication biomarkers** | Both (validate results across methods) |

## Best Practice Workflow

```{r best-practice-workflow, eval=FALSE}
# Step 1: Run parametric pipeline (fast, primary)
lm_res <- calculate_lm_interaction(ts_se, method = "gam", paired = TRUE)

# Step 2: Classify genes by q-dependency
classifications <- classify_q_dependency(
    detect_q_gene_interactions(entropy_long)
)

# Step 3: Validate with rank-based methods
rank_corr <- compute_rank_correlation_multiq(pvals_list)
rank_ci <- rank_correlation_bootstrap_ci(pvals_list, ci = "percentile")

# Step 4: Report on results meeting both criteria
# (good LMM p-value AND high rank correlation CI)
robust_genes <- which(
    lm_res$adj_p_interaction < 0.05 &
    rank_ci$interpretation$Includes_Zero == FALSE
)

# Step 5: Visualize and interpret
plot(rank_corr)  # Rank correlation heatmap
plot_tsallis_gene_profile(ts_se, gene = rownames(ts_se)[robust_genes[1]])
```

---

# Conclusion

Rank-based methods provide **robust validation** of TSENAT's main parametric pipeline:

- **Complementary**: Non-parametric alternatives require different assumptions
- **Reassuring**: High concordance between methods validates results
- **Diagnostic**: Discrepancies signal data violations needing investigation
- **Conservative**: Permutation-based p-values guarantee Type I control

For **publication-ready results**, report both parametric and rank-based findings, highlighting genes significant in both analyses.

---

# Session Information

```{r session-info}
sessionInfo()
```
