---
title: "TSENAT: Tsallis Entropy Analysis Toolbox"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5
)
```

# Overview

**TSENAT** (**Tsallis Entropy** Analysis Toolbox) quantifies **isoform heterogeneity** in RNA-seq data using **Tsallis entropy**—a **scale-dependent** diversity measure distinct from abundance-focused tools (DESeq2, Salmon) and differential transcript usage packages (*SplincingFactory*). By tuning a sensitivity parameter `q`, users examine diversity at different scales: rare variants (low `q`) or dominant isoforms (high `q`). This vignette demonstrates computing **Tsallis entropy** from transcript-level abundance estimates, comparing measures between groups, and visualizing **scale-dependent** differences via q-curves. All results integrate seamlessly with Bioconductor via `SummarizedExperiment`.

## Motivation and Bioconductor Contribution

Common RNA-seq tools, such as DESeq2 or differential transcript usage (DTU) packages like 
*SplincingFactory*, focus either on changes in **total gene abundance** or on **individual 
transcript usage proportions**. While powerful, these approaches can obscure an important 
biological phenomenon: genes often change *which* isoforms they express without large changes 
in overall abundance, and this isoform heterogeneity itself—independent of proportional shifts—
carries biological signal about cellular state and function.

**TSENAT** fills this gap by providing a **scale-dependent** **diversity** framework that captures 
isoform heterogeneity regardless of which specific isoforms dominate. Unlike *SplincingFactory*, 
which tests for shifts in transcript proportions between conditions, TSENAT quantifies the 
entropy or complexity of the isoform distribution itself. **Tsallis entropy** achieves this through a sensitivity parameter `q`: by tuning `q`, you slide a lens across abundance 
scales—zooming in on rare variants (low `q`) or focusing on dominant isoforms (high `q`)—
capturing scale-dependent isoform switching that ordinary summaries overlook.

Bioconductor inclusion reflects TSENAT's role as a complementary statistical framework 
for transcript-level analysis. The package introduces rigorous information-theoretic methods 
(Tsallis entropy with bootstrapped confidence intervals, paired testing, and functional 
principal component analysis) designed specifically for isoform-level RNA-seq. By integrating 
with Bioconductor's `SummarizedExperiment` standard and leveraging the ecosystem's mature 
statistical infrastructure, TSENAT empowers researchers to ask novel questions about 
isoform-level complexity that are inaccessible to existing DTU or abundance-focused tools.



## High-level workflow

1. Preprocess counts and filter low-abundance transcripts.
2. Compute relative transcript proportions within each gene and evaluate
Tsallis entropy across one or more `q` values.
3. Summarize per-gene values across samples (mean/median) and test for
differences between groups (Wilcoxon or permutations).
4. Visualize q-curves and inspect transcript-level counts for top genes.

Design assumption: This vignette assumes paired or longitudinal study designs
with a minimum of 6-8 samples per group for adequate statistical power.


## Installation

To install `TSENAT` from Bioconductor:

```{r installation, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TSENAT")
```

Then load the package:

```{r load-package, eval = FALSE}
library(TSENAT)
```

## Quick Start

For users who want to see results quickly, here is a minimal workflow:

```{r quickstart, eval = FALSE}
library(TSENAT)
library(SummarizedExperiment)

# Load example data
data("readcounts", package = "TSENAT")
readcounts <- as.matrix(salmon_dataset)

# Load sample metadata and annotation
metadata_df <- read.table(
  system.file("extdata", "metadata.tsv", package = "TSENAT"),
  header = TRUE, sep = "\t"
)
gff3_file <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build SummarizedExperiment and add metadata
se <- build_se(readcounts, gff3_file, metadata = metadata_df)

# Filter low-abundance transcripts
se <- filter_se(se, stringency = "medium")

# Compute Tsallis entropy (using single q value for quick start)
# Note: calculate_difference() only accepts single q values because 
# entropies at different q are mathematically dependent (AR(1) structure).
# For multi-q analysis accounting for correlation, use calculate_lm_interaction() instead.
ts_se <- calculate_diversity(se, q = 1, norm = TRUE, metadata = metadata_df)
```

## Data Structure Overview

`TSENAT` operates on transcript-level read counts rather than pre-aggregated gene-level summaries. Input data should be organized as a matrix with transcript IDs as rows and samples as columns, typically derived from quantification tools such as SALMON or kallisto. This matrix must be accompanied by transcript-to-gene mapping information, conventionally provided through a GFF3 annotation file.

All analysis results are organized into a `SummarizedExperiment` object—the standard Bioconductor container for high-throughput expression data. This unified structure facilitates seamless integration with the broader Bioconductor ecosystem. For more details on the `SummarizedExperiment` class and its capabilities, refer to the [SummarizedExperiment documentation](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html).

Within each **SummarizedExperiment** object, three key components work together: column metadata (`colData`) captures sample-level information including identifiers, group assignments, and pairing information; row metadata (`rowData`) stores transcript identifiers, gene assignments, and feature annotations; and assays contain the computed **diversity** measures themselves. The assay structure is flexible—typically one assay per `q` value, or a single assay when combining multiple q values—where each cell represents the normalized or unnormalized Tsallis entropy for a specific gene-sample combination.

# What is Entropy and Tsallis entropy?

**Tsallis entropy** is a one-parameter family of **diversity** measures that generalizes Shannon **entropy** [@Shannon1948] via a sensitivity parameter `q` [@Tsallis1988a]. Unlike Shannon entropy,
which treats all isoforms equally, **Tsallis entropy** lets you zoom in on rare isoforms (low q) or focus on dominant species (high q), capturing scale-dependent isoform
variation that ordinary summaries miss. For a discrete probability vector 
$p=(p_1,\dots,p_n)$ it is defined as:

\[
S_q(p) = \frac{1-\sum_{i=1}^n p_i^q}{q-1}.
\]

This parametric family unites diverse **entropy** concepts under a single framework:
- Generalization: Extends beyond classical Shannon **entropy** to capture **scale-dependent** phenomena
- Mathematical elegance: Reduces to well-known indices at specific `q`-values
- Practical flexibility: Allows data-driven exploration via **q-spectrum** curves

## Information-Theoretic Interpretation

In information theory terms [@Shannon1948, @CoverThomas2006], **entropy** measures the 
uncertainty or surprise when drawing a single item (transcript) from a probabilistic 
ensemble (isoform distribution). Higher entropy means the draw is less predictable 
(many similarly abundant isoforms contribute equally), while lower entropy means the 
distribution is dominated by one or a few highly abundant transcripts. This distinction 
is crucial: two genes may have identical total abundance but dramatically different 
isoform complexity.

**Tsallis entropy** generalizes this concept by introducing the `q` parameter as a sensitivity dial for **scale-dependent** emphasis:
- q < 1: Emphasizes rare, low-abundance isoforms
- q = 1: Recovers Shannon entropy (balanced across all abundance scales)
- q = 2: Yields the Gini-Simpson index, emphasizing dominant isoforms
- q > 2: Focuses almost exclusively on the most abundant species

This **Tsallis entropy** **scale-dependent** nature is the key innovation: by computing $S_q$ across a range of 
q-values (a "q-curve"), you obtain a complete picture of isoform heterogeneity 
across biological scales [@Jost2006, @Furuichi2006]. In practice, combining q-curves 
with resampling or paired designs yields a concise, statistically rigorous summary 
of isoform complexity.

### Essential Limiting Cases

Tsallis entropy exhibits several important mathematical limits and special cases:

- q→1 (Shannon entropy): \(\lim_{q\to1} S_q(p) = H(p) = -\sum_i p_i\log p_i\)
  - Classical information entropy (measured in bits if log base 2, or nats if natural log)
  - Treats all isoforms with equal weight regardless of abundance

- q = 0 (Richness): \(S_0 = m - 1\) where $m$ = number of observed isoforms
  - Counts expressed isoforms (pure species richness, log-scale)
  - Minimum assumption: just presence/absence

- q = 1 (Shannon): \(S_1(p) = -\sum_i p_i\log p_i\)
  - Standard information **entropy** [@Shannon1948]
  - Optimal estimator of information content and predictability

- q = 2 (Gini–Simpson / collision index): \(S_2 = 1 - \sum_i p_i^2\)
  - Probability that two randomly drawn transcripts differ [@Simpson1949]
  - Robust to rare variants (insensitive to very low-abundance isoforms)
  - Range: [0, 1 - 1/m] for m isoforms

- Uniform maximum: For $m$ equally expressed isoforms, maximum entropy is:
  \[S_{q,\max}(m) = \frac{1-m^{1-q}}{q-1}\]
  - Normalized **Tsallis entropy**: \(\tilde S_q = \frac{S_q(p)}{S_{q,\max}(m)} \in [0,1]\)
  - Enables comparison across genes with different isoform counts
  - Essential for distinguishing true biological heterogeneity from annotation artifacts

### Biological Potentiality: Why Tsallis Entropy Matters for Isoform Analysis

Standard RNA-seq analysis measures *whether* transcript abundance changes between 
conditions (fold change, differential expression). However, a critical complementary 
question remains underexplored: *how* does the diversity of isoforms change? 

A gene may show little change in total abundance while dramatically reshuffling its 
isoform repertoire—a phenomenon that current methods largely miss. **Tsallis entropy** captures precisely this: the complexity, richness, and balance of the **isoform heterogeneity** landscape. 
By varying `q`, researchers can examine whether changes are driven by shifts in rare variants or reorganization of dominant isoforms from **scale-dependent** analysis

The mathematical foundation of **Tsallis entropy** rests on Shannon's seminal information theory [@Shannon1948], which defined entropy as a measure of uncertainty. Building on 
this, Rényi [@Renyi1961] developed a parametrized family of entropy measures that 
generalize Shannon entropy. **Tsallis entropy** extends this framework further, providing 
a scale-dependent diversity measure where the `q` parameter controls emphasis: at $q \to 1$ 
it approaches Shannon entropy (balanced sensitivity), while $q = 2$ yields the Gini-Simpson 
index (focus on common species), and $q \to 0$ approaches richness (emphasis on rare species). 
Jost [@Jost2006] clarified how entropy families decouple "true diversity" (species richness) 
from abundance disparities—a principle directly transferable to isoforms: two genes with 
identical Shannon entropy may have very different functional diversity if one is dominated 
by a single isoform (low q sensitivity) while the other distributes evenly (high q sensitivity) 
[@Derian2022, @Seweryn2020].

### Applications of Tsallis Entropy in Genomics and Precision Medicine

**Tsallis entropy** has emerged as a powerful tool for analyzing isoform complexity across diverse
biological contexts, from single-cell transcriptomics [@Gandrillon2021, @Seweryn2020] to 
spatial transcriptomics [@Hinostroza2026]. Recent applications demonstrate its value:

**1. Disease biomarker discovery and cancer diagnosis**: Dysregulated isoform heterogeneity 
often precedes or accompanies disease phenotypes [@Chanda2020]. Cancer cells frequently exhibit 
abnormal splicing entropy signatures [@Seweryn2020, @Chanda2020], which may be detectable 
before bulk transcriptional changes. The ability to measure isoform complexity via Tsallis 
entropy at different q-values reveals whether disease-associated changes affect rare variants 
(low q) or dominant isoforms (high q), providing mechanistic insight into splicing dysregulation 
in malignant transformation.

**2. Cell state transitions and differentiation**: Developmental and differentiation processes 
involve systematic shifts in isoform complexity [@Gandrillon2021]. Maturation typically increases 
dominant isoform usage (lower entropy), while dedifferentiation or proliferation may raise entropy 
through stress-induced isoform switching. Single-cell transcriptomics with entropy measures 
[@Golomb2026] reveals how individual cells navigate state transitions, capturing both the 
deterministic aspects (dominant isoform selection) and stochastic heterogeneity (rare variant 
exploration) that define cellular plasticity.

**3. Adaptive and stress responses**: Rapid environmental challenges (heat stress, nutrient 
starvation, infection) trigger coordinated changes in splicing entropy [@Bajic2024] as cells 
modulate protein isoform ratios to meet functional demands. By measuring entropy at multiple 
q-values, researchers can distinguish between adaptive responses (systematic entropy changes) 
and noise, revealing the biological strategies cells employ for stress resilience.

**4. Tissue-specific regulation and functional specialization**: Different tissues maintain 
distinct entropy "set points" reflecting their functional requirements [@Golomb2026]. Brain tissues, 
for example, show higher isoform complexity than liver [@Golomb2026], reflecting the protein 
functional diversity needed for neural circuit complexity. Spatial transcriptomics now enables 
mapping of isoform entropy patterns within tissues [@Hinostroza2026], revealing how splicing 
regulation supports tissue architecture and function.

**5. Precision medicine and therapeutic stratification**: Isoform entropy signatures may 
predict patient response to therapy [@Derian2022, @Golomb2026]. Individual variation in 
basal isoform complexity, measured via Tsallis entropy, could stratify patients into responder 
and non-responder groups, enabling personalized treatment selection. This extends precision 
medicine beyond mutation detection to include functional complexity of the proteome.

The recent emphasis on information theory in computational biology [@Chanda2020, @Adami2004, 
@Bartal2022] reflects growing recognition that traditional measures (fold change, p-values) 
miss key aspects of biological organization. **Tsallis entropy**, by allowing exploration across 
the **q-spectrum**, captures organization at multiple biological scales simultaneously—a capability 
essential for unpacking the complexity of gene regulation in health and disease.


# Isoform switching workflow

Having established the mathematical and biological foundations of **Tsallis entropy**, we now demonstrate a complete workflow for detecting and characterizing **isoform switching**—the coordinated changes in transcript usage patterns between experimental conditions. Isoform switching occurs when cells reorganize their isoform landscape without necessarily changing total gene abundance, often reflecting strategic shifts in protein function driven by splicing regulation. This section walks through a practical analysis pipeline: from raw transcript counts to identifying genes with significant scale-dependent isoform complexity changes (via linear model interaction testing) and dissecting the specific transcripts driving those changes (via jackknife-based influence assessment). We'll use a paired experimental design where treated and control samples are linked, enabling detection of robust biological signals while controlling for subject-level variability. By the end of this workflow, you'll be able to identify not just *which genes* undergo isoform switching, but *which transcripts* are responsible and *at which diversity scales* the switching occurs.

## Load data and metadata

An example dataset is included for demonstration.

```{r setup}
# Load packages
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
})
```

Now we will load the example dataset and associated metadata:

```{r data}
# Load example dataset (lazy-loaded as 'readcounts' by default)
# This includes SALMON preprocessing outputs:
# - salmon_dataset: transcript-level NumReads (raw fragment counts)
# - salmon_tpm: TPM matrix (length- and library-normalized estimates)
# - salmon_effective_length: EffectiveLength vector (read-length corrected)
data(readcounts)

# Ensure numeric format (keep transcript IDs as rownames)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Load sample metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

# Get the GFF3.gz annotation file path
gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")
```

## Data preprocessing and filtering

Next, we will create a `SummarizedExperiment` data container, a standard 
Bioconductor class for storing high-throughput assay data along with associated 
metadata. You can read more about it in the [Bioconductor documentation]
(https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html).

We will use the GFF3.gz annotation file for extracting the transcript-to-gene 
mapping and building the SummarizedExperiment. We'll also include SALMON preprocessed 
data (TPM and EffectiveLength) for improved analysis:

```{r build-se}
## Build a `SummarizedExperiment` from readcounts + GFF3.gz annotation
# The gff3_dataset variable contains the path to annotation.gff3.gz
# SALMON preprocessed data (salmon_tpm, salmon_effective_length) is auto-detected
# Also map sample metadata for proper pairing detection in filtering
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
```

The `SummarizedExperiment` object contains transcript-level counts with gene annotations extracted from the GFF3 file. Let's examine its internal structure:

```{r examine-se-structure, include = FALSE}
# Display the structure of the SummarizedExperiment
cat("SummarizedExperiment dimensions:\n")
cat("  Transcripts (rows):", nrow(se), "\n")
cat("  Samples (columns):", ncol(se), "\n")
cat("  Assay (counts):", nrow(assay(se, "counts")), "×", ncol(assay(se, "counts")), "\n")
cat("\nRowData columns:", paste(colnames(rowData(se)), collapse = ", "), "\n")
cat("Sample metadata available:", exists("metadata_df"), "\n\n")

# Show rowData with gene information
rd <- as.data.frame(rowData(se)[1:5, ])
knitr::kable(rd, caption = "First 5 rows of rowData (gene information)", row.names = FALSE)
```

```{r examine-metadata, include=FALSE}
# Display sample metadata (before mapping to colData)
if (exists("metadata_df")) {
  cat("\nSample metadata (metadata_df):\n")
  cat("  Rows:", nrow(metadata_df), "samples\n")
  cat("  Columns:", paste(colnames(metadata_df), collapse = ", "), "\n\n")
  knitr::kable(metadata_df, caption = "Sample metadata (metadata_df)")
}
```

Before computing the Tsallis diversity, you might want to
filter out genes with a low overall expression or limit 
the analysis to transcripts with a sufficient minimum 
expression level. Expression estimates of transcript isoforms 
with zero or low expression might be highly variable.
For more details on the effect of transcript isoform
prefiltering on differential transcript usage, see 
[this paper](https://doi.org/10.1186/s13059-015-0862-3).

```{r readfilter}
## Filter lowly-expressed transcripts and report counts
# Use stringency = "medium" for balanced filtering:
# - Keeps transcripts expressed in ≥50% of samples
# - Uses median TPM threshold (Q2) as minimum abundance cutoff
# - Ensures reliable entropy estimation across genes
# Stringency options: "soft" (permissive), "medium" (balanced), "severe" (stringent)
se <- filter_se(se, stringency = "medium")
```

The `stringency = "medium"` parameter automatically determines filtering thresholds
based on your paired study design. It keeps transcripts present in at least 50% of 
samples with TPM values above the median of mean gene TPM. This balanced approach 
filters out low-abundance transcripts that contribute noise while retaining sufficient 
isoform diversity for meaningful entropy calculations.

## Compute Tsallis entropy

Now we will compute Tsallis entropy.

```{r tsallis-calc-sequence}
# compute Tsallis entropy for a sequence of values (normalized)
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(se,
    q = qvec, norm = TRUE, metadata = metadata_df
)
```

```{r kable-tsallis-calc-sequence, echo = FALSE}
# Create table with gene names instead of gene IDs
assay_data <- assay(ts_se)
ts_rowdata <- rowData(ts_se)
se_rowdata <- rowData(se)

# Select first 4 genes and first 3 columns
table_data <- assay_data[1:4, 1:3]

# Map gene IDs to gene names from original se
# First, get the gene IDs from ts_se rowData
ts_gene_id_col <- if ("gene_id" %in% colnames(ts_rowdata)) "gene_id" else if ("genes" %in% colnames(ts_rowdata)) "genes" else rownames(ts_se)[1:4]
if (is.character(ts_gene_id_col) && length(ts_gene_id_col) == 1) {
    gene_ids <- ts_rowdata[[ts_gene_id_col]][1:4]
} else {
    gene_ids <- rownames(ts_se)[1:4]
}

# Create mapping from se rowData
se_gene_id_col <- if ("gene_id" %in% colnames(se_rowdata)) "gene_id" else if ("genes" %in% colnames(se_rowdata)) "genes" else NULL
se_gene_name_col <- if ("gene_name" %in% colnames(se_rowdata)) "gene_name" else NULL

if (!is.null(se_gene_id_col) && !is.null(se_gene_name_col)) {
    gene_id_to_name <- setNames(
        as.character(se_rowdata[[se_gene_name_col]]),
        as.character(se_rowdata[[se_gene_id_col]])
    )
    # Remove duplicates, keeping first occurrence
    gene_id_to_name <- gene_id_to_name[!duplicated(names(gene_id_to_name))]
    
    row_labels <- unname(gene_id_to_name[as.character(gene_ids)])
    # Use gene IDs if names not found
    row_labels[is.na(row_labels)] <- as.character(gene_ids)[is.na(row_labels)]
} else {
    row_labels <- as.character(gene_ids)
}

rownames(table_data) <- row_labels

knitr::kable(
    table_data,
    caption = "Tsallis entropy for first 4 genes across q values",
    digits = 5,
    row.names = TRUE
)
```

Map once again the sample metadata into the multi-q 
`SummarizedExperiment` as before.

With the **q-spectrum** we can produce a q-curve per sample and gene.
These curves show how **diversity** emphasis shifts from rare to dominant
isoforms as `q` increases and form the basis for interaction tests. The
q-curve shows entropy as a function of `q`. Diverging curves between
groups indicate **scale-dependent** **diversity** differences: separation at low
`q` implies differences in rare isoforms, while separation at high `q`
signals differences in dominant isoforms.

```{r tsallis-q-curve, fig.width=10, fig.height=5.2}
# Plot overall q-curve (mean entropy across all genes)
p_qcurve <- plot_tsallis_q_curve(ts_se)
print(p_qcurve)
```

## Quality Control: Sample Influence Assessment

Before proceeding with group-level comparisons, we should assess whether individual samples exert disproportionate influence on our entropy estimates. To address this, we employ a leave-one-out influence assessment combined with robust M-estimation (iteratively re-weighted least squares with Huber loss). This approach quantifies how much each sample's removal affects the estimated location differences across the q-spectrum, providing a sample-level quality control metric independent of group assignment.

Samples with high influence scores exert disproportionate leverage on entropy estimates and may warrant deeper investigation for technical quality issues, subject-specific outliers, or genuine biological signals that require careful interpretation.

```{r qc-sample-check-multiQ, echo=TRUE}
# Perform multi-q sample influence analysis
# The m_estimate function handles:
# - Multi-q data collapsing (averaging/median across q values)
# - Leave-one-out sample influence assessment
# - Flagging samples based on influence_threshold (default: 0.75 = 75th percentile)
# Returns: Sample names, influence scores, and QC status

sample_qc <- m_estimate(ts_se, samples = "sample_type", 
                        loss_type = "huber", 
                        q_combine_method = "mean",
                        influence_threshold = 0.75,
                        paired = TRUE)
```

```{r qc-sample-display, echo=FALSE, results="asis"}
# Display QC results with condition, pair, influence, and centroid distance
base_display_cols <- c("Sample", "Condition", "Proportion_Affected", 
                       "Genes_Affected", "Distance_from_Centroid", "Status")

# Initialize with base columns
qc_table <- sample_qc[, base_display_cols]

# Add Pair_ID if available (check multiple possible column names)
pair_id_col <- NULL
if ("Pair_ID" %in% colnames(sample_qc)) {
  pair_id_col <- "Pair_ID"
} else if ("Pair" %in% colnames(sample_qc)) {
  pair_id_col <- "Pair"
} else if ("pair_id" %in% colnames(sample_qc)) {
  pair_id_col <- "pair_id"
}

if (!is.null(pair_id_col)) {
  qc_table <- sample_qc[, c("Sample", "Condition", pair_id_col, 
                            "Proportion_Affected", "Genes_Affected", 
                            "Distance_from_Centroid", "Status")]
  colnames(qc_table)[which(colnames(qc_table) == pair_id_col)] <- "Pair"
}

# Round numeric columns for better readability
qc_table$Proportion_Affected <- round(qc_table$Proportion_Affected, 3)
qc_table$Genes_Affected <- round(qc_table$Genes_Affected, 1)
qc_table$Distance_from_Centroid <- round(qc_table$Distance_from_Centroid, 4)

# Display as formatted table with automatic width and alignment
knitr::kable(
    qc_table,
    caption = "Sample Influence Assessment via M-estimation",
    digits = 4,
    row.names = FALSE,
    align = c("l", rep("c", ncol(qc_table) - 1))
)
```

## Linear-Model interaction

Once we have, we can test for interactions between `q` and
sample groups across a sequence of `q` values (`q-sequence`), that is,
whether the shape of the q-curve differs between groups.
This approach is a powerful way to detect scale-dependent
diversity differences that may not be apparent at a
single `q` value.

Compute **Tsallis entropy** across a sequence of `q` values 
is important because `q` acts as a sensitivity lens that 
shifts emphasis between rare and dominant events in data.
Thus,  the resulting q-curve can help us identify at which 
scales (i.e. for which `q` values) the **diversity** differences
between groups are most pronounced. The Tsallis `q-sequence` can 
also help with quality control and outlier detection. By 
examining the q-curves of individual samples, we can 
identify samples that deviate from expected patterns,
which may indicate technical issues or biological outliers.


Once we have the **q-spectrum** computed, we can test for interactions
between `q` and sample groups using one of four methods implemented
in `calculate_lm_interaction()`:

- linear: fits a simple linear model. This is
fast and interpretable and is appropriate when the q–response is
approximately linear and samples are independent.
- lmm: fits a linear mixed model which accounts
for subject-level random effects (paired or repeated measures).
- gam: fits a generalized additive model. This method allows
to capture nonlinear q–response shapes.
- fpca: uses a functional-data approach to reduce each gene's
q-curve to principal component scores. This is recommended when q
is densely sampled or when q-curves exhibit complex, high-dimensional
variation.

Below we demonstrate a practical `gam` (Generalized Additive Model) example.
This method flexibly captures nonlinear q-response patterns and is particularly
useful when the relationship between q and entropy is not linear (for example,
the q:group interaction with complex response shapes).

```{r lm-interaction-before-qcurve, message=FALSE}
## Linear-model interaction test across q values: detect q x group interaction
## with Hochberg FWER control (fast, closed-form procedure)
if (requireNamespace("mgcv", quietly = TRUE)) {
    # Compute and show top hits (by adjusted p-value)
    # Using Hochberg for FWER control via closed-form procedure
    lm_res <- calculate_lm_interaction(ts_se,
        method = "gam",
        paired = TRUE,
        multicorr = "hochberg",
        nthreads = 3,
        verbose = TRUE
    )
}
```
```{r kable-lm-interaction-before-qcurve, message=FALSE, echo = FALSE}
# show the top entries with proper p-value formatting
lm_res_display <- head(lm_res, 6)

# Select only key columns that exist in the data, including Shapiro-Wilk normality test results
col_select <- intersect(c("gene", "gene_name", "p_interaction", "adj_p_interaction", 
                          "family_used", "shapiro_p_value", "residuals_normal",
                          "n_observations", "n_subjects", "arima_transformation"), 
                        colnames(lm_res_display))
lm_res_display <- lm_res_display[, col_select]

# Use gene_name from lm_res (already populated by calculate_lm_interaction)
# If gene_name is available, display it; otherwise use gene (ID)
if ("gene_name" %in% colnames(lm_res_display) && "gene" %in% colnames(lm_res_display)) {
    lm_res_display$gene <- lm_res_display$gene_name
    lm_res_display$gene_name <- NULL  # Remove redundant column
}

# Select key columns for display
display_cols <- c("gene", "p_interaction", "adj_p_interaction", "family_used", "shapiro_p_value", "residuals_normal", "n_observations", "n_subjects", "arima_transformation")
lm_res_display <- lm_res_display[, intersect(display_cols, colnames(lm_res_display))]

# Format p-values in scientific notation to show actual values
if ("p_interaction" %in% colnames(lm_res_display)) {
    lm_res_display$p_interaction <- format(lm_res_display$p_interaction, scientific = TRUE, digits = 3)
}
if ("adj_p_interaction" %in% colnames(lm_res_display)) {
    lm_res_display$adj_p_interaction <- format(lm_res_display$adj_p_interaction, scientific = TRUE, digits = 3)
}
if ("shapiro_p_value" %in% colnames(lm_res_display)) {
    lm_res_display$shapiro_p_value <- format(lm_res_display$shapiro_p_value, scientific = TRUE, digits = 3)
}

knitr::kable(
    lm_res_display,
    caption = "Top q-interaction results from GAM analysis with conditional family selection (Gaussian/Gamma), Hochberg correction, and Shapiro-Wilk normality tests",
    digits = 5,
    row.names = FALSE
)
```

### Interpreting the Results: Understanding P-Values, Family Selection, and Residual Normality

The statistical analysis results in the table above provide several key outputs:

- **gene**: Gene identifier for each analysis result.

- **p_interaction**: The p-value from the GAM F-test comparing the null model (group + q main effects) to the alternative model (group × q interaction). Small p-values (< 0.05) indicate significant evidence that condition effects vary across the diversity scale q.

- **adj_p_interaction**: Adjusted p-value using Hochberg stepup multiple testing correction to control family-wise error rate across all genes.

- **family_used**: The statistical family selected for GAM modeling. **Gaussian(link='identity')** is the default for well-behaved entropy data, while **Gamma(link='log')** is selected when strong heteroscedasticity or boundary clustering is detected. The conditional selection logic ensures appropriate family choice for each gene's data characteristics.

- **residuals_normal**: A Boolean indicator (TRUE/FALSE) showing whether the GAM residuals pass the Shapiro-Wilk normality test at α = 0.05. When FALSE, residuals deviate from normality, which may occur due to:
  - Bounded entropy values (0 to 1 range) creating non-normal tails
  - Outliers or heteroscedasticity
  - Smaller sample sizes amplifying departure from normality
  
- **shapiro_p_value**: The exact p-value from the Shapiro-Wilk test. Values much less than 0.05 indicate strong evidence of non-normality.

**Note on robustness**: GAM models using Gaussian families are relatively robust to moderate violations of normality, especially with larger sample sizes. The TSENAT analysis uses the Gaussian family for entropy modeling to properly account for the effects of isoform heterogeneity structure. Even when residuals show non-normality, the p-values remain statistically valid under the GAM likelihood framework.

Now we will plot the q-curve profile for the top gene identified 
by the linear-model interaction test. We will start with `SNHG10`.   

```{r foxj2-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for the top 2 genes from lm_res with the most significant interactions
# Pass lm_res directly to the function to let it extract genes by significance
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "SNHG10")
print(plot_target)
```

Now let’s look at Now let’s look at FBXO3.

```{r pnpt1-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for top Now let’s look at FBXO3
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "FBXO3")
print(plot_target)
```


Finally, we can visualize the gene `PDE7A`.

```{r gsr-gene-qprofile, fig.width=10, fig.height=6.5}
# Plot q-curve profile for top PDE7A
plot_target <- plot_tsallis_gene_profile(ts_se, gene = "PDE7A")
print(plot_target)
```


## Transcript Switching Across Diversity Scales

The LM interaction test revealed that condition effects vary across the
diversity scale (q parameter). Here we assess whether these switching
patterns are consistent across different q values or vary by scale. This
bridges the gap between the multi-q LM test and the single-q jackknife
analysis.

```{r multi-q-switching-validation, echo=TRUE, results='hide'}

# Multi-Q analysis: Test switching at different q values
q_vector <- c(0.01, 0.5, 1.0, 1.5, 2.0)

# Single call for all q values at once (vectorized approach)
# Note: jackknife_isoform_switching now handles gene name to ID mapping automatically
multi_q_results <- jackknife_isoform_switching(
  se = se,
  condition_col = "sample_type",
  pair_col = "paired_samples",
  gene_col = "gene_id",
  isoform_col = "transcript_id",
  q = q_vector,
  norm = TRUE,
  n_bootstrap = 100,
  print_results = TRUE,
  threshold = 90,
  lm_results = lm_res,
  lm_p_threshold = 0.05,
  use_lm_fdr = TRUE
)

```

The tables below examine how transcript switching patterns vary across different diversity scales (q-values) for the four genes with the strongest q×condition interactions. The delta influence metric quantifies the difference in transcript importance between the two conditions: positive values indicate a transcript is more influential under the first condition, negative values indicate greater importance in the second condition. By comparing these values across q-values (0.01 to 2.0), we identify whether transcript switching is consistent across all scales or scale-dependent. The "Direction Consistency" column classifies each transcript's behavior: "Consistent positive/negative" indicates robust, scale-independent switching; "Mixed directions" indicates scale-dependent switching where the same transcript switches direction depending on the diversity scale analyzed. This analysis reveals whether isoform shifts are driven by the same transcripts across all q-values or whether different q-values emphasize switching in different transcripts.


```{r multi-q-switching-table, echo=TRUE, results='hide'}
# Use the helper function to prepare comparison tables across q-values
# Limiting to top 2 genes with strongest switching patterns
tables_result <- prepare_gene_switching_tables(
  lm_res = lm_res,
  multi_q_results = multi_q_results,
  n_top_genes = 2,
  verbose = FALSE
)
```

```{r multi-q-switching-display, echo=FALSE, results='asis'}

# Iterate over the prepared comparison tables and display top 2 genes
for (gene_idx in 1:min(2, length(tables_result$gene_headers))) {
  cat("#### Gene: ", tables_result$gene_headers[gene_idx], "\n\n", sep="")
  
  comparison_data <- tables_result$comparison_tables[[gene_idx]]
  q_metadata <- tables_result$q_metadata[[gene_idx]]
  
  if (!is.null(comparison_data) && !is.null(q_metadata)) {
    # Create display labels for column names (for knitr::kable)
    q_values_available <- q_metadata$q_values_available
    col_display <- c(
      "Transcript",
      sapply(q_values_available, function(qk) {
        q_num <- q_metadata$q_key_to_value[[qk]]
        paste0("q=", sprintf("%.2f", q_num))
      }),
      "",
      "Direction Consistency"
    )
    
    print(knitr::kable(comparison_data, digits = 3, col.names = col_display))
  }
  
  cat("\n")
}
```

### Delta Influence Across Diversity Scales

Visualize switching patterns for the top genes identified by the LM
interaction test. This shows which transcripts are switching in genes 
with significant q×condition interaction effects. The heatmaps below display
the **jackknife** delta influence (a measure of transcript switching robustness)
across different q-values (rows: 0.01 to 2.0, representing increasing diversity
scales) and transcripts (columns) for each gene.

```{r multi-q-heatmap, echo=TRUE, fig.width=18, fig.height=20}
# Generate multi-Q heatmaps using the dedicated function
# The function extracts all needed information from the multi_q_results object
heatmap_file <- NA

tryCatch({
  heatmap_file <- plot_multiq_delta_influence_heatmaps(
    switching_results = multi_q_results,
    n_genes = 4,
    lm_results = lm_res
  )
}, error = function(e) {
  cat("ERROR in plot_multiq_delta_influence_heatmaps: ", e$message, "\n")
})

# Display the generated heatmap if valid data was found
if (is.character(heatmap_file) && length(heatmap_file) > 0 && !is.na(heatmap_file)) {
  if (file.exists(heatmap_file)) {
    knitr::include_graphics(heatmap_file)
  } else {
    cat("⚠️  File was not created or path is invalid: ", heatmap_file, "\n")
  }
} else {
  cat("⚠️  No valid heatmap data generated.\n")
  cat("   Check the validation report above for details on which genes lack delta_influence data.\n")
  cat("   Possible causes:\n")
  cat("   - multi_q_results doesn't contain delta_influence values\n")
  cat("   - lm_res has no significant genes to display\n")
  cat("   - Data structure mismatch\n")
}
```

Interpretation:

- Rows: q-values from low (0.01) to high (2.0) diversity, showing scale-dependent effects
- Columns: Each transcript for the gene
- Red cells: Transcript more influential in the first condition
- Blue cells: Transcript more influential in the second condition
- Column patterns: Consistent color down a column indicates stable switching; varied colors indicate q-dependent switching


### Effect size analysis

To understand which diversity scales show the most pronounced biological differences 
between groups, we compute effect size metrics (Tsallis divergence D_q) across the q-spectrum while respecting the paired sample design. This reveals whether group differences are driven by rare 
isoforms (low q), dominant isoforms (high q), or uniformly across scales.

The approach uses linear mixed-effects regression (LMM) with q as continuous predictor to assess 
slope differences in entropy change across q-spectrum between treatment groups, with effect size computed 
as Tsallis divergence D_q(control||treatment) [@I002; @I003; @I004]. This information-theoretic measure 
automatically respects Tsallis entropy properties and adapts to each q value.

Theoretical Justification: **Tsallis entropy** theory [@I002, @I004] demonstrates that 
different q values reveal fundamentally different aspects of the isoform distribution. 
As documented in the foundational divergence literature [@I004], D₀.₅(P||Q) often differs 
substantially from D₂(P||Q), with q=0.5 emphasizing rare variants and q=2 emphasizing 
abundant isoforms. Computing effect sizes independently at each q value thus captures 
which diversity scales (rare vs. dominant isoforms) show the strongest biological 
differentiation between conditions.

Effect Size Interpretation: **Tsallis divergence** D_q quantifies the information-theoretic distance between 
two distributions' entropy patterns across the q-spectrum. For paired designs, divergence is computed 
separately for each pair and then averaged, accounting for within-pair correlation and reducing noise from 
between-pair variation. The mean divergence across q values serves as the effect size, automatically 
capturing how entropy distributions differ between control and treatment groups at all scales (rare to 
abundant isoforms). Values D > 0.1 indicate meaningful divergence, following information-theoretic 
thresholds. Genes with D > 0.1 demonstrate significant information-theoretic separation between conditions, 
revealing that isoform complexity patterns (captured by multi-q Tsallis entropy) fundamentally differ between 
treatment groups across all q-dependent scales.

```{r effect-size-multiq, message=FALSE, warning=FALSE, results='hide'}
# Effect size multi-q plot
if (requireNamespace("mgcv", quietly = TRUE) && exists("lm_res") && nrow(lm_res) > 0) {
  # Compute divergence effect sizes from raw transcript counts
  # calculate_divergence handles transcript-to-gene aggregation internally
  # and matches genes from lm_res against rowData gene_names column
  
  # Note: Set rownames of lm_res to gene names for later use in effect_sizes_divergence()
  rownames(lm_res) <- lm_res$gene
  
  # Calculate Tsallis divergence with bootstrap CIs (reduced iterations for speed)
  # divergence_results_se: SummarizedExperiment with divergence estimates for each gene-q combination
  divergence_results_se <- calculate_divergence(
      se = se,
      group_col = "sample_type",
      control_group = "normal",
      q = qvec,
      paired = TRUE,
      bootstrap = TRUE,
      nboot = 50,
      ci = 0.95,
      method = "percentile"
  )
  
  # Note: effect_sizes_divergence() will merge lm_res results with divergence_results_se,
  # automatically filtering lm_res to match genes in divergence_results_se
}
```


```{r divergence-assay-display, echo = FALSE}
# Display the raw divergence computation results (first 4 genes across 3 q values)
if (exists("divergence_results_se") && nrow(divergence_results_se) > 0) {
    div_assay <- assay(divergence_results_se)
    n_rows <- min(4, nrow(div_assay))
    n_cols <- min(3, ncol(div_assay))
    
    if (n_rows > 0 && n_cols > 0) {
        # Create table with gene names (rownames of divergence_results_se are already gene names)
        div_table <- div_assay[1:n_rows, 1:n_cols]
        
        # The rownames are set to gene_name by calculate_divergence(), so use them directly
        # No need for mapping since they're already gene names
        knitr::kable(
            div_table,
            caption = "Tsallis divergence estimates for first genes across q values",
            digits = 5,
            row.names = TRUE
        )
    }
}
```

```{r effect-size-merge, message=FALSE, warning=FALSE, results='hide'}
# Merge LMM results with divergence effect sizes
# The function automatically enriches results with per_q_pattern column
eff_res <- effect_sizes_divergence(
    lm_res = lm_res,
    divergence_results_se = divergence_results_se,
    significance_threshold = 0.05,
    enrich_per_q_pattern = TRUE,
    verbose = FALSE
)
```

```{r divergence-results-table, results='asis', echo = FALSE}
# Display divergence results for top genes by LM interaction significance
# Show top 5 genes sorted by adjusted p-value from LM interaction test
# NOTE: These estimates are specifically for q=1.0 (Shannon entropy / Kullback-Leibler divergence)
# The full per-q divergence spectrum (q=0.1 to 2.0) is available in the assay matrix
if (exists("divergence_results_se") && exists("lm_res")) {
    div_rd <- SummarizedExperiment::rowData(divergence_results_se)
    
    # Get divergence data with gene names
    div_df <- as.data.frame(div_rd)[!is.na(div_rd$estimate), 
                                     c("estimate", "lower_ci", "upper_ci"), 
                                     drop = FALSE]
    div_df$gene_name <- rownames(div_rd)[!is.na(div_rd$estimate)]
    
    # Merge with lm_res to get LM p-values
    lm_res_sub <- lm_res[, c("gene", "adj_p_interaction")]
    colnames(lm_res_sub) <- c("gene_name", "padj_lm")
    div_df <- merge(div_df, lm_res_sub, by = "gene_name", all.x = TRUE)
    
    # Sort by LM adjusted p-value (ascending = most significant first)
    div_df <- div_df[order(div_df$padj_lm, na.last = TRUE), ]
    
    # Get top 5 genes
    div_df <- head(div_df, 5)
    
    if (nrow(div_df) > 0) {
        # Reorder columns to put gene_name first
        div_df <- div_df[, c("gene_name", "estimate", "lower_ci", "upper_ci", "padj_lm")]
        # Format p-values in scientific notation
        div_df$padj_lm <- format(div_df$padj_lm, scientific = TRUE, digits = 3)
        # Format columns for display
        colnames(div_df) <- c("Gene", "KL Divergence (D_q=1.0)", "Lower CI", "Upper CI", "LM adj. p-value")
        
        knitr::kable(
            div_df,
            caption = "Top 5 genes by LM interaction significance with Kullback-Leibler divergence estimates and bootstrap 95% confidence intervals",
            digits = 5,
            row.names = FALSE
        )
    }
}
```

The effect sizes reveal which genes show the most information-theoretic separation 
across the q-spectrum **between paired treatment groups**. Genes with Tsallis divergence D > 0.1 show 
substantial divergence, indicating that entropy distributions differ fundamentally between conditions 
across the full q-spectrum.

**Biological Interpretation:** Following Tsallis entropy theory and information-theoretic principles [@I002; @I003; @I004], 
genes with large D values are those where isoform complexity distributions differ qualitatively between conditions 
when examined at all sensitivity scales (q = rare → abundant isoforms). For example, a gene might show 
high entropy at low q (emphasizing rare isoforms) in one condition but low entropy at high q (emphasizing 
abundant isoforms) in another, revealing scale-dependent regulatory mechanisms. These genes are candidates 
for investigation of condition-specific splicing architecture and dynamic isoform switching.

### Visualization Functions


```{r effect-size-plot-multiq, fig.width=10, fig.height=6}
# Visualize the distribution of Tsallis divergence effect sizes across genes
# The red dashed line marks D = 0.1 (information-theoretic significance threshold)
p <- plot_divergence_distribution(eff_res$interaction_results, 
    threshold = 0.1)
print(p)
```

### Classify Gene Pattern Type

Once you have per-q divergence values from `effect_sizes_divergence()` (which merges LMM p-values with divergence estimates), you can classify each gene's isoform switching mechanism:

```{r classify-pattern-compute, message=FALSE}
# Extract classification table directly from eff_res
# Function automatically handles NULL input and filters for valid per-q patterns
classification_tbl <- extract_classification_table(eff_res, top_n = 10, sort_by = 'padj')
```

```{r classify-pattern-table, results='asis', message=FALSE, echo = FALSE}
# Display classification table if valid genes are present
if (nrow(classification_tbl) > 0 && any(!is.na(classification_tbl$Pattern))) {
  print(knitr::kable(
    classification_tbl,
    caption = "Classification of gene q-patterns and per-q divergence values (top 10 genes)",
    digits = 4,
    row.names = FALSE
  ))
}
```

**Interpretation of Pattern Types:**

| Pattern Type | Signature | Biological Meaning |
|--------------|-----------|-------------------|
| Rare driven | D(q=0.5) >> D(q=2.0) | Low-abundance isoforms are condition-specific; treatment group preferentially expresses rare transcripts not seen in control |
| Abundant driver | D(q=0.5) << D(q=2.0) | High-abundance isoforms shift between conditions; treatment remodels the dominant transcript landscape without rare variants changing |
| Balanced | D(q=0.5) ≈ D(q=2.0) | All isoforms shift proportionally; no preferential weighting to rare or abundant forms; suggests systematic rebalancing |

### Visualize Q-Spectrum Curve

The q-spectrum is a curve showing how divergence varies across Tsallis sensitivity parameters. 
This curve is the full biological signature of **isoform switching**—the pattern encodes which 
abundance scales (rare vs. abundant) are affected.

```{r visualize-q-spectrum, fig.width=14, fig.height=6, results='hide'}
# Generate q-spectrum plot for SNHG10 using the new plot_gene_q_spectrum function
# This gene was identified as one of the top candidates by the LMM interaction test
plot_gene_q_spectrum(divergence_results_se, target_gene = "SNHG10", verbose = FALSE)
```

**Interpreting the Q-Spectrum Curve:**

- Shape matters more than single value: The **q-spectrum** curve is the complete biological story. 
  A flat curve (BALANCED) vs. declining curve (RARE_DRIVEN) vs. rising curve (ABUNDANT_DRIVEN) 
  encode fundamentally different mechanisms.

- q=1 (KL divergence): The middle point corresponds to ordinary Kullback-Leibler divergence, 
  which weights all isoforms equally. This is the "average" effect.

- q < 1 (rare isoforms): Emphasizes how much low-abundance transcripts differ between 
  groups. If D(q=0.5) >> D(q=1), the divergence is driven by changes in rare variants.

- q > 1 (abundant isoforms): Emphasizes dominant transcripts. If D(q=2) >> D(q=1), major 
  isoforms rebalance while rare ones stay similar.

- Bootstrap CI bands: Shaded region shows 95% confidence limits on per-q estimates. Narrow 
  bands indicate robust estimates; wide bands suggest noisy data or small sample sizes.

**Advanced Visualization: Multiple Gene Comparison**

```{r compare-q-spectra-plot, fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
# Compare q-spectra across multiple significant genes
p_multi <- plot_multi_gene_q_spectrum(
  eff_res = eff_res, 
  n_genes = 6, 
  ncol = 3,
  verbose = TRUE
)
print(p_multi)
```

**Statistical Validation During Interpretation:**

1. Bootstrap CI validity: Per-q CIs should narrow as q increases (higher q = aggregation 
   effect, less variance). If CIs grow with q, check for data quality issues.

2. Monotonicity check: By **Tsallis entropy** theory, entropy is monotone decreasing in q. Divergence 
   should NOT show erratic increases with q. Small fluctuations are normal, but large spikes 
   indicate numerical instability.

3. **Comparison with genome-wide patterns**: Compute q-spectra for housekeeping genes (GAPDH, 
   ACTB, etc.). These should show BALANCED patterns. If not, revisit normalization parameters.


## References

1. Adami, C. (2004). "Information Theory in Molecular Biology." *arXiv Preprint q-Bio/0405004*.

2. Bajić, D. (2024). "Information Theory, Living Systems, and Communication Engineering." *Entropy*, 26(5), 430. https://doi.org/10.3390/e26050430.

3. Bartal, A., & Jagodnik, K. M. (2022). "Progress in and Opportunities for Applying Information Theory to Computational Biology and Bioinformatics." *Entropy*, 24(7), 925. https://doi.org/10.3390/e24070925.

4. Chanda, P., Costa, E., Hu, J., Sukumar, S., Van Hemert, J., & Walia, R. (2020). "Information Theory in Computational Biology: Where We Stand Today." *Entropy*, 22(6), 627. https://doi.org/10.3390/e22060627.

5. Cover, T. M., & Thomas, J. A. (2006). *Elements of Information Theory* (2nd ed.). Wiley-Interscience.

6. Dérian, N., Pham, H.-P., Nehar-Belaid, D., et al. (2022). "The Tsallis Generalized Entropy Enhances the Interpretation of Transcriptomics Datasets." *PLOS ONE*, 17(4), e0266618. https://doi.org/10.1371/journal.pone.0266618.

7. Furuichi, S. (2006). "Information Theoretical Properties of Tsallis Entropies." *Journal of Mathematical Physics*, 47, 023302. https://doi.org/10.1063/1.2165744.

8. Gandrillon, O., Gaillard, M., Espinasse, T., et al. (2021). "Entropy as a Measure of Variability and Stemness in Single-Cell Transcriptomics." *Entropy*, 21(5), 450.

9. Golomb, R., Yoles, M., Fishilevich, S., et al. (2026). "An Information Content Principle Explains Regulatory Patterns of Gene Expression Across Human Tissues." *bioRxiv*, ahead of print. https://doi.org/10.64898/2026.02.19.706555.

10. Hinostroza Fuentes, V. G., Toledo Tan, M. J., Kapetanaki, M., Rashidi, P., Huang, K., & Benos, P. V. (2026). "Information-Theoretic Methods in Spatial Transcriptomics." *TechRxiv*, ahead of print. https://doi.org/10.36227/techrxiv.177155947.

11. Jost, L. (2006). "Entropy and Diversity." *Oikos*, 113(2), 363–375. https://doi.org/10.1111/j.2006.0030-1299.14714.x.

12. Rényi, A. (1961). "On Measures of Entropy and Information." *Proceedings of the Fourth Berkeley Symposium on Mathematical Statistics and Probability*, 1, 547–561.

13. Seweryn, M. T., Pietrzak, M., & Ma, Q. (2020). "Application of Information Theoretical Approaches to Assess Diversity and Similarity in Single-Cell Transcriptomics." *Computational and Structural Biotechnology Journal*, 18, 1830–1837. https://doi.org/10.1016/j.csbj.2020.05.006.

14. Shannon, C. E. (1948). "A Mathematical Theory of Communication." *The Bell System Technical Journal*, 27(3–4), 379–423.

15. Simpson, E. H. (1949). "Measurement of Diversity." *Nature*, 163, 688. https://doi.org/10.1038/163688a0.

16. Tsallis, C. (1988). "Possible Generalization of Boltzmann-Gibbs Statistics." *Journal of Statistical Physics*, 52(1), 479–487. https://doi.org/10.1007/BF01016429.

## Session Information

```{r session-info}
sessionInfo()
```
