---
title: "TSENAT Appendix B: Comparing Flexible Parametric (GAM) and Rank-Based (K-W) Methods"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: GAM vs K-W Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Flexible Parametric and Rank-Based Methods

## Purpose

This vignette validates q-value × group interaction detection using two complementary statistical approaches:

1. **Flexible Parametric (GAM):** Generalized Additive Models with ARIMA(1,1,0) transformation for handling q-ordered measurements
2. **Rank-Based (K-W):** Kruskal-Wallis with adaptive test selection based on data characteristics

Both methods now incorporate principled improvements to handle real-world data challenges:
- Heteroscedasticity detection and weighting
- Boundary condition handling
- Automatic test selection based on data properties
- Multiple testing correction accounting for AR(1) correlation in q-values

## Key Questions

1. Do both methods identify the same significant genes?
2. How much do results concordant vs. discordant?
3. Which genes are truly robust (significant in both methods)?
4. What data characteristics drive method disagreement?

---

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Subset for faster analysis (set to NULL for all genes)
n_genes_subset <- 200
if (!is.null(n_genes_subset) && nrow(readcounts) > n_genes_subset) {
  readcounts <- readcounts[1:n_genes_subset, ]
  cat("NOTE: Analyzing", n_genes_subset, "genes for speed. Set n_genes_subset=NULL for all genes.\n\n")
}

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build SummarizedExperiment and filter
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
se <- tryCatch(
  filter_se(se, stringency = "loose"),
  error = function(e) se
)

```

## Bayesian Pseudocount Regularization

This analysis incorporates **WLFC (Weighted Likelihood Fold Change)** empirical Bayesian regularization to handle sparse count data. The `estimate_wlfc_pseudocounts()` function:

1. **Fits an empirical Beta prior** from the observed count data across all genes
2. **Computes gene-specific WLFC pseudocounts** based on each gene's posterior distribution
3. **Returns a scalar pseudocount** (mean of per-gene estimates) for entropy regularization

This data-driven approach is more principled than fixed pseudocounts, as it learns the regularization strength from the actual data characteristics.

```{r bayesian-pseudocounts, message=FALSE}
# Estimate WLFC pseudocounts integrated workflow
result <- estimate_wlfc_pseudocounts(se, verbose = FALSE)
scalar_pseudocount <- result$scalar_pseudocount

# Calculate Tsallis entropy with Bayesian pseudocount regularization
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(
  se, 
  q = qvec, 
  norm = TRUE, 
  pseudocount = scalar_pseudocount,
  metadata = metadata_df,
  shrinkage = "empirical_bayes"
)

```


# Part 1: Rank-Based Approach (Kruskal-Wallis)

```{r kw-analysis, message=FALSE}
cat("\n=== RANK-BASED APPROACH (KRUSKAL-WALLIS) ===\n\n")

# Convert SummarizedExperiment to long-format data frame for rank-based analysis
entropy_matrix <- assay(ts_se)
ts_coldata <- colData(ts_se)
test_genes <- rownames(ts_se)
n_genes <- length(test_genes)
n_cols <- ncol(ts_se)

entropy_long <- data.frame(
    entropy = as.numeric(entropy_matrix),
    gene = rep(test_genes, n_cols),
    q = rep(ts_coldata$q, each = n_genes),
    stringsAsFactors = FALSE
)

# Run rank-based interaction test with conditional test selection
# Note: detect_q_gene_interactions automatically selects appropriate rank-based tests
# (Kruskal-Wallis, ART, Mood's Median) based on data characteristics
kw_results <- detect_q_gene_interactions(
    entropy_long,
    entropy_col = "entropy",
    q_col = "q",
    gene_col = "gene",
    multicorr = "hochberg",
    verbose = FALSE
)
```

```{r kw-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect_kw <- mean(kw_results$effect_size_eta2, na.rm = TRUE)
median_effect_kw <- median(kw_results$effect_size_eta2, na.rm = TRUE)

kw_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "NAs",
    "Mean effect size (η²)",
    "Median effect size (η²)",
    "Strong effect genes (η² > 10%)"
  ),
  Value = c(
    nrow(kw_results),
    sum(kw_results$p_value < 0.05, na.rm = TRUE),
    sum(is.na(kw_results$p_value)),
    if (is.na(mean_effect_kw) || !is.finite(mean_effect_kw)) "Not available" else sprintf("%.1f%%", mean_effect_kw * 100),
    if (is.na(median_effect_kw) || !is.finite(median_effect_kw)) "Not available" else sprintf("%.1f%%", median_effect_kw * 100),
    sum(kw_results$p_value < 0.05 & kw_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(kw_summary_stats, caption = "Rank-Based (K-W) Results Summary", align = c("l", "r"))

# Show top genes with effect size and test method
kw_sig <- kw_results[!is.na(kw_results$p_value), ]
if (nrow(kw_sig) > 0) {
  kw_sig <- kw_sig[order(kw_sig$p_value), ]
  kw_sig <- kw_sig[1:min(10, nrow(kw_sig)), ]
  
  # Select columns: gene, p-values, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(kw_sig))
  if (length(cols_to_show) > 0) {
    kw_display <- kw_sig[, cols_to_show, drop = FALSE]
    
    # Format p-values in scientific notation for readability
    for (col in c("p_value", "adj_p_value")) {
      if (col %in% colnames(kw_display)) {
        kw_display[[col]] <- sapply(kw_display[[col]], function(x) {
          if (is.na(x)) "NA" else sprintf("%.3e", x)
        })
      }
    }
    
    # Format effect size as percentage if available
    if ("effect_size_eta2" %in% colnames(kw_display)) {
      kw_display$effect_size_eta2 <- sapply(kw_display$effect_size_eta2, function(x) {
        if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
      })
    }
    
    # Improve column distribution: right-align numeric columns, left-align text
    # gene | p-value | adj_p | eta2 | test_method | interaction_class
    knitr::kable(kw_display, 
                 caption = "Top 10 Genes by K-W p-value (with Effect Size & Test Method)",
                 align = c("l", "r", "r", "r", "l", "l"),
                 format.args = list(big.mark = ","))
  }
}
```

---

# Part 2: Flexible Parametric Approach (GAM)

GAM analysis is computationally intensive:

- Per-gene fitting: Each gene gets a smooth GAM model fitted across all q-values.
- Model complexity: Generalized Additive Models use iterative smoothing optimization.
- Scale: With ~500 genes × 39 q-values = 19,500 model evaluations.
- Multiple testing correction: Additional computation for Hochberg adjustment.

Expected runtime: 2-5 minutes for 500 genes with 3 threads. For faster iteration during development, reduce `n_genes_subset` in the setup chunk.

```{r gam-analysis, message=FALSE}
# Flexible parametric approach (GAM)

gam_results_raw <- calculate_lm_interaction(
    ts_se,
    method = "gam",
    paired = TRUE,
    multicorr = "hochberg",
    nthreads = 3,
    verbose = FALSE
)

# Convert to data frame robustly
gam_results <- as.data.frame(gam_results_raw, stringsAsFactors = FALSE)
```

```{r gam-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"))

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format p-values in scientific notation
      for (col in c("p_interaction", "adj_p_interaction")) {
        if (col %in% colnames(gam_display)) {
          gam_display[[col]] <- sapply(gam_display[[col]], function(x) {
            if (is.na(x)) "NA" else sprintf("%.3e", x)
          })
        }
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size & Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"))
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

---



# Part 3: Method Concordance Analysis

```{r concordance-summary, message=FALSE}
# Create concordance summary statistics (show code and calculations)
# Check if we have valid results from both methods
comparison_df <- NULL
spearman_rho <- NA
high_conf <- NULL
agreement_table <- NULL

if (!("p_interaction" %in% colnames(gam_results)) || 
    !("p_value" %in% colnames(kw_results))) {
  cat("Note: Skipping concordance analysis - missing required columns\n")
} else {
  # Merge results using base R
  # Create matching sets - only genes with values in both methods
  gam_genes <- gam_results$gene[!is.na(gam_results$p_interaction)]
  kw_genes <- kw_results$gene[!is.na(kw_results$p_value)]
  common_genes <- intersect(gam_genes, kw_genes)
  
  if (length(common_genes) > 2) {
    # Extract matching rows
    gam_idx <- match(common_genes, gam_results$gene)
    kw_idx <- match(common_genes, kw_results$gene)
    
    comparison_df <- data.frame(
      gene = common_genes,
      p_gam = gam_results$p_interaction[gam_idx],
      padj_gam = gam_results$adj_p_interaction[gam_idx],
      effect_gam = if ("effect_size" %in% colnames(gam_results)) gam_results$effect_size[gam_idx] else NA,
      p_kw = kw_results$p_value[kw_idx],
      padj_kw = if ("adj_p_value" %in% colnames(kw_results)) kw_results$adj_p_value[kw_idx] else NA,
      effect_kw = if ("effect_size_eta2" %in% colnames(kw_results)) kw_results$effect_size_eta2[kw_idx] else NA,
      stringsAsFactors = FALSE
    )
    
    # Calculate correlation
    spearman_rho <- cor(comparison_df$p_gam, comparison_df$p_kw, 
                        method = "spearman", use = "complete.obs")
    
    # Categorize agreement
    comparison_df$gam_sig <- comparison_df$p_gam < 0.05
    comparison_df$kw_sig <- comparison_df$p_kw < 0.05
    comparison_df$agreement <- ifelse(
      comparison_df$gam_sig & comparison_df$kw_sig, "Both significant",
      ifelse(comparison_df$gam_sig & !comparison_df$kw_sig, "GAM only",
             ifelse(!comparison_df$gam_sig & comparison_df$kw_sig, "K-W only",
                    "Neither significant"))
    )
    
    # Calculate agreement table
    agreement_table <- table(comparison_df$agreement)
    
    # High-confidence genes (both methods)
    high_conf <- comparison_df[comparison_df$gam_sig & comparison_df$kw_sig, ]
    high_conf <- high_conf[order(pmax(high_conf$p_gam, high_conf$p_kw)), ]
  }
}
```

```{r concordance-output, message=FALSE, echo=FALSE}
if (is.null(comparison_df)) {
  # Silently skip - no output
} else {
  # Create agreement summary table
  agreement_stats <- data.frame(
    Category = names(agreement_table),
    Count = as.numeric(agreement_table),
    Proportion = as.numeric(prop.table(agreement_table)),
    stringsAsFactors = FALSE
  )
  
  # Format proportions as percentages
  agreement_stats$Proportion <- sprintf("%.1f%%", agreement_stats$Proportion * 100)
  
  knitr::kable(agreement_stats, 
               caption = "Agreement Summary between GAM and K-W Methods",
               align = c("l", "r", "r"))
  
  cat("\n")
  
  # High-confidence genes table
  if (nrow(high_conf) > 0) {
    high_conf_display <- high_conf[1:min(15, nrow(high_conf)), 
                                   c("gene", "padj_gam", "padj_kw", "effect_gam", "effect_kw")]
    rownames(high_conf_display) <- NULL
    
    # Format p-values
    high_conf_display$padj_gam <- sapply(high_conf_display$padj_gam, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3e", x)
    })
    high_conf_display$padj_kw <- sapply(high_conf_display$padj_kw, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3e", x)
    })
    
    # Format effect sizes
    high_conf_display$effect_gam <- sapply(high_conf_display$effect_gam, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3f", x)
    })
    high_conf_display$effect_kw <- sapply(high_conf_display$effect_kw, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3f", x)
    })
    
    # Rename columns for display
    colnames(high_conf_display) <- c("Gene", "GAM padj", "K-W padj", "GAM Effect", "K-W Effect")
    
    knitr::kable(high_conf_display, 
                 caption = "Top High-Confidence Genes Detected by Both Methods",
                 align = c("l", "r", "r", "r", "r"))
  }
}
```

## Visualization: Method Comparison

```{r concordance-plot, fig.width=10, fig.height=8}
# Create comparison plot only if we have valid comparison data
if (exists("comparison_df") && nrow(comparison_df) > 0) {
  
  # Create comparison plot
  p1 <- ggplot(comparison_df, aes(x = -log10(p_gam), y = -log10(p_kw), color = agreement)) +
    geom_point(size = 2.5, alpha = 0.6) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
    scale_color_manual(
      values = c("Both significant" = "#2ecc71", 
                 "GAM only" = "#3498db",
                 "K-W only" = "#e74c3c",
                 "Neither significant" = "#95a5a6"),
      breaks = c("Both significant", "GAM only", "K-W only", "Neither significant")
    ) +
    labs(
      title = "GAM vs Kruskal-Wallis: Method Concordance",
      x = "-log10(p-value, GAM)",
      y = "-log10(p-value, K-W)",
      color = "Significance"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      legend.position = "bottomright",
      panel.grid.major = element_line(color = "gray90")
    )
  
  # P-value distribution comparison
  p_long <- data.frame(
    p_value = c(comparison_df$p_gam, comparison_df$p_kw),
    method = c(rep("GAM", nrow(comparison_df)), rep("K-W", nrow(comparison_df)))
  )
  
  p2 <- ggplot(p_long, aes(x = p_value, fill = method)) +
    geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
    scale_fill_manual(values = c("GAM" = "#3498db", "K-W" = "#e74c3c")) +
    labs(
      title = "P-value Distributions",
      x = "P-value",
      y = "Frequency",
      fill = "Method"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12))
  
  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  cat("Visualization skipped: No valid comparison data available\n")
}
```

---

# Part 5: Recommendations

## Best Practices for Analysis

We recommend a **complementary two-method approach** for robust detection of q-value × group interactions:

**Primary analysis:** Start with GAM (flexible parametric method) as it offers superior power for detecting nonlinear q-entropy patterns. Use the default Hochberg multiple testing correction for stringent FWER control. **Validation step:** Confirm findings with rank-based Kruskal-Wallis analysis to ensure robustness against parametric assumptions. The conditional test selection automatically adapts to data characteristics (heteroscedasticity, boundary clustering, skewness), providing distribution-free validation without requiring assumptions about data normality.

**High-confidence genes** are those significant in *both* methods—these provide the strongest evidence of genuine q-value × group interactions and are recommended for downstream biological validation. Additionally, use **Bayesian pseudocount regularization (WLFC)** for robust entropy estimation on sparse count data, and leverage **multiple testing correction that accounts for AR(1) correlation** in q-value sequences (available via Hochberg or Benjamini-Yekutieli methods depending on whether you prioritize FWER or FDR control).

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References
