---
title: "TSENAT: Tsallis Entropy Analysis Toolbox"
author: "Cristobal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
  rmarkdown::pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
    latex_engine: xelatex
header-includes:
  - \usepackage{float}
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    echo = TRUE,
    eval = TRUE,
    results = 'markup',
    message = TRUE,
    warning = TRUE
)
suppressPackageStartupMessages(library(kableExtra))
```

# Overview

TSENAT (Tsallis Entropy Analysis Toolbox) quantifies **isoform heterogeneity** in RNA-seq data using Tsallis entropy-a scale-dependent diversity measure distinct from abundance-focused tools (DESeq2, Salmon) and differential transcript usage packages (*SplicingFactory*). By tuning a sensitivity parameter `q`, users examine diversity at different scales: rare variants (low `q`) or dominant isoforms (high `q`). This package enables computing Tsallis entropy from transcript-level abundance estimates, comparing measures between groups, and visualizing scale-dependent differences via q-curves. Results integrate seamlessly with Bioconductor's `SummarizedExperiment`, capturing scale-dependent isoform complexity that often signals cellular state changes independent of total abundance shifts.

## Motivation and Bioconductor Contribution

Common RNA-seq tools focus on either total gene abundance changes or individual transcript usage shifts. Yet genes often remodel their isoform landscapes without changing overall expression-a phenomenon missed by these approaches. TSENAT quantifies this *isoform complexity* directly via Tsallis entropy, a scale-dependent diversity framework tuned by parameter `q`. By sliding `q` across scales, researchers zoom between rare variants (low `q`) and dominant isoforms (high `q`), capturing biological signal invisible to abundance- or proportion-based summaries.

In this guide, we demonstrate the complete workflow: preprocessing transcript counts, computing entropy across q-values, testing for between-group differences, and visualizing scale-dependent complexity. All results integrate seamlessly with Bioconductor's `SummarizedExperiment`, making TSENAT a natural complement to existing DTU and abundance-focused tools in the ecosystem.



## High-level workflow

1. **Load and preprocess** transcript counts using `build_analysis_s4()`, then filter low-abundance transcripts with `filter_analysis_s4()`.
2. **Select q-values** tuned to your research question: low q (rare isoforms) vs. high q (dominant isoforms). Compute Tsallis entropy across these scales with `calculate_diversity_s4()`.
3. **Test for differences** in entropy between groups using `rank_test_q_condition_s4()` (Wilcoxon or permutation tests) [@K002; @K003].
4. **Visualize results** as q-curves and inspect transcript counts for genes with the strongest entropy shifts using `plot_tsallis_q_curve_s4()` and related functions.

**Design assumptions**: This guide assumes paired or longitudinal designs with >=6-8 samples per group for adequate power to detect entropy shifts while controlling false discovery rate. Smaller sample sizes may be underpowered to detect subtle isoform complexity changes.


## Installation

To install `TSENAT` from Bioconductor:

```{r install-package, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "http://cran.us.r-project.org")
BiocManager::install("TSENAT")
```

Then load the package:

```{r load-tsenat-package, eval = FALSE}
library(TSENAT)
```

## Quick Start

**⚠️ Important Parameter Note**: The `build_analysis_s4()` function has an optional `salmon_dir` parameter that comes before the required `tx2gene` parameter. Always use **named parameters** (`readcounts =`, `tx2gene =`, etc.) when calling this function. Using positional arguments may cause the annotation file to be misinterpreted as a Salmon directory, resulting in an error.

For users who want to see results quickly, here is a minimal workflow:

```{r quick-start-example, eval = FALSE}
library(TSENAT)
library(SummarizedExperiment)

# Load example data
data("readcounts", package = "TSENAT")
readcounts <- as.matrix(readcounts)

# Load sample metadata and annotation
metadata_df <- read.table(
  system.file("extdata", "metadata.tsv", package = "TSENAT"),
  header = TRUE, sep = "\t"
)
gff3_file <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Create analysis object with transcript counts, annotation, and metadata
# IMPORTANT: Use named parameters to avoid confusion with optional salmon_dir parameter
analysis <- build_analysis_s4(
  readcounts = readcounts,
  tx2gene = gff3_file,
  metadata = metadata_df
)

# Filter low-abundance transcripts
analysis <- filter_analysis_s4(analysis, stringency = "medium")

# Compute Tsallis entropy using S4 wrapper (using single q value for quick start)
analysis <- calculate_diversity_s4(analysis, q = 1, norm = TRUE)
```

## Data Structure Overview

**Input Data**: TSENAT expects transcript-level read counts (rows = transcripts, columns = samples) from quantification tools like SALMON or kallisto. A GFF3 annotation file maps transcripts to genes.

**Processing**: The `build_analysis_s4()` function accepts these inputs and creates a **TSENATAnalysis** S4 object, which is the central data container throughout your analysis.

**Organization**: The TSENATAnalysis object encapsulates:

- `@se` - `SummarizedExperiment` storing counts and metadata (access via `se(analysis)`)
- `@diversity_results` - Entropy values across q-values (accessor: `diversity(analysis)`)
- `@jackknife_results` - Jackknife confidence intervals and isoform switching results
- `@lm_results` - LM interaction statistics and rank test q-value effects (accessor: `lmResults(analysis)`)
- `@divergence_results` - Pairwise divergence metrics (accessor: `divergence(analysis)`)
- `@plots` - Generated visualizations

Use **accessor functions** like `diversity()`, `divergence()`, `lmResults()`, etc. for type-safe result extraction without direct slot manipulation.

For more details on the SummarizedExperiment class, see [SummarizedExperiment documentation](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html).

# What is Entropy and Tsallis Entropy?

## The Problem: Measuring Isoform Complexity

Standard RNA-seq analysis measures *whether* transcript abundance changes between conditions. But a critical complementary question remains underexplored: *how* does the diversity of isoforms change? A gene may show little change in total abundance while dramatically reshuffling its isoform repertoire-a phenomenon that current methods largely miss.

**Entropy** quantifies precisely this: the complexity, richness, and balance of isoform heterogeneity. By measuring entropy across different biological scales, researchers can detect whether changes are driven by shifts in rare variants or reorganization of dominant isoforms.

## Tsallis Entropy: A Scale-Dependent Diversity Measure

**Tsallis entropy** [@I002] is a one-parameter family of diversity measures that generalizes Shannon entropy [@I051]. Unlike Shannon entropy (which treats all isoforms equally), Tsallis entropy lets you tune a sensitivity parameter `q` to zoom into different aspects of isoform complexity:

- **q < 1**: Emphasizes rare, low-abundance isoforms (discovery mode)
- **q = 1**: Recovers Shannon entropy (balanced across all abundance scales)
- **q = 2**: Emphasizes dominant isoforms (robustness mode)
- **q > 2**: Focuses almost exclusively on the most abundant species

This is the key innovation: by computing entropy across a range of q-values (a "q-curve"), you obtain a complete picture of isoform heterogeneity [@I003]. In TSENAT, you'll explore multiple q-values using `calculate_diversity_s4()` and visualize results with `plot_tsallis_q_curve_s4()`.

## Mathematical Definition

For a discrete probability vector $p = (p_1, \ldots, p_n)$ representing isoform proportions within a gene, Tsallis entropy is:

\[
S_q(p) = \frac{1-\sum_{i=1}^n p_i^q}{q-1}.
\]

This parametric family unites diverse entropy concepts under a single framework:

- **Generalization**: Extends beyond Shannon entropy to capture scale-dependent phenomena
- **Mathematical elegance**: Reduces to well-known diversity indices at specific q-values
- **Practical flexibility**: Enables data-driven exploration across the full diversity spectrum

## Information-Theoretic Interpretation

From an information theory perspective [@I051; @I002], entropy measures the uncertainty or surprise when drawing a single transcript from an isoform distribution. Higher entropy means the draw is less predictable (many similarly abundant isoforms), while lower entropy means one or a few isoforms dominate. This distinction is crucial: two genes with identical total abundance may have dramatically different isoform complexity.

The q parameter acts as a **sensitivity dial**. Consider a simple example:
- A gene with 5 equally abundant isoforms: all q-values give high entropy
- A gene where 1 isoform dominates: low-q entropy stays high (rare variants matter), while high-q entropy drops (rare variants ignored)

This scale-dependent nature reveals biological signal invisible to abundance- or proportion-based summaries alone.

### Special Cases and Limiting Behavior

Tsallis entropy exhibits important special cases [@I013] that appear when you set specific q-values:

- **q = 0 (Richness)**: $S_0 = m - 1$ (number of expressed isoforms). Pure species count, most minimal assumption.
- **q = 1 (Shannon)**: $S_1(p) = -\sum_i p_i \log p_i$. Standard information entropy [@I051]; optimal estimator of information content.
- **q = 2 (Gini-Simpson)**: $S_2 = 1 - \sum_i p_i^2$. Probability that two randomly drawn transcripts differ [@I022]; robust to rare variants.

**Normalized entropy**: For genes with different isoform counts, normalize by maximum possible entropy: $\tilde{S}_q = \frac{S_q(p)}{S_{q,\max}(m)}$ where $S_{q,\max}(m) = \frac{1-m^{1-q}}{q-1}$. This enables fair comparison across genes.

## Biological Potentiality: Why TSENAT Matters

Tsallis entropy is grounded in Shannon's foundational information theory [@I051], generalized by Renyi's family of entropies [@I003], and extended by Tsallis [@I002]. The key innovation is a tunable parameter `q` that controls sensitivity to different aspects of the diversity distribution-which aspects of isoform heterogeneity become visible depends on the q-value chosen [@I005; @I007; @I009].

As stated explicitly in the mathematical literature [@I013]: "This introduces the formal possibility not to set rare and common events on the same footing, as in BG or Shannon statistics, but it enhances or depresses them according to the parameter chosen." This principle is formalized in the Hill numbers framework [@I021], which unifies diverse entropy-based measures (richness, Shannon, Simpson) as different manifestations of the same parametric family with varying sensitivity to abundance scales.

The flexibility is crucial for isoform analysis. The concept of "true diversity" [@I021] emphasizes that diversity can be decomposed into two independent components: species richness (how many distinct isoforms exist) and evenness (how evenly distributed they are across the population). Two genes can have identical Shannon entropy yet differ dramatically in isoform structure: one might be dominated by a single abundant isoform (maximizing apparent heterogeneity at low `q` where rare variants matter), while another distributes transcripts across many isoforms equally (maximizing heterogeneity across all `q` values). By tuning the q-parameter, we operationalize what Masi (2005) describes: depressing the weight on rare events (high q) reveals dominant isoform patterns, while enhancing rare-event weight (low q) reveals cryptic isoform complexity. TSENAT's multi-*q* approach captures this full spectrum of isoform diversity, enabling researchers to detect both rare isoform innovations and robust isoform usage patterns [@C113] by exploring the q-curve across multiple scales.

### Potential Applications in Genomics

The multi-scale nature of Tsallis entropy makes it suited for exploring isoform complexity across diverse biological contexts. By measuring information content at different q-values, researchers can detect patterns invisible to traditional transcript abundance measures alone:

**Isoform complexity as a biological signal**: Isoform switching-reorganization of the isoform landscape without necessarily changing total gene abundance-reflects strategic shifts in protein function driven by splicing regulation. Evidence from single-cell transcriptomics demonstrates that transcript-level complexity varies systematically across cell types and developmental states [@ISO010], validating that isoform heterogeneity is a genuine biological phenomenon rather than noise. Increased entropy in gene regulatory networks drives phenotypic heterogeneity and cellular plasticity [@Y006], suggesting that transcript-level entropy captures similar organizational principles. TSENAT enables detection of these changes through entropy-based approaches, which capture whether complexity is increasing (diversity spreading across isoforms) or decreasing (consolidation onto dominant isoforms).

**Scale-dependent organization**: Different biological processes may prioritize different scales of organization [@Y005]:
- Changes in rare isoform usage (revealed through low `q` sensitivity) might reflect exploratory or error-correction mechanisms
- Shifts in dominant isoform selection (revealed through high `q` sensitivity) might reflect functional specialization or robustness demands
- The full q-curve reveals whether cellular transitions involve wholesale reorganization or targeted adjustments, paralleling how systems biology approaches dissect biological disorder and organization at multiple scales

**Beyond abundance measures**: Traditional analysis focuses on fold-changes and differential abundance. Since isoform reorganization can occur independently of total abundance changes, entropy-based approaches complement classical methods by detecting complexity shifts that transcript-level statistics alone cannot reveal.

#### Evidence from Literature

The recent emphasis on information-theoretic approaches in computational biology [@I029, @I034] reflects broader recognition that complex biological systems encode information across multiple organizational scales. Peer-reviewed literature provides strong empirical support for these applications:

**Entropy and cancer biology**: Cancer cells accumulate genetic and epigenetic perturbations that systematically increase disorder in gene regulatory networks. As Tarabichi and colleagues demonstrate, "Increased entropy of signaling (or gene interaction networks) has been well studied as a cancer characteristic: Network entropy increases along with cancer progresses" [@Y005]. Systems biology approaches reveal that this entropy increase, rather than being an incidental feature, actively drives cancer progression through selection of cells with greater network flexibility and adaptability.

**Perturbation-driven mechanisms of heterogeneity**: Nijman's analysis of perturbation-driven entropy proposes a complementary mechanism: "cancer-associated perturbations collectively disrupt normal gene regulatory networks by increasing their entropy. Importantly, in this model both somatic driver and passenger alterations contribute to 'perturbation-driven entropy', thereby increasing phenotypic heterogeneity and evolvability" [@Y006]. This framework elegantly explains observed cancer heterogeneity without requiring that every genetic change confers a selective advantage-some mutations contribute entropy directly through network disruption.

**Single-cell validation of transcript diversity**: Cao and colleagues' landmark single-cell transcriptomics study provided empirical validation that transcript-level organization varies systematically across cell types: "expression levels of mRNA species are linked to cellular function and therefore can be used to classify cell types" [@ISO010]. Their comprehensive profiling of C. elegans demonstrates that individual cells maintain specific, consistent isoform compositions reflecting cellular identity-establishing that isoform complexity is not noise but a fundamental aspect of cellular differentiation and function.

These three perspectives-entropy as driver of cancer evolution, perturbations as network disruption mechanisms, and single-cell evidence for systematic isoform organization-converge on a framework where measuring entropy at multiple scales (via TSENAT's multi-*q* approach) captures biologically meaningful variation in cellular organization and adaptation. In cancer biology specifically, entropy-driven mechanisms explain how perturbations increase heterogeneity and plasticity [@Y006], demonstrating that entropy concepts have mechanistic relevance beyond abstract information theory. Tsallis entropy, through the parametric q-spectrum, provides a systematic framework for exploring this multi-scale organization at the transcript and isoform level.


# Isoform Switching Workflow

We now demonstrate a complete workflow for detecting **isoform switching**-when cells reorganize their isoform landscape without necessarily changing total gene abundance. This often reflects strategic shifts in protein function driven by splicing regulation.

In this workflow, you will:

1. Load transcript counts and sample metadata
2. Identify genes with significant scale-dependent isoform complexity changes (using linear model interaction testing)
3. Assess which individual transcripts drive those changes (using jackknife resampling [@S232])
4. Interpret results in the context of paired experimental designs

We'll work with a paired experimental design where treated and control samples are linked, enabling detection of robust biological signals while controlling for subject-level variability. By the end, you'll know not just *which genes* undergo isoform switching, but *which transcripts* are responsible and *at which diversity scales* the switching occurs.

## Load data and metadata

An example dataset is included for demonstration.

```{r workflow-setup}
# Load packages
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
})

# Setup random seed for reproducibility
set.seed(42)

```

Now we will load the example dataset and associated metadata:

```{r prepare-example-data}
# Load example dataset (lazy-loaded as 'readcounts' by default)
# This includes SALMON preprocessing outputs:
# - readcounts: transcript-level NumReads (raw fragment counts)
# - tpm: TPM matrix (length- and library-normalized estimates)
# - effective_length: EffectiveLength vector (read-length corrected)
data(readcounts)

readcounts <- as.matrix(readcounts)

metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_file <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")
```

## Data preprocessing and filtering

Next, we will create a `TSENATAnalysis` object containing a `SummarizedExperiment` 
data container (a standard Bioconductor class for storing high-throughput assay data 
along with associated metadata). You can read more about it in the [Bioconductor documentation](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html).

We will use the GFF3.gz annotation file for extracting the transcript-to-gene 
mapping and building the analysis object. We'll also include sample metadata for 
improved analysis. Following Bioconductor best practices, we create the configuration 
FIRST, then pass it to the builder function (fail-fast principle):

```{r initialize-analysis-object}
## Configure analysis parameters first (best practice: fail-fast principle)
## This validates all parameters against metadata before object creation
config <- tsenat_config(
  condition_col = "condition",
  subject_col = "paired_samples",
  q_values = seq(0, 2, by = 0.05),
  nthreads = 2,
  paired = TRUE,
  control = "normal",
  metadata = metadata_df
)

## Build a complete `TSENATAnalysis` object from readcounts + GFF3.gz annotation
## Pass config at construction (Bioconductor pattern): immutable object creation
analysis <- build_analysis_s4(
  config = config,
  readcounts = readcounts, 
  tx2gene = gff3_file, 
  tpm = tpm,
  effective_length = effective_length
)

```

The `TSENATAnalysis` object contains a `SummarizedExperiment` with transcript-level 
counts and gene annotations extracted from the GFF3 file.

To reduce noise and improve statistical power, we filter out lowly-expressed transcripts. 
Expression estimates of transcript isoforms with zero or low expression might be highly 
variable [@S202; @S203]. For more details on the effect of transcript isoform prefiltering 
on differential transcript usage, see [this paper](https://doi.org/10.1186/s13059-015-0862-3).

```{r filter-validate-main}
analysis <- filter_analysis_s4(analysis, stringency = "medium")
cat("Retained", nrow(se(analysis)), "transcripts after filtering\n")
```

The `stringency = "medium"` parameter keeps transcripts present in >=50% of samples 
with TPM values above the median of mean gene TPM. This balances noise reduction with 
preservation of isoform diversity needed for meaningful entropy calculations.

## Compute Tsallis Entropy

We now compute Tsallis entropy across your configured q-spectrum. Diversity measures 
provide a comprehensive framework for assessing isoform heterogeneity at multiple scales [@I007; @I010].
We normalize entropy to [0,1] range (`norm = TRUE`) for comparable cross-q assessment.

```{r tsallis-entropy-computation}
# Compute diversity using S4 wrapper with bootstrap confidence intervals [@S232; @S115]
analysis <- calculate_diversity_s4(
    analysis,
    norm = TRUE
)
```

```{r display-tsallis-table, echo = FALSE}
# Extract diversity results for three specific q values using the diversity() accessor
# Show entropy values for ONE sample across three q values (0.05, 1, 2)
# Note: q=0 (richness) is computed deterministically outside bootstrap to avoid discrete data artifacts

q_values_to_show <- c(0, 0.5, 1, 1.5, 2)
sample_data <- NULL
first_sample_name <- NULL

for (q_val in q_values_to_show) {
    div_q <- diversity(analysis, q = q_val)
    
    if (!is.null(div_q)) {
        # Extract matrix from SummarizedExperiment
        if (is(div_q, "SummarizedExperiment")) {
            mat <- assay(div_q)
        } else if (is.matrix(div_q)) {
            mat <- div_q
        } else {
            next
        }
        
        # Get first 4 genes and first sample only
        first_sample_idx <- 1
        first_sample_name <- colnames(mat)[first_sample_idx]
        mat_subset <- mat[1:min(4, nrow(mat)), first_sample_idx, drop = FALSE]
        
        # Add values for this q with descriptive column names
        if (is.null(sample_data)) {
            sample_data <- data.frame(Gene = rownames(mat_subset))
        }
        
        col_name <- paste0("q=", sprintf("%.1f", q_val))
        sample_data[[col_name]] <- as.numeric(mat_subset)
    }
}

if (!is.null(sample_data) && nrow(sample_data) > 0) {
    knitr::kable(
        sample_data,
        caption = paste0("**Table 1:** Tsallis entropy for first 4 genes across three diversity scales (sample: ", 
                         first_sample_name, ")"),
        digits = 5,
        row.names = FALSE,
        format = if (knitr::is_latex_output()) "latex" else "html",
        longtable = knitr::is_latex_output(),
        linesep = ""
    ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      latex_options = c("scale_down")
    )
}
```


With the q-spectrum we can produce a q-curve per sample and gene.
These curves show how diversity emphasis shifts from rare to dominant
isoforms as `q` increases and form the basis for interaction tests. The
q-curve shows entropy as a function of `q`. Diverging curves between
groups indicate scale-dependent diversity differences: separation at low
`q` implies differences in rare isoforms, while separation at high `q`
signals differences in dominant isoforms.

```{r fig-1-isoform-diversity-profiles, fig.width=10, fig.height=5.2, out.width="98%", fig.pos="H", fig.cap = "**Figure 1:** Isoform diversity profiles across q-values. Lines show normalized Tsallis entropy (0-1) for each sample, blue control and red treatment."}
# Plot overall q-curve
p_qcurve <- plot_tsallis_q_curve_s4(
    se = analysis
)
print(p_qcurve)
```

## Quality Control: Sample Influence Assessment

Before proceeding with group-level comparisons, we should assess whether individual samples exert disproportionate influence on our entropy estimates [@S115]. To address this, we employ a leave-one-out influence assessment combined with robust M-estimation [@S019] (iteratively re-weighted least squares with Huber loss). This approach quantifies how much each sample's removal affects the estimated location differences across the q-spectrum, providing a sample-level quality control metric independent of group assignment.

Samples with high influence scores exert disproportionate leverage on entropy estimates and may warrant deeper investigation for technical quality issues, subject-specific outliers, or genuine biological signals that require careful interpretation.

We configure three key parameters: `loss_type = "huber"` employs Huber M-estimation (robust to outliers with bounded influence), `q_combine_method = "mean"` averages influence metrics across the entire q-spectrum for a single score per sample, and `influence_threshold = 0.75` designates the percentile cutoff above which samples are flagged (such that the top 25% most influential samples are reviewed).

```{r qc-entropy-outlier-detection, echo=TRUE}
# Perform multi-q sample influence analysis via S4 wrapper
analysis <- m_estimate_s4(
    analysis,
    loss_type = "huber",
    q_combine_method = "mean",
    influence_threshold = 0.75
)

# Extract results from metadata using accessor function
sample_qc <- metadata(analysis, "m_estimate_results")
```

```{r display-qc-outlier-results, echo=FALSE, results="asis"}
if (!is.null(sample_qc) && nrow(sample_qc) > 0) {
  # Round numeric columns for better readability
  sample_qc$Proportion_Affected <- round(sample_qc$Proportion_Affected, 3)
  sample_qc$Genes_Affected <- round(sample_qc$Genes_Affected, 1)
  sample_qc$Distance_from_Centroid <- round(sample_qc$Distance_from_Centroid, 4)
  
  # Select columns: exclude Robustness_Weight and Pair_ID
  cols_to_show <- c("Sample", "Condition", "Proportion_Affected", "Genes_Affected", 
                    "Entropy_Mean", "Entropy_SD", "Distance_from_Centroid", "Status")
  cols_to_show <- cols_to_show[cols_to_show %in% colnames(sample_qc)]
  
  knitr::kable(
      sample_qc[, cols_to_show],
      caption = "**Table 2 | Sample Influence Assessment via M-estimation.** Ranked by magnitude of resampling-based influence values. Columns: Sample identifier; mean and standard deviation of Tsallis entropy; distance from centroid; outlier status. M-estimation identifies influential samples for sensitivity analysis.",
      digits = 4,
      row.names = FALSE,
      format = if (knitr::is_latex_output()) "latex" else "html",
      longtable = knitr::is_latex_output(),
      linesep = ""
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    latex_options = c("scale_down")
  )
}
```

**Interpretation:** Samples with distance from centroid >1.5 or proportion of affected genes >0.85 are flagged for quality control review. Flagged samples may reflect genuine biological heterogeneity or technical artifacts requiring careful examination before proceeding with downstream analysis.

## Linear-Model interaction

We can test for interactions between `q` and sample groups across the q-sequence
using one of five methods: linear, lmm (linear mixed model [@C018; @S019]), gam (generalized additive model [@C038; @C073]), gee (generalized estimating equations for paired data), 
or fpca (functional principal component analysis). The gam method flexibly captures nonlinear 
q-response patterns and is particularly useful for complex interaction structures.

```{r test-q-condition-interaction}
# Linear-model interaction test across q values using S4 wrapper
if (requireNamespace("mgcv", quietly = TRUE)) {
    analysis <- calculate_lm_interaction_s4(
        analysis,
        method = "gam",
        multicorr = "hochberg"
    )
}
```

```{r display-interaction-results, echo = FALSE}
lm_res <- lmResults(analysis)$lm_interaction
if (!is.null(lm_res) && is.data.frame(lm_res) && nrow(lm_res) > 0) {
    display <- head(lm_res, 6)
    
    # Select only most informative columns
    key_cols <- c("gene", "gene_name", "p_interaction", "adj_p_interaction", 
                  "effect_size", "test_statistic", "model_converged", "heteroscedasticity_detected")
    key_cols <- key_cols[key_cols %in% colnames(display)]
    display <- display[, key_cols]
    
    # Round numeric columns (but NOT p-value columns to preserve significance)
    numeric_cols <- sapply(display, is.numeric)
    pval_cols <- grepl("^p_|^adj_p_", colnames(display))
    display[, numeric_cols & !pval_cols] <- round(display[, numeric_cols & !pval_cols], 4)
    
    # Format p-values to show more precision (scientific notation for very small values)
    for (col in colnames(display)) {
        if (is.numeric(display[[col]]) && grepl("^p_|^adj_p_", col)) {
            display[[col]] <- format(display[[col]], scientific = TRUE, digits = 3)
        }
    }
    
    knitr::kable(
        display,
        caption = "**Table 3 | Leading genes with scale-dependent condition effects from generalized additive models.** We display the 6 genes with lowest adjusted p-values (*q* < 0.05). Benjamini-Hochberg correction applied; columns show gene identifier, effect size, test statistic, convergence status, and heteroscedasticity detection. Methods: GAMs with *q* and condition as smooth predictors [@S190].",
        col.names = c("Gene", "Gene Name", "P-value", "Adj. P-value", "Effect Size", 
                      "Test Statistic", "Model Converged", "Heteroscedasticity"),
        digits = 4,
        row.names = FALSE,
        format = if (knitr::is_latex_output()) "latex" else "html",
        longtable = knitr::is_latex_output(),
        linesep = ""
    ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      latex_options = c("scale_down")
    )
}
```

**Interpretation:** TRUE in the Heteroscedasticity column indicates that the model satisfies homogeneity-of-variance assumptions across the q-spectrum; FALSE values suggest variance heterogeneity and warrant caution in result interpretation.

Now we will plot the q-curve profile for the top genes identified 
by the linear-model interaction test.

```{r fig-2-scale-dependent-genes, fig.width=14, fig.height=12, out.width="98%", fig.pos="H", fig.cap = "**Figure 2:** Scale-dependent interaction analysis. GAM-identified genes showing significant q$\\times$condition effects (Benjamini-Hochberg q < 0.05)."}
# Plot q-curve profiles for the top 4 genes using the S4 wrapper
combined_plot <- plot_lm_interaction_gam_s4(
    analysis,
    n_top = 4
)
print(combined_plot)
```

## Transcript Switching Across Diversity Scales

### Why Scale-Dependent Analysis Matters

Biological systems operate through processes at fundamentally different organizational scales. A single gene's expression can be viewed through different "lenses"-from the perspective of dominant isoforms (high *q*, emphasizing abundant variants) to rare isoforms that may carry functional significance in specific contexts (low *q*, emphasizing rare variants). The LM interaction test revealed that condition effects vary across these diversity scales (q parameter), reflecting how biological reorganization can manifest differently depending on which aspect of isoform complexity we examine.

The theoretical foundation for scale-dependent analysis comes from generalized entropy measures. Renyi's entropy family [@I003] introduced a continuous spectrum of diversity measures parameterized by a sensitivity index, which Tsallis later extended and popularized [@I009]. Unlike Shannon entropy (which represents a single point on this spectrum), Tsallis entropy provides a family of measures where the parameter *q* acts as a "magnifying glass"-adjusting what organizational scales become visible. Mathematically, different *q*-values weight rare versus abundant isoforms differently:
- Low *q* (0-0.5): Emphasized rare isoforms, revealing exploratory or conditional splicing programs
- Mid *q* (0.5-1.5): Balanced sensitivity across the full isoform distribution  
- High *q* (1.5-2.0): Emphasized abundant isoforms, revealing core splicing decisions

In biological systems, these different scales often correspond to different regulatory mechanisms. A transcript may be consistently rare across conditions (no switching at low *q*) while being stably abundant at high *q*. Detecting such scale-specific switching patterns reveals the mechanistic structure of splicing regulation [@Y005].

### Bridging Theory: Multi-*q* Analysis to Fine-Scale Jackknife Testing

The current analysis bridges two complementary approaches:

1. **Multi-*q* LM interaction test**: Tests whether the condition effect on divergence depends on *q* (overall pattern across diversity scales)

2. **Single-*q* jackknife resampling**: Tests which individual transcripts drive those patterns at each *q*-value, with robustness assessment

This two-stage approach combines hypothesis testing with robustness validation [@S115]. The LM test identifies *where* (at which *q*-values) significant switching occurs. The jackknife-based delta influence then identifies *which transcripts* contribute to those patterns, weighting confidence by operational stability rather than simple frequency.

### Interpreting Scale-Dependent Switching Patterns

Here we assess whether these switching patterns are consistent across different q values or vary by scale. The key question is: **Do the same transcripts switch across all diversity scales, or does the identity of switching transcripts depend on which scale we examine?**

- **Consistent switching**: The same transcripts show significant delta influence across *q*-values suggests a robust, scale-independent splicing shift
- **Mixed/scale-dependent switching**: Different transcripts show importance at different *q*-values suggests regulatory complexity where the mechanism differs by scale

This classification reveals whether isoform reorganization is a coordinated, global phenomenon (same transcripts across scales) or a layered process where different regulatory inputs dominate at different organizational scales.

```{r identify-isoform-switching-transcripts, echo=TRUE, results='hide'}
# Multi-Q analysis: Test switching at different q values
analysis <- jackknife_isoform_switching_s4(
    analysis,
    q = c(0, 0.5, 1, 1.5, 2),
    nboot = 100,
    lm_p_threshold = 0.05,
    threshold = 90
)
```

We configure three key parameters for the switching analysis: `nboot = 100` specifies 100 bootstrap resamples to estimate robust confidence intervals for delta influence values [@S115]; `lm_p_threshold = 0.05` filters genes to those with significant q$\times$condition interaction effects (p < 0.05) before assessing transcript switching; and `threshold = 90` designates transcripts with support $\geq$ 90% across bootstrap samples as robust switches. These parameters balance sensitivity (detecting switching) with specificity (avoiding false positives from noise).

The tables below examine transcript switching patterns for the top genes with strongest q$\times$condition interactions. Delta influence, derived from jackknife resampling [@S115], quantifies how each transcript's relative importance changes between conditions, weighted by stability across bootstrap iterations (positive = more influential in first condition; negative = second condition). The `Direction Consistency` column classifies each transcript: "Consistent" indicates stable switching across q-values, while "Mixed" indicates scale-dependent switching. This reveals whether isoform shifts involve the same transcripts across scales or whether different q-values emphasize different transcripts.

```{r extract-switching-results, echo=TRUE, results='hide'}
# Prepare comparison tables for top genes across q-values using S4 accessor
tables_result <- prepare_gene_switching_tables_s4(
    analysis, 
    n_top_genes = 2
)
```

```{r display-switching-table, echo=FALSE, results='asis'}
# Only display tables if tables_result was successfully created
if (exists("tables_result") && is.list(tables_result) && 
    !is.null(tables_result$gene_headers) && length(tables_result$gene_headers) > 0) {
  # Iterate over the prepared comparison tables and display top 2 genes
  for (gene_idx in 1:min(2, length(tables_result$gene_headers))) {
    cat("#### Gene: ", tables_result$gene_headers[gene_idx], "\n\n", sep="")
    
    comparison_data <- tables_result$comparison_tables[[gene_idx]]
    q_metadata <- tables_result$q_metadata[[gene_idx]]
    
    if (!is.null(comparison_data) && !is.null(q_metadata)) {
      # Create display labels for column names (for knitr::kable)
      q_values_available <- q_metadata$q_values_available
      col_display <- c(
        "Transcript",
        sapply(q_values_available, function(qk) {
          q_num <- q_metadata$q_key_to_value[[qk]]
          paste0("q=", sprintf("%.2f", q_num))
        }),
        "",
        "Direction Consistency"
      )
      
      print(knitr::kable(comparison_data, 
                         digits = 3, 
                         col.names = col_display, 
                         format = if (knitr::is_latex_output()) "latex" else "html", 
                         longtable = knitr::is_latex_output(), 
                         linesep = "") %>%
            kableExtra::kable_styling(
              bootstrap_options = c("striped", "hover", "condensed"),
              latex_options = c("scale_down")
            ))
    }
    
    cat("\n")
  }
} else {
  cat("Multi-Q switching tables not available; skipping this section.\n")
}
```

### Delta Influence Across Diversity Scales

Visualize switching patterns for the top genes identified by the LM interaction test. This shows which transcripts are switching in genes with significant q * condition interaction effects. The heatmaps below display **jackknife delta influence** [@S232]-a resampling-based measure of how robustly each transcript's relative contribution changes between conditions-across different q-values (rows: 0 to 2.0, representing increasing diversity scales from rare to abundant isoforms) and transcripts (columns) for each gene. Delta influence values are computed from bootstrap resampling iterations, providing robust, non-parametric estimates of transcript switching significance weighted by consistency across replicates.

```{r fig-3-transcript-jackknife-influence, echo=TRUE, results='hide', fig.width=12, fig.height=12, eval=TRUE, out.width="98%", fig.pos="H", fig.cap = "**Figure 3:** Transcript-level switching patterns via jackknife resampling. Rows q-values (0-2.0), columns transcripts. Red/blue indicates condition influence."}
# Generate multi-Q heatmaps for top 4 genes using S4 wrapper
plot_multiq_delta_influence_heatmaps_s4(
    analysis, 
    n_genes = 4
)
```

**Interpretation:** Rows represent q-values across the diversity spectrum (0 to 2.0, with low q emphasizing rare isoforms and high q emphasizing abundant isoforms). Red cells indicate transcripts with strong positive delta influence in the first condition (meaning their relative importance increased); blue indicates negative influence in the second condition (meaning their importance decreased). **Patterns to look for:**

- **Strong color intensity (bright red or blue)**: Robust switching signal, consistently observed across bootstrap resamples
- **Weak color intensity (pale/white)**: Uncertain or noisy switching signal, high variability across resamples
- **Consistent colors down a column**: That transcript switches stably across all q-values (scale-independent switching)
- **Varied colors down a column**: That transcript shows scale-dependent switching (different roles at rare vs. abundant scales)
- **Color differences across rows**: Different q-values emphasize different transcripts, revealing how regulatory mechanisms vary by topological scale

This robustness-weighted visualization reveals not just *which* transcripts switch, but *how reliably* they switch, allowing distinction between robust biological signals and artifacts of sampling variation.

```{r fig-4-transcript-abundance-heatmap, echo=TRUE, results='hide', fig.width=12, fig.height=9, eval=TRUE, out.width="98%", fig.pos="H", fig.cap = "**Figure 4:** Transcript abundance heatmap with hierarchical clustering. Rows are transcripts, columns are conditions. Blue low, red high expression."}
# Generate transcript abundance heatmap for top 4 transcripts using hierarchical clustering
# metric = "median" aggregates expression values across paired conditions
plot_top_transcripts_s4(
    analysis, 
    top_n = 4, 
    metric = "median"
)
```

**Interpretation:** Transcripts (rows) are hierarchically clustered by expression similarity. Red indicates high expression, blue indicates low expression. Distinct horizontal bands reveal condition-specific isoforms; uniform coloring suggests constitutive expression. Clustering reveals which transcripts co-vary in expression patterns across conditions.

### Effect size analysis

To understand which diversity scales show the most pronounced biological differences 
between groups, we compute effect size metrics (Tsallis divergence $D_q$) across the q-spectrum while respecting the paired sample design. This reveals whether group differences are driven by rare 
isoforms (low q), dominant isoforms (high q), or uniformly across scales.

The approach uses linear mixed-effects regression (LMM) with q as continuous predictor to assess 
slope differences in entropy change across q-spectrum between treatment groups, with effect size computed 
as Tsallis divergence $D_q$(control||treatment) [@K001; @I002; @I003; @I004; @I059]. This information-theoretic measure 
automatically respects Tsallis entropy properties and adapts to each q value [@I061].

**Theoretical Justification**

Tsallis entropy theory [@I002, @I004; @I058] demonstrates that 
different q values reveal fundamentally different aspects of the isoform distribution. 
As documented in the foundational divergence literature [@I004; @I045; @I059; @I060], $D_{0.5}(P||Q)$ often differs 
substantially from $D_{2}(P||Q)$, with q=0.5 emphasizing rare variants and q=2 emphasizing 
abundant isoforms. Computing effect sizes independently at each q value thus captures 
which diversity scales (rare vs. dominant isoforms) show the strongest biological 
differentiation between conditions [@I048; @I060]. Re and Azad (2014) demonstrated that Tsallis divergence generalizations improve discrimination of genomic sequences, validating the multi-scale divergence approach for biological applications [@I060].

**Effect Size Interpretation**

**Tsallis divergence** $D_q$ quantifies the information-theoretic distance between 
two distributions' entropy patterns across the q-spectrum [@I011; @I012; @I038; @I040; @I059]. For paired designs, divergence is computed 
separately for each pair and then averaged, accounting for within-pair correlation and reducing noise from 
between-pair variation [@I057]. The mean divergence across q values serves as the effect size, automatically 
capturing how entropy distributions differ between control and treatment groups at all scales (rare to 
abundant isoforms). Values D > 0.1 indicate meaningful divergence, following effect size classification 
thresholds [@K002; @I021; @I059]. Genes with D > 0.1 demonstrate significant information-theoretic separation between conditions, 
revealing that isoform complexity patterns (captured by multi-q Tsallis entropy) fundamentally differ between 
treatment groups across all q-dependent scales [@I060].

```{r effect-size-multiq}
# Calculate divergence: computes pairwise information-theoretic distance (Tsallis divergence) between conditions
# using bootstrap confidence intervals across all configured q-values from getConfig(analysis)$q_values
analysis <- calculate_divergence_s4(
    analysis)
```


```{r divergence-assay-display, echo = FALSE}
# Display the raw divergence computation results using S4 accessor method
if (exists("analysis")) {
    div_results <- divergence(analysis)
    
    # Extract divergence data - format depends on how it's stored in divergence_results slot
    if (is.matrix(div_results) && nrow(div_results) > 0) {
        div_table <- div_results[1:min(4, nrow(div_results)), 1:min(3, ncol(div_results))]
        
        knitr::kable(
            div_table,
            caption = "**Table 3 | Pairwise Tsallis divergence estimates.** Divergence (distance) between conditions from Tsallis entropy framework. Columns: gene identifier; pairwise comparison; divergence value; confidence interval (95%). Ranked by magnitude.",
            digits = 5,
            row.names = TRUE,
            format = if (knitr::is_latex_output()) "latex" else "html",
            longtable = knitr::is_latex_output(),
            linesep = ""
        ) %>%
        kableExtra::kable_styling(
          bootstrap_options = c("striped", "hover", "condensed"),
          latex_options = c("scale_down")
        )
    } else if (is.list(div_results) && length(div_results) > 0) {
        # If stored as list per q-value, extract first q-value results
        first_q_result <- div_results[[1]]
        if (is.matrix(first_q_result)) {
            div_table <- first_q_result[1:min(4, nrow(first_q_result)), 1:min(3, ncol(first_q_result))]
            
            knitr::kable(
                div_table,
                caption = "**Table 4 | Pairwise Tsallis divergence from list-based results.** Alternative representation of divergence metrics for leading genes. Columns: gene identifier; pairwise comparison; divergence value; uncertainty. Asymmetry ($D(A \\to B) \\ne D(B \\to A)$) reflects information-theoretic properties. Ordered by divergence magnitude.",
                digits = 5,
                row.names = TRUE,
                format = if (knitr::is_latex_output()) "latex" else "html",
                longtable = knitr::is_latex_output(),
                linesep = ""
            ) %>%
            kableExtra::kable_styling(
              bootstrap_options = c("striped", "hover", "condensed"),
              latex_options = c("scale_down")
            )
        }
    }
}
```

```{r effect-size-merge}
# Compute effect sizes: significance_threshold = 0.05 filters genes to those with q*condition interaction p < 0.05;
# enrich_per_q_pattern = TRUE classifies each gene by its divergence pattern (Rare-driven/Balanced/Abundant-driven)
analysis <- effect_sizes_divergence_s4(
    analysis,
    significance_threshold = 0.05,
    enrich_per_q_pattern = TRUE
)
```

```{r extract-divergence-results, echo=FALSE, results='hide'}
# Extract results from effect_sizes_divergence_s4 metadata
divergence_results_se <- NULL
lm_res_results <- NULL

if (exists("analysis") && !is.null(metadata(analysis, "effect_sizes_divergence"))) {
    # Extract from effect sizes results (which contains merged LM + divergence data)
    effect_res <- metadata(analysis, "effect_sizes_divergence")
    if (!is.null(effect_res$interaction_results)) {
        lm_res_results <- effect_res$interaction_results
    }
}

# Also extract raw divergence using S4 accessor method (optional)
if (exists("analysis")) {
    divergence_results_se <- divergence(analysis)
}
```

```{r divergence-results-table, results='asis', echo = FALSE}
# Display top divergence results from effect_sizes_divergence_s4
if (!is.null(lm_res_results) && nrow(lm_res_results) > 0) {
    tryCatch({
        # Check for required columns
        if ("gene" %in% colnames(lm_res_results) && 
            ("adj_p_interaction" %in% colnames(lm_res_results) || "p_value_interaction" %in% colnames(lm_res_results))) {
            
            # Determine which p-value column to use
            p_col <- if ("adj_p_interaction" %in% colnames(lm_res_results)) {
                "adj_p_interaction"
            } else if ("p_value_interaction" %in% colnames(lm_res_results)) {
                "p_value_interaction"
            } else {
                NULL
            }
            
            if (!is.null(p_col)) {
                # Find effect size columns (like effect_size_D_q0_1, effect_size_D_q1_0, etc.)
                effect_cols <- grep("^effect_size_D", colnames(lm_res_results), value = TRUE)
                
                if (length(effect_cols) > 0) {
                    # Build display table
                    div_df <- data.frame(
                        Gene = lm_res_results$gene,
                        p_adj = lm_res_results[[p_col]],
                        stringsAsFactors = FALSE
                    )
                    
                    # Add first (or mean of) effect sizes for display
                    if (length(effect_cols) > 0) {
                        # Calculate mean effect size across q values for each gene
                        div_df$Mean_Effect_Size <- rowMeans(
                            lm_res_results[, effect_cols, drop = FALSE], na.rm = TRUE
                        )
                    }
                    
                    # Add pattern if available
                    if ("per_q_pattern" %in% colnames(lm_res_results)) {
                        div_df$Q_Pattern <- lm_res_results$per_q_pattern
                        # If Q_Pattern contains pattern labels, they'll display as-is
                        # If numeric fallback was used (old format), it would show commas
                    }
                    
                    # Add pattern metrics if available (div_rare_median, div_abundant_median, q_ratio)
                    if ("div_rare_median" %in% colnames(lm_res_results)) {
                        div_df$div_rare_median <- lm_res_results$div_rare_median
                    }
                    if ("div_abundant_median" %in% colnames(lm_res_results)) {
                        div_df$div_abundant_median <- lm_res_results$div_abundant_median
                    }
                    if ("q_ratio" %in% colnames(lm_res_results)) {
                        div_df$q_ratio <- lm_res_results$q_ratio
                    }
                    
                    # Sort and select top 10
                    div_df <- div_df[order(div_df$p_adj, na.last = TRUE), ]
                    div_df <- head(div_df, 10)
                    
                    # Remove rows with NA p-values
                    div_df <- div_df[!is.na(div_df$p_adj), ]
                    
                    if (nrow(div_df) > 0) {
                        # Format p-values
                        div_df$p_adj <- format(as.numeric(div_df$p_adj), scientific = TRUE, digits = 2)
                        
                        # Format effect sizes
                        if ("Mean_Effect_Size" %in% colnames(div_df)) {
                            div_df$Mean_Effect_Size <- round(as.numeric(div_df$Mean_Effect_Size), 4)
                        }
                        
                        # Format q-pattern metrics for clarity
                        if ("div_rare_median" %in% colnames(div_df)) {
                            div_df$div_rare_median <- round(as.numeric(div_df$div_rare_median), 4)
                        }
                        if ("div_abundant_median" %in% colnames(div_df)) {
                            div_df$div_abundant_median <- round(as.numeric(div_df$div_abundant_median), 4)
                        }
                        if ("q_ratio" %in% colnames(div_df)) {
                            div_df$q_ratio <- round(as.numeric(div_df$q_ratio), 2)
                        }
                        
                        # Rename columns for display
                        names(div_df)[names(div_df) == "p_adj"] <- "LM adj. p-value"
                        names(div_df)[names(div_df) == "Mean_Effect_Size"] <- "Mean Divergence"
                        names(div_df)[names(div_df) == "Q_Pattern"] <- "Q-Pattern"
                        names(div_df)[names(div_df) == "div_rare_median"] <- "D_rare"
                        names(div_df)[names(div_df) == "div_abundant_median"] <- "D_abundant"
                        names(div_df)[names(div_df) == "q_ratio"] <- "Ratio"
                        
                        # Reorder columns: Gene -> Effect -> Pattern details -> Significance
                        col_order <- c("Gene", "Mean Divergence", "Q-Pattern", "D_rare", "D_abundant", "Ratio", "LM adj. p-value")
                        col_order <- col_order[col_order %in% colnames(div_df)]
                        div_df <- div_df[, col_order, drop = FALSE]
                        
                        print(knitr::kable(div_df,
                            caption = "**Table 5 | Top genes by linear model significance with effect sizes and q-spectrum patterns.** Ranked by statistical significance (ascending *P*-values, Benjamini-Hochberg *q*-value < 0.05). Columns: gene identifier; effect size (mean divergence across q-spectrum); pattern classification (Rare-driven/Balanced/Abundant-driven); pattern metrics-D\\_rare: median divergence for rare isoforms (q<1), D\\_abundant: median divergence for abundant isoforms (q>1), Ratio: D\\_rare/D\\_abundant (>1.3 indicates rare-driven, <0.77 indicates abundant-driven); adjusted P-value for linear model interaction test.",
                            row.names = FALSE,
                            format = if (knitr::is_latex_output()) "latex" else "html",
                            longtable = knitr::is_latex_output(),
                            linesep = "") %>%
                        kableExtra::kable_styling(
                          bootstrap_options = c("striped", "hover", "condensed"),
                          latex_options = c("scale_down")
                        ))
                    } else {
                        cat("No valid divergence results to display.\n")
                    }
                } else {
                    cat("Note: No effect size columns found in results\n")
                }
            } else {
                cat("Note: No p-value column found in results\n")
            }
        } else {
            cat("Note: Required 'gene' and 'p-value' columns not found in results\n")
        }
    }, error = function(e) {
        cat("Note: Could not create divergence results table:", e$message, "\n")
    })
} else {
    cat("Note: LM results from effect_sizes_divergence_s4 not available\n")
}
```

Interpretation of Pattern Types:

| Pattern Type | Signature | Biological Meaning |
|--------------|-----------|-------------------|
| Rare driven | D(q=0.5) >> D(q=2.0) | Low-abundance isoforms are condition-specific; treatment group preferentially expresses rare transcripts not seen in control |
| Abundant driver | D(q=0.5) << D(q=2.0) | High-abundance isoforms shift between conditions; treatment remodels the dominant transcript landscape without rare variants changing |
| Balanced | D(q=0.5) ~= D(q=2.0) | All isoforms shift proportionally; no preferential weighting to rare or abundant forms; suggests systematic rebalancing |

The effect sizes reveal which genes show the most information-theoretic separation 
across the q-spectrum **between paired treatment groups**. Genes with Tsallis divergence D > 0.1 show 
substantial divergence, indicating that entropy distributions differ fundamentally between conditions 
across the full q-spectrum.

**Biological Interpretation:** Following Tsallis entropy theory and information-theoretic principles [@I002; @I003; @I004; @S204], 
effect size quantification [@S205] 
genes with large D values are those where isoform complexity distributions differ qualitatively between conditions 
when examined at all sensitivity scales (q = rare -> abundant isoforms). For example, a gene might show 
high entropy at low q (emphasizing rare isoforms) in one condition but low entropy at high q (emphasizing 
abundant isoforms) in another, revealing scale-dependent regulatory mechanisms. These genes are candidates 
for investigation of condition-specific splicing architecture and dynamic isoform switching.


```{r effect-size-plot-multiq, fig.width=10, fig.height=6, out.width="98%", fig.align='center', fig.pos="H"}
# Visualize the distribution of Tsallis divergence effect sizes across genes
# The red dashed line marks D = 0.1 (information-theoretic significance threshold)
# Generate divergence distribution plot (returns plot object)
# Optionally save to file for external use
plot_obj <- plot_divergence_distribution_s4(
    analysis, 
    threshold = 0.05
)

print(plot_obj)
```

```{r compare-q-spectra-plot, fig.width=10, fig.height=6, out.width='100%', fig.align='center', fig.cap='**Figure 5:** Q-spectrum curves for top genes by significance. Each line represents divergence as a function of q-value; shape reveals biological pattern (rare-driven, balanced, or abundant-driven).'}
# Visualize q-spectrum curves for top 4 genes by significance
# Plot is rendered directly by knitr (no file dependency)
p_multi <- plot_divergence_spectrum_s4(
    analysis, 
    n_genes = 4, 
    use_pvalue_ranking = TRUE
)

print(p_multi)
```


Interpreting the Q-Spectrum Curve:

- Shape matters more than single value: The **q-spectrum** curve is the complete biological story. 
  A flat curve (balanced) vs. declining curve (rare driven) vs. rising curve (abundant driven) 
  encode fundamentally different mechanisms.

- q=1 (KL divergence): The middle point corresponds to ordinary Kullback-Leibler divergence [@K001; @I060], 
  where in the limit q->1 the Tsallis divergence reduces to standard KL divergence [@I060]. 
  This weights all isoforms equally. This is the "average" effect.

- q < 1 (rare isoforms): Emphasizes how much low-abundance transcripts differ between 
  groups. If D(q=0.5) >> D(q=1), the divergence is driven by changes in rare variants.

- q > 1 (abundant isoforms): Emphasizes dominant transcripts. If D(q=2) >> D(q=1), major 
  isoforms rebalance while rare ones stay similar.

- Bootstrap CI bands: Shaded region shows 95% confidence limits on per-q estimates. Narrow 
  bands indicate robust estimates; wide bands suggest noisy data or small sample sizes.



We can also plot the global divergence spectrum across all genes. 
This curve is the full biological signature of **isoform switching**-the pattern encodes which 
abundance scales (rare vs. abundant) are affected.

```{r visualize-q-spectrum, fig.width=10, fig.height=5, out.width='100%', fig.align='center', fig.cap='**Figure 6:** Global divergence spectrum across all genes. Aggregated q-spectrum pattern shows which abundance scales (rare vs. abundant isoforms) are affected genome-wide.'}
# Visualize global divergence q-spectrum across all genes (aggregated)
# Plot is rendered directly by knitr (no file dependency)
p <- plot_divergence_spectrum_s4(
    analysis
)

print(p)
```

Statistical Validation During Interpretation:

1. Bootstrap CI validity: Per-q CIs should narrow as q increases [@S115] (higher q = aggregation 
   effect, less variance). If CIs grow with q, check for data quality issues.

2. Monotonicity check: By **Tsallis entropy** theory [@I002], entropy is monotone decreasing in q. Divergence 
   should NOT show erratic increases with q. Small fluctuations are normal, but large spikes 
   indicate numerical instability.

3. **Comparison with genome-wide patterns**: Compute q-spectra for housekeeping genes (GAPDH, 
   ACTB, etc.). These should show BALANCED patterns. If not, revisit normalization parameters.

## S4 Workflow: Unified Analysis via TSENATAnalysis Objects

TSENAT provides an integrated S4-based workflow for coordinated analysis following three principles:

1. **Configure once**: Use `tsenat_config()` to specify analysis parameters (q-values, metadata columns)
2. **Run pipeline**: Use `tsenat()` to orchestrate all analysis steps in sequence
3. **Access results**: Use S4 accessor methods for type-safe result retrieval

The `tsenat()` function accepts a configured `TSENATConfig` object and orchestrates a complete pipeline: diversity computation -> quality control -> statistical testing -> effect sizes (as specified by the `methods` parameter). Parameter `verbose = TRUE` outputs progress messages showing data filtering steps and method completion. This single-function approach ensures consistency between preprocessing and analysis steps, eliminating boilerplate and reducing risk of parameter mismatches.

**Accessor methods return types:** Most accessors return `SummarizedExperiment` objects (following Bioconductor conventions) containing results as assays and metadata:

- `diversity(analysis)` -> SummarizedExperiment (genes x samples, multiple assays for each q-value)

- `lmResults(analysis)` -> list containing data.frame of LM interaction results (or NULL if method not run)

- `jeoResults(analysis, q = 1.0)` -> SummarizedExperiment of jackknife entropy outlier confidence intervals (or NULL)

- `jisResults(analysis, q = 1.0)` -> list of jackknife isoform switching results (or NULL)

- `rankResults(analysis)` -> data.frame of rank test results (or NULL)

- `divergence(analysis)` -> SummarizedExperiment of pairwise divergence metrics

If a requested method was not run during `tsenat()` orchestration, the accessor returns NULL-users should check results before downstream use.

```{r s4-workflow-demo, include = TRUE, echo = TRUE, eval = FALSE}
# Configure analysis parameters once (best practice for reproducibility)
tsenat_config <- tsenat_config(
  q_values = seq(0, 2, by = 0.05),
  sample_type_col = "sample_type",
  subject_col = "paired_samples"
)

# Run complete pipeline with one function call orchestrating all specified methods
# methods parameter: vector of analysis steps to execute ("diversity", "lm_interaction", "jackknife", "divergence")
# verbose = TRUE: prints progress messages showing gene filtering, step completion, and result summaries
analysis <- tsenat(
  se = se,
  config = tsenat_config,
  methods = c("diversity", "lm_interaction", "jackknife", "divergence"),
  verbose = TRUE
)

# Access results via consistent S4 methods (type-safe, mutually-exclusive access instead of nested list indexing)
diversity_results <- diversity(analysis)
lm_table <- lmResults(analysis)
jk_entropy <- jeoResults(analysis, q = 1.0)
jk_iso <- jisResults(analysis, q = 1.0)
rank_test <- rankResults(analysis)
divergence_table <- divergence(analysis)
```

For detailed step-by-step implementation, see the sections above demonstrating each analysis component and its accessor patterns.


# Appendices

This main vignette is complemented by two comprehensive appendices:

### Appendix A: Equivalence Validation - TSENAT vs SplicingFactory

Validates that TSENAT's Shannon and Simpson entropy implementations are mathematically equivalent to SplicingFactory. Includes:

- Mathematical proof of equivalence for q=1 (Shannon) and q=2 (Simpson)
- Benchmarking on TCGA BRCA RNA-seq data
- Detailed comparison tables and visualizations
- When to use TSENAT vs SplicingFactory

**[View Appendix A](TSENAT_appendix_A.html)**

### TSENAT Appendix B: Non-Parametric Validation of Linear Model Results via GAM and Rank-Based Methods

**Objective**: Validate q-value * group interaction detection results from linear models using complementary non-parametric statistical approaches.

TSENAT's default linear modeling approach (via `calculate_difference_s4()`) provides parametric tests for detecting scale-dependent differences in Tsallis entropy across q-values and experimental groups. To ensure robustness and generalization, Appendix B presents two independent, non-parametric alternatives that serve as validation methods:

**Approach 1: Generalized Additive Model (GAM) with ARIMA-Ordered Time Series**
- Treats q-values as explicit sequential measurements (ordered from low to high)
- Applies ARIMA(1,1,0) differencing to account for autocorrelation between adjacent q-points
- Fits smooth additive functions of q and group effects (non-parametric basis functions)
- Automatically detects heteroscedasticity and applies optimal weighting
- Reference: Goude (2024)

**Approach 2: Rank-Based Friedman Test with Hochberg Correction**
- Non-parametric alternative requiring no distributional assumptions
- Friedman test for detecting differences across multiple q-values within each group
- Accounts for paired/blocked structure in q-ordered measurements
- Hochberg step-up procedure for controlling family-wise error rate (FWER)
- References: Efron & Tibshirani (1993); Benjamini & Hochberg (1995)

Both methods incorporate principled advances to handle real-world RNA-seq data:
- **Heteroscedasticity detection and weighting**: Entropy variance often increases with q-value; both approaches adaptively weight measurements
- **Boundary condition handling**: Low and high q-values exhibit different variance properties; explicit boundary adaptation improves reliability
- **Automatic test selection**: Model diagnostics trigger appropriate method switches based on data structure

Appendix B provides:
- Detailed statistical methodology for both approaches
- Step-by-step implementation examples
- Comparison of linear model results vs. GAM vs. rank-based findings
- Concordance tables showing agreement across methods
- Recommendations for when each method is most appropriate
- Power analysis and sample size guidance for different designs

**Key Finding**: High concordance (typically >90% agreement) across methods validates that linear model discoveries are robust to model assumptions and generalize to non-parametric frameworks.

**[View Appendix B](TSENAT_appendix_B.html)**

## References

1. Adami, C. (2004). "Information Theory in Molecular Biology." *arXiv Preprint q-Bio/0405004*.

2. Bajic, D. (2024). "Information Theory, Living Systems, and Communication Engineering." *Entropy*, 26(5), 430. https://doi.org/10.3390/e26050430.

3. Bartal, A., & Jagodnik, K. M. (2022). "Progress in and Opportunities for Applying Information Theory to Computational Biology and Bioinformatics." *Entropy*, 24(7), 925. https://doi.org/10.3390/e24070925.

4. Chanda, P., Costa, E., Hu, J., Sukumar, S., Van Hemert, J., & Walia, R. (2020). "Information Theory in Computational Biology: Where We Stand Today." *Entropy*, 22(6), 627. https://doi.org/10.3390/e22060627.

5. Cover, T. M., & Thomas, J. A. (2006). *Elements of Information Theory* (2nd ed.). Wiley-Interscience.

6. Derian, N., Pham, H.-P., Nehar-Belaid, D., et al. (2022). "The Tsallis Generalized Entropy Enhances the Interpretation of Transcriptomics Datasets." *PLOS ONE*, 17(4), e0266618. https://doi.org/10.1371/journal.pone.0266618.

7. Furuichi, S. (2006). "Information Theoretical Properties of Tsallis Entropies." *Journal of Mathematical Physics*, 47, 023302. https://doi.org/10.1063/1.2165744.

8. Gandrillon, O., Gaillard, M., Espinasse, T., et al. (2021). "Entropy as a Measure of Variability and Stemness in Single-Cell Transcriptomics." *Entropy*, 21(5), 450.

9. Golomb, R., Yoles, M., Fishilevich, S., et al. (2026). "An Information Content Principle Explains Regulatory Patterns of Gene Expression Across Human Tissues." *bioRxiv*, ahead of print. https://doi.org/10.64898/2026.02.19.706555.

<!-- Commented out: TechRxiv preprint no longer available (404)
10. Hinostroza Fuentes, V. G., Toledo Tan, M. J., Kapetanaki, M., Rashidi, P., Huang, K., & Benos, P. V. (2026). "Information-Theoretic Methods in Spatial Transcriptomics." *TechRxiv*, ahead of print. https://doi.org/10.36227/techrxiv.177155947.
-->

11. Jost, L. (2006). "Entropy and Diversity." *Oikos*, 113(2), 363-375. https://doi.org/10.1111/j.2006.0030-1299.14714.x.

12. Renyi, A. (1961). "On Measures of Entropy and Information." *Proceedings of the Fourth Berkeley Symposium on Mathematical Statistics and Probability*, 1, 547-561.

13. Seweryn, M. T., Pietrzak, M., & Ma, Q. (2020). "Application of Information Theoretical Approaches to Assess Diversity and Similarity in Single-Cell Transcriptomics." *Computational and Structural Biotechnology Journal*, 18, 1830-1837. https://doi.org/10.1016/j.csbj.2020.05.006.

14. Shannon, C. E. (1948). "A Mathematical Theory of Communication." *The Bell System Technical Journal*, 27(3-4), 379-423.

15. Simpson, E. H. (1949). "Measurement of Diversity." *Nature*, 163, 688. https://doi.org/10.1038/163688a0.

16. Tsallis, C. (1988). "Possible Generalization of Boltzmann-Gibbs Statistics." *Journal of Statistical Physics*, 52(1), 479-487. https://doi.org/10.1007/BF01016429.

## Session Information

```{r session-info}
sessionInfo()
```
