---
title: "TSENAT Appendix B: Comparing Flexible Parametric (GAM) and Rank-Based (K-W) Methods"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: GAM vs K-W Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Flexible Parametric and Rank-Based Methods

## Purpose

This vignette validates q-value × group interaction detection using two complementary statistical approaches:

1. **Flexible Parametric (GAM):** Generalized Additive Models with ARIMA(1,1,0) transformation for handling q-ordered measurements
2. **Rank-Based (K-W):** Kruskal-Wallis with adaptive test selection based on data characteristics

Both methods now incorporate principled improvements to handle real-world data challenges:
- Heteroscedasticity detection and weighting
- Boundary condition handling
- Automatic test selection based on data properties
- Multiple testing correction accounting for AR(1) correlation in q-values

## Key Questions

1. Do both methods identify the same significant genes?
2. How much do results concordant vs. discordant?
3. Which genes are truly robust (significant in both methods)?
4. What data characteristics drive method disagreement?

## Interpretation Guide

- **Both methods significant:** Highest confidence findings (robust across statistical frameworks)
- **GAM only:** Method detects nonlinear patterns; K-W loses power on boundary data or ties
- **K-W only:** Rare; method detects rank patterns K-W captures differently
- **Neither significant:** Insufficient evidence for q×group interaction

---

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Subset for faster analysis (set to NULL for all genes)
n_genes_subset <- 500
if (!is.null(n_genes_subset) && nrow(readcounts) > n_genes_subset) {
  readcounts <- readcounts[1:n_genes_subset, ]
  cat("NOTE: Analyzing", n_genes_subset, "genes for speed. Set n_genes_subset=NULL for all genes.\n\n")
}

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build SummarizedExperiment and filter
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
se <- tryCatch(
  filter_se(se, stringency = "loose"),
  error = function(e) se
)

cat("Data after filtering - Genes:", nrow(se), "\n")
```

## Bayesian Pseudocount Regularization

```{r bayesian-pseudocounts, message=FALSE}
cat("\n=== BAYESIAN PSEUDOCOUNT REGULARIZATION (WLFC) ===\n\n")

raw_counts <- SummarizedExperiment::assay(se)

# Diagnostic: Check for duplicate rownames in SE before processing
cat("Diagnostic: SE structure\n")
cat("  Rows:", nrow(se), "\n")
cat("  Unique rownames:", length(unique(rownames(se))), "\n")
if (length(unique(rownames(se))) < nrow(se)) {
  dup_genes <- names(table(rownames(se))[table(rownames(se)) > 1])
  cat("  WARNING: Duplicate genes found:", paste(head(dup_genes, 5), collapse=", "), "\n")
}

# Fit empirical Beta prior
prior_params <- fit_empirical_beta_prior(raw_counts)
alpha_prior <- prior_params$alpha
beta_prior <- prior_params$beta

cat("\nEmpirical Beta prior fitted:\n")
cat(sprintf("  α (alpha):  %.4f\n", alpha_prior))
cat(sprintf("  β (beta):   %.4f\n", beta_prior))

# Compute WLFC pseudocounts
wlfc_pseudocounts <- numeric(nrow(raw_counts))
names(wlfc_pseudocounts) <- rownames(raw_counts)

for (i in seq_len(nrow(raw_counts))) {
  gene_counts <- raw_counts[i, ]
  wlfc_pseudocounts[i] <- compute_wlfc_pseudocounts(
    counts = gene_counts,
    alpha = alpha_prior,
    beta = beta_prior
  )
}

cat("\nWLFC Pseudocount Distribution:\n")
cat(sprintf("  Mean:       %.6f\n", mean(wlfc_pseudocounts)))
cat(sprintf("  Median:     %.6f\n", median(wlfc_pseudocounts)))
cat(sprintf("  Range:      [%.6f, %.6f]\n", min(wlfc_pseudocounts), max(wlfc_pseudocounts)))

# Use mean as scalar regularization
scalar_pseudocount <- mean(wlfc_pseudocounts)

# Calculate Tsallis entropy with regularization
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(
  se, 
  q = qvec, 
  norm = TRUE, 
  pseudocount = scalar_pseudocount,
  metadata = metadata_df,
  shrinkage = "empirical_bayes"
)

cat("\nTsallis Entropy Data Summary:\n")
cat("  Genes:", nrow(ts_se), "\n")
cat("  Samples:", length(unique(colData(ts_se)$samples)), "\n")
cat("  Q-values:", length(unique(colData(ts_se)$q)), "\n")
cat("  Total observations:", ncol(ts_se), "\n")
```

---

# Part 1: Flexible Parametric Approach (GAM)

## Computational Note

GAM analysis is computationally intensive because:
- **Per-gene fitting:** Each gene gets a smooth GAM model fitted across all q-values
- **Model complexity:** Generalized Additive Models use iterative smoothing optimization
- **Scale:** With ~500 genes × 39 q-values = 19,500 model evaluations
- **Multiple testing correction:** Additional computation for Hochberg adjustment

Expected runtime: 2-5 minutes for 500 genes with 3 threads. For faster iteration during development, reduce `n_genes_subset` in the setup chunk.

```{r gam-analysis, message=FALSE}
cat("\n=== FLEXIBLE PARAMETRIC APPROACH (GAM) ===\n\n")

gam_results <- calculate_lm_interaction(
    ts_se,
    method = "gam",
    paired = TRUE,
    multicorr = "hochberg",
    nthreads = 3,
    verbose = FALSE
)

cat("GAM Results Summary:\n")
cat("  Genes tested:", nrow(gam_results), "\n")
cat("  Significant (p < 0.05):", sum(gam_results$p_interaction < 0.05, na.rm = TRUE), "\n")
cat("  NAs:", sum(is.na(gam_results$p_interaction)), "\n")

# Show top genes
gam_sig <- gam_results %>%
  filter(!is.na(p_interaction)) %>%
  arrange(p_interaction) %>%
  head(10)

cat("\nTop 10 Genes (GAM):\n")
print(gam_sig[, c("gene", "p_interaction")])
```

---

# Part 2: Rank-Based Approach (Kruskal-Wallis)

```{r kw-analysis, message=FALSE}
cat("\n=== RANK-BASED APPROACH (KRUSKAL-WALLIS) ===\n\n")

# Convert SummarizedExperiment to long-format data frame for rank-based analysis
entropy_matrix <- assay(ts_se)
ts_coldata <- colData(ts_se)
test_genes <- rownames(ts_se)
n_genes <- length(test_genes)
n_cols <- ncol(ts_se)

entropy_long <- data.frame(
    entropy = as.numeric(entropy_matrix),
    gene = rep(test_genes, n_cols),
    q = rep(ts_coldata$q, each = n_genes),
    stringsAsFactors = FALSE
)

# Run rank-based interaction test with conditional test selection
# Note: detect_q_gene_interactions automatically selects appropriate rank-based tests
# (Kruskal-Wallis, ART, Mood's Median) based on data characteristics
kw_results <- detect_q_gene_interactions(
    entropy_long,
    entropy_col = "entropy",
    q_col = "q",
    gene_col = "gene",
    multicorr = "hochberg",
    verbose = FALSE
)

cat("K-W Results Summary:\n")
cat("  Genes tested:", nrow(kw_results), "\n")
cat("  Significant (p < 0.05):", sum(kw_results$p_value < 0.05, na.rm = TRUE), "\n")
cat("  NAs:", sum(is.na(kw_results$p_value)), "\n")

# Show top genes
kw_sig <- kw_results %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value) %>%
  head(10)

cat("\nTop 10 Genes (K-W):\n")
print(kw_sig[, c("gene", "p_value")])
```

---

# Part 3: Method Concordance Analysis

```{r concordance, message=FALSE}
cat("\n=== METHOD CONCORDANCE ANALYSIS ===\n\n")

# Merge results
# Note: GAM returns p_interaction, K-W returns p_value
comparison_df <- gam_results %>%
  select(gene, p_interaction) %>%
  rename(p_gam = p_interaction) %>%
  left_join(
    kw_results %>% select(gene, p_value) %>% rename(p_kw = p_value),
    by = "gene"
  ) %>%
  filter(!is.na(p_gam) & !is.na(p_kw))

# Calculate correlation
if (nrow(comparison_df) > 2) {
  spearman_rho <- cor(comparison_df$p_gam, comparison_df$p_kw, 
                      method = "spearman")
  cat(sprintf("Spearman correlation: ρ = %.4f\n", spearman_rho))
}

# Categorize agreement
comparison_df <- comparison_df %>%
  mutate(
    gam_sig = p_gam < 0.05,
    kw_sig = p_kw < 0.05,
    agreement = case_when(
      gam_sig & kw_sig ~ "Both significant",
      gam_sig & !kw_sig ~ "GAM only",
      !gam_sig & kw_sig ~ "K-W only",
      TRUE ~ "Neither significant"
    )
  )

cat("\nAgreement Summary:\n")
agreement_table <- table(comparison_df$agreement)
print(agreement_table)
print(prop.table(agreement_table))

# High-confidence genes (both methods)
high_conf <- comparison_df %>%
  filter(gam_sig & kw_sig) %>%
  arrange(pmax(p_gam, p_kw))

cat(sprintf("\n\nHIGH-CONFIDENCE GENES (significant in both methods): %d\n", nrow(high_conf)))
if (nrow(high_conf) > 0) {
  print(head(high_conf[, c("gene", "p_gam", "p_kw", "agreement")], 15))
}
```

## Visualization: Method Comparison

```{r concordance-plot, fig.width=10, fig.height=8}
# Create comparison plot
p1 <- ggplot(comparison_df, aes(x = -log10(p_gam), y = -log10(p_kw), color = agreement)) +
  geom_point(size = 2.5, alpha = 0.6) +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
  scale_color_manual(
    values = c("Both significant" = "#2ecc71", 
               "GAM only" = "#3498db",
               "K-W only" = "#e74c3c",
               "Neither significant" = "#95a5a6")
  ) +
  labs(
    title = "GAM vs Kruskal-Wallis: Method Concordance",
    x = "-log10(p-value, GAM)",
    y = "-log10(p-value, K-W)",
    color = "Significance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "bottomright",
    panel.grid.major = element_line(color = "gray90")
  )

# P-value distribution comparison
p2 <- ggplot(comparison_df %>% select(p_gam, p_kw) %>% 
             pivot_longer(everything(), names_to = "method", values_to = "p_value"),
             aes(x = p_value, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("p_gam" = "#3498db", "p_kw" = "#e74c3c")) +
  labs(
    title = "P-value Distributions",
    x = "P-value",
    y = "Frequency",
    fill = "Method"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

# Part 5: Recommendations

```{r recommendations, message=FALSE}
cat("\n=== RECOMMENDATIONS ===\n\n")

cat("1. PRIMARY ANALYSIS: Use GAM (Flexible Parametric)\n")
cat("   - Reason: Better power for detecting nonlinear q-entropy patterns\n")
cat("   - Default: multicorr = 'hochberg' (FWER control)\n")
cat("   - Recommended: Hochberg correction for high-confidence findings\n\n")

cat("2. VALIDATION: Confirm with K-W (Rank-Based)\n")
cat("   - Purpose: Robustness check using distribution-free method\n")
cat("   - Default: Conditional test selection adapts to data characteristics\n")
cat("   - Advantage: No parametric assumptions needed\n\n")

cat("3. HIGH-CONFIDENCE GENES\n")
cat(sprintf("   - %d genes significant in BOTH methods\n", nrow(high_conf)))
cat("   - Recommended for downstream biological validation\n")
cat("   - Strong evidence of genuine q-value × group interaction\n\n")

cat("4. MULTIPLE TESTING CORRECTION\n")
cat("   - Default: Hochberg (FWER, more stringent)\n")
cat("   - Alternative: Benjamini-Yekutieli (FDR, more lenient)\n")
cat("   - Both: Account for AR(1) correlation in q-values\n\n")

cat("5. PARAMETER SETTINGS\n")
cat("   - GAM method: 'gam' with paired=TRUE (accounts for subject structure)\n")
cat("   - K-W method: 'conditional' test selection\n")
cat("   - Pseudocount: WLFC for robust entropy estimation\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References
