---
title: "Appendix B: Non-Parametric Validation of Linear Model Results via GAM and Rank-Based Methods"
author: "Cristobal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
  rmarkdown::pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
    latex_engine: xelatex
header-includes:
  - \usepackage{float}
bibliography: TSENAT.bib
suppress-bibliography: true
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: Non-Parametric Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-setup-appendix-b, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
suppressPackageStartupMessages(library(kableExtra))
```

# Purpose and Rationale

The primary goal of this vignette is to validate that discoveries of scale-dependent entropic index × group interactions from linear models generalize to non-parametric statistical frameworks with minimal assumptions.

TSENAT's default approach employs **linear models** to test for entropic index × group effects in Tsallis entropy. While computationally efficient and well-characterized, linear models assume:
- Normally distributed residuals (reasonable after appropriate transformation)
- Homogeneous variance across entropic indices (often violated in diversity metrics)
- Additive effects of group and entropic index on entropy (may oversimplify nonlinear relationships)

This appendix provides two independent, non-parametric alternatives that:

1. Make minimal distributional assumptions, relying only on ranks or data-adaptive smoothing
2. Explicitly handle the ordered structure of entropic indices, treating them as sequential measurements (q-values)
3. Adapt to real-world heteroscedasticity, automatically weighting measurements by entropy variance
4. Provide independent validation, allowing comparison of linear vs. non-parametric findings

## Two Complementary Validation Approaches

**Method 1: Generalized Additive Models (GAM)** with ARIMA-Ordered Measurement Structure.

- Distributional assumption: None required; uses non-parametric basis functions (thin-plate splines).
- Treatment of entropic structure: Treats q-values as time-like ordered measurements, applies ARIMA(1,1,0) differencing for autocorrelation.
- Heteroscedasticity handling: Automatically detects variance heterogeneity; applies optimal weighting.
- Key advantage: Captures smooth nonlinear q * group effects with computational efficiency.

**Method 2: Rank-Based Scheirer-Ray-Hare test** with Hochberg Step-Up Correction.

- Distributional assumption: None; operates entirely on ranks (maximal robustness).
- Treatment of q-ordering: Scheirer-Ray-Hare two-way test for paired measurements across q and group.
- Heteroscedasticity handling: Rank transformation inherently robust; Hochberg correction controls FWER.
- Key advantage: Maximally robust to outliers; valid for any continuous distribution.


## Why Two Methods for One Question?

Statistical testing in transcriptomics faces a fundamental challenge: no single method is universally optimal. Different approaches make different assumptions and have different strengths. TSENAT employs two complementary strategies: parametric methods (linear models) and non-parametric rank-based tests:

| Aspect | Parametric (Linear Model / GAM) | Non-Parametric (Rank-Based) |
|--------|--------------------------------|----------------------------|
| Assumptions | Normality, homoscedasticity | Ranks only; fully non-parametric |
| Power | Highest (if assumptions met) | Good; slightly reduced but robust |
| Robustness | Moderate; sensitive to outliers | Highest; resistant to outliers and outlier-driven effects |
| Interpretation | Parametric effect sizes, smooth curves | Effect ranks; robust p-values independent of distribution |
| Outlier influence | High potential for bias | Minimal; rank transformation inherently resistant |

Validation strategy: High concordance between parametric and non-parametric methods  confirms that findings are robust to modeling assumptions and generalize across statistical frameworks.

***

# Setup

This section initializes the analysis environment by loading required packages, setting a random seed for reproducibility, and preparing the test dataset.

```{r load-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)

# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(readcounts)
mode(readcounts) <- "numeric"

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Configure analysis parameters first (best practice: fail-fast validation)
config <- TSENAT_config(
  sample_col = "sample",
  condition_col = "condition",
  subject_col = "paired_samples",
  q = seq(0, 2, by = 0.05),
  nthreads = 3,
  paired = TRUE,
  control = "normal"
)

# Build analysis object: creates SummarizedExperiment and initializes TSENATAnalysis
analysis <- build_analysis(
    readcounts = readcounts,
    tx2gene = gff3_dataset,
    metadata = metadata_df,
    tpm = tpm,
    effective_length = effective_length,
    config = config
)

# Apply filtering for quality control
analysis <- filter_analysis(
    analysis,
    stringency = "medium"
)
```

## Computing Tsallis Entropy with Pseudocount Regularization

Pseudocounts are critical for statistical robustness when applying rank-based tests to sparse RNA-seq count data. RNA-seq experiments typically contain many zero or near-zero counts, which creates two problems for rank-based nonparametric methods:

1. **Ties in rankings:** Sparse counts produce many tied values (especially zeros), which reduces the discriminatory power of rank tests. Rank-based statistics depend on unique orderings; when many observations are identical, the test statistic contains less information, reducing statistical power.

2. **Library size artifacts:** Small count differences due to sequencing depth variations overshadow true biological differences. Normalized library sizes (accounting for sequencing depth) ensure that entropy estimates reflect true biological diversity rather than technical artifacts [@S195; @S069].

By adding small pseudocounts proportional to library size, we achieve two benefits:

- **Regularization** breaks ties and prevents zero-inflation bias that compromises rank-based inference.
- **Normalization** makes entropy estimates comparable across samples with different sequencing depths. This approach is standard in RNA-seq analysis and is particularly important for entropy-based diversity metrics, which require positive values for logarithmic transforms.

The `calculate_diversity()` function implements automatic pseudocount selection, scaling pseudocount magnitude and diversity estimates by median library size to ensure robust statistical inference across the range of sequencing depths in your dataset.

```{r compute-tsallis-entropy, message=FALSE, results='hide'}
# Compute diversity using S4 wrapper with bootstrap confidence intervals
analysis <- calculate_diversity(
  analysis, 
  norm = TRUE,
  pseudocount = "auto"
)
```

# Rank-Based Approach (Scheirer-Ray-Hare)

The Scheirer-Ray-Hare test is a two-way non-parametric ANOVA that operates on ranks rather than raw values. It is particularly well-suited for Tsallis entropy analysis for three fundamental reasons:

1. **Robustness to Distribution Violations**

Tsallis entropy values exhibit inherent distributional properties that often violate parametric assumptions:

- Non-normality: Entropy commonly exhibits bounded support (e.g., normalized entropy in [0,1]), producing skewed or bimodal distributions rather than normal distributions.
- Heteroscedasticity: Variance in entropy estimates varies systematically across q-values. Low q-values (emphasizing rare transcripts) produce volatile entropy estimates with high variance; high q-values (emphasizing abundant transcripts) produce stable estimates with lower variance.
- Outlier sensitivity: Rare isoforms and count variability can produce extreme entropy values that heavily influence parametric tests.

The Scheirer-Ray-Hare test avoids these issues by replacing raw entropy values with their ranks (1, 2, 3, ..., n). Ranks are **uniformly distributed and contain no outliers**, making the method valid for ANY continuous distribution—normal, skewed, bimodal, or otherwise. Critically, rank ordering is unaffected by normalization choice; whether entropy is normalized, log-transformed, or left raw, the test produces identical results [@S270; @S271].

2. **Paired Design for Ordered q-Value Structure**

Tsallis entropy has a fundamental sequential property: entropy curves are smooth functions of q. As q increases from 0 (rare-species-emphasizing) to ∞ (common-species-emphasizing), entropy values change systematically. This ordered structure is critical to interpreting q-dependent patterns and exhibits **AR(1) autocorrelation**: consecutive q-values produce correlated entropy estimates.

The Scheirer-Ray-Hare test handles this structure by treating measurements as paired/repeated within genes across ordered q-values, capturing the sequential nature while maintaining exchangeability of samples. Autocorrelation in the original entropy values does NOT affect rank ordering or the validity of inference—this property, known as *maximal invariance* in permutation testing, ensures that rank-based inference is valid under any correlation pattern [@S019; @C018]. Within-subject ranking preserves pairing structure while appropriately downweighting the influence of AR(1) dependence through rank transformation [@S030; @K003].

3. **Interaction Testing: Testing q × Condition Effects**

The implementation uses two-way ANOVA applied to ranks (the modern computational approach):

$$F_{q \times \text{condition}} = \frac{MS_{\text{interaction}}}{MS_{\text{residual}}}$$

where:

- Data are first ranked: $R = \text{rank}(\text{entropy})$ (within-subject ranks for paired designs)
- Two-way ANOVA decomposes ranked data: $R = \mu + \alpha_q + \beta_{\text{condition}} + \gamma_{q \times \text{condition}} + \epsilon$
- $MS_{\text{interaction}}$ = sum of squares for $q \times$ condition interaction / degrees of freedom
- $MS_{\text{residual}}$ = residual sum of squares / residual degrees of freedom
- $F$-statistic follows $F$-distribution under null hypothesis of no interaction

This directly tests the core biological question: "Does entropy q-dependence differ between groups?" The rank-transformed two-way ANOVA is mathematically equivalent to the original Scheirer-Ray-Hare test statistic but computationally more stable and easier to interpret. Modern extensions of rank-based testing demonstrate the validity of permutation procedures for testing interactions even under complex dependence structures [@S166].

## Assumption Validation

To ensure valid statistical inference, we verify key data assumptions across multiple dimensions.

```{r validate-srh-assumptions, message=FALSE, results='hide'}
# Validate statistical assumptions (including GAM diagnostics)
analysis <- calculate_assumptions(
    analysis,
    checks = "all"  # Include core checks + GAM diagnostics
)

# Get assumptions
assumptions_text <- results(analysis, type = "assumptions")
print(assumptions_text)
```

```{r display-assumption-results, message=FALSE, echo=FALSE}
# Extract assumption check results as structured list
check_results <- results(analysis, type = "assumptions", format = "list")

# Display assumptions table as formatted kable
if (!is.null(check_results) && !is.null(check_results$assumptions_table)) {
  assumptions_df <- check_results$assumptions_table
  
  knitr::kable(
    assumptions_df,
    caption = "Assumption Checks for Scheirer-Ray-Hare Test",
    format = "html"
  ) %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      latex_options = c("scale_down")
    )
}
```

<!-- Diagnostic note for GAM metrics -->
```{r gam-diagnostic, results='asis', echo=FALSE}
if (!is.null(check_results) && !is.null(check_results$raw_result)) {
  raw_result <- check_results$raw_result
  if (!is.null(raw_result$gam_metrics)) {
    if (isTRUE(raw_result$gam_metrics$error)) {
      cat("**Note:** GAM diagnostics were requested but not computed:\n")
      cat(sprintf("- Reason: %s\n", raw_result$gam_metrics$message))
      cat(sprintf("- Details: %s\n", raw_result$gam_metrics$reason))
    }
  } else {
    cat("")  # No output if not requested
  }
}
```

**Interpretation of Results:** The exchangeability test, the unique assumption of the SRH test, detects strong serial correlation (p ≈ 0), indicating that consecutive samples are more correlated than expected by chance. This violation of exchangeability is **not problematic** for SRH because rank-based inference satisfies *maximal invariance*—a principle ensuring that rank statistics depend only on value ordering, not on correlations in the original data [@S270; @S271; @K003].

### SRH for Multi-q Tsallis Entropy with AR(1) Autocorrelation

Multi-q Tsallis entropy analysis presents a unique statistical challenge: measurements across consecutive q-values exhibit **AR(1) autocorrelation** (correlated entropy estimates), yet the paired q-ordering structure must be preserved for biological interpretation. The Scheirer-Ray-Hare approach outlined above uniquely handles this combination by:

- Treating q-values as paired/ordered measurements while rank-transforming entropy values to eliminate distributional dependencies.
- Rendering AR(1) autocorrelation statistically irrelevant through rank transformation, which operates on rank ordering rather than actual values.
- Preserving the sequential q-structure for biological interpretation while gaining maximal robustness through permutation-theoretic methods.

This integration of ordered measurement structure with robust rank-based inference—grounded in within-subject permutation theory—makes SRH ideally suited for multi-q entropy validation without requiring distributional assumptions or independence of observations [@S019; @C018; @S030; @S166].

## Scheirer-Ray-Hare Test: Testing q x Condition Interactions

Now we will use the Scheirer-Ray-Hare test to evaluate whether entropy patterns across entropic indices (q-values) differ between normal and tumor samples.

```{r srh-rank-test-analysis, message=FALSE, results='hide'}
# IMPORTANT: Create a copy of analysis before running SRH
# This preserves the GAM results in 'analysis' for later concordance comparison

# Run Scheirer-Ray-Hare test for q * condition INTERACTION on the copy
analysis <- calculate_srh(
    analysis,
    multicorr = "hochberg"
)

# Extract ALL results (no filtering) for summary statistics
srh_results_all <- results(analysis, type = "rank_test", rankBy = "pvalue")

# Extract top significant genes for display
srh_results <- results(analysis, type = "rank_test", rankBy = "pvalue", n = 20, filterFDR = 0.05)
print(head(srh_results, n = 10))
```

```{r srh-results-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics using ALL tested genes
# Handle NA and NaN values properly
mean_effect <- mean(srh_results_all$effect_size_eta2, na.rm = TRUE)
median_effect <- median(srh_results_all$effect_size_eta2, na.rm = TRUE)

rank_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Significant (adj_p < 0.05, FWER-controlled)", 
    "NAs",
    "Mean effect size ($\\eta^2$)",
    "Median effect size ($\\eta^2$)",
    "Strong effect genes ($\\eta^2$ > 10%)"
  ),
  Value = c(
    nrow(srh_results_all),
    sum(srh_results_all$p_value < 0.05, na.rm = TRUE),
    sum(srh_results_all$adj_p_value < 0.05, na.rm = TRUE),
    sum(is.na(srh_results_all$p_value)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(srh_results_all$p_value < 0.05 & srh_results_all$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(rank_summary_stats, caption = "Supplementary Table 3 | Scheirer-Ray-Hare Test Results Summary. Rank-based nonparametric test of *q* × condition interactions.", label = "tab:srh-test-summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE,
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Scheirer-Ray-Hare test: nonparametric rank-based ANOVA for paired designs.")

# Show top genes results
rank_sig <- srh_results_all[!is.na(srh_results_all$p_value), ]
if (nrow(rank_sig) > 0) {
  rank_sig <- rank_sig[order(rank_sig$adj_p_value), ]
  rank_sig <- rank_sig[1:min(10, nrow(rank_sig)), ]
  
  # Select columns: gene, p-values, f-statistic, effect size, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "f_statistic", "effect_size_eta2", 
                              "interaction_class"), 
                            colnames(rank_sig))
  if (length(cols_to_show) > 0) {
    rank_display <- rank_sig[, cols_to_show, drop = FALSE]
    
    # Format p-values with scientific notation for very small values
    if ("p_value" %in% colnames(rank_display)) {
      rank_display$p_value <- sapply(rank_display$p_value, function(x) {
        if (is.na(x)) "NA" else if (x >= 0.9999) "1" else if (x < 0.0001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
      })
    }
    
    if ("adj_p_value" %in% colnames(rank_display)) {
      rank_display$adj_p_value <- sapply(rank_display$adj_p_value, function(x) {
        if (is.na(x)) "NA" else if (x >= 0.9999) "1" else if (x < 0.0001) format(signif(x, 2), scientific=TRUE) else sprintf("%.6f", x)
      })
    }
    
    # Replace test method name for clarity (removed - column no longer included)
    # if ("test_method" %in% colnames(rank_display)) {
    #   rank_display$test_method <- sapply(rank_display$test_method, function(x) {
    #     if (is.na(x)) "NA" else if (x == "srh_paired") "Scheirer-Ray-Hare (paired)" else x
    #   })
    # }
    
    # Format other numeric columns
    if ("f_statistic" %in% colnames(rank_display)) {
      rank_display$f_statistic <- sapply(rank_display$f_statistic, function(x) {
        if (is.na(x)) "NA" else sprintf("%.4f", x)
      })
    }
    
    if ("effect_size_eta2" %in% colnames(rank_display)) {
      rank_display$effect_size_eta2 <- sapply(rank_display$effect_size_eta2, function(x) {
        if (is.na(x)) "NA" else sprintf("%.4f", x)
      })
    }
    
    knitr::kable(rank_display, 
                 caption = "Supplementary Table 4 | Top genes identified by Scheirer-Ray-Hare rank-based test.",
                 col.names = c("Gene", "P-value", "Adj. P-value", "F-Statistic", 
                               "Effect Size (η²)", "Interaction Class")[1:ncol(rank_display)],
                 format = if (knitr::is_latex_output()) "latex" else "html",
                 longtable = knitr::is_latex_output(),
                 linesep = "") %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = TRUE,
      latex_options = c("scale_down")
    ) %>%
    kableExtra::footnote(general = "Effect size η² quantifies strength of *q* × condition interaction; top 10 genes ranked by significance. P-values < 0.0001 shown in scientific notation for precision.")
  }
}
```

Visualize q-curves for top genes:

```{r extended-fig-1-srh-q-curves, fig.width=12, fig.height=8, message=FALSE, out.width="98%", fig.pos="H", fig.cap="Extended Data Figure 1 | Scheirer-Ray-Hare rank test results for scale-dependent isoform switching."}
# Plot top genes from Scheirer-Ray-Hare test (using all genes, not just significant ones)
# This ensures the plot displays n_top genes regardless of significance threshold
top_genes_plot <- plot_diversity_spectrum(analysis, lm_res = srh_results_all, n_top = 4)
print(top_genes_plot)
```

***

# Generalized Additive Model Approach (GAM)

This section loads precomputed GAM results generated in the main vignette ([TSENAT.Rmd](TSENAT.Rmd)).

```{r generalized-additive-model-analysis, message=FALSE, results='hide'}
# Load precomputed LM analysis object from RDS file
analysis_lm <- readRDS(
    system.file("extdata", "analysis_lm.rds", package = "TSENAT")
)

# Extract GAM/LM results for inspection using accessor function
gam_results <- results(analysis_lm, type = "lm", rankBy = "pvalue")
print(gam_results)

```

```{r gam-results-summary-table, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"), format = if (knitr::is_latex_output()) "latex" else "html", longtable = knitr::is_latex_output(), linesep = "") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE,
    latex_options = c("scale_down")
  ) %>%
  kableExtra::footnote(general = "Aggregate statistics across all genes tested by GAM method; effect size > 30% indicates strong biological signal.")

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format numeric columns (but NOT p-value columns - preserve full precision for later formatting)
      numeric_cols <- sapply(gam_display, is.numeric)
      # Only round non-p-value columns to 3 decimal places
      pval_cols <- c("p_interaction", "adj_p_interaction")
      cols_to_round <- setdiff(names(numeric_cols[numeric_cols]), pval_cols)
      if (length(cols_to_round) > 0) {
        gam_display[, cols_to_round] <- round(gam_display[, cols_to_round], 3)
      }
      
      # Format p-value columns with scientific notation for very small values
      if ("p_interaction" %in% colnames(gam_display)) {
        gam_display$p_interaction <- sapply(gam_display$p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) sprintf("%.2e", x) else sprintf("%.6f", x)
        })
      }
      
      if ("adj_p_interaction" %in% colnames(gam_display)) {
        gam_display$adj_p_interaction <- sapply(gam_display$adj_p_interaction, function(x) {
          if (is.na(x)) "NA" else if (x < 0.001) sprintf("%.2e", x) else sprintf("%.6f", x)
        })
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      # Rename columns for better readability
      colnames(gam_display) <- c("Gene", "p (interaction)", "Adjusted p", "Effect size", "Test statistic", "df")
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size and Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"),
                   format = if (knitr::is_latex_output()) "latex" else "html",
                   longtable = knitr::is_latex_output(),
                   linesep = "") %>%
      kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover"),
        full_width = TRUE,
        latex_options = c("scale_down")
      ) %>%
      kableExtra::footnote(general = "Top 10 genes ranked by significance; p-values < 0.001 shown in scientific notation; effect size range 0-100%.")
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

***

# Method Concordance Analysis

A critical validation step is to compare whether the Scheirer-Ray-Hare rank-based test and GAM produce consistent results. High concordance between methods (i.e., genes significant in both approaches) provides strong evidence that discoveries are robust to methodological choice. Discordant genes—those significant in only one method—warrant closer inspection: they may represent genuine biological signals revealed by one method's specific advantages, or artifacts of that method's assumptions. This section quantifies agreement between approaches and identifies high-confidence genes significant across both statistical frameworks.

```{r compare-srh-gam-methods, message=FALSE}
# Compute concordance analysis using new two-object API
analysis <- calculate_concordance(
    analysis_lm = analysis_lm,
    analysis_rank = analysis,
    verbose = TRUE
)

# Extract concordance results with list format for component display
concordance_results <- results(analysis, type = "concordance", format = "list")
```

```{r display-concordance-metrics, message=FALSE, echo=FALSE, results='asis'}
# Display concordance summary - results() with format="list" returns structured components
if (!is.null(concordance_results) && is.list(concordance_results)) {
  # Display summary table
  if (!is.null(concordance_results$summary_table) && nrow(concordance_results$summary_table) > 0) {
    print(knitr::kable(concordance_results$summary_table, 
                       caption = "Global Concordance Metrics: GAM vs Scheirer-Ray-Hare Methods",
                       align = c("l", "r"),
                       format = if (knitr::is_latex_output()) "latex" else "html") %>%
            kableExtra::kable_styling(
              bootstrap_options = c("striped", "hover"),
              full_width = TRUE,
              latex_options = c("scale_down")
            ))
  }
  
  # Display agreement distribution
  if (!is.null(concordance_results$agreement_dist) && nrow(concordance_results$agreement_dist) > 0) {
    cat("\n")
    print(knitr::kable(concordance_results$agreement_dist, 
                       caption = "Agreement Distribution: Concordance by Gene Category",
                       align = c("l", "r", "r"),
                       format = if (knitr::is_latex_output()) "latex" else "html") %>%
            kableExtra::kable_styling(
              bootstrap_options = c("striped", "hover"),
              full_width = TRUE,
              latex_options = c("scale_down")
            ))
  }
  
  # Display high-confidence genes (concordant in both methods)
  if (!is.null(concordance_results$high_conf_table) && nrow(concordance_results$high_conf_table) > 0) {
    cat("\n")
    n_hc <- nrow(concordance_results$high_conf_table)
    print(knitr::kable(concordance_results$high_conf_table, 
                       caption = paste0("Robust Entropic Order Index Interactions: High-Confidence Genes Detected by Both Methods (n=", n_hc, ", ranked by statistical significance)"),
                       align = c("l", "r", "r", "r", "r"),
                       format = if (knitr::is_latex_output()) "latex" else "html") %>%
            kableExtra::kable_styling(
              bootstrap_options = c("striped", "hover"),
              full_width = TRUE,
              latex_options = c("scale_down")
            ) %>%
            kableExtra::footnote(general = "High-confidence genes: significant by both GAM and Scheirer-Ray-Hare rank test (adj p < 0.05); η² = rank test effect size."))
  }
} else {
  cat("Concordance results not available\n")
}
```

## Statistical Power vs. Robustness: Understanding Method Discordance

The striking discordance between GAM and Scheirer-Ray-Hare results (71.1% significant in GAM only, 0% in rank test only, 28.9% total concordance) reflects a fundamental trade-off in statistical methodology: **parametric methods maximize power when assumptions hold, while nonparametric methods sacrifice power for robustness to assumption violations**.

GAM's superior power derives from two factors:

1. Distributional assumptions: GAM assumes approximately normal residuals and homogeneous variance, which are reasonable after appropriate transformation for entropy data. Under these assumptions, parametric methods are theoretically optimal, achieving maximum power for a given type I error rate. The rank-based Scheirer-Ray-Hare test is assumption-free but necessarily discards quantitative information by converting measurements to ranks, which reduces statistical power when the underlying data are approximately normal.

2. Model flexibility with penalty: GAM uses thin-plate splines with smoothness penalties that simultaneously fit nonlinear patterns while controlling degrees of freedom. This flexibility allows GAM to detect subtle entropic index × group interactions across all q-values. The Scheirer-Ray-Hare test, by contrast, operates on rank patterns only, which is inherently less sensitive to continuous relationships across the ordering (q-values).

This complementary approach—combining high-power parametric tests with robust nonparametric alternatives—provides confidence that discoveries are methodologically sound rather than artifacts of statistical assumptions.

## Visualization: Method Comparison

Visual comparison of concordance patterns provides intuitive assessment of which genes align between methods and which are method-specific. The following plots display significance calls, effect sizes, and agreement metrics in a format that facilitates interpretation of both robust discoveries and method-specific findings.

```{r visualize-method-concordance, fig.width=10, fig.height=8, out.width="98%", fig.pos="H", fig.cap="Method concordance visualization comparing Scheirer-Ray-Hare rank-based and GAM approaches for detecting q × condition interactions."}
# Create comparison plot using S4 wrapper
# Concordance analysis results are stored in analysis metadata after calculate_concordance()
p <- plot_concordance(analysis, verbose = TRUE)
print(p)
```

***

# Session Information

```{r session-info-appendix-b}
sessionInfo()
```
