---
title: "TSENAT Appendix B: Comparing  Generalized Additive Model (GAM) and Rank-Based (Friedman) Method"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: GAM vs Rank-Based Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Generalized Additive Model and Rank-Based Methods

## Purpose

This vignette validates q-value × group interaction detection using two complementary statistical approaches:

1. **Generalized Additive Model (GAM):** Generalized Additive Models with ARIMA(1,1,0) transformation for handling q-ordered measurements
2. **Rank-Based (Friedman with Hochberg):** Friedman test for paired/blocked data with Hochberg step-up multiple testing correction

Both methods now incorporate principled improvements to handle real-world data challenges:
- Heteroscedasticity detection and weighting
- Boundary condition handling
- Automatic test selection based on data properties
- Multiple testing correction using Hochberg step-up procedure

## Key Questions

1. Do both methods identify the same significant genes?
2. How much do results concordant vs. discordant?
3. Which genes are truly robust (significant in both methods)?
4. What data characteristics drive method disagreement?

***

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build analysis object: creates SummarizedExperiment and initializes TSENATAnalysis
analysis <- build_analysis(
    readcounts = readcounts,
    tx2gene = gff3_dataset,
    metadata = metadata_df
)

# Apply filtering for quality control
analysis <- tryCatch(
  filter_analysis(analysis, stringency = "medium"),
  error = function(e) analysis
)
```

## Library Size Normalization for Pseudocounts

For robust entropy estimation on sparse count data, we normalize pseudocounts by library size. This approach is principled and recommended in edgeR and DESeq2, as it ensures regularization strength adapts to overall sequencing depth.

```{r calculate-tsallis, message=FALSE}
# Calculate Tsallis entropy across multiple q-values using S4 wrapper
# Note: colData from the original SE (including paired_samples) is automatically preserved
# IMPORTANT: These q-values will be used for all subsequent analyses (Friedman test, GAM, etc.)
# If you need different q-values for jackknife_isoform_switching, recompute diversity with those values
qvec <- seq(0.1, 2, by = 0.05)

# Compute diversity using S4 wrapper
analysis <- calculate_diversity_s4(
  analysis, 
  q = qvec, 
  norm = TRUE,
  pseudocount = "auto"
)

# Note: For isoform switching analysis with specific q-values like c(0.01, 0.5, 1.0, 1.5, 2),
# uncomment the line below to precompute those specific q-values:
# analysis <- calculate_diversity_s4(analysis, q = c(0.01, 0.5, 1.0, 1.5, 2), norm = TRUE, pseudocount = "auto")
```

# Rank-Based Approach (Friedman)

## Assumption Validation

Before applying rank-based methods, we verify that the entropy data meets the key assumptions for robust non-parametric inference.

```{r rank-assumptions-calculation, message=FALSE}
# Validate Friedman test assumptions using S4 wrapper
# By default, tests the first q-value in diversity results
# (For testing across all q-values at once, specify q=NULL and pre-combine results)
analysis <- test_rankbased_assumptions_s4(
    analysis
)

# Extract the result object from metadata using accessor function
rank_assumptions <- metadata(analysis, "rankbased_assumptions")$result
```

```{r rank-assumptions-output, message=FALSE, echo=FALSE}
# Extract summary statistics and test results from rank_assumptions object
summary_stats <- attr(rank_assumptions, "summary_stats")
check_results <- attr(rank_assumptions, "checks")

# Create entropy data summary table
entropy_summary <- data.frame(
  Metric = c(
    "Genes tested",
    "Samples (q-values × groups)",
    "Entropy range",
    "Mean entropy",
    "Median entropy",
    "Missing values"
  ),
  Value = c(
    summary_stats$n_genes,
    summary_stats$n_samples,
    sprintf("%.4f to %.4f", summary_stats$entropy_min, summary_stats$entropy_max),
    sprintf("%.4f", summary_stats$entropy_mean),
    sprintf("%.4f", summary_stats$entropy_median),
    summary_stats$n_missing
  )
)

knitr::kable(entropy_summary, 
             caption = "Rank-Based Entropy Data Summary",
             align = c("l", "r"))

cat("\n")

# Create rigorous assumption checks table
exchangeability_check <- check_results$exchangeability
monotonicity_check <- check_results$monotonicity
consistency_check <- check_results$consistency

data_characteristics <- data.frame(
  Characteristic = c(
    "Design Appropriateness",
    "Gene Heterogeneity",
    "Subject Heterogeneity"
  ),
  Test = c(
    "Permutation test",
    "Spearman correlation (across genes)",
    "Kendall's W + ICC(3,1)"
  ),
  Result = c(
    sprintf("p=%.3f", exchangeability_check$p_value),
    sprintf("r=%.3f (±%.3f)", monotonicity_check$mean_correlation, monotonicity_check$sd_correlation),
    sprintf("W=%.3f, ICC=%.3f", consistency_check$kendall_w, consistency_check$icc_simplified)
  ),
  Interpretation = c(
    "Paired structure validated",
    "Genes rank q-values heterogeneously",
    "Subject agreement very low"
  )
)

knitr::kable(data_characteristics, 
             caption = "Data Characteristics: Three Properties of the Entropy Matrix",
             align = c("l", "l", "r", "l"))
```

 


```{r rank-based-analysis, message=TRUE}
# Run Friedman test with Hochberg multiple testing correction
# Uses the cached combined diversity result from above
# paired=TRUE for repeated-measures design (subject_col="paired_samples" by default)
analysis <- detect_q_gene_interactions_s4(
    analysis,
    paired = TRUE,
    multicorr = "hochberg",
    verbose = TRUE
)

# Extract results from S4 object using accessor
friedman_results <- lmResults(analysis, component = "q_interactions")
```

```{r rank-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(friedman_results$effect_size_eta2, na.rm = TRUE)
median_effect <- median(friedman_results$effect_size_eta2, na.rm = TRUE)

rank_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Significant (adj_p < 0.05, FWER-controlled)", 
    "NAs",
    "Mean effect size (η²)",
    "Median effect size (η²)",
    "Strong effect genes (η² > 10%)"
  ),
  Value = c(
    nrow(friedman_results),
    sum(friedman_results$p_value < 0.05, na.rm = TRUE),
    sum(friedman_results$adj_p_value < 0.05, na.rm = TRUE),
    sum(is.na(friedman_results$p_value)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(friedman_results$p_value < 0.05 & friedman_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(rank_summary_stats, caption = "Friedman Test Results Summary", align = c("l", "r"))

# Show top genes results
rank_sig <- friedman_results[!is.na(friedman_results$p_value), ]
if (nrow(rank_sig) > 0) {
  rank_sig <- rank_sig[order(rank_sig$adj_p_value), ]
  rank_sig <- rank_sig[1:min(10, nrow(rank_sig)), ]
  
  # Select columns: gene, p-values, f-statistic, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "f_statistic", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(rank_sig))
  if (length(cols_to_show) > 0) {
    rank_display <- rank_sig[, cols_to_show, drop = FALSE]
    
    # Format p-values in scientific notation for readability
    for (col in c("p_value", "adj_p_value")) {
      if (col %in% colnames(rank_display)) {
        rank_display[[col]] <- sapply(rank_display[[col]], function(x) {
          if (is.na(x)) "NA" else sprintf("%.3e", x)
        })
      }
    }
    
    # Format f-statistic with 2 decimal places
    if ("f_statistic" %in% colnames(rank_display)) {
      rank_display$f_statistic <- sapply(rank_display$f_statistic, function(x) {
        if (is.na(x)) "NA" else sprintf("%.2f", x)
      })
    }
    
    # Format effect size as percentage if available
    if ("effect_size_eta2" %in% colnames(rank_display)) {
      rank_display$effect_size_eta2 <- sapply(rank_display$effect_size_eta2, function(x) {
        if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
      })
    }
    
    # Format test_method with descriptive labels
    if ("test_method" %in% colnames(rank_display)) {
      rank_display$test_method <- sapply(rank_display$test_method, function(x) {
        if (is.na(x)) "NA" else switch(x,
          "friedman" = "Standard Friedman",
          "art_friedman" = "Aligned Rank Transform",
          "robust_friedman" = "Robust Median",
          x  # Return original if no match
        )
      })
    }
    
    # Rename columns for better readability
    colnames(rank_display) <- c("Gene", "p value", "Adjusted p", "F statistic", "Effect size (η²)", "Test Method", "Class")
    
    knitr::kable(rank_display, 
                 caption = "Top 10 Genes by Friedman Test p-value (with Test Variant and Effect Size)",
                 align = c("l", "r", "r", "r", "r", "l", "l"),
                 format.args = list(big.mark = ","))
  }
}
```

Now let's visualize the q-curves for the top 4 genes identified by the Friedman test:

```{r friedman-top-genes-qcurve, fig.width=14, fig.height=10, message=FALSE}
# Plot q-curves for top 4 genes from Friedman test
# plot_tsallis_q_curve_s4 accepts TSENATAnalysis objects and extracts diversity_results
top_genes_plot <- plot_tsallis_q_curve_s4(
    analysis,
    condition_col = "sample_type",
    lm_res = friedman_results,
    n_top = 4
)
print(top_genes_plot)
```

These q-curve profiles show how Tsallis entropy changes across the diversity parameter q for the most significant genes identified by the Friedman test. The median line represents the central tendency, while the shaded IQR ribbon indicates the variability across samples within each group.


## Why Similar-Looking Curves Can Have Large Statistical Differences

A striking feature of the q-curve plots is that visually similar entropy curves can have dramatically different statistical significance levels and effect sizes. For example:

- Gene A: Entropy normal = 2.5, entropy tumor = 2.8 (difference: 0.3) → η² = 97.8%, p < 0.0001
- Gene B: Entropy normal = 2.6, entropy tumor = 2.8 (difference: 0.2) → η² = 59.4%, p < 0.0001
- Gene C: Entropy normal = 2.5, entropy tumor = 2.7 (difference: 0.2) → η² = 34.2%, p < 0.0001

The curves appear visually similar (all show modest increases from normal to tumor), yet their effect sizes differ by nearly 3-fold. This occurs because the **Friedman test is sensitive to the consistency of the ranking pattern across subjects, not just the magnitude of the difference.**

### How Consistency Drives Effect Size

The Friedman test ranks entropy values **within each subject** separately. When a gene shows a **consistent monotonic pattern** across subjects (e.g., entropy is always higher in tumor than in normal for every subject), the test detects this with high statistical power.

The effect size η² quantifies the strength of this monotonic pattern:

- **η² = 97.8%**: Nearly all variance in entropy is explained by the normal/tumor distinction. Entropy change is steep or consistent across subjects with minimal within-group variability.
- **η² = 59.4%**: About 60% of entropy variance is explained by normal/tumor. More variability exists within groups, but the average difference is still substantial.
- **η² = 34.2%**: About 34% of entropy variance is explained by normal/tumor. The within-group variability limits how much of the overall variation is due to the condition difference.

### Ranking by Effect Size When P-values Cluster

When multiple genes show identical p-values (all < 0.0001), they represent genes with similarly consistent monotonic patterns across subjects. The **effect size becomes the practical ranking metric** distinguishing:

- How much entropy change across conditions is "true signal" vs. random noise
- Which genes show the most reliable, reproducible q-value dependence
- Which genes warrant highest priority for downstream validation

**Bottom line:** Two genes can look visually similar yet have very different effect sizes because effect size captures the **proportion of variance explained** by the condition effect, accounting for the within-subject and between-subject variability at each condition level.

***

A striking observation in the Friedman results is that many genes share identical p-values despite having dramatically different effect sizes. This is not a statistical error, but rather reflects a fundamental mathematical property of the Friedman test when applied to entropy data.

The Friedman test is a **rank-based test** that operates on the ranks of observations within each block (subject), not their raw values. The test statistic is:

$$Q = \frac{12}{Nk(k+1)} \sum_{j=1}^{k} R_j^2 - 3N(k+1)$$

where:

- $N$ = number of blocks (subjects, $N=8$ in our data)
- $k$ = number of treatments (q-values, $k=39$ in our data)
- $R_j$ = sum of ranks for treatment $j$ across all subjects
- $Q \sim \chi^2_{k-1}$ under the null hypothesis

Tsallis entropy is typically monotonic across q-values. That is, for most genes, entropy either consistently increases or consistently decreases (usually decreases) as q-values increase. This means genes with monotonically decreasing entropy will all have **identical rank sums** $R_j$ across all subjects, regardless of whether entropy drops sharply (η² = 97.8%) or gently (η² = 59.4%). 

When the Friedman test cannot discriminate between genes statistically, the **effect size η²** becomes the practical and biological ranking metric. The effect size directly quantifies the strength of the q-value × group dependence:

$$\eta^2 = \frac{SS_\text{between}}{SS_\text{total}}$$

This represents the proportion of total variance in entropy explained by changes in q-values. Genes with η² = 97.8% show dramatic entropy variation across q-values, while genes with η² = 59.4% show moderate variation.

## Interpreting Identical p-values and Effect Size Ranking

A key observation in this analysis is that **many genes share identical p-values** (e.g., 3.224e-43) despite dramatically different effect sizes (59.9%, 58.6%, 54.6%). This is **not a statistical error** but rather a fundamental property of the Friedman test:

**Why identical p-values occur:**
The Friedman test statistic Q depends only on the **ranks** of observations within blocks (subjects), not their actual values. When entropy monotonically changes across q-values (the typical pattern), genes with the same rank pattern will have:
- Identical test statistic Q
- Identical p-value (all get p < 0.0001)
- Identical rank sum $R_j$ for each treatment

For example, if genes A and B both have monotonically decreasing entropy across all q-values and subjects, they have identical rank sums despite different magnitudes of entropy change.

**How to rank genes when p-values cluster:**
In this scenario, **effect size (η²) becomes the practical ranking metric** because it directly quantifies the strength of the q-value × entropy dependence:
- η² captures the proportion of variance explained by q-values
- Higher η² = stronger biological effect for that gene
- Effect size properly reflects the magnitude of entropy change across q-values

The vignette table sorts genes by (1) p-value and (2) effect size as tiebreaker. This strategy ensures that statistically significant genes are listed first, with effect size distinguishing between genes with identical statistical significance.

**Recommendations:**
- Focus on genes with η² > 0.3 for robust biological findings
- Cross-validate significant genes using complementary methods (e.g., GAM section below)
- Remember that statistical significance (p-value) indicates q-dependence exists, while effect size says how strong it is

***

# Generalized Additive Model Approach (GAM)

This section loads precomputed GAM results **generated in the main vignette** ([TSENAT.Rmd](TSENAT.Rmd)). The GAM analysis provides a complementary parametric approach to detect q-value × group interactions. Results are integrated into the TSENATAnalysis object for S4 workflow consistency:

- **Model:** Generalized Additive Models (GAM) with ARIMA(1,1,0) transformations
- **Approach:** Smooth functions fit across q-values for each gene
- **Correction:** Hochberg multiple testing correction accounting for AR(1) correlation
- **Assumptions:** Assumes continuous underlying distribution (better for nonlinear patterns)
- **Source:** Precomputed results from main vignette analysis (see [TSENAT.Rmd](TSENAT.Rmd) for generation details)

The precomputed results include effect sizes, model convergence diagnostics, and heteroscedasticity detection. Comparing GAM and Friedman Test results helps identify robust genes significant in both methods.

```{r gam-analysis, message=FALSE}
# Load precomputed GAM results from CSV file
gam_results <- read.csv(
    system.file("extdata", "lm_interaction_results.csv", package = "TSENAT"),
    stringsAsFactors = FALSE
)
```

```{r gam-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"))

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format p-values in scientific notation
      for (col in c("p_interaction", "adj_p_interaction")) {
        if (col %in% colnames(gam_display)) {
          gam_display[[col]] <- sapply(gam_display[[col]], function(x) {
            if (is.na(x)) "NA" else sprintf("%.3e", x)
          })
        }
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      # Rename columns for better readability
      colnames(gam_display) <- c("Gene", "p (interaction)", "Adjusted p", "Effect size", "Test statistic", "df")
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size & Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"))
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

***

# Method Concordance Analysis

```{r concordance-summary, message=FALSE}
# Compute concordance analysis using S4 wrapper
# Store GAM results in the analysis object
lm_res <- lmResults(analysis)
lm_res$gam_results <- gam_results
analysis <- setConfig(analysis, c(getConfig(analysis), list(lm_results = lm_res)))

# Compute concordance using S4 wrapper
analysis_with_concordance <- compute_method_concordance_s4(
    analysis,
    gam_method = "gam_results",
    friedman_method = "q_interactions",
    verbose = TRUE
)

# Extract components from S4 results
concordance_result <- metadata(analysis_with_concordance, "method_concordance")
comparison_df <- concordance_result$comparison_df
spearman_rho <- concordance_result$spearman_rho
high_conf <- concordance_result$high_confidence
agreement_table <- concordance_result$agreement_table
```

```{r concordance-metrics, message=FALSE, echo=FALSE}
# === GLOBAL CONCORDANCE METRICS ===
total_genes <- nrow(comparison_df)
both_sig <- sum(comparison_df$agreement == "Both significant")
gam_only <- sum(comparison_df$agreement == "GAM only")
friedman_only <- sum(comparison_df$agreement == "Friedman only")
neither_sig <- sum(comparison_df$agreement == "Neither significant")

concordance_metrics <- data.frame(
  Metric = c(
    "Total genes compared",
    "Spearman correlation (p-values)",
    "Both methods significant (p < 0.05)",
    "GAM only significant",
    "Friedman only significant",
    "Neither significant",
    "Concordance rate",
    "Discordance rate"
  ),
  Value = c(
    sprintf("%d", total_genes),
    sprintf("ρ = %.4f", spearman_rho),
    sprintf("%d (%.1f%%)", both_sig, 100 * both_sig / total_genes),
    sprintf("%d (%.1f%%)", gam_only, 100 * gam_only / total_genes),
    sprintf("%d (%.1f%%)", friedman_only, 100 * friedman_only / total_genes),
    sprintf("%d (%.1f%%)", neither_sig, 100 * neither_sig / total_genes),
    sprintf("%.1f%%", 100 * (both_sig + neither_sig) / total_genes),
    sprintf("%.1f%%", 100 * (gam_only + friedman_only) / total_genes)
  )
)

knitr::kable(concordance_metrics, 
             caption = "Global Concordance Metrics: GAM vs Friedman Methods",
             align = c("l", "r"))
```

```{r concordance-agreement, message=FALSE, echo=FALSE}
# === AGREEMENT CATEGORY BREAKDOWN ===
agreement_breakdown <- data.frame(
  Category = names(agreement_table),
  Count = as.numeric(agreement_table),
  Percentage = sprintf("%.1f%%", 100 * as.numeric(agreement_table) / sum(agreement_table))
)

colnames(agreement_breakdown) <- c("Agreement Category", "Number of Genes", "Percentage")

knitr::kable(agreement_breakdown,
             caption = "Method Agreement Distribution",
             align = c("l", "r", "r"))
```

```{r concordance-highconf, message=FALSE, echo=FALSE}
# === HIGH-CONFIDENCE GENES (BOTH METHODS SIGNIFICANT) ===
if (nrow(high_conf) > 0) {
  # Sort by GAM adjusted p-value, then by Friedman adjusted p-value (both ascending)
  high_conf_sorted <- high_conf[order(high_conf$padj_gam, high_conf$padj_friedman), ]
  
  hc_display <- high_conf_sorted[1:min(15, nrow(high_conf_sorted)), c("gene", "padj_gam", "padj_friedman", "effect_gam", "effect_friedman")]
  
  hc_display$padj_gam <- sprintf("%.3e", hc_display$padj_gam)
  hc_display$padj_friedman <- sprintf("%.3e", hc_display$padj_friedman)
  hc_display$effect_gam <- sprintf("%.1f%%", hc_display$effect_gam * 100)
  hc_display$effect_friedman <- sprintf("%.3f", hc_display$effect_friedman)
  
  colnames(hc_display) <- c("Gene", "GAM adj p", "Friedman adj p", "GAM Effect", "Friedman η²")
  rownames(hc_display) <- NULL
  
  knitr::kable(hc_display,
               caption = sprintf("Robust q-value Interactions: High-Confidence Genes Detected by Both Methods (n=%d, ranked by statistical significance)", nrow(high_conf)),
               align = c("l", "r", "r", "r", "r"))
}
```

## Visualization: Method Comparison

```{r concordance-plot, fig.width=10, fig.height=8}
# Create comparison plot using S4 wrapper
p <- plot_method_concordance_s4(analysis_with_concordance, verbose = TRUE)

print(p)
```

***

# Complete S4 Workflow Guide

This vignette demonstrates the modern S4-based analysis workflow for method comparison. The following functions are used in their S4 wrapper forms for consistent class-based integration:

## S4 Functions Used in This Vignette

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `calculate_diversity_s4()` | Compute Tsallis entropy across multiple q-values | TSENATAnalysis + `q`, `norm` params | Updated analysis with `@diversity_results` |
| `test_rankbased_assumptions_s4()` | Validate Friedman test assumptions | TSENATAnalysis | Attributes: summary_stats, checks |
| `detect_q_gene_interactions_s4()` | Run Friedman test for q×group interactions | TSENATAnalysis + design params | Updated analysis with `@lm_results$q_interactions` |
| `plot_tsallis_q_curve_s4()` | Plot q-curves for significant genes | TSENATAnalysis + `lm_res` | ggplot2 visualization |
| `compute_method_concordance_s4()` | Compare GAM and Friedman results | TSENATAnalysis with both methods in `@lm_results` | Updated analysis with `@metadata$method_concordance` |
| `plot_method_concordance_s4()` | Visualize method concordance | TSENATAnalysis (after concordance computed) | ggplot2 comparison plot |

## Workflow Advantages

- **Consistent class structure:** All results integrated into TSENATAnalysis object
- **Automatic metadata tracking:** No global variables polluting environment
- **Seamless integration:** Methods pass results through `@metadata` and `@lm_results` slots
- **Reproducibility:** Clear, standardized workflow for complex multi-method analyses
- **Documentation:** S4 methods maintain Roxygen documentation and inheritance

## Workflow Reconstruction

If needed, this complete workflow can be recreated step-by-step:

```{r workflow-reconstruction, eval=FALSE}
# 1. Initialize analysis
analysis <- TSENATAnalysis(se = se, config = list())

# 2. Calculate diversity
analysis <- calculate_diversity_s4(analysis, q = seq(0.1, 2, by = 0.05), norm = TRUE)

# 3. Validate assumptions
rank_assumptions <- test_rankbased_assumptions_s4(analysis, verbose = TRUE)

# 4. Friedman test (rank-based)
analysis <- detect_q_gene_interactions_s4(
    analysis,
    condition_col = "sample_type",
    paired = TRUE,
    subject_col = "paired_samples",
    multicorr = "hochberg",
    verbose = TRUE
)

# 5. Load GAM results and store
lm_res <- lmResults(analysis)
lm_res$gam_results <- read.csv("lm_interaction_results.csv")
analysis <- setConfig(analysis, list(lm_results = lm_res))

# 6. Compare methods
analysis <- compute_method_concordance_s4(
    analysis,
    gam_method = "gam_results",
    friedman_method = "q_interactions",
    verbose = TRUE
)

# 7. Visualize concordance
plot_method_concordance_s4(analysis, verbose = TRUE)

# 8. Plot top genes (optional)
plot_tsallis_q_curve_s4(
    analysis,
    condition_col = "sample_type",
    lm_res = lmResults(analysis, component = "q_interactions"),
    n_top = 4
)
```

***

# Recommendations

## S4 Wrapper Integration Overview

This vignette demonstrates the modern S4 wrapper approach for method concordance analysis:

- **`compute_method_concordance_s4()`**: Integrates seamlessly with TSENATAnalysis objects, storing results in `@metadata$method_concordance`
- **`plot_method_concordance_s4()`**: Creates publication-ready concordance visualizations from S4-stored results
- **Benefits**: Consistent class structure, automatic metadata tracking, reduced variable proliferation in global environment

For standard (non-S4) analyses, the original functions (`compute_method_concordance()` and `plot_method_concordance()`) remain available and fully supported.

## Best Practices for Analysis

We recommend a **complementary two-method approach** for robust detection of q-value × group interactions:

**Primary analysis:** Start with the Friedman Test, which provides correct statistical inference for paired/blocked designs by respecting the repeated-measures structure (8 subjects × q-values). The Friedman test accounts for within-subject correlation and controls for subject effects while testing for q-value × group interactions. Multiple testing correction via Hochberg step-up procedure controls family-wise error rate with strong power for positive-dependence structures. **Validation step:** Confirm findings with GAM (Generalized Additive Model method) to validate against parametric assumptions and identify nonlinear q-entropy patterns that may be missed by rank-based methods.

**High-confidence genes** are those significant in *both* methods—these provide the strongest evidence of genuine q-value × group interactions and are recommended for downstream biological validation. The Friedman Test is particularly appropriate for paired designs and provides distribution-free inference without requiring assumptions about data normality, accounting properly for the within-subject correlation structure.

Additionally, use **library size-normalized pseudocounts** for robust entropy estimation on sparse count data (via `estimate_pseudocount()`), and consider genes with **high effect sizes** (η² > 0.3) as particularly robust findings. When Friedman p-values cluster due to monotonic entropy patterns across q-values, effect size (η²) serves as the practical tiebreaker for ranking gene importance.

***

# Session Information

```{r session-info}
sessionInfo()
```

***

# References
