---
title: "TSENAT Appendix B: Comparing Flexible Parametric (GAM) and Rank-Based (K-W) Methods"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
bibliography: TSENAT.bib
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: GAM vs K-W Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Flexible Parametric and Rank-Based Methods

## Purpose

This vignette validates q-value × group interaction detection using two complementary statistical approaches:

1. **Flexible Parametric (GAM):** Generalized Additive Models with ARIMA(1,1,0) transformation for handling q-ordered measurements
2. **Rank-Based (K-W):** Kruskal-Wallis with adaptive test selection based on data characteristics

Both methods now incorporate principled improvements to handle real-world data challenges:
- Heteroscedasticity detection and weighting
- Boundary condition handling
- Automatic test selection based on data properties
- Multiple testing correction accounting for AR(1) correlation in q-values

## Key Questions

1. Do both methods identify the same significant genes?
2. How much do results concordant vs. discordant?
3. Which genes are truly robust (significant in both methods)?
4. What data characteristics drive method disagreement?

---

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(salmon_dataset)
mode(readcounts) <- "numeric"

# Subset for faster analysis (set to NULL for all genes)
n_genes_subset <- 200
if (!is.null(n_genes_subset) && nrow(readcounts) > n_genes_subset) {
  readcounts <- readcounts[1:n_genes_subset, ]
  cat("NOTE: Analyzing", n_genes_subset, "genes for speed. Set n_genes_subset=NULL for all genes.\n\n")
}

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Build SummarizedExperiment and filter
se <- build_se(readcounts, gff3_dataset, metadata = metadata_df)
se <- tryCatch(
  filter_se(se, stringency = "loose"),
  error = function(e) se
)

```

## Bayesian Pseudocount Regularization

This analysis incorporates **WLFC (Weighted Likelihood Fold Change)** empirical Bayesian regularization to handle sparse count data. The `estimate_wlfc_pseudocounts()` function:

1. **Fits an empirical Beta prior** from the observed count data across all genes
2. **Computes gene-specific WLFC pseudocounts** based on each gene's posterior distribution
3. **Returns a scalar pseudocount** (mean of per-gene estimates) for entropy regularization

This data-driven approach is more principled than fixed pseudocounts, as it learns the regularization strength from the actual data characteristics.

```{r bayesian-pseudocounts, message=FALSE}
# Estimate WLFC pseudocounts integrated workflow
result <- estimate_wlfc_pseudocounts(se, verbose = FALSE)
scalar_pseudocount <- result$scalar_pseudocount

# Calculate Tsallis entropy with Bayesian pseudocount regularization
qvec <- seq(0.1, 2, by = 0.05)
ts_se <- calculate_diversity(
  se, 
  q = qvec, 
  norm = TRUE, 
  pseudocount = scalar_pseudocount,
  metadata = metadata_df,
  shrinkage = "empirical_bayes"
)

```

# Rank-Based Approach (Kruskal-Wallis)

## Assumption Validation

Before applying rank-based methods, we verify that the entropy data meets the key assumptions for robust non-parametric inference.

```{r rank-assumptions-calculation, message=FALSE}
# Extract entropy matrix for assumption validation
entropy_matrix <- assay(ts_se)

# Test rank-based assumptions:
# 1. Exchangeability: samples are independent and identically distributed
# 2. Monotonicity: rank ordering is preserved across observations
# 3. Consistency: ranks are consistent across replicate measurements
rank_assumptions <- test_rankbased_assumptions(entropy_matrix)
```

```{r rank-assumptions-output, message=FALSE, echo=FALSE}
# Extract summary statistics and test results from rank_assumptions object
summary_stats <- attr(rank_assumptions, "summary_stats")
check_results <- attr(rank_assumptions, "checks")

# Create entropy data summary table
entropy_summary <- data.frame(
  Metric = c(
    "Genes tested",
    "Samples (q-values × groups)",
    "Entropy range",
    "Mean entropy",
    "Median entropy",
    "Missing values"
  ),
  Value = c(
    summary_stats$n_genes,
    summary_stats$n_samples,
    sprintf("%.4f to %.4f", summary_stats$entropy_min, summary_stats$entropy_max),
    sprintf("%.4f", summary_stats$entropy_mean),
    sprintf("%.4f", summary_stats$entropy_median),
    summary_stats$n_missing
  )
)

knitr::kable(entropy_summary, 
             caption = "Rank-Based Entropy Data Summary",
             align = c("l", "r"))

cat("\n")

# Create rigorous assumption checks table
exchangeability_check <- check_results$exchangeability
monotonicity_check <- check_results$monotonicity
consistency_check <- check_results$consistency

assumptions_checks <- data.frame(
  Assumption = c(
    "Exchangeability",
    "Monotonicity",
    "Consistency"
  ),
  Test = c(
    "Permutation test",
    "Spearman correlation",
    "Kendall's W + ICC"
  ),
  Result = c(
    sprintf("p=%.3f", exchangeability_check$p_value),
    sprintf("r=%.3f (±%.3f)", monotonicity_check$mean_correlation, monotonicity_check$sd_correlation),
    sprintf("W=%.3f, ICC=%.3f", consistency_check$kendall_w, consistency_check$icc_simplified)
  ),
  Status = c(
    exchangeability_check$status,
    monotonicity_check$status,
    consistency_check$status
  )
)

knitr::kable(assumptions_checks, 
             caption = "Rigorous Statistical Tests for Rank-Based Method Assumptions",
             align = c("l", "l", "r", "r"))
```

```{r kw-analysis, message=FALSE}
# Run rank-based interaction test with automatic SummarizedExperiment conversion
# detect_q_gene_interactions() handles the data transformation internally:
# 1. Extracts entropy matrix from ts_se assay
# 2. Retrieves q-values from colData
# 3. Extracts gene identifiers from rownames
# 4. Converts to long-format data frame
# 5. Selects appropriate rank-based test (Kruskal-Wallis, ART, Mood's Median)
#    based on data characteristics
# 6. Applies multiple testing correction with AR(1) correlation awareness

kw_results <- detect_q_gene_interactions(
    ts_se,
    multicorr = "hochberg",
    verbose = FALSE
)
```

```{r kw-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect_kw <- mean(kw_results$effect_size_eta2, na.rm = TRUE)
median_effect_kw <- median(kw_results$effect_size_eta2, na.rm = TRUE)

kw_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "NAs",
    "Mean effect size (η²)",
    "Median effect size (η²)",
    "Strong effect genes (η² > 10%)"
  ),
  Value = c(
    nrow(kw_results),
    sum(kw_results$p_value < 0.05, na.rm = TRUE),
    sum(is.na(kw_results$p_value)),
    if (is.na(mean_effect_kw) || !is.finite(mean_effect_kw)) "Not available" else sprintf("%.1f%%", mean_effect_kw * 100),
    if (is.na(median_effect_kw) || !is.finite(median_effect_kw)) "Not available" else sprintf("%.1f%%", median_effect_kw * 100),
    sum(kw_results$p_value < 0.05 & kw_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(kw_summary_stats, caption = "Rank-Based (K-W) Results Summary", align = c("l", "r"))

# Show top genes with effect size and test method
kw_sig <- kw_results[!is.na(kw_results$p_value), ]
if (nrow(kw_sig) > 0) {
  kw_sig <- kw_sig[order(kw_sig$p_value), ]
  kw_sig <- kw_sig[1:min(10, nrow(kw_sig)), ]
  
  # Select columns: gene, p-values, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(kw_sig))
  if (length(cols_to_show) > 0) {
    kw_display <- kw_sig[, cols_to_show, drop = FALSE]
    
    # Format p-values in scientific notation for readability
    for (col in c("p_value", "adj_p_value")) {
      if (col %in% colnames(kw_display)) {
        kw_display[[col]] <- sapply(kw_display[[col]], function(x) {
          if (is.na(x)) "NA" else sprintf("%.3e", x)
        })
      }
    }
    
    # Format effect size as percentage if available
    if ("effect_size_eta2" %in% colnames(kw_display)) {
      kw_display$effect_size_eta2 <- sapply(kw_display$effect_size_eta2, function(x) {
        if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
      })
    }
    
    # Improve column distribution: right-align numeric columns, left-align text
    # gene | p-value | adj_p | eta2 | test_method | interaction_class
    knitr::kable(kw_display, 
                 caption = "Top 10 Genes by K-W p-value (with Effect Size & Test Method)",
                 align = c("l", "r", "r", "r", "l", "l"),
                 format.args = list(big.mark = ","))
  }
}
```

---

# Flexible Parametric Approach (GAM)

This section loads precomputed GAM results from `lm_interaction_results.csv`. The GAM analysis uses flexible parametric modeling to detect q-value × group interactions:

- **Model:** Generalized Additive Models (GAM) with ARIMA(1,1,0) transformations
- **Approach:** Smooth functions fit across q-values for each gene
- **Correction:** Hochberg multiple testing correction accounting for AR(1) correlation
- **Precomputation:** Results computed separately for efficiency

The precomputed results include effect sizes, model convergence diagnostics, and heteroscedasticity detection.

```{r gam-analysis, message=FALSE}
# Load precomputed GAM results from CSV file
gam_results <- read.csv(
    system.file("extdata", "lm_interaction_results.csv", package = "TSENAT"),
    stringsAsFactors = FALSE
)
```

```{r gam-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"))

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format p-values in scientific notation
      for (col in c("p_interaction", "adj_p_interaction")) {
        if (col %in% colnames(gam_display)) {
          gam_display[[col]] <- sapply(gam_display[[col]], function(x) {
            if (is.na(x)) "NA" else sprintf("%.3e", x)
          })
        }
      }
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size & Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"))
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

---

# Method Concordance Analysis

```{r concordance-summary, message=FALSE}
# Compute concordance analysis using dedicated function
concordance_result <- compute_method_concordance(gam_results, kw_results)

# Extract components
comparison_df <- concordance_result$comparison_df
spearman_rho <- concordance_result$spearman_rho
high_conf <- concordance_result$high_conf
agreement_table <- concordance_result$agreement_table
```

```{r concordance-output, message=FALSE, echo=FALSE}
if (is.null(comparison_df)) {
  # Silently skip - no output
} else {
  # Create agreement summary table
  agreement_stats <- data.frame(
    Category = names(agreement_table),
    Count = as.numeric(agreement_table),
    Proportion = as.numeric(prop.table(agreement_table)),
    stringsAsFactors = FALSE
  )
  
  # Format proportions as percentages
  agreement_stats$Proportion <- sprintf("%.1f%%", agreement_stats$Proportion * 100)
  
  knitr::kable(agreement_stats, 
               caption = "Agreement Summary between GAM and K-W Methods",
               align = c("l", "r", "r"))
  
  cat("\n")
  
  # High-confidence genes table
  if (nrow(high_conf) > 0) {
    high_conf_display <- high_conf[1:min(15, nrow(high_conf)), 
                                   c("gene", "padj_gam", "padj_kw", "effect_gam", "effect_kw")]
    rownames(high_conf_display) <- NULL
    
    # Format p-values
    high_conf_display$padj_gam <- sapply(high_conf_display$padj_gam, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3e", x)
    })
    high_conf_display$padj_kw <- sapply(high_conf_display$padj_kw, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3e", x)
    })
    
    # Format effect sizes
    high_conf_display$effect_gam <- sapply(high_conf_display$effect_gam, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3f", x)
    })
    high_conf_display$effect_kw <- sapply(high_conf_display$effect_kw, function(x) {
      if (is.na(x)) "NA" else sprintf("%.3f", x)
    })
    
    # Rename columns for display
    colnames(high_conf_display) <- c("Gene", "GAM padj", "K-W padj", "GAM Effect", "K-W Effect")
    
    knitr::kable(high_conf_display, 
                 caption = "Top High-Confidence Genes Detected by Both Methods",
                 align = c("l", "r", "r", "r", "r"))
  }
}
```

## Visualization: Method Comparison

```{r concordance-plot, fig.width=10, fig.height=8}
# Create comparison plot only if we have valid comparison data
if (exists("comparison_df") && nrow(comparison_df) > 0) {
  plot_method_concordance(comparison_df)
} else {
  cat("Visualization skipped: No valid comparison data available\n")
}
```

---

# Recommendations

## Best Practices for Analysis

We recommend a **complementary two-method approach** for robust detection of q-value × group interactions:

**Primary analysis:** Start with GAM (flexible parametric method) as it offers superior power for detecting nonlinear q-entropy patterns. Use the default Hochberg multiple testing correction for stringent FWER control. **Validation step:** Confirm findings with rank-based Kruskal-Wallis analysis to ensure robustness against parametric assumptions. The conditional test selection automatically adapts to data characteristics (heteroscedasticity, boundary clustering, skewness), providing distribution-free validation without requiring assumptions about data normality.

**High-confidence genes** are those significant in *both* methods—these provide the strongest evidence of genuine q-value × group interactions and are recommended for downstream biological validation. Additionally, use **Bayesian pseudocount regularization (WLFC)** for robust entropy estimation on sparse count data, and leverage **multiple testing correction that accounts for AR(1) correlation** in q-value sequences (available via Hochberg or Benjamini-Yekutieli methods depending on whether you prioritize FWER or FDR control).

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References
