---
title: "Appendix A: Equivalence Validation - TSENAT vs SplicingFactory"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_document:
        toc: true
        fig_width: 10
        fig_height: 6.5
bibliography: TSENAT.bib
vignette: |
    %\VignetteIndexEntry{TSENAT Validation: SplicingFactory Equivalence}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5
)
```


# Introduction

This appendix validates that **TSENAT's transcript diversity measures (Shannon entropy, Simpson index) are mathematically equivalent to SplicingFactory's implementations**. We benchmark both packages on TCGA BRCA RNA-seq data to demonstrate:

- Shannon equivalence: TSENAT Tsallis q=1 ≈ SplicingFactory Shannon
- Simpson equivalence: TSENAT Tsallis q=2 ≈ SplicingFactory Simpson  
- Identical statistical results: Same gene rankings, p-values, fold changes

**Who needs this appendix?** Anyone evaluating TSENAT for transcript diversity analysis or validating cross-package comparisons.

***

## Document Structure

- Background: Conceptual overview of transcript diversity and the Tsallis framework
- Benchmarking (main section): Hands-on comparison using both packages
- Results: Side-by-side analysis tables (Shannon and Simpson)
- Analysis: Technical explanation of equivalence, normalization differences, and when to use each method
- Technical Note: Gene filtering implementation difference
- Conclusion: Summary and recommendations

***

## Background: Transcript Diversity Concepts

**Transcript diversity** measures how variable isoform expression is within genes. Two standard approaches:

- Shannon entropy: Overall diversity (number and evenness of isoforms)
- Simpson index: Dominance (whether one isoform monopolizes expression)

TSENAT's approach: Generalized Tsallis entropy framework where q=1 gives Shannon and q=2 gives Simpson. This allows sensitivity analysis across the q continuum while maintaining equivalence with established methods.

**Key question**: Does TSENAT's mathematical generalization produce identical statistical results?



***

## Dataset and Methods

We use TCGA BRCA RNA-seq data as our validation dataset:
- Public, reproducible resource
- Multiple transcript isoforms per gene
- Real biological signal (Normal vs. Tumor phenotypes)
- Variable sequencing depth and isoform abundance

Analysis pipeline:
1. Load transcript-level read counts
2. Filter genes (minimum 6 expressed isoforms across all samples)
3. Calculate diversity using both SplicingFactory and TSENAT
4. Statistical testing (Wilcoxon rank-sum) for group differences
5. Compare results (concordance, effect sizes, rankings)

***

# Benchmarking

### Importing example data

```{r setup, message=FALSE, warning=FALSE}
library("SplicingFactory")
library(TSENAT)
library("SummarizedExperiment")

# Load dataset
data(tcga_brca_luma_dataset)

# Extract gene names
genes <- tcga_brca_luma_dataset[, 1]

# Extract read count data without gene names
readcounts <- tcga_brca_luma_dataset[, -1]
```

### Data filtering and preprocessing

```{r readfilter}
tokeep <- rowSums(readcounts > 5) > 5
readcounts <- readcounts[tokeep, ]
genes      <- genes[tokeep]

# Create SummarizedExperiment object for cleaner data handling
# Important: Set gene names as rownames for compatibility with SplicingFactory
se_data <- SummarizedExperiment(
  assays = list(counts = as.matrix(readcounts)),
  rowData = DataFrame(gene = genes)
)
rownames(se_data) <- genes
```


```{r helper_functions, message=FALSE, warning=FALSE, include=FALSE}
#' Format Top Genes Table
#' 
#' @description Extract and format top differentially expressed genes for display,
#' with consistent column naming and numerical formatting.
#'
#' @param results_df Data frame containing differential analysis results
#' @param gene_col Column name containing gene identifiers
#' @param padj_col Column name containing adjusted p-values
#' @param n_top Number of top genes to extract (default: 10)
#' @param select_cols Vector of column names to select (default: standard columns)
#' @param col_names Vector of display names for columns (for knitr::kable)
#'
#' @return Data frame with formatted top genes, ready for display
format_top_genes <- function(results_df, gene_col, padj_col, n_top = 10, 
                            select_cols = NULL, col_names = NULL) {
  # Default columns to display
  if (is.null(select_cols)) {
    select_cols <- c(gene_col, "Normal_mean", "Tumor_mean", "mean_difference", 
                    "log2_fold_change", "pvalue", padj_col)
  }
  
  top_genes <- results_df %>%
    arrange(across(all_of(padj_col))) %>%
    slice(1:min(n_top, nrow(results_df))) %>%
    select(all_of(intersect(select_cols, colnames(results_df)))) %>%
    mutate(
      across(where(is.numeric) & !matches("abundance|mean|fold|stat"), 
             ~ format(., scientific = TRUE, digits = 3)),
      across(matches("_mean$|mean_"), ~ round(., 4)),
      across(matches("fold_change|difference"), ~ round(., 4))
    )
  
  # Rename columns if provided
  if (!is.null(col_names) && length(col_names) == ncol(top_genes)) {
    colnames(top_genes) <- col_names
  }
  
  top_genes
}

# Create summary statistics comparison table
#' Create Summary Statistics Comparison
#'
#' @description Compare differential analysis results between two methods,
#' computing concordance metrics and statistical thresholds.
#'
#' @param method1_results Data frame with first method's results
#' @param method2_results Data frame with second method's results
#' @param method1_name Display name for first method
#' @param method2_name Display name for second method
#' @param padj_col1 Adjusted p-value column name in method1 results
#' @param padj_col2 Adjusted p-value column name in method2 results
#'
#' @return Data frame summarizing comparison statistics
create_summary_stats <- function(method1_results, method2_results, 
                               method1_name, method2_name,
                               padj_col1 = "adjusted_p_values", 
                               padj_col2 = "padj") {
  data.frame(
    Method = c(method1_name, method2_name),
    "Genes Tested" = c(nrow(method1_results), nrow(method2_results)),
    "Significant padj<0.05" = c(
      sum(method1_results[[padj_col1]] < 0.05, na.rm = TRUE),
      sum(method2_results[[padj_col2]] < 0.05, na.rm = TRUE)
    ),
    "Significant padj<0.01" = c(
      sum(method1_results[[padj_col1]] < 0.01, na.rm = TRUE),
      sum(method2_results[[padj_col2]] < 0.01, na.rm = TRUE)
    ),
    "Mean log₂FC" = c(
      round(mean(method1_results$log2_fold_change, na.rm = TRUE), 3),
      round(mean(method2_results$log2_fold_change, na.rm = TRUE), 3)
    ),
    "Median log₂FC" = c(
      round(median(method1_results$log2_fold_change, na.rm = TRUE), 3),
      round(median(method2_results$log2_fold_change, na.rm = TRUE), 3)
    ),
    "Min padj" = c(
      format(min(method1_results[[padj_col1]], na.rm = TRUE), scientific = TRUE, digits = 3),
      format(min(method2_results[[padj_col2]], na.rm = TRUE), scientific = TRUE, digits = 3)
    ),
    stringsAsFactors = FALSE
  )
}

# Calculate gene overlap statistics (both in padj < threshold)
#' Calculate Gene Overlap Statistics
#'
#' @description Quantify agreement between two differential analysis methods,
#' showing genes detected by both, only by first, or only by second method.
#'
#' @param results1 Data frame with first method's results
#' @param results2 Data frame with second method's results
#' @param gene_col1 Column name for genes in results1
#' @param gene_col2 Column name for genes in results2
#' @param padj_col1 Adjusted p-value column in results1
#' @param padj_col2 Adjusted p-value column in results2
#' @param threshold Significance threshold (default: 0.05)
#'
#' @return Data frame showing overlap statistics with counts and percentages
gene_overlap_stats <- function(results1, results2, gene_col1, gene_col2, 
                              padj_col1, padj_col2, threshold = 0.05) {
  sig1 <- results1[[gene_col1]][results1[[padj_col1]] < threshold]
  sig2 <- results2[[gene_col2]][results2[[padj_col2]] < threshold]
  
  both <- length(intersect(sig1, sig2))
  only1 <- length(setdiff(sig1, sig2))
  only2 <- length(setdiff(sig2, sig1))
  total <- length(union(sig1, sig2))
  
  data.frame(
    Category = c("Detected by both methods", "Only by first method", "Only by second method"),
    Count = c(both, only1, only2),
    Percentage = c(
      round(100 * both / total, 1),
      round(100 * only1 / total, 1),
      round(100 * only2 / total, 1)
    )
  )
}
```

### Transcript diversity calculation

To calculate transcript diversity measures (Shannon entropy, Simpson index, and Tsallis entropy at q=1 and q=2), use:

```{r diversity_calculations}
# Create sample group classification based on sample names
# (Normal samples end with _N, Tumor samples don't)
sample_group <- ifelse(grepl("_N$", colnames(se_data)), "Normal", "Tumor")

# Add group metadata to se_data for all downstream analyses
colData(se_data)$group <- sample_group

# SplicingFactory: Shannon entropy (Laplace method)
# Note: SplicingFactory extracts genes from rownames(se_data) automatically
shannon_entropy <- SplicingFactory::calculate_diversity(
  x = se_data,
  method = "naive",
  norm = TRUE,
  verbose = FALSE
)

# SplicingFactory: Simpson index
simpson_index <- SplicingFactory::calculate_diversity(
  x = se_data,
  method = "simpson",
  norm = TRUE,
  verbose = FALSE
)

# Add sample_type and group metadata to SplicingFactory objects
# (sample_type matches SplicingFactory naming; group matches TSENAT naming for consistency)
colData(shannon_entropy)$sample_type <- sample_group
colData(shannon_entropy)$group <- sample_group
colData(simpson_index)$sample_type <- sample_group
colData(simpson_index)$group <- sample_group

# TSENAT: Create TSENATAnalysis objects for S4 wrapper approach
# This maintains consistent workflow with S4 infrastructure
tsenat_analysis_q1 <- TSENATAnalysis(se_data)
tsenat_analysis_q2 <- TSENATAnalysis(se_data)

# Set condition column in config for downstream differential analysis
tsenat_analysis_q1 <- setConfigValue(tsenat_analysis_q1, "condition_col", "group")
tsenat_analysis_q2 <- setConfigValue(tsenat_analysis_q2, "condition_col", "group")

# TSENAT: Tsallis entropy at q=1 (equals Shannon entropy) - using S4 wrapper
tsenat_analysis_q1 <- calculate_diversity_s4(
  analysis = tsenat_analysis_q1,
  q = 1,
  norm = TRUE,
  verbose = FALSE
)
tsenat_q1_values <- diversity(tsenat_analysis_q1, q = 1)

# TSENAT: Tsallis entropy at q=2 (equals Simpson) - using S4 wrapper
tsenat_analysis_q2 <- calculate_diversity_s4(
  analysis = tsenat_analysis_q2,
  q = 2,
  norm = TRUE,
  verbose = FALSE
)
tsenat_q2_values <- diversity(tsenat_analysis_q2, q = 2)
```

Both packages return a `SummarizedExperiment` object, that you can investigate further with the `assay` function.


### Differential analysis

```{r differential_analysis}
# SplicingFactory: Differential analysis for Shannon entropy
entropy_significance <- SplicingFactory::calculate_difference(
  x = shannon_entropy,
  samples = "sample_type",
  control = "Normal",
  method = "mean",
  test = "wilcoxon",
  verbose = FALSE
)

# SplicingFactory: Differential analysis for Simpson index
simpson_significance <- SplicingFactory::calculate_difference(
  x = simpson_index,
  samples = "sample_type",
  control = "Normal",
  method = "mean",
  test = "wilcoxon",
  verbose = FALSE
)

# TSENAT: Differential analysis for Tsallis q=1
# Ensure the grouping column is set in config for the wrapper
tsenat_analysis_q1 <- setConfigValue(tsenat_analysis_q1, "condition_col", "group")
tsenat_analysis_q1 <- calculate_difference_s4(
  analysis = tsenat_analysis_q1,
  control = "Normal",
  method = "mean",
  test = "wilcoxon",
  verbose = FALSE
)
# Extract results for downstream functions
tsenat_shannon_diff <- lmResults(tsenat_analysis_q1, component = "difference")

# TSENAT: Differential analysis for Tsallis q=2
# Ensure the grouping column is set in config for the wrapper
tsenat_analysis_q2 <- setConfigValue(tsenat_analysis_q2, "condition_col", "group")
tsenat_analysis_q2 <- calculate_difference_s4(
  analysis = tsenat_analysis_q2,
  control = "Normal",
  method = "mean",
  test = "wilcoxon",
  verbose = FALSE
)
# Extract results for downstream functions
tsenat_simpson_diff <- lmResults(tsenat_analysis_q2, component = "difference")
```


For the benchmark comparison with TSENAT, we will compute Tsallis entropy 
at q=1 (equivalent to Shannon) and q=2 (equivalent to Simpson). This 
allows direct comparison with the SplicingFactory results shown above.

# Comparison with TSENAT: Wilcoxon Differential Analysis

Compare diversity analysis results between SplicingFactory and TSENAT using Wilcoxon 
tests. We examine Shannon entropy (SplicingFactory) versus Tsallis q=1 (TSENAT), 
and Simpson index (SplicingFactory) versus Tsallis q=2 (TSENAT). The theoretical 
equivalence between these measures allows direct comparison:

- q = 1: Tsallis entropy equals Shannon entropy
- q = 2: Tsallis entropy equals Simpson index

## Benchmark Results Summary

### Shannon Entropy (SplicingFactory Laplace vs TSENAT Tsallis q=1)

```{r shannon_summary_stats, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(knitr)

# Summary statistics using helper function
shannon_summary <- create_summary_stats(
  entropy_significance, tsenat_shannon_diff,
  "SplicingFactory (Laplace)", "TSENAT (Tsallis q=1)",
  padj_col1 = "adjusted_p_values", padj_col2 = "padj"
)

knitr::kable(shannon_summary, caption = "Shannon Entropy Analysis Summary Statistics", align = c("l", "r", "r", "r", "r", "r", "r"))
```

**Interpretation**: Both methods similarly identify significantly different transcript diversity between Normal and Tumor samples. The comparable number of significant genes (padj<0.05) and mean log₂FC values demonstrate equivalent statistical power and effect size detection.

#### Top 10 Significant Genes - SplicingFactory (Laplace)
```{r shannon_sf_top_genes, message=FALSE, warning=FALSE, echo=FALSE}
shannon_sf_top <- format_top_genes(
  entropy_significance %>% select(genes, Normal_mean, Tumor_mean, mean_difference, 
                                  log2_fold_change, raw_p_values, adjusted_p_values),
  gene_col = "genes",
  padj_col = "adjusted_p_values",
  n_top = 10,
  select_cols = c("genes", "Normal_mean", "Tumor_mean", "mean_difference", "log2_fold_change", "raw_p_values", "adjusted_p_values"),
  col_names = c("Gene", "Normal Mean", "Tumor Mean", "Mean Diff", "log2FC", "p-value", "adj p-value")
)
knitr::kable(shannon_sf_top, row.names = FALSE, caption = "SplicingFactory (Laplace) - Top 10 Significant Genes", align = c("l", "r", "r", "r", "r", "r", "r"))
```

#### Top 10 Significant Genes - TSENAT (Tsallis q=1)
```{r shannon_tsenat_top_genes, message=FALSE, warning=FALSE, echo=FALSE}
shannon_tsenat_top <- format_top_genes(
  tsenat_shannon_diff %>% select(gene_id, Normal_mean, Tumor_mean, mean_difference, 
                                 log2_fold_change, pvalue, padj),
  gene_col = "gene_id",
  padj_col = "padj",
  n_top = 10,
  select_cols = c("gene_id", "Normal_mean", "Tumor_mean", "mean_difference", "log2_fold_change", "pvalue", "padj"),
  col_names = c("Gene", "Normal Mean", "Tumor Mean", "Mean Diff", "log2FC", "p-value", "adj p-value")
)
knitr::kable(shannon_tsenat_top, row.names = FALSE, caption = "TSENAT (Tsallis q=1) - Top 10 Significant Genes", align = c("l", "r", "r", "r", "r", "r", "r"))
```

```{r comparison_volcano_ma_plots, fig.width=14, fig.height=7, message=FALSE, warning=FALSE}
# Create comparison plots: Volcano and MA plots side-by-side for each method
library(cowplot)

# Create individual volcano and MA plots for Shannon/q=1
# SplicingFactory method for comparison
shannon_sf <- TSENAT:::plot_volcano_ma_grid(
  diff_df = entropy_significance,
  x_col = "mean_difference",
  padj_col = "adjusted_p_values",
  sig_alpha = 0.05,
  top_n = 5,
  title_volcano = "",
  title_ma = ""
)

# TSENAT S4 wrapper method
shannon_tsenat <- plot_volcano_ma_grid_s4(
  analysis = tsenat_analysis_q1,
  x_col = "mean_difference",
  padj_col = "padj",
  sig_alpha = 0.05,
  top_n = 5,
  title_volcano = "",
  title_ma = "",
  verbose = FALSE
)

# Create title labels for each grid
shannon_sf_title <- cowplot::ggdraw() + 
  cowplot::draw_label("Shannon (SplicingFactory)",
                     fontface = "bold", size = 16, x = 0.5, y = 0.5)

shannon_tsenat_title <- cowplot::ggdraw() + 
  cowplot::draw_label("Tsallis q=1 (TSENAT)",
                     fontface = "bold", size = 16, x = 0.5, y = 0.5)

# Create and combine all Shannon comparison plots
shannon_comparison <- cowplot::plot_grid(
  cowplot::ggdraw() + 
    cowplot::draw_label("Volcano and MA plot Comparison: SplicingFactory vs TSENAT",
                       fontface = "bold", size = 20, x = 0.5, y = 0.5),
  cowplot::plot_grid(
    cowplot::plot_grid(shannon_sf_title, shannon_sf, nrow = 2, rel_heights = c(0.12, 1)),
    cowplot::plot_grid(shannon_tsenat_title, shannon_tsenat, nrow = 2, rel_heights = c(0.12, 1)),
    ncol = 2
  ),
  nrow = 2,
  rel_heights = c(0.08, 1)
)

print(shannon_comparison)
```

***

### Simpson Index (SplicingFactory vs TSENAT Tsallis q=2)

```{r simpson_summary_stats, message=FALSE, warning=FALSE, echo=FALSE}
# Summary statistics using helper function
simpson_summary <- create_summary_stats(
  simpson_significance, tsenat_simpson_diff,
  "SplicingFactory (Simpson)", "TSENAT (Tsallis q=2)",
  padj_col1 = "adjusted_p_values", padj_col2 = "padj"
)

knitr::kable(simpson_summary, caption = "Simpson Index Analysis Summary Statistics", align = c("l", "r", "r", "r", "r", "r", "r"))
```

**Interpretation**: Simpson index results parallel Shannon entropy findings, with both methods identifying similar numbers of significantly different isoform dominance patterns. The agreement across both diversity measures provides strong evidence for TSENAT's mathematical and statistical equivalence to SplicingFactory.

#### Top 10 Significant Genes - SplicingFactory (Simpson)
```{r simpson_sf_top_genes, message=FALSE, warning=FALSE, echo=FALSE}
simpson_sf_top <- format_top_genes(
  simpson_significance %>% select(genes, Normal_mean, Tumor_mean, mean_difference, 
                                  log2_fold_change, raw_p_values, adjusted_p_values),
  gene_col = "genes",
  padj_col = "adjusted_p_values",
  n_top = 10,
  select_cols = c("genes", "Normal_mean", "Tumor_mean", "mean_difference", "log2_fold_change", "raw_p_values", "adjusted_p_values"),
  col_names = c("Gene", "Normal Mean", "Tumor Mean", "Mean Diff", "log2FC", "p-value", "adj p-value")
)
knitr::kable(simpson_sf_top, row.names = FALSE, caption = "SplicingFactory (Simpson) - Top 10 Significant Genes", align = c("l", "r", "r", "r", "r", "r", "r"))
```

#### Top 10 Significant Genes - TSENAT (Tsallis q=2)
```{r simpson_tsenat_top_genes, message=FALSE, warning=FALSE, echo=FALSE}
simpson_tsenat_top <- format_top_genes(
  tsenat_simpson_diff %>% select(gene_id, Normal_mean, Tumor_mean, mean_difference, 
                                 log2_fold_change, pvalue, padj),
  gene_col = "gene_id",
  padj_col = "padj",
  n_top = 10,
  select_cols = c("gene_id", "Normal_mean", "Tumor_mean", "mean_difference", "log2_fold_change", "pvalue", "padj"),
  col_names = c("Gene", "Normal Mean", "Tumor Mean", "Mean Diff", "log2FC", "p-value", "adj p-value")
)
knitr::kable(simpson_tsenat_top, row.names = FALSE, caption = "TSENAT (Tsallis q=2) - Top 10 Significant Genes", align = c("l", "r", "r", "r", "r", "r", "r"))
```


```{r comparison_ma_plots, fig.width=14, fig.height=7, message=FALSE, warning=FALSE}
# Create individual volcano and MA plots for Simpson/q=2
# SplicingFactory method for comparison
simpson_sf <- TSENAT:::plot_volcano_ma_grid(
  diff_df = simpson_significance,
  x_col = "mean_difference",
  padj_col = "adjusted_p_values",
  sig_alpha = 0.05,
  top_n = 5,
  title_volcano = "",
  title_ma = ""
)

# TSENAT S4 wrapper method
simpson_tsenat <- plot_volcano_ma_grid_s4(
  analysis = tsenat_analysis_q2,
  x_col = "mean_difference",
  padj_col = "padj",
  sig_alpha = 0.05,
  top_n = 5,
  title_volcano = "",
  title_ma = "",
  verbose = FALSE
)

# Create title labels for each grid
simpson_sf_title <- cowplot::ggdraw() + 
  cowplot::draw_label("Simpson (SplicingFactory)",
                     fontface = "bold", size = 16, x = 0.5, y = 0.5)

simpson_tsenat_title <- cowplot::ggdraw() + 
  cowplot::draw_label("Tsallis q=2 (TSENAT)",
                     fontface = "bold", size = 16, x = 0.5, y = 0.5)

# Create and combine all Simpson comparison plots
simpson_comparison <- cowplot::plot_grid(
  cowplot::ggdraw() + 
    cowplot::draw_label("Volcano and MA plot Comparison: SplicingFactory vs TSENAT",
                       fontface = "bold", size = 20, x = 0.5, y = 0.5),
  cowplot::plot_grid(
    cowplot::plot_grid(simpson_sf_title, simpson_sf, nrow = 2, rel_heights = c(0.12, 1)),
    cowplot::plot_grid(simpson_tsenat_title, simpson_tsenat, nrow = 2, rel_heights = c(0.12, 1)),
    ncol = 2
  ),
  nrow = 2,
  rel_heights = c(0.08, 1)
)

print(simpson_comparison)
```

***

## Analysis: Simpson Index and Tsallis Entropy at q=2

Both methods share the same mathematical formula:

$$Simpson = Tsallis_{q=2} = 1 - \sum_{i=1}^n p_i^2$$

where $p_i = \frac{x_i}{\sum_i x_i}$ (transcript proportions).

Howerver, SplicingFactory returns raw values, while TSENAT normalizes by the theoretical maximum $(1-1/n)$:

| | SplicingFactory | TSENAT |
|---|---|---|
| Formula | $1 - \sum p_i^2$ | $\frac{1 - \sum p_i^2}{1 - 1/n}$ |
| Range | 0 to ~1 | 0 to 1 |
| Scale difference | Raw | ~2x for 2-isoform genes |

This explains observed differences like *C1orf213* Normal: 0.3852 (SplicingFactory) vs 0.7703 (TSENAT). Since C1orf213 has 2 expressed isoforms, the normalization factor is $1 - 1/2 = 0.5$. The calculation: $0.3852 / 0.5 = 0.7704$ confirms this scaling factor. Despite raw value differences, statistical analysis yields identical results:

1. log2 fold changes: Normalization factor cancels in ratios: $\log_2\left(\frac{S_{norm,t}}{S_{norm,n}}\right) = \log_2\left(\frac{S_{raw,t}/K}{S_{raw,n}/K}\right) = \log_2\left(\frac{S_{raw,t}}{S_{raw,n}}\right)$

2. P-values and rankings: Wilcoxon rank-sum tests depend only on gene ranks, not absolute values, so both methods identify the same significant genes

3. Statistical validity: Both approaches are correct—SplicingFactory favors unbounded diversity, TSENAT favors interpretable [0,1] scaling

***

## Technical Note: Gene Filtering Difference

The 1-gene difference between SplicingFactory (214 genes tested) and TSENAT (213 genes tested) is due to a subtle but important difference in how the two packages filter genes before statistical testing.

Implementation Difference:

SplicingFactory's Implementation:
```{r sf_gene_filtering, eval=FALSE}
# SplicingFactory/R/calculate_difference.R (lines 163-164)
x$cond_1 <- apply(x[grep(unique(samples)[1], samples) + 1], 1, 
                   function(x) sum(!is.na(x)))
x$cond_2 <- apply(x[grep(unique(samples)[2], samples) + 1], 1, 
                   function(x) sum(!is.na(x)))

if (test == "wilcoxon") {
    y <- x[x$cond_1 >= 3 & x$cond_2 >= 3 & sum(x$cond_1, x$cond_2) >= 8, ]
    #                                         ^^^^^^^^^^^^^^^^^^^^^^
    #                                         This computes a SCALAR
}
```

The problematic line uses `sum(x$cond_1, x$cond_2)`, which sums ALL values in both column vectors together, producing a single scalar (approximately 1,040 in our case). Since this global sum is always ≥ 8, the condition is essentially always TRUE, and the filtering is ineffective.

TSENAT's Implementation:
```{r tsenat_gene_filtering, eval=FALSE}
# TSENAT/R/difference_helpers.R (lines 57-58)
df$cond_1 <- rowSums(!is.na(df[, idx1 + 1, drop = FALSE]))
df$cond_2 <- rowSums(!is.na(df[, idx2 + 1, drop = FALSE]))

if (test == "wilcoxon") {
    keep_mask <- (df$cond_1 >= 3 & df$cond_2 >= 3 & (df$cond_1 + df$cond_2) >= 8)
    #                                                 ^^^^^^^^^^^^^^^^
    #                                                 Element-wise addition per gene
}
```

TSENAT uses `(df$cond_1 + df$cond_2)`, which performs element-wise addition to check each individual gene's total observation count.

The consequence is that TSENAT's filter is stricter: it requires each gene individually to have at least 8 total observations (3 in one group, 3 in the other, plus 2 more from either group). SplicingFactory's filter effectively doesn't work for the Wilcoxon test because the global sum condition is almost never FALSE.

***

# Conclusion: Equivalence Validation Summary

This appendix has demonstrated that TSENAT's implementation of transcript diversity measures is mathematically and statistically equivalent to SplicingFactory's established methods for both Shannon entropy and Simpson index calculations.

***

## Session Information

```{r session_info, include=TRUE}
sessionInfo()
```

```{r cleanup, include=FALSE}
if ("package:SplicingFactory" %in% search()) {
    detach("package:SplicingFactory", unload = TRUE)
}
```

