---
title: "TSENAT Appendix B: Comparing  Generalized Additive Model (GAM) and Rank-Based (Friedman) Method"
author: "Cristóbal Gallardo <gallardoalba@pm.me>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
  rmarkdown::pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6.5
    toc_depth: 2
    latex_engine: xelatex
bibliography: TSENAT.bib
vignette: |
  %\VignetteIndexEntry{TSENAT Appendix B: GAM vs Rank-Based Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-global, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 10,
    fig.height = 6.5,
    message = FALSE,
    warning = FALSE
)
```

# Overview: Comparing Generalized Additive Model and Rank-Based Methods

## Purpose

This vignette validates q-value × group interaction detection using two complementary statistical approaches:

1. **Generalized Additive Model (GAM):** Generalized Additive Models with ARIMA(1,1,0) transformation for handling q-ordered measurements [@C073]
2. **Rank-Based (Friedman with Hochberg):** Friedman test for paired/blocked data [@S115; @S205] with Hochberg step-up multiple testing correction [@BenjaminiHochberg1995]

Both methods now incorporate principled improvements to handle real-world data challenges:
- Heteroscedasticity detection and weighting
- Boundary condition handling
- Automatic test selection based on data properties
- Multiple testing correction using Hochberg step-up procedure

## Key Questions

1. Do both methods identify the same significant genes?
2. How much do results concordant vs. discordant?
3. Which genes are truly robust (significant in both methods)?
4. What data characteristics drive method disagreement?

***

# Setup

```{r setup-packages, message=FALSE}
suppressPackageStartupMessages({
    library(TSENAT)
    library(ggplot2)
    library(SummarizedExperiment)
    library(dplyr)
    library(gridExtra)
})

set.seed(42)
```

```{r setup-data, message=FALSE}
# Load preprocessed dataset
data(readcounts)
readcounts <- as.matrix(readcounts)
mode(readcounts) <- "numeric"

# Load metadata
metadata_df <- read.table(
    system.file("extdata", "metadata.tsv", package = "TSENAT"),
    header = TRUE, sep = "\t"
)

gff3_dataset <- system.file("extdata", "annotation.gff3.gz", package = "TSENAT")

# Configure analysis parameters first (best practice: fail-fast validation)
config <- tsenat_config(
  condition_col = "condition",
  subject_col = "paired_samples",
  q_values = seq(0, 2, by = 0.05),
  nthreads = 3,
  paired = TRUE,
  control = "normal"
)

# Build analysis object: creates SummarizedExperiment and initializes TSENATAnalysis
# Pass config at construction (Bioconductor pattern)
analysis <- build_analysis_s4(
    readcounts = readcounts,
    tx2gene = gff3_dataset,
    metadata = metadata_df,
    tpm = tpm,
    effective_length = effective_length,
    config = config
)

# Apply filtering for quality control
analysis <- filter_analysis_s4(
    analysis,
    stringency = "medium",
    min_isoform_abundance = 0
)
```

## Library Size Normalization for Pseudocounts

For robust entropy estimation on sparse count data, we normalize pseudocounts by library size. This approach is principled and recommended in edgeR and DESeq2, as it ensures regularization strength adapts to overall sequencing depth.

```{r calculate-tsallis, message=FALSE}
# Compute diversity using S4 wrapper with bootstrap confidence intervals
analysis <- calculate_diversity_s4(
  analysis, 
  norm = TRUE,
  pseudocount = "auto"
)
```

# Rank-Based Approach (Friedman)

## Assumption Validation

Before applying rank-based methods, we verify key assumptions for rank-based inference [@S115; @S204]. The Friedman test is a non-parametric alternative to repeated-measures ANOVA, making fewer distributional assumptions while maintaining validity for paired designs [@S044]:

```{r rank-assumptions-calculation, message=FALSE}
# Validate Friedman test assumptions using S4 wrapper
analysis <- test_rankbased_assumptions_s4(
    analysis
)

# Extract the result object from metadata using accessor function
rank_assumptions <- metadata(analysis, "rankbased_assumptions")$result
```

```{r rank-assumptions-output, message=FALSE, echo=FALSE}
# Extract summary statistics and test results from rank_assumptions object
summary_stats <- attr(rank_assumptions, "summary_stats")
check_results <- attr(rank_assumptions, "checks")

# Create entropy data summary table
entropy_summary <- data.frame(
  Metric = c(
    "Genes tested",
    "Samples (q-values × groups)",
    "Entropy range",
    "Mean entropy",
    "Median entropy",
    "Missing values"
  ),
  Value = c(
    summary_stats$n_genes,
    summary_stats$n_samples,
    sprintf("%.4f to %.4f", summary_stats$entropy_min, summary_stats$entropy_max),
    sprintf("%.4f", summary_stats$entropy_mean),
    sprintf("%.4f", summary_stats$entropy_median),
    summary_stats$n_missing
  )
)

knitr::kable(entropy_summary, 
             caption = "Rank-Based Entropy Data Summary",
             align = c("l", "r"))

cat("\n")

# Create assumption checks table
exchangeability_check <- check_results$exchangeability
monotonicity_check <- check_results$monotonicity
consistency_check <- check_results$consistency

data_characteristics <- data.frame(
  Characteristic = c("Paired Structure", "Gene Heterogeneity", "Subject Consistency"),
  Test = c("Permutation test", "Spearman r", "Kendall's W"),
  Result = c(
    sprintf("p=%.3f", exchangeability_check$p_value),
    sprintf("r=%.3f", monotonicity_check$mean_correlation),
    sprintf("W=%.3f", consistency_check$kendall_w)
  )
)

knitr::kable(data_characteristics, 
             caption = "Data Characteristics: Three Properties of the Entropy Matrix",
             align = c("l", "l", "r", "l"))
```

 


```{r rank-based-analysis, message=TRUE}
# Run Friedman test for q × condition INTERACTION (FIXED - March 2026)
# IMPORTANT: Now tests whether q-effect differs between conditions (tumor vs normal)
# PREVIOUS BEHAVIOR (INCORRECT): Tested only q main effect, ignoring condition
# NEW BEHAVIOR (CORRECT): Tests if entropy pattern across q-values differs by condition
# 
# With condition_col="sample_type", automatically uses:
#  - Two-way Friedman test (paired=TRUE: q within-subjects, condition between-subjects)
#  - Westfall-Young permutation correction respects AR(1) q-correlation + condition structure
analysis <- rank_test_q_condition_s4(
    analysis,
    multicorr = "hochberg"
)

# Extract results from S4 object using accessor
friedman_results <- rankResults(analysis)
```

```{r rank-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(friedman_results$effect_size_eta2, na.rm = TRUE)
median_effect <- median(friedman_results$effect_size_eta2, na.rm = TRUE)

rank_summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Significant (adj_p < 0.05, FWER-controlled)", 
    "NAs",
    "Mean effect size (η²)",
    "Median effect size (η²)",
    "Strong effect genes (η² > 10%)"
  ),
  Value = c(
    nrow(friedman_results),
    sum(friedman_results$p_value < 0.05, na.rm = TRUE),
    sum(friedman_results$adj_p_value < 0.05, na.rm = TRUE),
    sum(is.na(friedman_results$p_value)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(friedman_results$p_value < 0.05 & friedman_results$effect_size_eta2 > 0.10, na.rm = TRUE)
  )
)

knitr::kable(rank_summary_stats, caption = "Friedman Test Results Summary", align = c("l", "r"))

# Show top genes results
rank_sig <- friedman_results[!is.na(friedman_results$p_value), ]
if (nrow(rank_sig) > 0) {
  rank_sig <- rank_sig[order(rank_sig$adj_p_value), ]
  rank_sig <- rank_sig[1:min(10, nrow(rank_sig)), ]
  
  # Select columns: gene, p-values, f-statistic, effect size, test method, interaction class
  cols_to_show <- intersect(c("gene", "p_value", "adj_p_value", "f_statistic", "effect_size_eta2", 
                              "test_method", "interaction_class"), 
                            colnames(rank_sig))
  if (length(cols_to_show) > 0) {
    rank_display <- rank_sig[, cols_to_show, drop = FALSE]
    
    # Format numeric columns
    numeric_cols <- sapply(rank_display, is.numeric)
    rank_display[, numeric_cols] <- round(rank_display[, numeric_cols], 3)
    
    knitr::kable(rank_display, 
                 caption = "Top 10 Genes by Friedman Test",
                 digits = 3)
  }
}
```

Visualize q-curves for top genes:

```{r friedman-top-genes-qcurve, fig.width=12, fig.height=8, message=FALSE, fig.cap="Top 4 genes identified by Friedman rank-based test, showing q-curves (Tsallis entropy vs sensitivity parameter q). Curves depict entropy across q-values (0-2) for each group (typically control vs treatment). Friedman test identifies genes where rank patterns differ significantly between paired samples. Q-curves reveal whether group differences manifest uniformly across all diversity scales or are scale-specific (more pronounced at low q or high q)."}
top_genes_plot <- plot_tsallis_q_curve_s4(analysis, lm_res = friedman_results, n_top = 4)
print(top_genes_plot)
```


## Friedman Test: Testing q × Condition Interactions

The Friedman test evaluates whether entropy patterns across q-values differ between normal and tumor samples [@S205]. It ranks observations within each subject, preserving the paired structure critical for valid inference in paired designs [@S044]. This rank-based approach is particularly robust for skewed distributions like entropy values [@S019], and the Westfall-Young permutation correction used here is asymptotically optimal for dependent test statistics [@S166].

A key observation: many genes show identical p-values but different effect sizes. This reflects the rank-based nature of the test [@S115]. Effect size η² directly quantifies the strength of the q-value × condition dependence and serves as the practical ranking metric when p-values cluster [@S205]:

$$\eta^2 = \frac{SS_\text{between}}{SS_\text{total}}$$

This represents the proportion of total variance in entropy explained by changes in q-values. Genes with η² = 97.8% show dramatic entropy variation across q-values, while genes with η² = 59.4% show moderate variation.

## Interpreting Identical p-values and Effect Size Ranking

A key observation in this analysis is that **many genes share identical p-values** (e.g., 3.224e-43) despite dramatically different effect sizes (59.9%, 58.6%, 54.6%). This is **not a statistical error** but rather a fundamental property of the two-way Friedman test applied to q × condition interaction:

**Why identical p-values occur in two-way Friedman:**
The test statistic depends on the **rank patterns** of entropy values across q-levels within each condition and subject [@S205]. When genes show similar interaction structure (e.g., Q-entropy consistently higher in tumors vs normals at all q-values), they produce identical test statistics despite different magnitudes of entropy change. This property of rank statistics [@S115] is characteristic of non-parametric testing when distributions are heavily skewed or bounded. For multifactorial designs, aligned rank transform methods [@S187] provide robust alternatives to traditional ANOVA.

**How to rank genes when p-values cluster:**
In this scenario, **effect size (η²) becomes the practical ranking metric** [@S205] because it directly quantifies the strength of the q-value × entropy dependence:
- η² captures the proportion of variance explained by q-values [@S205]
- Higher η² = stronger biological effect for that gene
- Effect size properly reflects the magnitude of entropy change across q-values

The vignette table sorts genes by (1) p-value and (2) effect size as tiebreaker. This strategy ensures that statistically significant genes are listed first, with effect size distinguishing between genes with identical statistical significance.

**Recommendations:**
- Focus on genes with η² > 0.3 for robust biological findings
- Cross-validate significant genes using complementary methods (e.g., GAM section below)
- Remember that statistical significance (p-value) indicates q-dependence exists, while effect size says how strong it is

***

# Generalized Additive Model Approach (GAM)

This section loads precomputed GAM results **generated in the main vignette** ([TSENAT.Rmd](TSENAT.Rmd)). The GAM analysis provides a complementary parametric approach to detect q-value × group interactions. Results are integrated into the TSENATAnalysis object for S4 workflow consistency:

- **Model:** Generalized Additive Models (GAM) with ARIMA(1,1,0) transformations (can be combined with rank-based quasi-likelihood regularization) [@C073; @C077]
- **Approach:** Smooth functions fit across q-values for each gene
- **Correction:** Hochberg multiple testing correction accounting for AR(1) correlation
- **Assumptions:** Assumes continuous underlying distribution (better for nonlinear patterns)
- **Source:** Precomputed results from main vignette analysis (see [TSENAT.Rmd](TSENAT.Rmd) for generation details)

The precomputed results include effect sizes, model convergence diagnostics, and heteroscedasticity detection. Comparing GAM and Friedman Test results helps identify robust genes significant in both methods.

```{r gam-analysis, message=FALSE}
# Load precomputed GAM results from TSV file
gam_results <- read.delim(
    system.file("extdata", "lm_interaction_results.tsv", package = "TSENAT"),
    stringsAsFactors = FALSE
)
```

```{r gam-summary, message=FALSE, echo=FALSE}
# Create summary statistics table with additional metrics
# Handle NA and NaN values properly
mean_effect <- mean(gam_results$effect_size, na.rm = TRUE)
median_effect <- median(gam_results$effect_size, na.rm = TRUE)

summary_stats <- data.frame(
  Metric = c(
    "Genes tested", 
    "Significant (p < 0.05)", 
    "Concordant (p < 0.05 AND adj_p < 0.05)",
    "NAs",
    "Mean effect size",
    "Median effect size",
    "Strong effect genes (effect_size > 30%)",
    "Model convergence rate"
  ),
  Value = c(
    nrow(gam_results),
    sum(gam_results$p_interaction < 0.05, na.rm = TRUE),
    sum(gam_results$p_interaction < 0.05 & gam_results$adj_p_interaction < 0.05, na.rm = TRUE),
    sum(is.na(gam_results$p_interaction)),
    if (is.na(mean_effect) || !is.finite(mean_effect)) "Not available" else sprintf("%.1f%%", mean_effect * 100),
    if (is.na(median_effect) || !is.finite(median_effect)) "Not available" else sprintf("%.1f%%", median_effect * 100),
    sum(gam_results$p_interaction < 0.05 & gam_results$effect_size > 0.30, na.rm = TRUE),
    sprintf("%.1f%%", mean(gam_results$model_converged, na.rm = TRUE) * 100)
  )
)

knitr::kable(summary_stats, caption = "GAM Results Summary", align = c("l", "r"))

# Show top genes with additional information
if ("p_interaction" %in% colnames(gam_results)) {
  gam_sig <- gam_results[!is.na(gam_results$p_interaction), ]
  
  if (nrow(gam_sig) > 0) {
    gam_sig <- gam_sig[order(gam_sig$p_interaction), ]
    gam_sig <- gam_sig[1:min(10, nrow(gam_sig)), ]
    
    # Select columns for display: gene, p-values, effect size, test statistic, df
    cols_to_show <- intersect(c("gene", "p_interaction", "adj_p_interaction", "effect_size", 
                                "test_statistic", "df_residual"), 
                              colnames(gam_sig))
    if (length(cols_to_show) > 0) {
      gam_display <- gam_sig[, cols_to_show, drop = FALSE]
      
      # Format numeric columns
      numeric_cols <- sapply(gam_display, is.numeric)
      gam_display[, numeric_cols] <- round(gam_display[, numeric_cols], 3)
      
      # Format effect size as percentage
      if ("effect_size" %in% colnames(gam_display)) {
        gam_display$effect_size <- sapply(gam_display$effect_size, function(x) {
          if (is.na(x)) "NA" else sprintf("%.1f%%", x * 100)
        })
      }
      
      # Format test statistic
      if ("test_statistic" %in% colnames(gam_display)) {
        gam_display$test_statistic <- sapply(gam_display$test_statistic, function(x) {
          if (is.na(x)) "NA" else sprintf("%.2f", x)
        })
      }
      
      # Format df as integer
      if ("df_residual" %in% colnames(gam_display)) {
        gam_display$df_residual <- sapply(gam_display$df_residual, function(x) {
          if (is.na(x)) "NA" else as.integer(x)
        })
      }
      
      # Rename columns for better readability
      colnames(gam_display) <- c("Gene", "p (interaction)", "Adjusted p", "Effect size", "Test statistic", "df")
      
      knitr::kable(gam_display, 
                   caption = "Top 10 Genes by GAM p-value (with Effect Size & Test Statistics)",
                   align = c("l", "r", "r", "r", "r", "r"))
    }
  }
} else {
  cat("WARNING: p_interaction column not found in results\n")
  cat("Available columns:", paste(colnames(gam_results), collapse=", "), "\n")
}
```

***

# Method Concordance Analysis

```{r concordance-summary, message=FALSE}
# Compute concordance analysis using S4 wrapper
# Pass GAM results directly - the function handles storage automatically
analysis_with_concordance <- compute_method_concordance_s4(
    analysis,
    gam_method = "gam_results",
    friedman_method = "q_interactions",
    gam_results = gam_results,
    verbose = TRUE
)

# Extract components from S4 results
concordance_result <- metadata(analysis_with_concordance, "method_concordance")
comparison_df <- concordance_result$comparison_df
spearman_rho <- concordance_result$spearman_rho
high_conf <- concordance_result$high_confidence
agreement_table <- concordance_result$agreement_table
```

```{r concordance-metrics, message=FALSE, echo=FALSE}
# === GLOBAL CONCORDANCE METRICS ===
total_genes <- nrow(comparison_df)
both_sig <- sum(comparison_df$agreement == "Both significant")
gam_only <- sum(comparison_df$agreement == "GAM only")
friedman_only <- sum(comparison_df$agreement == "Friedman only")
neither_sig <- sum(comparison_df$agreement == "Neither significant")

concordance_metrics <- data.frame(
  Metric = c(
    "Total genes compared",
    "Spearman correlation (p-values)",
    "Both methods significant (p < 0.05)",
    "GAM only significant",
    "Friedman only significant",
    "Neither significant",
    "Concordance rate",
    "Discordance rate"
  ),
  Value = c(
    sprintf("%d", total_genes),
    sprintf("ρ = %.4f", spearman_rho),
    sprintf("%d (%.1f%%)", both_sig, 100 * both_sig / total_genes),
    sprintf("%d (%.1f%%)", gam_only, 100 * gam_only / total_genes),
    sprintf("%d (%.1f%%)", friedman_only, 100 * friedman_only / total_genes),
    sprintf("%d (%.1f%%)", neither_sig, 100 * neither_sig / total_genes),
    sprintf("%.1f%%", 100 * (both_sig + neither_sig) / total_genes),
    sprintf("%.1f%%", 100 * (gam_only + friedman_only) / total_genes)
  )
)

knitr::kable(concordance_metrics, 
             caption = "Global Concordance Metrics: GAM vs Friedman Methods",
             align = c("l", "r"))
```

```{r concordance-agreement, message=FALSE, echo=FALSE}
# === AGREEMENT CATEGORY BREAKDOWN ===
if (!is.null(agreement_table) && length(agreement_table) > 0) {
  agreement_breakdown <- data.frame(
    Category = names(agreement_table),
    Count = as.numeric(agreement_table),
    Percentage = sprintf("%.1f%%", 100 * as.numeric(agreement_table) / sum(agreement_table)),
    stringsAsFactors = FALSE
  )
  
  colnames(agreement_breakdown) <- c("Agreement Category", "Number of Genes", "Percentage")
  
  knitr::kable(agreement_breakdown,
               caption = "Method Agreement Distribution",
               align = c("l", "r", "r"))
} else {
  cat("No agreement table available from concordance analysis.")
}
```

```{r concordance-highconf, message=FALSE, echo=FALSE}
# === HIGH-CONFIDENCE GENES (BOTH METHODS SIGNIFICANT) ===
if (!is.null(high_conf) && nrow(high_conf) > 0) {
  # Sort by GAM adjusted p-value, then by Friedman adjusted p-value (both ascending)
  high_conf_sorted <- high_conf[order(high_conf$padj_gam, high_conf$padj_friedman), ]
  
  hc_display <- high_conf_sorted[1:min(15, nrow(high_conf_sorted)), c("gene", "padj_gam", "padj_friedman", "effect_gam", "effect_friedman")]
  
  hc_display$padj_gam <- sprintf("%.3e", hc_display$padj_gam)
  hc_display$padj_friedman <- sprintf("%.3e", hc_display$padj_friedman)
  hc_display$effect_gam <- sprintf("%.1f%%", hc_display$effect_gam * 100)
  hc_display$effect_friedman <- sprintf("%.3f", hc_display$effect_friedman)
  
  colnames(hc_display) <- c("Gene", "GAM adj p", "Friedman adj p", "GAM Effect", "Friedman η²")
  rownames(hc_display) <- NULL
  
  knitr::kable(hc_display,
               caption = sprintf("Robust q-value Interactions: High-Confidence Genes Detected by Both Methods (n=%d, ranked by statistical significance)", nrow(high_conf)),
               align = c("l", "r", "r", "r", "r"))
} else {
  cat("No high-confidence genes detected by both methods (p < 0.05 in both GAM and Friedman tests).")
}
```

## Visualization: Method Comparison

```{r concordance-plot, fig.width=10, fig.height=8, fig.cap="Method concordance visualization comparing Friedman rank-based and GAM approaches for detecting q×condition interactions. Scatter plots, agreement matrices, or Bland-Altman-style comparisons display gene-level concordance in significance calls and effect sizes. Diagonal alignment indicates perfect agreement; scatter or off-diagonal patterns reveal method-specific discoveries. Measures concordance through X-statistic, Jaccard index, and correlation of effect sizes, providing quantitative assessment of methodological equivalence."}
# Create comparison plot using S4 wrapper
if (!is.null(comparison_df) && nrow(comparison_df) > 0) {
  p <- plot_method_concordance_s4(analysis_with_concordance, verbose = TRUE)
  print(p)
} else {
  cat("No data available for concordance visualization. Ensure both GAM and Friedman results have sufficient genes for comparison.")
}
```

***

# Complete S4 Workflow Guide

This vignette demonstrates the modern S4-based analysis workflow for method comparison. The following functions are used in their S4 wrapper forms for consistent class-based integration:

## S4 Functions Used in This Vignette

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `calculate_diversity_s4()` | Compute Tsallis entropy across multiple q-values | TSENATAnalysis + `q`, `norm` params | Updated analysis with diversity results |
| `test_rankbased_assumptions_s4()` | Validate Friedman test assumptions | TSENATAnalysis | Stores results in metadata via `metadata(analysis, "rankbased_assumptions")` |
| `rank_test_q_condition_s4()` | Run Friedman test for q×group interactions | TSENATAnalysis + design params | Updated analysis with results via `rankResults()` |
| `plot_tsallis_q_curve_s4()` | Plot q-curves for significant genes | TSENATAnalysis + `lm_res` | ggplot2 visualization |
| `compute_method_concordance_s4()` | Compare GAM and Friedman results | TSENATAnalysis with LM results via `lmResults()` and rank test results via `rankResults()` | Updated analysis with results via `metadata()` |
| `plot_method_concordance_s4()` | Visualize method concordance | TSENATAnalysis (after concordance computed) | ggplot2 comparison plot |

## Workflow Advantages

- **Consistent class structure:** All results integrated into TSENATAnalysis object
- **Automatic metadata tracking:** No global variables polluting environment
- **Seamless integration:** Methods pass results through accessor methods `metadata()`, `lmResults()`, and `rankResults()`
- **Reproducibility:** Clear, standardized workflow for complex multi-method analyses
- **Documentation:** S4 methods maintain Roxygen documentation and inheritance

## Workflow Reconstruction

If needed, this complete workflow can be recreated step-by-step:

```{r workflow-reconstruction, eval=FALSE}
# 1. Initialize analysis
analysis <- TSENATAnalysis(se = se, config = list())

# 2. Calculate diversity
analysis <- calculate_diversity_s4(analysis, q = seq(0.1, 2, by = 0.05), norm = TRUE)

# 3. Validate assumptions
rank_assumptions <- test_rankbased_assumptions_s4(analysis, verbose = TRUE)

# 4. Friedman test (rank-based)
analysis <- rank_test_q_condition_s4(
    analysis,
    condition_col = "condition",
    paired = TRUE,
    subject_col = "paired_samples",
    multicorr = "hochberg",
    verbose = TRUE
)

# 5. Load GAM results and pass directly to concordance function
gam_results_loaded <- read.csv("lm_interaction_results.csv")

# 6. Compare methods (gam_results parameter handles storage automatically)
analysis <- compute_method_concordance_s4(
    analysis,
    gam_method = "gam_results",
    friedman_method = "q_interactions",
    gam_results = gam_results_loaded,
    verbose = TRUE
)

# 7. Visualize concordance
plot_method_concordance_s4(analysis, verbose = TRUE)

# 8. Plot top genes (optional)
plot_tsallis_q_curve_s4(
    analysis,
    condition_col = "condition",
    rank_res = rankResults(analysis),
    n_top = 4
)
```

***

# Recommendations

## S4 Wrapper Integration Overview

This vignette demonstrates the modern S4 wrapper approach for method concordance analysis:

- **`compute_method_concordance_s4()`**: Integrates seamlessly with TSENATAnalysis objects, storing results accessed via `metadata(analysis, "method_concordance")`
- **`plot_method_concordance_s4()`**: Creates publication-ready concordance visualizations from S4-stored results
- **Benefits**: Consistent class structure, automatic metadata tracking via accessor methods, reduced variable proliferation in global environment

For standard (non-S4) analyses, the original functions (`compute_method_concordance()` and `plot_method_concordance()`) remain available and fully supported.

## Best Practices for Analysis

We recommend a **complementary two-method approach** for robust detection of q-value × group interactions:

**Primary analysis:** Start with the Friedman Test for q × condition interaction, which provides correct statistical inference for paired/blocked designs by:
1. **Testing the interaction**: Does the q-effect differ between conditions? (Not just whether entropy varies across q-values)
2. **Respecting the repeated-measures structure**: 8 paired tumor-normal samples, testing if entropy patterns across q-values differ by condition
3. **Accounting for within-subject correlation**: Friedman test properly handles paired structure
4. **Multiple testing correction**: Westfall-Young permutation respects AR(1) correlation structure across q-values
5. **Condition-aware**: Uses two-way Friedman (q within-subjects, condition between-subjects)

**Validation step:** Confirm findings with GAM to validate against parametric assumptions and identify condition-specific nonlinear q-entropy patterns that may be missed by rank-based methods.

**High-confidence genes** are those significant in *both* methods—these provide the strongest evidence of genuine q × condition interactions and are recommended for downstream biological validation. The Friedman Test is particularly appropriate for paired designs and provides distribution-free inference without requiring assumptions about data normality, while properly accounting for the q × condition interaction structure.

***

# Session Information

```{r session-info}
sessionInfo()
```

***

# References
